---
title: "Conditional RMSD for Rasch item fit, Part A"
subtitle: "Null behaviour, stability in n, leakage, discrimination and targeting, without a bootstrap"
author: "Magnus Johansson"
date: today
---

## Purpose

In large samples, a calibrated item fit test detects misfit too small to
matter. This study looks for an effect size that stays put as n grows. The
candidate is a conditional RMSD (cRMSD), the root mean squared deviation
between the observed and the model-implied conditional item characteristic
curve, with the total score as the conditioning variable and CML item
parameters. It is the numeric summary of what `RMitemICCPlot()` draws. The
literature RMSD is marginal (MML posterior pseudo-counts on a theta grid). A
sum-score conditional version was not found. Background and the literature
summary are in `conditional-rmsd-design.md`.

Part A computes statistics only, with no bootstrap. Calibrated tests are
Part B.

## Definitions

For item $i$ with maximum score $m_i$, let $s$ be the total score, $n_s$ the
number of respondents with score $s$, $N$ the number with a non-extreme total
score, $O_{is}$ the observed mean item score in group $s$, and $E_{is}$,
$V_{is}$ the conditional mean and variance under the Rasch model
(`.cond_moments()`, from the elementary symmetric functions and the CML
thresholds). With weights $w_s = n_s/N$ over non-extreme scores,

$$
\text{MSD}_i = \frac{1}{m_i^2}\sum_s w_s (O_{is} - E_{is})^2, \qquad
\text{cRMSD}_i = \sqrt{\text{MSD}_i}.
$$

Under the model, responses within a score group are independent, so
$E[(O_{is} - E_{is})^2] = V_{is}/n_s$ and the bias of $\text{MSD}_i$ is

$$
B_i = \frac{1}{N m_i^2}\sum_{s \in G} V_{is},
$$

with $G$ the observed non-extreme scores. The bias-corrected value is
$\text{cRMSD}^{bc}_i = \sqrt{\max(\text{MSD}_i - B_i, 0)}$. Estimating the
item location removes about one degree of freedom, so $B_i$ should slightly
overstate the null mean of $\text{MSD}_i$. The null condition checks this.

Two exact properties under CML, both checked on every dataset:

1. The signed mean deviation $\sum_s n_s (O_{is} - E_{is})$ is 0 (the CML
   score equation). A uniform shift is absorbed into the item location, so
   cRMSD measures the shape of the conditional curve only.
2. For dichotomous items, conditional infit is a linear contrast of the same
   residuals,
   $$
   \text{Infit}_i - 1 = \frac{\sum_s n_s (O_{is} - E_{is})(1 - 2E_{is})}{\sum_s n_s V_{is}}.
   $$
   Infit is a signed slope contrast, and cRMSD is the quadratic omnibus
   summary of the same residuals.

Two secondary variants are recorded:

- **Pooled.** Adjacent scores are merged until each group has at least 50
  respondents. Pooling reduces noise but can hide deviations within a group.
- **Information-weighted.** Weights $w_s \propto n_s V_{is}$, which emphasise
  score groups where the item is informative, as infit does.

## Predictions, stated before running

- **P1.** Under second-dimension misfit, infit discriminates misfitting from
  fitting items at least as well as cRMSD, and better at rho = .75. Infit is
  the directed contrast for a flatter curve.
- **P2.** Bias-corrected cRMSD and infit MSQ are both stable in n. Raw cRMSD
  falls with n.
- **P3.** Leakage into fitting items is small relative to the misfitting items
  at rho = .50, and closer at rho = .75.
- **P4.** cRMSD of the misfitting item at -2 logits is smaller than at 0
  logits, relative to infit.
- **P5.** Under asymmetric-IRF misfit (LPE), cRMSD discriminates better than
  infit, because an asymmetric deviation is roughly orthogonal to the slope
  contrast.
- **P6.** With three misfitting items at rho = .15 and .50, where detection
  saturates, cRMSD and RMSEA separate misfitting from fitting items better
  than |infit - 1| (lower false-positive rate at 95 percent detection, Q4),
  as S-X2 did in `sx2_comparison_v2.qmd`. The leakage into fitting items
  (Q3) is a smaller share of the misfitting items' value for cRMSD than for
  |infit - 1|.

## Design

- **Items.** 20 dichotomous items with locations evenly spaced on [-2, 2]. The
  misfitting items are the ones nearest 0, -1 and -2 logits. With one
  misfitting item it is the one nearest 0.
- **Persons.** Theta ~ N(0, 1.5^2).
- **Conditions.**
  - `null`: Rasch model, no misfit.
  - `xdim`: misfitting items respond to a second dimension, correlation
    rho = .50 or .75 with the first, both SD 1.5, with one or three
    misfitting items. Also rho = .15 with three misfitting items, the
    strongest cascade in `sx2_comparison_v2.qmd`. The generator, the item locations and the choice of misfitting
    items are the same as in `sx2_comparison_v2.R`, so the cells at
    n = 500 to 2,000 match that study.
  - `lpe`: misfitting items follow a logistic positive exponent IRF with a
    slope adjustment,
    $P(\theta) = \text{logistic}(a(\theta - b'))^{\xi}$, with $b'$ chosen so
    that $P(b) = .5$. The curve is asymmetric but monotone. Without the
    adjustment ($a = 1$), the exponent also changes the effective slope
    (in a smoke test, $\xi = 4$ gave infit 0.85), so infit would detect it as
    slope misfit. $a$ is chosen so that the item's mean infit is 1 at
    n = 20,000, separately for each misfit position. What remains is
    asymmetry with no net slope contrast, which is the case P5 is about.
    $\xi$ and $a$ are set in the calibration chunks below. One or three
    misfitting items.
- **Sample sizes.** 500, 1,000, 2,000, 5,000, 10,000, 20,000. The values at
  n = 20,000, averaged over replications, serve as the population values.
- **Replications.** 300 per cell, each a fresh draw (no subsampling of a
  master dataset).
- **Statistics per item.** Raw and bias-corrected cRMSD (primary: distribution
  weighted, unpooled), the pooled and information-weighted variants,
  conditional infit MSQ (`iarm::out_infit()`), the item-restscore gamma
  difference (observed minus expected, `iarm::item_restscore()`), and S-X2
  with RMSEA (`mirt`). S-X2 uses mirt's df + 1, the correction found in
  `sx2_null_diagnostic.qmd`, and RMSEA is recomputed from S-X2, the corrected
  df and N.

**Relation to the S-X2 studies.** `sx2_itemfit_sim.qmd` is superseded and is
not used here. `sx2_comparison_v2.qmd` compared tests at a common family-wise
error rate up to n = 2,000. Conditional infit with Westfall-Young detected
more. The S-X2 bootstrap flagged far fewer fitting items when three items
misfit at n of 1,000 or more, at nearly the same detection when misfit was
strong or moderate (rho = .50, n = 1,000: fitting items flagged in 3.5
against 39.5 percent of datasets, detection 93.8 against 98).

v2 attributes this to the bootstrap reference, because asymptotic S-X2 with
BH cascaded more. A reanalysis of its cache (2026-10-05, p-values as stored)
shows that the difference is BH, not the reference. BH controls the false
discovery rate and loosens its threshold once the misfitting items are
rejected. With Holm, which controls the family-wise rate like Westfall-Young,
df-corrected asymptotic S-X2 flags fitting items about as rarely as the
bootstrap, at the same detection:

| rho | n | infit WY | infit Holm | S-X2 boot WY | S-X2 df+1 Holm | S-X2 df+1 BH |
|---|---|---|---|---|---|---|
| .15 | 1,000 | 97.0 | 90.5 | 6.5 | 11.0 | 34.0 |
| .15 | 2,000 | 100 | 100 | 47.5 | 55.0 | 93.0 |
| .50 | 1,000 | 39.5 | 28.5 | 3.5 | 4.5 | 14.5 |
| .50 | 2,000 | 87.0 | 72.5 | 7.5 | 9.0 | 28.0 |

Datasets with at least one fitting item flagged (%), three misfitting items.

So the cascade resistance is a property of the S-X2 statistic. A likely
mechanism follows from the infit identity above. The cascade puts small slope
deviations on the fitting items. Infit is a one-df slope contrast and is
sensitive to them. S-X2 spreads the same deviation over about 17 df and is not
sensitive to it. Against the large deviation of a misfitting item both
saturate. cRMSD is an omnibus summary like S-X2, so it may share this
property (P6).

Part A asks a different question from v2: how the statistics behave as effect
sizes, up to n = 20,000, with no test. The S-X2 columns here are therefore
descriptive (RMSEA as an effect size, and the ROC on it), not a rerun of that
comparison.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)  # for .cond_moments()
  library(psychotools)
  library(iarm)
  library(mirt)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

NCORES      <- 10
R_REPS      <- 300
NS          <- c(500, 1000, 2000, 5000, 10000, 20000)
K_ITEMS     <- 20
THETA_SD    <- 1.5
POOL_MIN    <- 50
INCLUDE_SX2 <- TRUE
CACHE_DIR   <- "conditional_rmsd"
dir.create(CACHE_DIR, showWarnings = FALSE)

locs <- seq(-2, 2, length.out = K_ITEMS)
misfit_idx <- integer(0)
for (tg in c(0, -1, -2)) {
  misfit_idx <- c(misfit_idx, setdiff(order(abs(locs - tg)), misfit_idx)[1])
}
data.frame(item = paste0("V", misfit_idx), target = c(0, -1, -2),
           location = round(locs[misfit_idx], 3))
```

### Data generation

```{r gen}
# LPE item: P = logistic(a * (theta - b'))^xi, with b' = b - shift / a so
# that P(b) = .5.
lpe_shift <- function(xi) qlogis(0.5^(1 / xi))

# Slope that gives the LPE curve the Rasch slope (1/4) at P = .5. The
# derivative there is xi * a * 0.5 * (1 - 0.5^(1/xi)). Starting value for the
# numeric calibration below.
lpe_a_start <- function(xi) 0.5 / (xi * (1 - 0.5^(1 / xi)))

lpe_prob <- function(th, b, xi, a) {
  plogis(a * (th - (b - lpe_shift(xi) / a)))^xi
}

# `a_lpe` holds one slope per misfit position (0, -1, -2 logits). It is set
# by the calibration chunk and defaults to the analytic start.
a_lpe <- NULL

# `pos` selects which misfit positions are used (1 = 0 logits, 2 = -1,
# 3 = -2). The default is the first k. The calibration uses single positions.
gen_data <- function(type, k, par, n, a = a_lpe, pos = seq_len(k)) {
  use <- misfit_idx[pos]
  th <- rnorm(n, 0, THETA_SD)
  if (type == "xdim") {
    # Same generator as sx2_comparison_v2.R: fitting items respond to theta1,
    # misfitting items to theta2, both SD 1.5, correlation `par`.
    TH <- matrix(th, n, K_ITEMS)
    if (k > 0) {
      th2 <- par * th + sqrt(1 - par^2) * rnorm(n, 0, THETA_SD)
      TH[, use] <- th2
    }
    P <- plogis(TH - matrix(locs, n, K_ITEMS, byrow = TRUE))
    X <- (matrix(runif(n * K_ITEMS), n) < P) * 1L
  } else {
    P <- plogis(outer(th, locs, "-"))
    if (type == "lpe" && k > 0) {
      if (is.null(a)) a <- rep(lpe_a_start(par), 3)
      for (p in pos) {
        j <- misfit_idx[p]
        P[, j] <- lpe_prob(th, locs[j], par, a[p])
      }
    }
    X <- (matrix(runif(n * K_ITEMS), n) < P) * 1L
  }
  X <- as.data.frame(X)
  names(X) <- paste0("V", seq_len(K_ITEMS))
  X
}
```

### Statistics

```{r stats}
# Merge adjacent score groups until each has at least `min_n` respondents.
pool_groups <- function(n_g, min_n) {
  id <- integer(length(n_g)); g <- 1L; acc <- 0
  for (j in seq_along(n_g)) {
    id[j] <- g; acc <- acc + n_g[j]
    if (acc >= min_n && j < length(n_g)) { g <- g + 1L; acc <- 0 }
  }
  if (acc < min_n && g > 1L) id[id == g] <- g - 1L  # fold a short tail back
  id
}

compute_stats <- function(X) {
  m <- psychotools::pcmodel(X)
  thr <- lapply(psychotools::threshpar(m), as.numeric)
  mom <- .cond_moments(thr)
  m_i <- vapply(thr, length, integer(1))
  s <- rowSums(X)
  keep <- s > 0 & s < mom$max_score
  N <- sum(keep)
  ss <- s[keep]
  S_sum <- rowsum(as.matrix(X[keep, ]), ss)       # groups x items
  g <- as.integer(rownames(S_sum))
  n_g <- as.numeric(table(ss)[as.character(g)])
  O <- S_sum / n_g
  E <- mom$E[g + 1, , drop = FALSE]
  V <- mom$V[g + 1, , drop = FALSE]
  D <- O - E

  # primary: distribution weighted, unpooled
  msd <- colSums(n_g * D^2) / N / m_i^2
  B   <- colSums(V) / N / m_i^2
  # information weighted
  wv    <- n_g * V
  msd_w <- colSums(wv * D^2) / colSums(wv) / m_i^2
  B_w   <- colSums(V^2) / colSums(wv) / m_i^2
  # pooled
  pid <- pool_groups(n_g, POOL_MIN)
  n_p <- as.numeric(tapply(n_g, pid, sum))
  O_p <- rowsum(S_sum, pid) / n_p
  E_p <- rowsum(n_g * E, pid) / n_p
  B_p <- colSums(rowsum(n_g * V, pid) / n_p) / N / m_i^2
  msd_p <- colSums(n_p * (O_p - E_p)^2) / N / m_i^2

  infit <- iarm::out_infit(m)$Infit
  infit_lin <- 1 + colSums(n_g * D * (1 - 2 * E)) / colSums(n_g * V)

  rs <- iarm::item_restscore(m)
  gdiff <- as.numeric(rs[, "observed"]) - as.numeric(rs[, "expected"])

  out <- data.frame(
    item = seq_len(ncol(X)),
    msd = msd, B = B,
    crmsd_raw = sqrt(msd),
    crmsd_bc  = sqrt(pmax(msd - B, 0)),
    crmsd_w_bc = sqrt(pmax(msd_w - B_w, 0)),
    crmsd_p_bc = sqrt(pmax(msd_p - B_p, 0)),
    infit = infit,
    gamma_diff = gdiff,
    md = colSums(n_g * D) / N,
    infit_identity_err = abs(infit - infit_lin),
    n_groups = length(g), n_pooled = length(n_p)
  )

  if (INCLUDE_SX2) {
    sx <- tryCatch({
      mm <- mirt::mirt(X, 1, itemtype = "Rasch", verbose = FALSE)
      f <- mirt::itemfit(mm, fit_stats = "S_X2")
      # mirt removes one df per item for the estimated latent variance,
      # which is one too many for the Rasch model (sx2_null_diagnostic.qmd).
      # Use df + 1 for both the p-value and RMSEA.
      data.frame(sx2 = f$S_X2, sx2_df = f$df.S_X2 + 1)
    }, error = function(e) data.frame(sx2 = NA, sx2_df = NA))
    out <- cbind(out, sx[seq_len(ncol(X)), ])
    out$sx2_p <- pchisq(out$sx2, out$sx2_df, lower.tail = FALSE)
    out$sx2_rmsea <- sqrt(pmax(0, (out$sx2 - out$sx2_df) /
                                  (out$sx2_df * (nrow(X) - 1))))
  }
  out
}
```

### Calibrating the LPE item

The LPE condition is meant as misfit of moderate size with no net slope
contrast, so that infit is near 1 while the conditional curve is still
wrong. Two steps.

**Step 1: choose $\xi$.** For several values of $\xi$, with the analytic
slope `lpe_a_start(xi)` (the Rasch slope at $P = .5$), the table gives the
bias-corrected cRMSD and the infit of a single LPE item at 0 logits, at
n = 20,000 (5 datasets each). Choose `XI` so that cRMSD is roughly .03 to .05
(small to medium on the Köhler et al., 2020, scale). Record the choice and
the reason here **before** the main run.

```{r lpe-xi}
cal_file <- file.path(CACHE_DIR, "lpe_xi_table.rds")
if (file.exists(cal_file)) {
  cal <- readRDS(cal_file)
} else {
  cal <- do.call(rbind, lapply(c(0.1, 0.25, 0.5, 2, 4, 8, 16), function(xi) {
    a0 <- lpe_a_start(xi)
    do.call(rbind, lapply(1:5, function(r) {
      set.seed(9e5 + 100 * xi + r)
      X <- gen_data("lpe", 1, xi, 20000, a = rep(a0, 3), pos = 1)
      st <- compute_stats(X)
      data.frame(xi = xi, a_start = a0, rep = r,
                 st[misfit_idx[1], c("crmsd_bc", "infit", "gamma_diff")])
    }))
  }))
  saveRDS(cal, cal_file)
}
cal |>
  group_by(xi) |>
  summarise(across(c(a_start, crmsd_bc, infit, gamma_diff),
                   \(x) round(mean(x), 3)))

XI <- 0.1
```

**Choice, recorded 2026-10-05 before the main run.** With the analytic slope,
only $\xi = 0.1$ reached the target range (cRMSD .039, infit 0.97). The other
values gave .006 to .026. The slope adjustment removes most of what an
exponent does to the curve, so a strongly asymmetric curve is needed for a
cRMSD in the small-to-medium range. $\xi = 0.1$ makes the curve rise
gradually at low theta and steeply near and above its location.

**Step 2: set the slope for each position.** For the chosen `XI`, find
the `a` that makes the item's mean infit equal 1 at n = 20,000, separately
for the items at 0, -1 and -2 logits, each as the only LPE item. The same
three seeds are used at every candidate `a` (common random numbers), so
mean infit is a smooth function of `a` and `uniroot()` can solve for it. The
check table then reports the item's infit and cRMSD at the calibrated slope
on five new datasets. In the three-item condition the other LPE items shift
the total score slightly, so infit there will be close to 1 but not exactly
1. Changing `XI` or these slopes later means deleting the `lpe_*` cache
files.

```{r lpe-slope}
slope_file <- file.path(CACHE_DIR, sprintf("lpe_slopes_xi%s.rds", XI))
if (file.exists(slope_file)) {
  slope_cal <- readRDS(slope_file)
} else {
  slope_cal <- do.call(rbind, lapply(1:3, function(p) {
    j <- misfit_idx[p]
    mean_infit <- function(a) {
      aa <- rep(lpe_a_start(XI), 3); aa[p] <- a
      mean(vapply(1:3, function(r) {
        set.seed(8e5 + 10 * p + r)
        compute_stats(gen_data("lpe", 1, XI, 20000, a = aa, pos = p))$infit[j]
      }, numeric(1)))
    }
    a0 <- lpe_a_start(XI)
    root <- uniroot(function(a) mean_infit(a) - 1,
                    interval = c(a0 / 2, a0 * 2), tol = 1e-3)
    data.frame(position = p, target = c(0, -1, -2)[p], a_start = a0,
               a = root$root, iterations = root$iter)
  }))
  saveRDS(slope_cal, slope_file)
}
a_lpe <- slope_cal$a

check_file <- file.path(CACHE_DIR, sprintf("lpe_slope_check_xi%s.rds", XI))
if (file.exists(check_file)) {
  slope_check <- readRDS(check_file)
} else {
  slope_check <- do.call(rbind, lapply(1:3, function(p) {
    do.call(rbind, lapply(1:5, function(r) {
      set.seed(7e5 + 10 * p + r)
      st <- compute_stats(gen_data("lpe", 1, XI, 20000, pos = p))
      data.frame(position = p, st[misfit_idx[p], c("crmsd_bc", "infit",
                                                    "gamma_diff")])
    }))
  }))
  saveRDS(slope_check, check_file)
}
slope_cal |> mutate(across(c(a_start, a), \(x) round(x, 3)))
slope_check |>
  group_by(position) |>
  summarise(across(everything(), \(x) round(mean(x), 3)))
```

If `uniroot()` fails because no slope in [a/2, 2a] gives infit 1, choose
another `XI`.

### Run

One cache file per condition and sample size. Delete a file to rerun that
cell.

```{r conditions}
conds <- tribble(
  ~cond,        ~type,  ~k, ~par,
  "null",       "null",  0,  NA,
  "xdim15_k3",  "xdim",  3,  0.15,
  "xdim50_k1",  "xdim",  1,  0.50,
  "xdim50_k3",  "xdim",  3,  0.50,
  "xdim75_k1",  "xdim",  1,  0.75,
  "xdim75_k3",  "xdim",  3,  0.75,
  "lpe_k1",     "lpe",   1,  XI,
  "lpe_k3",     "lpe",   3,  XI
)
conds
```

```{r run}
run_cell <- function(ci, n) {
  cc <- conds[ci, ]
  f <- file.path(CACHE_DIR, sprintf("%s_n%d.rds", cc$cond, n))
  if (file.exists(f)) return(readRDS(f))
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_REPS), function(r) {
    set.seed(1e7 * ci + 100 * (n %/% 100) + r)
    out <- tryCatch({
      X <- gen_data(cc$type, cc$k, cc$par, n)
      st <- compute_stats(X)
      st$rep <- r
      st
    }, error = function(e) conditionMessage(e))
    out
  }, mc.cores = NCORES)
  ok <- vapply(res, is.data.frame, logical(1))
  df <- do.call(rbind, res[ok])
  df$cond <- cc$cond; df$type <- cc$type; df$k <- cc$k; df$par <- cc$par
  df$n <- n
  attr(df, "failed") <- unlist(res[!ok])
  attr(df, "minutes") <- as.numeric(Sys.time() - t0, units = "mins")
  saveRDS(df, f)
  df
}

cells <- expand.grid(ci = seq_len(nrow(conds)), n = NS)
all_res <- lapply(seq_len(nrow(cells)), function(j) run_cell(cells$ci[j],
                                                             cells$n[j]))
timing <- data.frame(
  cond = conds$cond[cells$ci], n = cells$n,
  failed = vapply(all_res, function(d) length(attr(d, "failed")), integer(1)),
  minutes = round(vapply(all_res, function(d) attr(d, "minutes") %||% NA_real_,
                         numeric(1)), 2)
)
res <- bind_rows(all_res) |>
  mutate(misfit = mapply(function(k, it) it %in% misfit_idx[seq_len(k)],
                         k, item),
         target = c("0", "-1", "-2")[match(item, misfit_idx)],
         abs_infit = abs(infit - 1),
         abs_gamma = abs(gamma_diff),
         # Untruncated scores for the ROC. cRMSD_bc and RMSEA are 0 for most
         # null items, and the ties make thresholds at 0 flag every item.
         # Both scores rank items like the truncated values where those are
         # positive.
         crmsd_score = msd - B,
         sx2_score = if (INCLUDE_SX2) (sx2 - sx2_df) / sx2_df else NA,
         n = factor(n, levels = NS))
```

## Checks

Failed datasets, the CML score equation (signed mean deviation should be 0
up to the CML convergence tolerance), and the infit identity (difference from
`iarm::out_infit()`, which returns infit unrounded).

```{r checks}
timing |> filter(failed > 0)
res |>
  group_by(cond) |>
  summarise(max_abs_md = signif(max(abs(md)), 2),
            max_infit_identity_err = signif(max(infit_identity_err), 2),
            datasets = n_distinct(paste(n, rep)))
```

## Q1. Null behaviour

Raw and bias-corrected cRMSD under a true Rasch model, and the ratio of the
mean MSD to the mean analytic bias $B_i$. A ratio below 1 is the degree of
freedom taken by the item location.

```{r q1}
res |>
  filter(cond == "null") |>
  group_by(n) |>
  summarise(raw_mean = mean(crmsd_raw), bc_mean = mean(crmsd_bc),
            bc_p95 = quantile(crmsd_bc, .95), bc_zero = mean(crmsd_bc == 0),
            msd_over_B = mean(msd) / mean(B),
            pooled_bc_mean = mean(crmsd_p_bc), info_bc_mean = mean(crmsd_w_bc),
            infit_sd = sd(infit), .groups = "drop") |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

## Q2. Stability in n

Mean values for the misfitting items by sample size. A statistic is stable if
its mean does not drift with n. The n = 20,000 row is the population proxy.

```{r q2}
q2 <- res |>
  filter(misfit) |>
  group_by(cond, n) |>
  summarise(crmsd_raw = mean(crmsd_raw), crmsd_bc = mean(crmsd_bc),
            crmsd_p_bc = mean(crmsd_p_bc), infit = mean(infit),
            gamma_diff = mean(gamma_diff),
            sx2_rmsea = if (INCLUDE_SX2) mean(sx2_rmsea) else NA,
            .groups = "drop")
q2 |> mutate(across(where(is.numeric), \(x) round(x, 3)))
```

```{r q2-fig, fig.height = 6}
q2 |>
  pivot_longer(c(crmsd_raw, crmsd_bc, infit, gamma_diff, sx2_rmsea),
               names_to = "stat") |>
  group_by(cond, stat) |>
  mutate(rel = value / value[n == "20000"]) |>
  ggplot(aes(n, rel, colour = stat, group = stat)) +
  geom_hline(yintercept = 1, linetype = 2) +
  geom_line() + geom_point() +
  facet_wrap(~cond) +
  labs(x = "Sample size", y = "Mean relative to n = 20,000", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Q3. Leakage into fitting items

Mean values for the *fitting* items, by condition, at each n, against the
null. The last table compares fitting and misfitting items at n = 20,000,
using each item's mean over replications as its population value. It gives
the mean over items and the most extreme item in each group: the largest
fitting item and the smallest misfitting item. If the smallest misfitting
item stands well above the largest fitting item, an effect-size floor between
the two can stop the cascade at any n.

```{r q3}
res |>
  filter(!misfit) |>
  group_by(cond, n) |>
  summarise(crmsd_bc = mean(crmsd_bc), infit_signed = mean(infit - 1),
            abs_infit = mean(abs_infit), abs_gamma = mean(abs_gamma),
            .groups = "drop") |>
  filter(n %in% c("500", "2000", "20000")) |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))

res |>
  filter(n == "20000", cond != "null") |>
  group_by(cond, misfit, item) |>
  summarise(crmsd_bc = mean(crmsd_bc), abs_infit = mean(abs_infit),
            abs_gamma = mean(abs_gamma), .groups = "drop") |>
  group_by(cond, misfit) |>
  summarise(crmsd_mean = mean(crmsd_bc),
            crmsd_extreme = if (first(misfit)) min(crmsd_bc) else max(crmsd_bc),
            abs_infit_mean = mean(abs_infit),
            abs_infit_extreme = if (first(misfit)) min(abs_infit) else max(abs_infit),
            abs_gamma_mean = mean(abs_gamma),
            abs_gamma_extreme = if (first(misfit)) min(abs_gamma) else max(abs_gamma),
            .groups = "drop") |>
  mutate(across(where(is.numeric), \(x) round(x, 4)))
```

## Q4. Discrimination

Misfitting against fitting items in the same datasets, for each statistic.
All statistics are made direction-free: cRMSD, |infit - 1|, |gamma
difference|, and S-X2. cRMSD enters as MSD - B and S-X2 as (S-X2 - df)/df,
the untruncated quantities behind bias-corrected cRMSD and RMSEA. They rank
items like the truncated values where those are positive, without the ties at
0. The pooled and information-weighted cRMSD variants are left out of the ROC
for the same reason, since only their truncated values were stored. For each
cell: the AUC, the false-positive rate on
fitting items at a matched detection rate of the misfitting items (50, 80 and
95 percent), and the partial AUC up to a false-positive rate of .05,
normalised to [0, 1]. Items within a dataset are not independent, so treat
small differences with caution.

```{r q4}
roc_summ <- function(pos, neg) {
  auc <- mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
  fpr_at <- function(tpr) mean(neg >= quantile(pos, 1 - tpr, type = 1))
  thr <- sort(unique(c(pos, neg)), decreasing = TRUE)
  fpr <- vapply(thr, function(t) mean(neg >= t), numeric(1))
  tpr <- vapply(thr, function(t) mean(pos >= t), numeric(1))
  sel <- fpr <= 0.05
  pauc <- if (sum(sel) > 1) sum(diff(fpr[sel]) * head(tpr[sel], -1)) / 0.05 else 0
  data.frame(auc = auc, fpr_tpr50 = fpr_at(.5), fpr_tpr80 = fpr_at(.8),
             fpr_tpr95 = fpr_at(.95), pauc05 = pauc)
}
stats_q4 <- c("crmsd_score", "abs_infit", "abs_gamma",
              if (INCLUDE_SX2) "sx2_score")

q4 <- res |>
  filter(cond != "null") |>
  group_by(cond, n) |>
  group_modify(function(d, key) {
    bind_rows(lapply(stats_q4, function(st) {
      cbind(stat = st, roc_summ(d[[st]][d$misfit], d[[st]][!d$misfit]))
    }))
  }) |>
  ungroup()

q4 |>
  select(cond, n, stat, auc, fpr_tpr80, fpr_tpr95, pauc05) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  pivot_wider(names_from = stat, values_from = c(auc, fpr_tpr80, fpr_tpr95, pauc05))
```

```{r q4-fig, fig.height = 6}
q4 |>
  ggplot(aes(n, fpr_tpr80, colour = stat, group = stat)) +
  geom_line() + geom_point() +
  facet_wrap(~cond, scales = "free_y") +
  labs(x = "Sample size",
       y = "False-positive rate at 80% detection", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Q5. Targeting

Values for each misfitting item at 0, -1 and -2 logits, in the three-item
conditions, at n = 2,000 and n = 20,000.

```{r q5}
res |>
  filter(misfit, k == 3, n %in% c("2000", "20000")) |>
  group_by(cond, n, target) |>
  summarise(crmsd_bc = mean(crmsd_bc), infit = mean(infit),
            gamma_diff = mean(gamma_diff),
            sx2_rmsea = if (INCLUDE_SX2) mean(sx2_rmsea) else NA,
            .groups = "drop") |>
  arrange(cond, n, factor(target, levels = c("0", "-1", "-2"))) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Run time

```{r timing}
timing |> group_by(n) |> summarise(minutes = sum(minutes, na.rm = TRUE))
```

## Interpretation

Written 2026-10-05 after the run. 14,400 datasets, none failed, 9.9 minutes on
10 cores. The CML mean deviation was at most 2e-4 and the infit identity held
to 3e-13 in every dataset.

**1. The bias correction works, with the expected limits.** Under the null,
raw cRMSD falls from .068 at n = 500 to .011 at n = 20,000. The
bias-corrected value is 0 for about 64 percent of items at every n, and its
mean (.013 to .002) is what truncation at zero leaves. The mean MSD is 0.92 of
the analytic bias $B_i$ at every n, the one degree of freedom taken by the
item location. Under misfit, bias-corrected cRMSD is within .003 of its
n = 20,000 value from n = 1,000 when the population value is .06 or more. For
smaller values it is biased downward at small n (.027 at n = 500 against .037
for the single LPE item), as Robitzsch (2022) found for the marginal RMSD.

**2. Infit MSQ and the gamma difference are stable effect sizes at every n
(P2 holds for infit, and is wrong about the gamma difference).** Their means
for the misfitting items move by at most .01 between n = 500 and n = 20,000
in any condition. Both are differences or linear contrasts, not squares, so neither
carries the positive bias of an RMSD. RMSEA drifts down by 9 to 16 percent
over the same range.

**3. Under second-dimension misfit, cRMSD and infit carry the same
information.** For every item whose infit departs from 1 by more than .05,
cRMSD is 0.34 to 0.41 times |infit − 1|. Their leakage shares are about the
same. The share of the misfitting items' value that shows up on fitting items
is 18 to 19 percent for both with three misfitting items at rho = .15 and .50,
and 5 to 6 percent with one misfitting item. Their fall-off with targeting is
also the same: the item at −2 logits keeps about 0.62 of the value of the
item at 0 on both scales (P4 does not hold). The gamma difference keeps 0.83 to
0.92, so it is the least sensitive to targeting. RMSEA keeps 0.66 to 0.73.

**4. Infit and the gamma difference rank items better than cRMSD at small n
(P1 holds).** Where the ranking is not perfect (rho = .75 with three
misfitting items, n up to 2,000, and rho = .75 with one at n = 500), |infit −
1| and |gamma difference| have the highest AUC and partial AUC, S-X2 is next
and cRMSD last. At n = 2,000 and rho = .75 with three misfitting items, the
false-positive rate at 95 percent detection is .047 for infit, .025 for the
gamma difference, .161 for S-X2 and .301 for cRMSD.

**5. As effect sizes, all four separate misfitting items from cascade-affected
fitting items (P6 does not hold).** In the saturated cells (rho = .15 and .50)
every statistic reaches AUC 1. At n = 20,000 the smallest misfitting item
exceeds the largest fitting item by a factor of 3.1 to 3.8 for cRMSD, 3.4 to
3.5 for |infit − 1| and 3.9 to 4.3 for the gamma difference, with three
misfitting items. The leakage is real and does not shrink with n. At
rho = .15, fitting items have mean infit 0.93 at every n, which is why an
infit *test* flags them as n grows. But it stays well below the misfitting
items. So S-X2's resistance to the cascade in `sx2_comparison_v2.qmd` belongs
to testing, not to ranking. An omnibus test spends its power over many df and
does not reach significance for the small slope deviations of the fitting
items. As an effect size, infit ranks the items as well as S-X2 does.

**6. cRMSD earns its place on misfit that is not a slope (P5 holds).** With the
LPE items calibrated to infit 1, |infit − 1| has an AUC of .34 to .51 at every
n, which is no better than chance. cRMSD and S-X2 reach AUC .94 to .97 at
n = 2,000 with one LPE item and 1.00 from n = 5,000. S-X2 is a little better
than cRMSD at moderate n. The gamma difference picks up part of it with three
LPE items (AUC .62 to .77). That misfit is small, though: cRMSD .024 to .037.
A monotone curve whose slope matches the Rasch curve can only depart from it
by a little, so misfit that infit cannot see is likely to be small too.

**What this means for large samples.** The large-sample problem is a testing
problem. The effect sizes are stable, and under second-dimension misfit they
put the misfitting items well above the cascade-affected ones at every n. An
effect-size floor would therefore stop the cascade. The floor can sit on infit
MSQ, which needs no bias correction and ranks best, or on the gamma
difference, which is the least sensitive to targeting. Within a condition the
gap is wide. Across conditions it is narrower: the smallest misfitting item at
rho = .75 (|infit − 1| = .093) is close to the largest fitting item at
rho = .15 (.078). So one fixed floor for every scale would not work, and the
floor should come from impact (option D in `HANDOFF-large-sample-fit.md`).
cRMSD adds probability units and sensitivity to non-slope misfit. It does not
replace infit.

**Limits.** Dichotomous items only, 20 items, normal theta, and two misfit
mechanisms. The LPE condition was calibrated to infit 1 by design, so P5 tests
detection of non-slope misfit when it exists, not how often it occurs. Items
within a dataset are not independent, so small ROC differences are uncertain.
The pooled and information-weighted cRMSD variants were stored only in
truncated form and are not in the ROC.

## References

Köhler, C., Robitzsch, A., & Hartig, J. (2020). A bias-corrected RMSD item fit
statistic: An evaluation and comparison to alternatives. *Journal of
Educational and Behavioral Statistics, 45*(3), 251-273.

Robitzsch, A. (2022). Statistical properties of estimators of the RMSD item
fit statistic. *Foundations, 2*, 488-503.
