---
title: "Uniform-shift CML for DIF"
subtitle: "Validation of the prototype estimator behind the planned RMdifImpact()"
author: "Magnus Johansson"
date: today
params:
  run_recovery: false
---

## Purpose

The planned DIF impact function (`dev/dif-impact-design.md`, Part B) needs a
split model in which each DIF item has its own location in each group while
the remaining items anchor the scale. `psychotools::pcmodel()` cannot fit an
item-split dataset (see `dev/psychotools-na-bug/`), and a free split has two
further drawbacks. It adds one parameter per threshold instead of one per item,
and it fails when a category is empty in one group.

The prototype in `dev/dif_shift_cml.R` fits a **uniform-shift** model instead.
Item thresholds are shared, and group 2 has every threshold of a DIF item
shifted by $d_i$. With two groups the conditional likelihood is the sum of the
two groups' PCM conditional likelihoods, so it is built from category counts
and score distributions per group and the elementary symmetric functions from
`psychotools::elementary_symmetric_functions()`.

This note checks four things before the estimator is used anywhere else:

1. With no DIF items it reproduces `pcmodel()`.
2. The analytic gradient matches numerical derivatives.
3. With one DIF item and a large sample it agrees with a free split (patched
   `pcmodel()` and `eRm::PCM()`), and it fits a category that is empty in one
   group, where the free split cannot.
4. Recovery: bias, SE calibration and CI coverage of $\hat d$ by group size
   and item location. This part is a simulation and runs only with
   `params$run_recovery = TRUE`.

```{r}
#| label: setup
library(psychotools)
source("dif_shift_cml.R")
set.seed(20261003)

thr_items <- function(locs, steps) {
  out <- lapply(locs, function(l) l + steps)
  names(out) <- paste0("I", seq_along(locs))
  out
}
```

## 1. No DIF items: agreement with `pcmodel()`

With an empty DIF set the shift model is the pooled PCM, and the group split
does not change the conditional likelihood. The fit must match `pcmodel()`
to optimiser tolerance.

```{r}
#| label: v1
v1_cases <- list(
  dich_k10_n300 = list(n = 300, thr = thr_items(seq(-2, 2, length.out = 10), 0)),
  poly3_k8_n500 = list(n = 500, thr = thr_items(seq(-1.5, 1.5, length.out = 8), c(-0.5, 0.5))),
  poly5_k6_n200 = list(n = 200, thr = thr_items(seq(-1, 1, length.out = 6), c(-1.2, -0.4, 0.4, 1.2))),
  mixed_k7_n400 = list(n = 400, thr = c(
    thr_items(c(-1, 0, 1), 0),
    setNames(thr_items(c(-0.5, 0.5), c(-0.6, 0.6)), c("I4", "I5")),
    setNames(thr_items(c(-0.8, 0.8), c(-1, 0, 1)), c("I6", "I7"))
  ))
)
v1 <- do.call(rbind, lapply(names(v1_cases), function(nm) {
  cs <- v1_cases[[nm]]
  s <- sim_dif(cs$n / 2, cs$n / 2, cs$thr)
  f <- shift_cml(s$X, s$g, hessian = FALSE)
  p <- pcmodel(s$X, nullcats = "downcode")
  data.frame(
    case = nm,
    loglik_diff = f$loglik - as.numeric(logLik(p)),
    max_par_diff = max(abs(f$psi[-1] - coef(p))),
    convergence = f$convergence
  )
}))
v1
```

## 2. Gradient check

Central differences with step $10^{-6}$ at random parameter values, with two
DIF items, for dichotomous and polytomous data.

```{r}
#| label: v4
grad_check <- function(s, dif) {
  prep <- .shift_prep(s$X, s$g, dif)
  par <- rnorm(prep$P - 1L + length(prep$dif_items), 0, 0.5)
  h <- 1e-6
  num <- vapply(seq_along(par), function(i) {
    e <- replace(numeric(length(par)), i, h)
    (.shift_negll(par + e, prep) - .shift_negll(par - e, prep)) / (2 * h)
  }, numeric(1))
  ana <- .shift_grad(par, prep)
  c(max_abs = max(abs(num - ana)), max_rel = max(abs(num - ana) / pmax(1, abs(ana))))
}
s_d <- sim_dif(300, 300, v1_cases$dich_k10_n300$thr, dif = c(I3 = 0.5))
s_p <- sim_dif(300, 300, v1_cases$poly5_k6_n200$thr, dif = c(I2 = 0.5))
rbind(
  dich = grad_check(s_d, c("I3", "I7")),
  poly = grad_check(s_p, c("I2", "I5"))
)
```

## 3. Agreement with a free split, and the sparse case

### Large sample, uniform DIF

Eight items with three categories, $n$ = 2000 per group, group 2 0.5 SD
higher, $d = 0.8$ on I4. The free split fits I4 as two columns. Its mean
threshold difference should agree closely with $\hat d$, and its
log-likelihood can only be higher, since the shift model is nested in it.

```{r}
#| label: v2
thr8 <- thr_items(seq(-1.5, 1.5, length.out = 8), c(-0.5, 0.5))
s <- sim_dif(2000, 2000, thr8, dif = c(I4 = 0.8), mu2 = 0.5)
f_shift <- shift_cml(s$X, s$g, "I4")
sp <- split_items(s$X, s$g, "I4")
f_free <- pcmodel_fixed(sp)
tp <- threshpar(f_free)
e_free <- eRm::PCM(sp)
te <- eRm::thresholds(e_free)$threshtable[[1]]
lr <- 2 * (as.numeric(logLik(f_free)) - f_shift$loglik)
data.frame(
  estimate = c("shift d-hat", "free split, pcmodel (patched)", "free split, eRm"),
  d = c(f_shift$d, mean(tp$I4_g2 - tp$I4_g1), mean(te["I4_g2", -1] - te["I4_g1", -1])),
  se = c(f_shift$se_d, NA, NA),
  loglik = c(f_shift$loglik, as.numeric(logLik(f_free)), as.numeric(e_free$loglik))
)
c(LR_free_vs_shift = lr, df = 1, p = pchisq(lr, df = 1, lower.tail = FALSE))
```

The eRm log-likelihood is on a different constant from psychotools and is
shown for completeness only. The LR statistic compares the two psychotools
fits. Under a uniform-shift generating model it should be unremarkable.

### Large sample, threshold-specific DIF

The same design with only the second threshold of I4 shifted (by 0.8). Now
the shift model is wrong for I4, the free split is right, and the LR test
should detect it. This is the DSF case left for a later version, and this
check shows what the shift model averages over.

```{r}
#| label: v2b
thr_dsf <- thr8
s2 <- sim_dif(2000, 2000, thr8, mu2 = 0.5)
# Rebuild group 2 for I4 with only the upper threshold shifted.
g2 <- s2$g == "g2"
s2$X[g2, "I4"] <- sim_pcm(s2$theta[g2], list(I4 = thr8$I4 + c(0, 0.8)))
f_shift2 <- shift_cml(s2$X, s2$g, "I4")
f_free2 <- pcmodel_fixed(split_items(s2$X, s2$g, "I4"))
tp2 <- threshpar(f_free2)
lr2 <- 2 * (as.numeric(logLik(f_free2)) - f_shift2$loglik)
list(
  shift_d = f_shift2$d,
  free_threshold_diffs = tp2$I4_g2 - tp2$I4_g1,
  LR = c(stat = lr2, p = pchisq(lr2, df = 1, lower.tail = FALSE))
)
```

### A category empty in one group

A small focal group (n = 60) on a DIF item with four categories, where the top
category is never used in group 2. Here the emptiness is forced by recoding
group 2's top responses down one step, to make the demonstration
deterministic. The shift model should still return a finite $\hat d$ with an
SE, while the free split has no information on the top threshold in group 2.

```{r}
#| label: sparse
thr_sp <- c(thr_items(seq(-1, 1, length.out = 7), c(-0.5, 0.5)),
            list(I8 = 0.5 + c(-1, 0, 1)))
s3 <- sim_dif(300, 60, thr_sp, dif = c(I8 = 0.4), mu2 = 0)
g2 <- s3$g == "g2"
s3$X[g2 & s3$X[, "I8"] == 3, "I8"] <- 2L
table(group = s3$g, I8 = s3$X[, "I8"])

f_sp <- shift_cml(s3$X, s3$g, "I8")
c(d = f_sp$d, se = f_sp$se_d, convergence = f_sp$convergence)

sp3 <- split_items(s3$X, s3$g, "I8")
free_pc <- tryCatch(threshpar(pcmodel_fixed(sp3))[c("I8_g1", "I8_g2")],
                    error = function(e) conditionMessage(e), warning = function(w) conditionMessage(w))
free_erm <- tryCatch({
  te3 <- eRm::thresholds(eRm::PCM(sp3))$threshtable[[1]]
  te3[c("I8_g1", "I8_g2"), ]
}, error = function(e) conditionMessage(e))
list(pcmodel_free_split = free_pc, eRm_free_split = free_erm)
```

## 4. Recovery

```{r}
#| label: recovery-design
rec_thr <- thr_items(seq(-2, 2, length.out = 8), c(-0.6, 0.6))
rec_grid <- expand.grid(
  sizes = c("150/150", "400/400", "600/150", "100/50"),
  item = c("I5", "I8"), # I5 near the centre, I8 at +2 logits
  d = c(0, 0.5),
  stringsAsFactors = FALSE
)
rec_reps <- 500
rec_file <- "dif_shift_cml_recovery.rds"
```

The design crosses four group-size splits (balanced and unbalanced), a DIF
item near the centre of the person distribution (I5, location 0.29) or at
+2 logits (I8), and $d \in \{0, 0.5\}$. Group 1 has θ ~ N(0, 1), group 2
θ ~ N(0.5, 1). Each of the `r nrow(rec_grid)` cells has `r rec_reps`
replications. Outcomes: bias and empirical SD of $\hat d$, mean model SE and
its ratio to the empirical SD, 95% Wald coverage, the share of replications
where a category of the DIF item is empty in group 2, and failures (no
convergence or no finite SE).

```{r}
#| label: recovery-run
#| eval: !expr params$run_recovery
one_rep <- function(cell, seed) {
  set.seed(seed)
  n <- as.integer(strsplit(cell$sizes, "/")[[1]])
  dif <- setNames(cell$d, cell$item)
  s <- sim_dif(n[1], n[2], rec_thr, dif = dif, mu2 = 0.5)
  empty <- any(tabulate(s$X[s$g == "g2", cell$item] + 1L, nbins = 3L) == 0L)
  f <- tryCatch(shift_cml(s$X, s$g, cell$item), error = function(e) NULL)
  if (is.null(f)) return(c(d_hat = NA, se = NA, ok = 0, empty = empty))
  c(d_hat = unname(f$d), se = unname(f$se_d),
    ok = as.numeric(f$convergence == 0 && is.finite(f$se_d)), empty = empty)
}
t0 <- Sys.time()
rec <- do.call(rbind, lapply(seq_len(nrow(rec_grid)), function(i) {
  cell <- rec_grid[i, ]
  res <- parallel::mclapply(seq_len(rec_reps), function(r) one_rep(cell, 1e5 * i + r),
                            mc.cores = 10L)
  data.frame(cell, do.call(rbind, res))
}))
attr(rec, "runtime") <- Sys.time() - t0
saveRDS(rec, rec_file)
```

```{r}
#| label: recovery-summary
if (file.exists(rec_file)) {
  rec <- readRDS(rec_file)
  agg <- do.call(rbind, lapply(split(rec, rec[c("sizes", "item", "d")], drop = TRUE), function(x) {
    y <- x[x$ok == 1, ]
    data.frame(
      sizes = x$sizes[1], item = x$item[1], d = x$d[1],
      bias = mean(y$d_hat - y$d),
      emp_sd = sd(y$d_hat),
      mean_se = mean(y$se),
      se_ratio = mean(y$se) / sd(y$d_hat),
      coverage = mean(abs(y$d_hat - y$d) <= 1.96 * y$se),
      empty_cat = mean(x$empty),
      failed = mean(x$ok != 1)
    )
  }))
  agg <- agg[order(agg$item, agg$d, agg$sizes), ]
  rownames(agg) <- NULL
  num <- c("bias", "emp_sd", "mean_se", "se_ratio", "coverage", "empty_cat", "failed")
  agg[num] <- lapply(agg[num], round, 3)
  names(agg)[names(agg) == "empty_cat"] <- "empty"
  agg
} else {
  cat("Recovery not run. Render with -P run_recovery:true to run it.\n")
}
```

## Findings

Recovery run 2026-10-03, 16 cells × 500 replications, 13 seconds on 10 cores.
The Monte Carlo SE of a coverage rate near .95 is about .010, and of the bias
about SD/22.

1. **The estimator reproduces `pcmodel()` and its gradient is right.** With no
   DIF items the largest parameter difference is below $10^{-6}$ in all four
   cases, and the analytic gradient agrees with central differences to below
   $10^{-6}$.
2. **It agrees with a free split when DIF is uniform.** At 2000 per group,
   $\hat d$ = 0.746 against 0.745 from both the patched `pcmodel()` and eRm
   free splits, and the LR test of the shift model against the free split is
   unremarkable. When only one threshold carries the DIF, the LR test detects
   it, so the shift model's adequacy can be checked against a free split
   whenever the free split is estimable.
3. **It handles a category that is empty in one group.** It returns a finite
   $\hat d$ with an SE, while eRm returns `NA` for that threshold and the
   patched `pcmodel()` drops it without notice. In the recovery run a
   category of the off-target item was empty in the focal group in 12.4% of
   replications at 100/50 with $d$ = 0.5, with no failures.
4. **Recovery is good down to 100/50.** No replication failed. Bias is within
   0.012 of zero for the central item. For the off-target item it reaches
   0.050 at 100/50 and 0.024 at 600/150 with $d$ = 0.5, about 2.5 Monte Carlo
   SEs, a small upward bias where the focal group is small and the item is
   far from it. Model SEs track the empirical SD (ratio 0.93 to 1.10), and
   Wald coverage is .928 to .968, within about 2.3 Monte Carlo SEs of .95
   everywhere.

The prototype is fit for use in the impact study. Two things to carry
forward: the small-sample bias off target is the same direction as the
noise-floor question (study item 3), and the SEs are good enough that a Wald
interval for $\hat d$ can be reported directly.
