---
title: "Partial gamma DIF null with unequal groups"
subtitle: "Does random group assignment in RMdifGammaCutoff() hold the error rate when groups differ in θ?"
author: "Magnus Johansson"
date: today
---

## Purpose

`RMdifGammaCutoff()` builds its null by resampling WLE estimates from the
pooled sample and assigning simulated respondents to groups **at random**,
with the observed proportions. The simulated groups therefore always share
one θ distribution. Real groups often differ in mean θ, which is the
situation where DIF analysis matters most.

Under the Rasch model the item response is independent of group given the
total score, so the null hypothesis holds whatever the groups' θ
distributions. But the null **distribution** of partial gamma depends on how
each group is spread across score strata. With a large group difference, the
extreme strata hold mostly one group and contribute few comparable pairs. A
null built from equal groups may then have the wrong spread.
`dev/dif-impact-design.md`, Part A, question 6, leaves the choice of null to
this study.

## Four nulls

1. **random**: the current method. Pooled WLE pool, groups assigned at random
   with multinomial sizes, CML thresholds fixed, iterations with an empty
   category (polytomous) or fewer than 8 positive responses (dichotomous)
   dropped, as in `run_single_partgam_sim()`.
2. **groupwise**: each group resampled from its own WLE estimates under the
   pooled CML thresholds, group sizes fixed at the observed sizes, the same
   validity check.
3. **conditional**: every respondent keeps their observed group and total
   score, and a new response pattern is drawn from the Rasch conditional
   distribution given that score (`.sim_cond_dataset()`, the package's
   conditional DGP).
4. **permutation**: group labels permuted within strata of the total score.
   Given the score, the Rasch model makes response patterns exchangeable
   across groups, so this null is exact and needs no item parameters. The
   same permutation is applied to all items, which keeps their dependence
   for the Westfall-Young step-down.

Nulls 3 and 4 condition on the observed (group, score) pairs, so in theory
they are unaffected by the group difference. Null 2 reproduces the group
difference without conditioning. Null 1 ignores it.

## Design

- **Items.** `poly9`: 9 items with 4 categories, locations −1.5 to 1.5,
  thresholds at location ± 0.8 and the location. `dich15`: 15 dichotomous
  items, locations −2 to 2.
- **Persons.** Group 1 θ ~ N(0, 1.5²), group 2 θ ~ N(Δ · 1.5, 1.5²), with
  Δ ∈ {0, 0.5, 1} SD.
- **Group sizes.** 150/150 and 480/120.
- **No DIF.** Every flag is a false positive.
- **Replications.** R = 300 datasets per condition, B = 400 iterations per
  null (the approved new default).
- **Statistic.** The vectorised partial gamma (`dif_partgam_fast.R`),
  identical to `iarm::partgam_DIF()`. p-values from the package's
  `.bootstrap_pvalues()`, two-sided, as `RMdifGamma()` computes them.
- **Outcomes.**
  - Per-item marginal rejection at α = .05, pooled over items and datasets.
  - Family-wise rate: share of datasets with any Westfall-Young adjusted
    p < .05.
  - Spread: each null's SD of gamma per item, divided by the true sampling
    SD of gamma under the generating model (2000 datasets per condition),
    median over datasets and averaged over items. A ratio below 1 means the
    null is too narrow and the test anti-conservative.

The Monte Carlo SE of a family-wise rate near .05 at R = 300 is about .013.
The pooled per-item rate rests on 300 × 9 or 300 × 15 tests, but the items
within a dataset are dependent, so its effective precision is lower than
that count suggests.

```{r}
#| label: setup
suppressMessages(devtools::load_all("..", quiet = TRUE))
source("dif_partgam_fast.R")
library(parallel)

out_dir <- "dif_gamma_null_groups"
dir.create(out_dir, showWarnings = FALSE)
R_reps <- 300L
B_iter <- 400L
n_true <- 2000L
cores <- 10L

item_sets <- list(
  poly9 = lapply(seq(-1.5, 1.5, length.out = 9), function(l) l + c(-0.8, 0, 0.8)),
  dich15 = as.list(seq(-2, 2, length.out = 15))
)
for (nm in names(item_sets)) {
  names(item_sets[[nm]]) <- paste0("I", seq_along(item_sets[[nm]]))
}
grid <- expand.grid(
  items = names(item_sets),
  delta = c(0, 0.5, 1),
  sizes = c("150/150", "480/120"),
  stringsAsFactors = FALSE
)
grid$id <- sprintf("%s_d%s_%s", grid$items, grid$delta, sub("/", "-", grid$sizes))
```

```{r}
#| label: generators
# Vectorised PCM simulation (distributionally identical to the package's
# sim_partial_score() and psychotools::rrm(), much faster).
sim_pcm_fast <- function(theta, thr_list) {
  n <- length(theta)
  vapply(thr_list, function(thr) {
    m <- length(thr)
    cum <- outer(theta, 0:m) - matrix(c(0, cumsum(thr)), n, m + 1L, byrow = TRUE)
    p <- exp(cum - apply(cum, 1, max))
    cp <- t(apply(p / rowSums(p), 1, cumsum))
    as.integer(rowSums(stats::runif(n) > cp[, -(m + 1L), drop = FALSE]))
  }, integer(n))
}

valid_sim <- function(X, n_cats) {
  if (all(n_cats == 2L)) return(all(colSums(X) >= 8L))
  all(vapply(seq_len(ncol(X)), function(j) all(tabulate(X[, j] + 1L, nbins = n_cats[j]) > 0L), logical(1)))
}

gen_observed <- function(cond) {
  n <- as.integer(strsplit(cond$sizes, "/")[[1]])
  thr <- item_sets[[cond$items]]
  theta <- c(stats::rnorm(n[1], 0, 1.5), stats::rnorm(n[2], cond$delta * 1.5, 1.5))
  g <- factor(rep(c("g1", "g2"), n))
  X <- sim_pcm_fast(theta, thr)
  colnames(X) <- names(thr)
  list(X = X, g = g)
}

gam <- function(X, g) {
  out <- partgam_dif_fast(X, g)
  stats::setNames(out$gamma, out$Item)
}
```

```{r}
#| label: nulls
# One dataset: observed gamma plus B draws from each null.
null_draws <- function(obs, B) {
  X <- obs$X
  g <- obs$g
  n <- nrow(X)
  n_cats <- apply(X, 2, max) + 1L
  pool <- .wle_theta_pool(X)
  thr <- pool$thr_list
  wle <- .estimate_thetas(X, thr, method = "WLE")$theta
  props <- as.numeric(table(g)) / n
  lev <- levels(g)
  wle_g <- lapply(lev, function(l) { w <- wle[g == l]; w[is.finite(w)] })
  n_g <- as.integer(table(g))
  cond_dl <- list(thr_list = thr, cond_groups = .cond_groups(X, thr),
                  sample_n = n, n_items = ncol(X))
  score <- rowSums(X)
  strata <- split(seq_len(n), score)

  draw <- function(method) {
    repeat {
      if (method == "random") {
        th <- sample(pool$thetas, n, replace = TRUE)
        Xs <- sim_pcm_fast(th, thr)
        gs <- factor(sample(lev, n, replace = TRUE, prob = props), levels = lev)
        if (valid_sim(Xs, n_cats)) break
      } else if (method == "groupwise") {
        th <- unlist(lapply(seq_along(lev), function(k) sample(wle_g[[k]], n_g[k], replace = TRUE)))
        Xs <- sim_pcm_fast(th, thr)
        gs <- factor(rep(lev, n_g), levels = lev)
        if (valid_sim(Xs, n_cats)) break
      } else if (method == "conditional") {
        Xs <- as.matrix(.sim_cond_dataset(cond_dl))
        gs <- g
        break
      } else {
        Xs <- X
        gs <- g
        for (s in strata) if (length(s) > 1L) gs[s] <- gs[s[sample.int(length(s))]]
        break
      }
    }
    colnames(Xs) <- colnames(X)
    gam(Xs, gs)
  }
  methods <- c("random", "groupwise", "conditional", "permutation")
  sims <- lapply(stats::setNames(methods, methods), function(mt) {
    do.call(rbind, lapply(seq_len(B), function(b) draw(mt)))
  })
  list(observed = gam(X, g), sims = sims)
}

summarise_dataset <- function(nd) {
  do.call(rbind, lapply(names(nd$sims), function(mt) {
    pv <- .bootstrap_pvalues(nd$observed, nd$sims[[mt]], correction = "fwer", tail = "two.sided")
    data.frame(
      method = mt,
      item = pv$name,
      p = pv$p,
      padj = pv$padj,
      null_sd = apply(nd$sims[[mt]], 2, stats::sd)[pv$name],
      stringsAsFactors = FALSE
    )
  }))
}
```

```{r}
#| label: run
RNGkind("L'Ecuyer-CMRG")
for (i in seq_len(nrow(grid))) {
  cond <- grid[i, ]
  f <- file.path(out_dir, paste0(cond$id, ".rds"))
  if (file.exists(f)) next
  t0 <- Sys.time()
  set.seed(1000L + i)
  true_g <- do.call(rbind, mclapply(seq_len(n_true), function(r) {
    o <- gen_observed(cond)
    gam(o$X, o$g)
  }, mc.cores = cores))
  true_sd <- apply(true_g, 2, stats::sd)
  res <- mclapply(seq_len(R_reps), function(r) {
    o <- gen_observed(cond)
    out <- tryCatch(summarise_dataset(null_draws(o, B_iter)), error = function(e) NULL)
    if (!is.null(out)) out$rep <- r
    out
  }, mc.cores = cores)
  res <- do.call(rbind, res)
  res$true_sd <- true_sd[res$item]
  res <- cbind(cond[rep(1, nrow(res)), c("items", "delta", "sizes")], res)
  attr(res, "runtime") <- Sys.time() - t0
  saveRDS(res, f)
}
```

## Results

```{r}
#| label: results
all_res <- do.call(rbind, lapply(file.path(out_dir, paste0(grid$id, ".rds")), readRDS))
key <- c("items", "sizes", "delta", "method")
summ <- do.call(rbind, lapply(split(all_res, all_res[key], drop = TRUE), function(x) {
  fw <- tapply(x$padj, x$rep, function(p) any(p < 0.05, na.rm = TRUE))
  sd_ratio <- tapply(x$null_sd / x$true_sd, x$item, stats::median, na.rm = TRUE)
  data.frame(
    items = x$items[1], sizes = x$sizes[1], delta = x$delta[1], method = x$method[1],
    per_item = round(mean(x$p < 0.05, na.rm = TRUE), 3),
    familywise = round(mean(fw), 3),
    sd_ratio = round(mean(sd_ratio), 3),
    reps = length(fw)
  )
}))
summ$method <- factor(summ$method, c("random", "groupwise", "conditional", "permutation"))
summ <- summ[order(summ$items, summ$sizes, summ$delta, summ$method), ]
rownames(summ) <- NULL
summ
```

```{r}
#| label: fig-rates
#| fig-cap: "Family-wise false-positive rate (Westfall-Young, α = .05) by group difference, null and design. Dashed line at .05, band ±2 Monte Carlo SE."
library(ggplot2)
mcse <- sqrt(0.05 * 0.95 / R_reps)
ggplot(summ, aes(delta, familywise, colour = method)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.05 - 2 * mcse, ymax = 0.05 + 2 * mcse,
           alpha = 0.15) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_line() +
  geom_point() +
  facet_grid(items ~ sizes) +
  labs(x = "Group difference in θ (SD units)", y = "Family-wise rate", colour = "Null") +
  theme_minimal()
```

```{r}
#| label: runtime
vapply(file.path(out_dir, paste0(grid$id, ".rds")),
       function(f) as.numeric(attr(readRDS(f), "runtime"), units = "mins"), numeric(1)) |>
  stats::setNames(grid$id) |>
  round(1)
```

## Findings

Run 2026-10-03, about two hours on 10 cores. The Monte Carlo SE of a
family-wise rate near .05 at R = 300 is .013, so rates between about .025 and
.075 are compatible with nominal.

1. **The current null is miscalibrated when the groups differ by 1 SD and
   are unequal in size.** With 480/120 and Δ = 1, the random-assignment null
   gives a family-wise rate of .137 (dichotomous) and .147 (polytomous), about
   three times nominal, and per-item rates of .075 and .089. With equal
   groups at Δ = 1 the excess is smaller (per item .063 and .059,
   family-wise .060 and .073). At Δ ≤ 0.5 it is within Monte Carlo error.
2. **The mechanism is a null that is too narrow, and it is item-specific.**
   At 480/120 and Δ = 1 the random null's SD is 0.75 to 0.77 of the true
   sampling SD for the easiest item and rises towards 1 for items near the
   centre (I1 is the easiest item in both sets). The likely reason is the
   group composition of the score strata: with a 1 SD difference, the strata
   that carry information about an easy item hold mostly one group, so fewer
   comparable pairs exist than in a null with equal groups. This explanation
   is not tested directly here. The cutoff band from
   `RMdifGammaCutoff()` is affected in the same way, since it is built from
   the same simulated values.
3. **All three alternatives hold the error rate in every condition.**
   Family-wise rates are .023 to .063 for groupwise, .030 to .067 for
   conditional and .030 to .067 for permutation, and per-item rates .040 to
   .058. Their null SDs are within 2% of the true SD on average and
   never below 0.958 for any item.
4. **Choosing among them.** The three are equally well calibrated here, so
   the choice rests on other grounds:
   - *Permutation* within score strata is exact under the Rasch model, needs
     no item parameters or person estimates, never produces an invalid
     simulated dataset, and is the cheapest by far (no data generation). It
     is a direct test of item ⊥ group | score, which is the hypothesis
     partial gamma measures. It also gives a no-DIF reference for the impact
     function (Part B), since it removes only the item-group association.
   - *Conditional* is the package's existing `dgp = "conditional"`, already
     the default in `RMitemRestscoreCutoff()`, so it would be the more
     consistent choice across functions.
   - *Groupwise* is the smallest change to the current code but is the only
     one of the three that depends on WLE pools and is not exact.

Recommendation for Part A question 6: replace random assignment. Permutation
as the default, with the conditional DGP as the alternative if consistency
with the restscore cutoff is preferred. Random assignment should not remain
an option, since it fails in a common design (a minority group with a
different mean).
