---
title: "Does it matter whether misfitting items share a dimension?"
subtitle: "Attenuation of item fit as a function of the correlation between misfit sources"
author: "Magnus Johansson"
date: today
---

## Question

When two items misfit, does each one's apparent misfit depend on **whether they
misfit for the same reason**? Specifically: two items loading on a single second
dimension, versus one item on a second dimension and one on a third.

Informal earlier work suggested little or no difference. This checks that, and
identifies the parameter that actually governs it.

The practical stake is the one-at-a-time removal advice in Johansson (2025):
if a second misfitting item attenuates the first item's fit statistic, each
departure is being measured against a reference (the conditioning score and the
CML parameters) that has already partly absorbed it.

## The structural constraint

This turns out to be the key point. Let the main dimension be `theta1` and let
the two misfitting items load on `theta2` and `theta3`, each correlated `rho`
with `theta1`. Write `c = cor(theta2, theta3)`.

The correlation matrix

```
[ 1    rho  rho ]
[ rho  1    c   ]
[ rho  c    1   ]
```

has determinant `1 + 2*rho^2*c - 2*rho^2 - c^2`, which is non-negative only for

$$c \ \ge\ 2\rho^2 - 1$$

So **two "separate" dimensions cannot be made independent** when both load
substantially on the main dimension. At `rho = 0.85` the minimum possible
correlation between them is **0.445**, and under conditional independence given
`theta1` it is `rho^2 = 0.72`. Only when the misfit is strong (low `rho`) does
genuine separation become available.

That predicts the informal finding directly: at realistic misfit levels,
"different dimensions" is nearly the same condition as "same dimension".

## Design

Rather than three discrete conditions, `c` is swept across its whole admissible
range, so the gradient is measured rather than inferred. The sweep is
parameterised as a fraction `t` of the available range:

$$c = c_{\min} + t\,(1 - c_{\min}), \qquad c_{\min} = \max(0,\ 2\rho^2 - 1)$$

`t = 1` is always "same dimension" (`c = 1`); `t = 0` is as separate as the
geometry permits. This keeps conditions comparable across `rho`.

- Arm: **phq9** — 9 items, CML parameters from the packaged real responses,
  latent SD estimated from the same data
- Misfitting items: **q5** (-0.15, on target) and **q8** (+1.40, off target),
  as in the other studies in this series
- `rho` in {0.85 (calibrated to real phq9 misfit), 0.50}
- `t` in {0, 0.25, 0.5, 0.75, 1}
- **Baseline**: only q5 misfits (k = 1), the reference against which attenuation
  is measured
- **Noise anchor**: both items misfit through independent random responding —
  a genuine `c = 0` condition reachable at any `rho`, calibrated to match the
  `rho = 0.85` baseline infit
- n in {250, 1000}; 300 replications

Only item fit is computed — no bootstrap — so this runs in minutes. Detection
consequences would need cutoffs and are left as a follow-up.

```{r setup}
suppressMessages({
  pkgload::load_all("..", quiet = TRUE)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

RHOS    <- c(0.85, 0.50)
TS      <- c(0, 0.25, 0.5, 0.75, 1)
NS      <- c(250, 1000)
R_REPS  <- 300
NOISE_P <- 0.15          # matches rho = 0.85 baseline infit (see mechanism study)
NCORES  <- 10
RES     <- "infit_misfit_source_correlation.rds"
```

```{r arm}
load("../data/phq9.rda")
PH_OBS <- as.matrix(phq9[, paste0("q", 1:9)])
PH_OBS <- PH_OBS[complete.cases(PH_OBS), ]
PH_THR  <- .fit_cml_thresholds(PH_OBS)
PH_LOCS <- vapply(PH_THR, mean, numeric(1))
.g <- seq(-6, 6, length.out = 81)
PH_SD <- .estimate_prior_sd(
  .grid_loglik(PH_OBS, .logp_tables(PH_THR, .g), .g), .g, 0)
ITEM_A <- 5L   # on target
ITEM_B <- 8L   # off target
MAXCAT <- 3L

c_min <- function(rho) max(0, 2 * rho^2 - 1)
c_of  <- function(rho, t) c_min(rho) + t * (1 - c_min(rho))

# admissible range and the conditional-independence value, per rho
data.frame(rho = RHOS) |>
  mutate(c_min = round(sapply(rho, c_min), 3),
         c_cond_indep = round(rho^2, 3),
         c_same_dim = 1)
```

```{r sim}
# cond: "base" (only q5 misfits) | "sweep" (both, cor = c) | "noise" (both, independent)
simdat <- function(n, cond, rho, t, seed) {
  set.seed(seed)
  if (cond == "noise") {
    th <- rnorm(n, 0, PH_SD)
    d <- sim_partial_score(PH_THR, th)
    for (it in c(ITEM_A, ITEM_B)) {
      fl <- runif(n) < NOISE_P
      if (any(fl)) d[fl, it] <- sample(0:MAXCAT, sum(fl), replace = TRUE)
    }
  } else if (cond == "base") {
    S <- PH_SD^2 * matrix(c(1, rho, rho, 1), 2, 2)
    th <- MASS::mvrnorm(n, c(0, 0), S)
    d <- sim_partial_score(PH_THR, th[, 1])
    d[, ITEM_A] <- sim_partial_score(PH_THR[ITEM_A], th[, 2])[, 1]
  } else {
    cc <- c_of(rho, t)
    S <- PH_SD^2 * matrix(c(1, rho, rho,
                            rho, 1, cc,
                            rho, cc, 1), 3, 3)
    th <- MASS::mvrnorm(n, c(0, 0, 0), S)
    d <- sim_partial_score(PH_THR, th[, 1])
    d[, ITEM_A] <- sim_partial_score(PH_THR[ITEM_A], th[, 2])[, 1]
    d[, ITEM_B] <- sim_partial_score(PH_THR[ITEM_B], th[, 3])[, 1]
  }
  d <- as.data.frame(d); names(d) <- paste0("q", 1:9); d
}

one <- function(g) {
  out <- try({
    # `t` is NA for the baseline and `rho`/`t` are NA for the noise condition;
    # substitute 0 so the seed arithmetic stays finite.
    rho_s <- if (is.na(g$rho)) 0 else g$rho
    t_s   <- if (is.na(g$t))   0 else g$t
    d <- simdat(g$n, g$cond, g$rho, g$t,
                seed = 4e8 + 1e6 * match(g$cond, c("base", "sweep", "noise")) +
                  1e5 * round(rho_s * 100) + 1e4 * round(t_s * 100) +
                  1e3 * g$n + g$rep)
    x <- RMitemInfit(d, output = "dataframe")
    data.frame(infit_A = x$Infit_MSQ[ITEM_A], infit_B = x$Infit_MSQ[ITEM_B])
  }, silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

grid <- rbind(
  expand.grid(rep = 1:R_REPS, n = NS, rho = RHOS, t = NA_real_,
              cond = "base", stringsAsFactors = FALSE),
  expand.grid(rep = 1:R_REPS, n = NS, rho = RHOS, t = TS,
              cond = "sweep", stringsAsFactors = FALSE),
  expand.grid(rep = 1:R_REPS, n = NS, rho = NA_real_, t = NA_real_,
              cond = "noise", stringsAsFactors = FALSE)
)[, c("cond", "rho", "t", "n", "rep")]

if (file.exists(RES)) {
  res <- readRDS(RES)
} else {
  t0 <- Sys.time()
  out <- mclapply(seq_len(nrow(grid)), function(i) one(grid[i, , drop = FALSE]),
                  mc.cores = NCORES)
  res <- list(grid = grid, out = out,
              elapsed_min = as.numeric(Sys.time() - t0, "mins"))
  saveRDS(res, RES)
}

D <- cbind(res$grid,
           do.call(rbind, lapply(res$out, function(x)
             if (is.null(x)) data.frame(infit_A = NA, infit_B = NA) else x)))
cat(nrow(D), "datasets |", round(res$elapsed_min, 1), "min | failed:",
    sum(is.na(D$infit_A)), "\n")
```

## Attenuation of the on-target item (q5)

q5 misfits in every condition. The baseline is q5 misfitting **alone**; every
other row adds a second misfitting item (q8) and differs only in how correlated
the two misfit sources are.

```{r attenuation}
base <- D |>
  filter(cond == "base") |>
  group_by(rho, n) |>
  summarise(baseline = mean(infit_A, na.rm = TRUE), .groups = "drop")

D |>
  filter(cond == "sweep") |>
  group_by(rho, t, n) |>
  summarise(q5 = mean(infit_A, na.rm = TRUE),
            q8 = mean(infit_B, na.rm = TRUE), .groups = "drop") |>
  left_join(base, by = c("rho", "n")) |>
  mutate(cor_sources = round(c_of(rho, t), 3),
         attenuation = round(q5 - baseline, 3),
         across(c(q5, q8, baseline), \(x) round(x, 3))) |>
  select(rho, n, t, cor_sources, baseline, q5, q8, attenuation) |>
  arrange(rho, n, t)
```

The `c = 0` anchor, from independent random responding:

```{r noise_anchor}
D |>
  filter(cond == "noise") |>
  group_by(n) |>
  summarise(q5 = round(mean(infit_A, na.rm = TRUE), 3),
            q8 = round(mean(infit_B, na.rm = TRUE), 3), .groups = "drop") |>
  mutate(note = "both items misfit independently (cor = 0)")
```

```{r fig, fig.width = 8, fig.height = 4.5}
plotd <- D |>
  filter(cond == "sweep") |>
  group_by(rho, t, n) |>
  summarise(q5 = mean(infit_A, na.rm = TRUE), .groups = "drop") |>
  mutate(cor_sources = c_of(rho, t))

ggplot(plotd, aes(cor_sources, q5, colour = factor(rho))) +
  geom_line() + geom_point() +
  geom_hline(data = base, aes(yintercept = baseline, colour = factor(rho)),
             linetype = 2) +
  facet_wrap(~n, labeller = label_both) +
  labs(x = "Correlation between the two misfit sources",
       y = "Mean infit of q5",
       colour = "rho",
       caption = "Dashed line = q5 misfitting alone (baseline)") +
  theme_minimal(base_size = 11)
```

## Detection follow-up

Does the attenuation above translate into a detection difference? This section
adds bootstrap cutoffs to the same conditions.

**Sample sizes are n = 150 and 250, not 1000.** At n = 1000 detection of a
`rho = 0.85` misfitting item is already ~92 percent, so there is no room for an
0.08 shift in infit to show; the informative range is where detection sits
around 25-50 percent. The attenuation itself was shown above to be
n-invariant, so nothing is lost by measuring its consequences at the sample
sizes where they are visible.

B = 2000 (converged), 120 replications, HDCI `w = .999` and the FWER p-value.

```{r detsetup}
DET_NS   <- c(150, 250)
DET_B    <- 2000
DET_REPS <- 120
ALPHA    <- 0.05
DET_RES  <- "infit_misfit_source_correlation_detection.rds"
FITTING  <- setdiff(1:9, c(ITEM_A, ITEM_B))
```

```{r detrun}
library(ggdist)

bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}

one_det <- function(g) {
  out <- try({
    rho_s <- if (is.na(g$rho)) 0 else g$rho
    t_s   <- if (is.na(g$t))   0 else g$t
    d <- simdat(g$n, g$cond, g$rho, g$t,
                seed = 6e8 + 1e6 * match(g$cond, c("base", "sweep", "noise")) +
                  1e5 * round(rho_s * 100) + 1e4 * round(t_s * 100) +
                  1e3 * g$n + g$rep)
    items <- names(d)
    obs <- RMitemInfit(d, output = "dataframe")
    msq <- obs$Infit_MSQ[match(items, obs$Item)]
    co <- RMitemInfitCutoff(d, iterations = DET_B, parallel = FALSE,
                            seed = g$rep * 100 + 7, verbose = FALSE)
    b <- bounds_at(co$results, 0.999, items)
    hdci <- as.integer(msq > b[2, ] | msq < b[1, ])
    fw <- suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                  alpha = ALPHA, output = "dataframe"))
    fwer <- as.integer(fw$padj_infit[match(items, fw$Item)] < ALPHA)
    data.frame(item = seq_along(items), hdci = hdci, fwer = fwer)
  }, silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

det_grid <- rbind(
  expand.grid(rep = 1:DET_REPS, n = DET_NS, rho = RHOS, t = NA_real_,
              cond = "base", stringsAsFactors = FALSE),
  expand.grid(rep = 1:DET_REPS, n = DET_NS, rho = RHOS, t = TS,
              cond = "sweep", stringsAsFactors = FALSE),
  expand.grid(rep = 1:DET_REPS, n = DET_NS, rho = NA_real_, t = NA_real_,
              cond = "noise", stringsAsFactors = FALSE)
)[, c("cond", "rho", "t", "n", "rep")]
det_grid <- det_grid[order(det_grid$n), ]
rownames(det_grid) <- NULL
cat("detection cells:", nrow(det_grid), "\n")

run_resumable <- function(res_file, grid, keys, fn, ncores = NCORES,
                          chunk_size = 100) {
  key_of <- function(g) if (nrow(g) == 0L) character(0) else
    do.call(paste, c(unname(as.list(g[keys])), sep = "|"))
  acc <- if (file.exists(res_file)) readRDS(res_file) else
    list(res = list(), grid = grid[0, , drop = FALSE], elapsed_min = 0)
  todo <- which(!(key_of(grid) %in% key_of(acc$grid)))
  if (length(todo) == 0L) { message("cached: nothing to run"); return(acc) }
  message("running ", length(todo), " new cells")
  for (s in seq(1, length(todo), by = chunk_size)) {
    idx <- todo[s:min(s + chunk_size - 1L, length(todo))]
    t0 <- Sys.time()
    o <- mclapply(idx, function(i) fn(grid[i, , drop = FALSE]), mc.cores = ncores)
    acc$res <- c(acc$res, o)
    acc$grid <- rbind(acc$grid, grid[idx, , drop = FALSE])
    acc$elapsed_min <- acc$elapsed_min + as.numeric(Sys.time() - t0, "mins")
    saveRDS(acc, res_file)
    message("  checkpoint ", nrow(acc$grid), " / ", nrow(grid))
  }
  acc
}

ds <- run_resumable(DET_RES, det_grid, c("cond", "rho", "t", "n", "rep"), one_det)

DET <- do.call(rbind, lapply(seq_len(nrow(ds$grid)), function(i) {
  r <- ds$res[[i]]
  if (is.null(r)) return(NULL)
  cbind(ds$grid[i, c("cond", "rho", "t", "n", "rep")], r)
}))
cat(nrow(ds$grid), "cells |", round(ds$elapsed_min, 1), "min | failed:",
    sum(sapply(ds$res, is.null)), "\n")
```

### Detection of q5, across the correlation sweep

q5 misfits in every condition, so this is a clean within-item comparison. The
baseline row is q5 misfitting alone.

```{r detq5}
det_base <- DET |>
  filter(cond == "base", item == ITEM_A) |>
  group_by(rho, n) |>
  summarise(base_hdci = 100 * mean(hdci), base_fwer = 100 * mean(fwer),
            .groups = "drop")

DET |>
  filter(cond == "sweep", item == ITEM_A) |>
  group_by(rho, t, n) |>
  summarise(hdci = 100 * mean(hdci), fwer = 100 * mean(fwer), .groups = "drop") |>
  left_join(det_base, by = c("rho", "n")) |>
  mutate(cor_src = round(c_of(rho, t), 3),
         d_hdci = round(hdci - base_hdci, 1),
         d_fwer = round(fwer - base_fwer, 1),
         across(c(hdci, fwer, base_hdci, base_fwer), \(x) round(x, 1))) |>
  select(rho, n, cor_src, base_hdci, hdci, d_hdci, base_fwer, fwer, d_fwer) |>
  arrange(rho, n, cor_src)
```

### False positives on the seven fitting items

```{r detfp}
DET |>
  filter(cond == "sweep", item %in% FITTING) |>
  group_by(rho, t, n) |>
  summarise(hdci = round(100 * mean(hdci), 1),
            fwer = round(100 * mean(fwer), 1), .groups = "drop") |>
  mutate(cor_src = round(c_of(rho, t), 3)) |>
  select(rho, n, cor_src, hdci, fwer) |>
  arrange(rho, n, cor_src)
```

### The c = 0 anchor

```{r detnoise}
DET |>
  filter(cond == "noise") |>
  mutate(grp = case_when(item == ITEM_A ~ "q5 (misfit)",
                         item == ITEM_B ~ "q8 (misfit)",
                         TRUE ~ "fitting items")) |>
  group_by(n, grp) |>
  summarise(hdci = round(100 * mean(hdci), 1),
            fwer = round(100 * mean(fwer), 1), .groups = "drop")
```

```{r detfig, fig.width = 8, fig.height = 4.5}
DET |>
  filter(cond == "sweep", item == ITEM_A) |>
  group_by(rho, t, n) |>
  summarise(fwer = 100 * mean(fwer), .groups = "drop") |>
  mutate(cor_src = c_of(rho, t)) |>
  ggplot(aes(cor_src, fwer, colour = factor(rho))) +
  geom_line() + geom_point() +
  geom_hline(data = det_base, aes(yintercept = base_fwer, colour = factor(rho)),
             linetype = 2) +
  facet_wrap(~n, labeller = label_both) +
  labs(x = "Correlation between the two misfit sources",
       y = "Detection of q5 (%), FWER",
       colour = "rho",
       caption = "Dashed line = q5 misfitting alone") +
  theme_minimal(base_size = 11)
```

## Notes for the write-up

Two points to carry into the paper:

1. **Attenuation is graded by the correlation between misfit sources**, not by
   whether they are nominally "the same dimension". Conditional infit conditions
   on the total score, so a second misfitting item makes the first look less
   deviant only to the extent that its deviation is correlated with the first
   item's — which is why independent random responding produces no attenuation
   at all.

2. **The admissible range of that correlation is bounded below by
   `2*rho^2 - 1`.** At realistic misfit levels nearly the whole range is
   compressed toward 1, so same-dimension and separate-dimension designs are
   close to indistinguishable. This is a property of the geometry, not of the
   fit statistic, and it explains why earlier informal comparisons found little
   difference.

Detection consequences are not measured here — that requires bootstrap cutoffs
and is a separate run.

## References

Johansson, M. (2025). Detecting Item Misfit in Rasch Models.
*Educational Methods & Psychometrics, 3*(18).
