---
title: "Decision rules on a short scale: five polytomous items"
subtitle: "Does the width-to-error-rate relation hold below nine items?"
author: "Magnus Johansson"
date: today
---

## Scope

The width and iterations studies verified the Sidak relation `1 - w^k` at
k = 9, 15 and 20 items, in both item formats. Nothing below nine items was
tested, and the manuscript says so in its Limitations: each item is a larger
share of the total score on a short scale, so the dependence among item fit
statistics is stronger, and the Sidak bound could be looser there than it is at
k = 20.

This study fills that gap at **k = 5** and reruns the decision-rule comparison
under otherwise identical conditions.

Two things make short scales worth a separate look.

**The Sidak prediction is much lower.** At k = 5 the predicted familywise error
rates are 41.0, 22.6, 4.9, 2.5 and 0.5 percent for the five widths, against
18.2 percent at w = .99 for k = 20. If positive dependence among item
statistics makes the bound conservative, the effect should be most visible
where the predicted rate is small.

**The Benjamini-Hochberg floor stops binding.** Its reachability floor is
`m/alpha - 1`, which is 399 at 20 items but only **99 at 5 items**. So BH can
flag a standout item at every value of B tested here, which is not true in any
other study in this series. This is the one setting where BH competes on equal
terms.

### Misfit calibration

Misfit severity is anchored on the infit scale, as everywhere else. `rho` is
not portable across item counts any more than across item formats: at 15
polytomous items `rho` = .85 injects infit around 1.32, while at 5 items it
injects only 1.20. A short scale needs stronger misfit for the same infit,
because the misfitting item is a larger part of the total score it is
conditioned on and therefore attenuates its own statistic more.

A scan of 80 replications at n = 600 gives:

| rho | injected infit, k = 1 |
|---|---|
| 0.75 | 1.31 |
| 0.85 | 1.20 |
| 0.90 | 1.14 |
| 0.95 | 1.06 |

`rho` = **0.75** is used, which lands inside the 1.23 to 1.32 band that the
real phq9 items reach. The study reports the achieved value so the anchoring
is verifiable from the run.

### Design

Five polytomous items with four response categories, locations evenly spaced
between -2 and 2 logits, thresholds spread one logit around each location, and
person locations with a standard deviation of 1.5. This is the `sim15` recipe
at k = 5. The misfitting item is the one at 0 logits, on target relative to the
person mean. Sample sizes are 250 and 600, iterations 100, 400 and 2000, and
all five interval widths and all four decision rules are computed from the same
bootstrap within each replication, so every comparison is paired.

```{r setup}
suppressMessages({
  pkgload::load_all("..", quiet = TRUE)   # internal helpers
  library(ggdist)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

WIDTHS <- c(0.90, 0.95, 0.99, 0.995, 0.999)
NS     <- c(250, 600)
BS     <- c(100, 400, 2000)
RHO    <- 0.75          # injects infit ~1.31 at k = 5, see the scan above
KS     <- c(0L, 1L)     # 0 = null (error rates), 1 = one misfitting item
REPS   <- 250
NCORES <- 10
ALPHA  <- 0.05
CHUNK  <- 50
RES    <- "infit_short_scale.rds"

LOCS <- seq(-2, 2, length.out = 5)
THR  <- lapply(LOCS, function(L) L + seq(-1, 1, length.out = 3))
MIS  <- which.min(abs(LOCS))    # item 3, at 0 logits
SD   <- 1.5
NM   <- paste0("V", seq_along(LOCS))
```

```{r runner}
run_resumable <- function(res_file, grid, keys, fn,
                          ncores = NCORES, chunk_size = CHUNK,
                          extra = list()) {
  key_of <- function(g) {
    if (nrow(g) == 0L) return(character(0))
    do.call(paste, c(unname(as.list(g[keys])), sep = "|"))
  }
  acc <- if (file.exists(res_file)) readRDS(res_file) else
    c(list(res = list(), grid = grid[0, , drop = FALSE], elapsed_min = 0), extra)
  todo <- which(!(key_of(grid) %in% key_of(acc$grid)))
  if (length(todo) == 0L) {
    message("cached: ", nrow(acc$grid), " cells, nothing to run"); return(acc)
  }
  message("running ", length(todo), " new cells (", nrow(acc$grid), " cached)")
  for (s in seq(1, length(todo), by = chunk_size)) {
    idx <- todo[s:min(s + chunk_size - 1L, length(todo))]
    t0 <- Sys.time()
    out <- mclapply(idx, function(i) fn(grid[i, , drop = FALSE]), mc.cores = ncores)
    acc$res <- c(acc$res, out)
    acc$grid <- rbind(acc$grid, grid[idx, , drop = FALSE])
    acc$elapsed_min <- acc$elapsed_min + as.numeric(Sys.time() - t0, "mins")
    acc <- utils::modifyList(acc, extra)
    saveRDS(acc, res_file)
    message("  checkpoint ", nrow(acc$grid), " / ", nrow(grid),
            "  (", round(acc$elapsed_min, 1), " min)")
  }
  acc
}

bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}
flag_code <- function(msq, lo, hi) ifelse(msq > hi, 2L, ifelse(msq < lo, 1L, 0L))
```

```{r simfun}
sim_short <- function(n, k, rho, seed) {
  set.seed(seed)
  S <- SD^2 * matrix(c(1, rho, rho, 1), 2, 2)
  th <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = S)
  d <- sim_partial_score(THR, th[, 1])
  if (k > 0L) d[, MIS] <- sim_partial_score(THR[MIS], th[, 2])[, 1]
  d <- as.data.frame(d)
  names(d) <- NM
  d
}
```

### Run

```{r run}
one_rep <- function(g) {
  out <- try({
    k <- g$k; n <- g$n; rep <- g$rep
    d <- sim_short(n, k, RHO, seed = 6e8 + 1e6 * k + 1e3 * n + rep)
    items <- names(d)
    obs <- RMitemInfit(d, output = "dataframe")
    msq <- obs$Infit_MSQ[match(items, obs$Item)]
    hd <- list(); pv <- list()
    for (j in seq_along(BS)) {
      co <- try(RMitemInfitCutoff(d, iterations = BS[j], parallel = FALSE,
                                  seed = rep * 100 + j, verbose = FALSE),
                silent = TRUE)
      if (inherits(co, "try-error")) next
      hd[[length(hd) + 1L]] <- do.call(rbind, lapply(WIDTHS, function(w) {
        b <- bounds_at(co$results, w, items)
        data.frame(B = BS[j], w = w, item = seq_along(items),
                   code = flag_code(msq, b[1, ], b[2, ]))
      }))
      fw <- try(suppressWarnings(
        RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                    alpha = ALPHA, output = "dataframe")), silent = TRUE)
      bh <- try(suppressWarnings(
        RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fdr_bh",
                    alpha = ALPHA, output = "dataframe")), silent = TRUE)
      if (!inherits(fw, "try-error") && !inherits(bh, "try-error")) {
        i1 <- match(items, fw$Item); i2 <- match(items, bh$Item)
        pv[[length(pv) + 1L]] <- data.frame(
          B = BS[j], item = seq_along(items),
          msq = msq,
          p_marg = fw$p_infit[i1], padj_fwer = fw$padj_infit[i1],
          padj_bh = bh$padj_infit[i2])
      }
    }
    list(hdci = do.call(rbind, hd), pval = do.call(rbind, pv))
  }, silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

grid <- expand.grid(rep = 1:REPS, n = NS, k = KS)[, c("k", "n", "rep")]
grid <- grid[order(grid$k, grid$n), ]
rownames(grid) <- NULL
cat("cells to run:", nrow(grid), "\n")

s <- run_resumable(RES, grid, c("k", "n", "rep"), one_rep,
                   extra = list(locs = LOCS, mis = MIS, rho = RHO,
                                widths = WIDTHS, bs = BS, alpha = ALPHA))

bind_part <- function(part) {
  do.call(rbind, lapply(seq_len(nrow(s$grid)), function(i) {
    r <- s$res[[i]][[part]]
    if (is.null(r)) return(NULL)
    g <- s$grid[i, c("k", "n", "rep"), drop = FALSE]
    rownames(g) <- NULL
    cbind(g, r)
  }))
}
H <- bind_part("hdci") |>
  mutate(is_mis = k > 0 & item == MIS, flag = as.integer(code > 0),
         w = sprintf("%.3f", w))
P <- bind_part("pval") |> mutate(is_mis = k > 0 & item == MIS)
cat(nrow(s$grid), "cells complete |", round(s$elapsed_min, 1), "min |",
    "failed:", sum(vapply(s$res, is.null, logical(1))), "\n")
```

## Achieved misfit

```{r calib}
P |>
  filter(is_mis, B == 2000) |>
  summarise(infit = round(mean(msq), 3), sd = round(sd(msq), 3), .by = n)
```

## The Sidak relation at k = 5

Predicted familywise error rate, `1 - w^5`, against the observed rate under the
null. The interval rule flags an item when its observed infit falls outside the
band, so under the null any flag is a false positive.

```{r sidak}
pred <- data.frame(w = sprintf("%.3f", WIDTHS),
                   predicted = round(100 * (1 - WIDTHS^5), 1))
H |>
  filter(k == 0) |>
  summarise(any = as.integer(sum(flag) > 0), .by = c(w, B, n, rep)) |>
  summarise(observed = round(100 * mean(any), 1), .by = c(w, B, n)) |>
  pivot_wider(names_from = n, values_from = observed, names_prefix = "n = ") |>
  left_join(pred, by = "w") |>
  arrange(w, B) |>
  relocate(predicted, .after = w)
```

## Error rates of the p-value rules under the null

A replication is dropped when its bootstrap yields no usable p-values, which
happens when the simulated statistics have no variance to studentise by. This
occurred once in 1000 cells, at `n` = 250 and `B` = 400, and `dropped` counts
it so the loss is visible rather than silently absorbed.

```{r null-rules}
P |>
  filter(k == 0) |>
  summarise(usable = !all(is.na(p_marg)),
            marginal = as.integer(sum(p_marg < ALPHA, na.rm = TRUE) > 0),
            wy = as.integer(sum(padj_fwer < ALPHA, na.rm = TRUE) > 0),
            bh = as.integer(sum(padj_bh < ALPHA, na.rm = TRUE) > 0),
            .by = c(B, n, rep)) |>
  summarise(dropped = sum(!usable),
            across(c(marginal, wy, bh),
                   \(x) round(100 * mean(x[usable]), 1)),
            .by = c(B, n)) |>
  arrange(n, B)
```

## Detection and false positives, all four rules

```{r rules}
bind_rows(
  H |>
    filter(k > 0, w == "0.999") |>
    summarise(rule = "Interval .999",
              det = 100 * mean(flag[is_mis]),
              fp = 100 * mean(flag[!is_mis]), .by = c(n, B)),
  H |>
    filter(k > 0, w == "0.950") |>
    summarise(rule = "Interval .95",
              det = 100 * mean(flag[is_mis]),
              fp = 100 * mean(flag[!is_mis]), .by = c(n, B)),
  P |>
    filter(k > 0) |>
    summarise(rule = "Westfall-Young",
              det = 100 * mean(padj_fwer[is_mis] < ALPHA, na.rm = TRUE),
              fp = 100 * mean(padj_fwer[!is_mis] < ALPHA, na.rm = TRUE), .by = c(n, B)),
  P |>
    filter(k > 0) |>
    summarise(rule = "Benjamini-Hochberg",
              det = 100 * mean(padj_bh[is_mis] < ALPHA, na.rm = TRUE),
              fp = 100 * mean(padj_bh[!is_mis] < ALPHA, na.rm = TRUE), .by = c(n, B))
) |>
  mutate(across(c(det, fp), \(x) round(x, 1))) |>
  arrange(n, rule, B)
```

```{r fig-rules, fig.width = 8, fig.height = 4}
bind_rows(
  H |> filter(k > 0, w == "0.999") |>
    summarise(det = 100 * mean(flag[is_mis]), .by = c(n, B)) |>
    mutate(rule = "Interval .999"),
  P |> filter(k > 0) |>
    summarise(det = 100 * mean(padj_fwer[is_mis] < ALPHA, na.rm = TRUE), .by = c(n, B)) |>
    mutate(rule = "Westfall-Young"),
  P |> filter(k > 0) |>
    summarise(det = 100 * mean(padj_bh[is_mis] < ALPHA, na.rm = TRUE), .by = c(n, B)) |>
    mutate(rule = "Benjamini-Hochberg")
) |>
  ggplot(aes(factor(B), det, colour = rule, group = rule)) +
  geom_line() + geom_point(size = 2) +
  facet_wrap(~ n, labeller = label_both) +
  ylim(0, 100) +
  labs(x = "Bootstrap iterations (B)", y = "Detection (%)", colour = "Rule") +
  theme_minimal(base_size = 10)
```

## Benjamini-Hochberg reachability

At five items the floor is `m/alpha - 1` = 99 iterations, so BH is reachable at
every B tested. The smallest attainable adjusted p-value is shown against
alpha for reference.

```{r bh-floor}
data.frame(B = BS) |>
  mutate(floor_single = 5 / (B + 1),
         reachable = floor_single < ALPHA)
```
