---
title: "Bootstrap iterations for infit MSQ: FWER p-values vs simulation cutoffs"
subtitle: "Extending the 100/200/400 comparison in Johansson (2025) to FWER-adjusted p-values"
author: "Magnus Johansson"
date: today
---

## Purpose

Johansson (2025), Study 1, compared 100, 200 and 400 bootstrap iterations for
determining simulation-based **cutoffs** for conditional infit MSQ, and found
that *more* iterations gave *lower* detection rates (80.0 / 76.8 / 69.4 percent
at n = 150, on-target misfitting item), with false positives appearing only at
n >= 500. The recommendation was 100 iterations for n <= 250 and 200-400 for
larger samples.

easyRasch2 has since added an alternative decision rule for the same bootstrap:
Monte-Carlo **p-values** with multiplicity correction, via
`RMitemInfit(p_value = TRUE, correction = "fwer")` — the Westfall-Young
studentised-max step-down. Its documentation currently states that at least
1000 iterations are recommended, that the FWER correction is "liberal when the
simulation is small", and that marginal p-values are "well calibrated even at a
few hundred iterations".

That guidance has not been established by an iteration sweep. `dev/infit_dgp_fwer.R`
fixed B = 1000 and varied the data-generating process instead. This document
sweeps B directly.

### Why the two rules should behave differently

The mechanisms are not the same, and this predicts opposite directions:

- **HDCI cutoffs** need the *extreme quantiles* of the null distribution
  (`hdci_width = 0.999` implies the .0005 and .9995 quantiles). With B = 100
  draws those quantiles cannot be estimated — they are effectively the observed
  min and max — so the interval is too narrow and the rule over-flags. More
  iterations widen the interval, which *reduces* detection. That is the
  artifact visible in Johansson (2025), not a benefit of few iterations.

- **Monte-Carlo p-values** use the form `(1 + count) / (B + 1)`, which is
  exactly valid for any B under exchangeability. Increasing B should therefore
  *not* change the error rate, only reduce the variance of the estimate. The
  floor is `1 / (B + 1)`: .0099, .0050, .0025, .0010 for B = 100, 200, 400,
  1000 — all below alpha = .05, so no hard floor problem at conventional alpha.

- **The FWER step-down** is the exception. It studentises by the bootstrap mean
  and SD, both estimated from the same B draws, and then takes a maximum over
  items. Noise in the studentisation inflates the simulated maxima
  inconsistently, which is the stated reason for expecting liberal behaviour at
  small B.

Two things are therefore measured: the **error rate** as a function of B
(Study A), and the **Monte-Carlo stability of the decision** for a fixed
dataset (Study B), which the error rate alone hides.

```{r setup}
suppressMessages({
  library(easyRasch2)
  library(eRm)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

BS       <- c(100, 200, 400, 1000)
NS       <- c(150, 250, 500, 1000)
NMISFITS <- c(0, 1, 3)
R_REPS   <- 100
NCORES   <- 12          # performance cores only (M4 Max)
ALPHA    <- 0.05
```

### Data generation

Identical to `sx2_itemfit_sim.qmd`, which follows Johansson (2025): 20
dichotomous items, locations `runif(20, -2, 2)`, person locations bivariate
normal with SD 1.5 and inter-dimensional correlation 0.15, misfit induced by
loading items on the second dimension, misfitting items at approximately 0, -1
and -2 logits.

```{r gen}
make_master <- function(n_misfit, N = 10000, seed = 20250727) {
  set.seed(seed)
  locs <- runif(20, -2, 2)
  misfit_idx <- integer(0)
  for (tg in c(0, -1, -2)) {
    cand <- setdiff(order(abs(locs - tg)), misfit_idx)[1]
    misfit_idx <- c(misfit_idx, cand)
  }
  use <- misfit_idx[seq_len(n_misfit)]
  W <- cbind(rep(1, 20), rep(0, 20))
  if (n_misfit > 0) W[use, ] <- matrix(c(0, 1), nrow = length(use), ncol = 2,
                                       byrow = TRUE)
  Sigma <- matrix(c(1.5^2, .15 * 1.5^2, .15 * 1.5^2, 1.5^2), 2, 2)
  set.seed(seed + n_misfit)
  th <- MASS::mvrnorm(N, mu = c(0, 0), Sigma = Sigma)
  d <- as.data.frame(eRm::sim.xdim(persons = th, items = locs,
                                   Sigma = Sigma, weightmat = W))
  names(d) <- paste0("V", 1:20)
  list(data = d, locs = locs, misfit = use)
}

masters <- lapply(NMISFITS, make_master)
names(masters) <- as.character(NMISFITS)
misfit_all <- masters[["3"]]$misfit
```

## Study A: error rate and detection as a function of B

All four decision rules are read off the **same** bootstrap run, so differences
are attributable to the rule and not to simulation noise between rules.

```{r studyA}
RES_A <- "infit_fwer_iterations_A.rds"

one_rep_A <- function(rep, k, n, B) {
  set.seed(1e7 * k + 1e4 * n + 1e2 * rep + log10(B))
  md <- masters[[as.character(k)]]$data
  d  <- md[sample(nrow(md), n), ]
  out <- data.frame(item = 1:20, hdci = NA_integer_, fwer = NA_integer_,
                    bh = NA_integer_, uncorr = NA_integer_,
                    padj_fwer = NA_real_, p_marg = NA_real_)
  try({
    co <- RMitemInfitCutoff(d, iterations = B, parallel = FALSE,
                            seed = rep, verbose = FALSE)
    a <- RMitemInfit(d, cutoff = co, output = "dataframe")
    out$hdci <- as.integer(a$Flagged[match(names(d), a$Item)] != "")
    f <- suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                  alpha = ALPHA, output = "dataframe"))
    i <- match(names(d), f$Item)
    out$fwer      <- as.integer(f$padj_infit[i] < ALPHA)
    out$padj_fwer <- f$padj_infit[i]
    out$p_marg    <- f$p_infit[i]
    out$uncorr    <- as.integer(f$p_infit[i] < ALPHA)
    bh <- suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fdr_bh",
                  alpha = ALPHA, output = "dataframe"))
    out$bh <- as.integer(bh$padj_infit[match(names(d), bh$Item)] < ALPHA)
  }, silent = TRUE)
  out
}

gridA <- expand.grid(rep = 1:R_REPS, B = BS, n = NS, k = NMISFITS)

if (file.exists(RES_A)) {
  simA <- readRDS(RES_A)
} else {
  t0 <- Sys.time()
  resA <- mclapply(seq_len(nrow(gridA)),
                   function(i) one_rep_A(gridA$rep[i], gridA$k[i],
                                         gridA$n[i], gridA$B[i]),
                   mc.cores = NCORES)
  simA <- list(res = resA, grid = gridA, misfit = misfit_all,
               elapsed_min = as.numeric(Sys.time() - t0, "mins"))
  saveRDS(simA, RES_A)
}

longA <- do.call(rbind, lapply(seq_len(nrow(simA$grid)),
                               function(i) cbind(simA$grid[i, ], simA$res[[i]]))) |>
  mutate(is_misfit = mapply(function(k, it) it %in% simA$misfit[seq_len(k)],
                            k, item))
cat("elapsed:", round(simA$elapsed_min, 1), "min | failed:",
    sum(is.na(longA$hdci)), "of", nrow(longA), "\n")
```

### Family-wise error rate under a true Rasch model (k = 0)

The quantity the FWER correction is meant to control is
P(**any** item flagged) per dataset, which should be at or below alpha = .05.

```{r fwerA}
longA |>
  filter(k == 0) |>
  pivot_longer(c(hdci, fwer, bh, uncorr), names_to = "rule", values_to = "flag") |>
  group_by(rule, B, n, rep) |>
  summarise(any_flag = as.integer(sum(flag, na.rm = TRUE) > 0), .groups = "drop") |>
  group_by(rule, B, n) |>
  summarise(v = 100 * mean(any_flag), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rule, B)
```

Per-item false positive rate over the same datasets:

```{r fpA}
longA |>
  filter(k == 0) |>
  pivot_longer(c(hdci, fwer, bh, uncorr), names_to = "rule", values_to = "flag") |>
  group_by(rule, B, n) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  arrange(rule, B)
```

### Detection, one misfitting on-target item (k = 1)

Directly comparable to Figure 2 in Johansson (2025).

```{r detA1}
longA |>
  filter(k == 1, is_misfit) |>
  pivot_longer(c(hdci, fwer, bh, uncorr), names_to = "rule", values_to = "flag") |>
  group_by(rule, B, n) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rule, B)
```

### Detection and false positives, three misfitting items (k = 3)

```{r detA3}
longA |>
  filter(k == 3) |>
  pivot_longer(c(hdci, fwer, bh, uncorr), names_to = "rule", values_to = "flag") |>
  mutate(type = ifelse(is_misfit, "detection", "false_pos")) |>
  group_by(rule, B, n, type) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(type, rule, B)
```

### Figure

```{r figA, fig.width = 10, fig.height = 6}
longA |>
  filter(k %in% c(0, 1)) |>
  pivot_longer(c(hdci, fwer, bh, uncorr), names_to = "rule", values_to = "flag") |>
  mutate(panel = ifelse(k == 0, "Type I error (per item, true model)",
                        "Detection (misfitting item)")) |>
  filter(k == 0 | is_misfit) |>
  group_by(rule, B, n, panel) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(factor(B), v, colour = rule, group = rule)) +
  geom_line() + geom_point() +
  facet_grid(panel ~ n, scales = "free_y", labeller = label_both) +
  labs(x = "Bootstrap iterations (B)", y = "Percent", colour = "Rule") +
  theme_minimal(base_size = 11)
```

## Study B: Monte-Carlo stability of the decision

Error rates averaged over datasets can look fine while individual decisions
remain unstable. Here the *dataset is held fixed* and only the bootstrap seed
varies, isolating Monte-Carlo noise from sampling noise. If an analyst re-runs
`RMitemInfitCutoff()` on the same data with a different seed, how often does
the flag change?

```{r studyB}
RES_B <- "infit_fwer_iterations_B.rds"
N_SEEDS <- 40

fixed_sets <- list()
for (nn in c(250, 500)) {
  set.seed(4242 + nn)
  fixed_sets[[as.character(nn)]] <- masters[["3"]]$data[sample(10000, nn), ]
}

gridB <- expand.grid(seed = 1:N_SEEDS, B = BS, n = c(250, 500))

one_rep_B <- function(seed, B, n) {
  d <- fixed_sets[[as.character(n)]]
  out <- data.frame(item = 1:20, padj_fwer = NA_real_, p_marg = NA_real_,
                    hdci = NA_integer_)
  try({
    co <- RMitemInfitCutoff(d, iterations = B, parallel = FALSE,
                            seed = seed, verbose = FALSE)
    a <- RMitemInfit(d, cutoff = co, output = "dataframe")
    out$hdci <- as.integer(a$Flagged[match(names(d), a$Item)] != "")
    f <- suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                  alpha = ALPHA, output = "dataframe"))
    i <- match(names(d), f$Item)
    out$padj_fwer <- f$padj_infit[i]
    out$p_marg    <- f$p_infit[i]
  }, silent = TRUE)
  out
}

if (file.exists(RES_B)) {
  simB <- readRDS(RES_B)
} else {
  resB <- mclapply(seq_len(nrow(gridB)),
                   function(i) one_rep_B(gridB$seed[i], gridB$B[i], gridB$n[i]),
                   mc.cores = NCORES)
  simB <- list(res = resB, grid = gridB, misfit = misfit_all)
  saveRDS(simB, RES_B)
}

longB <- do.call(rbind, lapply(seq_len(nrow(simB$grid)),
                               function(i) cbind(simB$grid[i, ], simB$res[[i]]))) |>
  mutate(is_misfit = item %in% simB$misfit)
```

### Decision instability

Percentage of items whose FWER flag is **not unanimous** across the 40 bootstrap
seeds — i.e. the same data and the same rule give different answers depending
only on the seed.

```{r flipB}
longB |>
  group_by(n, B, item) |>
  summarise(rate = mean(padj_fwer < ALPHA, na.rm = TRUE), .groups = "drop") |>
  group_by(n, B) |>
  summarise(unstable_items = 100 * mean(rate > 0 & rate < 1), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = unstable_items, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

Same, for the HDCI cutoff rule, as a reference point:

```{r flipB2}
longB |>
  group_by(n, B, item) |>
  summarise(rate = mean(hdci, na.rm = TRUE), .groups = "drop") |>
  group_by(n, B) |>
  summarise(unstable_items = 100 * mean(rate > 0 & rate < 1), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = unstable_items, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Spread of the adjusted p-value across seeds

Median across items of the within-item SD of `padj_infit` over the 40 seeds.

```{r sdB}
longB |>
  group_by(n, B, item) |>
  summarise(sd_padj = sd(padj_fwer, na.rm = TRUE), .groups = "drop") |>
  group_by(n, B) |>
  summarise(median_sd = round(median(sd_padj, na.rm = TRUE), 4), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = median_sd, names_prefix = "n=")
```

```{r figB, fig.width = 9, fig.height = 5}
longB |>
  filter(n == 500) |>
  mutate(item_lab = factor(paste0("V", item), levels = paste0("V", 1:20)),
         type = ifelse(is_misfit, "misfitting", "fitting")) |>
  ggplot(aes(item_lab, padj_fwer, colour = type)) +
  geom_hline(yintercept = ALPHA, linetype = 2) +
  geom_jitter(width = .15, alpha = .4, size = .8) +
  facet_wrap(~B, labeller = label_both, ncol = 1) +
  labs(x = NULL, y = "FWER-adjusted p", colour = NULL,
       title = "Spread of FWER p across 40 bootstrap seeds, same dataset (n = 500)") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5))
```

## Study C: the small-sample trade, and buying detection honestly

Johansson (2025) recommended 100 iterations for n <= 250 because detection was
highest there. Study A shows that setting carries a 23-38 percent chance of at
least one false flag. But the detection loss at B = 1000 is real, not an
illusion, so "just use 1000" is only defensible if the loss is either small or
recoverable.

The key observation is that **B and `hdci_width` are confounded**. At B = 100
the .0005 and .9995 quantiles cannot be estimated, so the realised interval is
narrower than the nominal 99.9 percent and the effective per-item alpha is
unknown — it drifts with B. If wider flagging is wanted, the principled way to
get it is to *choose* a wider alpha explicitly at large B, which is
reproducible and has a stated error rate.

This study tests that: one B = 1000 bootstrap per replication, with the HDCI
recomputed at several widths from the stored draws (`$results`), against the
paper's B = 100 / width .999 setting. Direction of each flag is recorded,
because a false "overfit" flag and a false "underfit" flag have very different
consequences in an iterative item-removal workflow.

```{r studyC}
RES_C <- "infit_fwer_iterations_C.rds"
WIDTHS <- c(0.999, 0.99, 0.95, 0.90)
R_C <- 200

flag_dir <- function(msq, lo, hi) {
  ifelse(msq > hi, "underfit", ifelse(msq < lo, "overfit", ""))
}

bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}

one_rep_C <- function(rep, k, n) {
  set.seed(3e7 + 1e6 * k + 1e3 * n + rep)
  md <- masters[[as.character(k)]]$data
  d  <- md[sample(nrow(md), n), ]
  items <- names(d)
  obs <- RMitemInfit(d, output = "dataframe")
  msq <- obs$Infit_MSQ[match(items, obs$Item)]
  out <- data.frame(item = 1:20, rule = NA_character_, dir = NA_character_)[0, ]
  try({
    # paper's setting: B = 100, width .999
    c100 <- RMitemInfitCutoff(d, iterations = 100, parallel = FALSE,
                              seed = rep, verbose = FALSE)
    b <- bounds_at(c100$results, 0.999, items)
    out <- rbind(out, data.frame(item = 1:20, rule = "B=100 w=.999",
                                 dir = flag_dir(msq, b[1, ], b[2, ])))
    # one B = 1000 bootstrap, several widths
    c1000 <- RMitemInfitCutoff(d, iterations = 1000, parallel = FALSE,
                               seed = rep, verbose = FALSE)
    for (w in WIDTHS) {
      b <- bounds_at(c1000$results, w, items)
      out <- rbind(out, data.frame(item = 1:20,
                                   rule = paste0("B=1000 w=", sub("^0", "", w)),
                                   dir = flag_dir(msq, b[1, ], b[2, ])))
    }
  }, silent = TRUE)
  out
}

gridC <- expand.grid(rep = 1:R_C, n = c(150, 250), k = c(0, 3))

if (file.exists(RES_C)) {
  simC <- readRDS(RES_C)
} else {
  resC <- mclapply(seq_len(nrow(gridC)),
                   function(i) one_rep_C(gridC$rep[i], gridC$k[i], gridC$n[i]),
                   mc.cores = NCORES)
  simC <- list(res = resC, grid = gridC, misfit = misfit_all)
  saveRDS(simC, RES_C)
}

longC <- do.call(rbind, lapply(seq_len(nrow(simC$grid)), function(i) {
  r <- simC$res[[i]]
  if (nrow(r) == 0) return(NULL)
  cbind(simC$grid[i, ], r)
})) |>
  mutate(is_misfit = mapply(function(k, it) it %in% simC$misfit[seq_len(k)],
                            k, item),
         flag = as.integer(dir != ""))
```

### Detection and false positives at n = 150 and 250 (k = 3)

```{r detC}
longC |>
  filter(k == 3) |>
  mutate(type = ifelse(is_misfit, "detection", "false_pos")) |>
  group_by(rule, n, type) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = c(type, n), values_from = v) |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Family-wise false positive rate under a true model (k = 0)

```{r fweC}
longC |>
  filter(k == 0) |>
  group_by(rule, n, rep) |>
  summarise(any = as.integer(sum(flag) > 0), .groups = "drop") |>
  group_by(rule, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Direction of the false flags (k = 3, fitting items only)

```{r dirC}
dir_split <- function(kk) {
  longC |>
    filter(k == kk, if (kk == 3) !is_misfit else TRUE, dir != "") |>
    group_by(rule, sample_n = n, dir) |>
    summarise(ct = n(), .groups = "drop") |>
    group_by(rule, sample_n) |>
    mutate(pct = round(100 * ct / sum(ct), 1)) |>
    ungroup() |>
    select(rule, sample_n, dir, pct) |>
    pivot_wider(names_from = dir, values_from = pct, values_fill = 0)
}
dir_split(3)
```

The same split under a **true model** (k = 0), where every flag is a false
positive:

```{r dirC0}
dir_split(0)
```

### Expected flags per analysis (k = 3)

The quantity an analyst actually faces: how many of the 3 misfitting items are
flagged, and how many of the 17 fitting items are flagged, in a single run.

```{r expC}
longC |>
  filter(k == 3) |>
  group_by(rule, n, rep) |>
  summarise(true_flags = sum(flag[is_misfit]),
            false_flags = sum(flag[!is_misfit]), .groups = "drop") |>
  group_by(rule, n) |>
  summarise(true = round(mean(true_flags), 2),
            false = round(mean(false_flags), 2), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = c(true, false))
```

## Conclusions

**1. Marginal p-values are calibrated at every B — including 100.** Per-item
false positives under a true model were 4.25-5.15 percent at B = 100 and
4.25-4.70 at B = 1000, with no trend. The documentation claim that marginal
p-values are "well calibrated even at a few hundred iterations" is confirmed,
and can be stated more strongly: 100 iterations is enough for calibration.

**2. FWER is close to B-invariant in both error rate and power.** Detection of
a single on-target misfitting item was 99-100 percent at *every* B from 100 to
1000. The family-wise error rate fell only mildly, from 9/3/2/10 percent
(n = 150/250/500/1000) at B = 100 to 2/2/3/6 at B = 1000. The documented
"liberal when the simulation is small" behaviour is real but small — a few
percentage points, partly within Monte-Carlo noise at 100 replications. It is
not the main reason to run more iterations.

**3. The real cost of small B is decision instability, which averaged error
rates hide.** Holding the dataset fixed and varying only the bootstrap seed,
the percentage of items whose FWER flag was *not unanimous* across 40 seeds was
15 percent at B = 100 and 0 percent at B = 1000 (n = 500). The median
within-item SD of the adjusted p-value fell from .068 to .019. Critically, the
three genuinely misfitting items were flagged in 100 percent of seeds at every
B — **all of the instability was in fitting items flipping in and out of
significance.** Small B therefore costs reproducibility and produces spurious
flags, not missed detections. This is the argument for >= 1000 iterations, and
it is a different argument from the one currently in the documentation.

**4. BH has a hard reachability floor that FWER does not.** The smallest
attainable BH-adjusted p-value is `m / (B + 1)` for m comparisons, so BH cannot
flag a single comparison unless `B >= m/alpha - 1`. With m = 20 items at
alpha = .05 that is B >= 399. This is arithmetic, not power, and it is visible
in the results: BH detection of one misfitting item was 1 / 8 / 99 / 100
percent at B = 100 / 200 / 400 / 1000. The jump at B = 400 is exactly where
`20/401 = .0499` crosses below .05. FWER has no such floor — the studentised-max
p-value needs only `B >= 1/alpha - 1 = 19`.

**5. HDCI cutoffs are strongly B-dependent, and B = 100 is badly calibrated.**
Under a true model the probability of at least one spuriously flagged item was
38 / 23 / 33 / 28 percent at B = 100, falling to 2 / 3 / 3 / 5 percent at
B = 1000. Per-item this is 2.30 percent versus 0.10 percent at n = 150. This
explains the Johansson (2025) Study 1 finding that detection *falls* as
iterations increase (80.0 / 76.8 / 69.4 percent at n = 150): at small B the
.0005 and .9995 quantiles cannot be estimated, the HDCI band is too narrow, and
the rule over-flags. The apparent power advantage of 100 iterations is that
artifact, not genuine sensitivity. The paper's per-item false-positive rates at
B = 100 (roughly 1.4-2.3 percent here) fall below the 5 percent threshold used
for highlighting in its figures, but aggregate to a 23-38 percent chance of at
least one false flag across 20 items.

**6. The small-sample detection loss is real, but it is a `hdci_width`
question, not an iterations question.** Study C shows that `B = 1000` with
`hdci_width = 0.99` reproduces the detection rate of the Johansson (2025)
setting (`B = 100`, width .999) while being better on every other measure:

| Rule | det n=150 | fp n=150 | det n=250 | fp n=250 | family-wise FP (true model) n=150 / n=250 |
|---|---|---|---|---|---|
| B=100 w=.999 | 77.5 | 4.8 | 92.8 | 7.2 | 30.0 / 26.0 |
| B=1000 w=.99 | 78.5 | 2.9 | 93.2 | 5.1 | 20.0 / 14.0 |
| B=1000 w=.999 | 64.3 | 0.6 | 86.5 | 1.4 | 5.5 / 2.0 |

Expected flags per analysis at n = 150: 2.33 true / 0.81 false for the B = 100
setting, versus 2.36 true / 0.50 false at B = 1000 with width .99. Equal
detection, roughly 40 percent fewer false flags, a *stated* error rate, and a
reproducible result. The two settings are not a trade-off — one dominates.

**7. The "false positives are mostly overfit, so they are benign" argument does
not extend to the null case.** With three misfitting items present, 97-100
percent of false flags were *overfit*, reproducing Table 2 of Johansson (2025)
and consistent with the overfit cascade. But under a **true model** the split
was roughly 50/50 overfit/underfit at every width. Table 2 was computed on
datasets that *contained* misfitting items, so it does not license the general
claim that false flags are harmless. About half the flags raised by an
over-liberal rule on a well-fitting scale point the wrong way in the direction
that triggers item deletion.

### Practical recommendations

| Use | Iterations |
|---|---|
| HDCI cutoffs (`RMitemInfit` with `cutoff`) | >= 1000 always; tune `hdci_width`, not B |
| Small samples (n <= 250) wanting more sensitivity | `iterations = 1000, hdci_width = 0.99` — dominates `iterations = 100, hdci_width = 0.999` |
| Marginal p-value, single pre-specified item | 100 is adequate; 1000 for a stable reported value |
| FWER (`correction = "fwer"`) | >= 1000, for reproducibility rather than calibration |
| BH (`correction = "fdr_bh"`) | `B >= m/alpha - 1`, i.e. >= 400 for 20 items — and far more for pairwise comparisons |

The existing >= 1000 warning threshold is well chosen. The stated *reason*
should change: the dominant problem at small B is Monte-Carlo instability of
the decision (and, for BH, unreachability), not liberal FWER calibration.

The reason B and `hdci_width` must not be conflated: at B = 100 the .0005 and
.9995 quantiles cannot be estimated — the HDCI endpoints are effectively the
observed min and max of 100 draws — so the realised interval is narrower than
nominal and the effective per-item alpha is unknown and drifts with B. That is
visible in the family-wise rate at B = 100, which wanders (38 / 23 / 33 / 28
percent across n in Study A) instead of settling. Choosing a wider
`hdci_width` at large B gives the same sensitivity with an error rate that is
stated, stable across n, and reproducible across seeds.

### Note on other functions using this helper

`.bootstrap_pvalues()` is shared with the Q3, partial-gamma DIF/LD, and CFA
cutoff functions, so point 4 applies to them with much larger m. Verified
directly for Q3: with 20 items (190 pairs) at B = 1000, on data with three
genuinely locally dependent pairs sitting at the marginal p floor of
`1/1001 = .000999`, `correction = "fdr_bh"` returns adjusted p = .0633 for all
three and flags **none**, while `correction = "fwer"` on the same bootstrap
flags all three. A single pair would need `B >= 190/.05 - 1 = 3799`. The
package default is `"fwer"`, so the default path is unaffected, but the BH
option is unusable at the documented iteration counts for pairwise comparisons.

## References

Ferreira, J. A. (2024). The Benjamini-Hochberg method with infinitely many
discoveries. *Journal of Statistical Planning and Inference*.

Johansson, M. (2025). Detecting Item Misfit in Rasch Models.
*Educational Methods & Psychometrics, 3*(18). <https://dx.doi.org/10.61186/emp.2025.5>

Westfall, P. H., & Young, S. S. (1993). *Resampling-Based Multiple Testing*.
Wiley.
