---
title: "Interval width and bootstrap iterations: polytomous items"
subtitle: "Generalisation check for the follow-up to Johansson (2025)"
author: "Magnus Johansson"
date: today
---

## Scope

Companion to `infit_width_iterations.qmd`, which used 20 **dichotomous** items.
Two claims established there are model-agnostic in principle but were only
measured on dichotomous data:

1. The HDCI rule's family-wise error is approximately `1 - w^k` (the Sidak
   relation), so `hdci_width` is the error-rate parameter.
2. The Westfall-Young FWER p-value matches or beats the width-based rule at
   equal error, and needs a smaller B because it estimates no extreme quantile.

This document repeats the design with **polytomous (PCM) items**.

### Matched to the dichotomous study

Item locations are generated with the *same seed and call* as the dichotomous
study (`set.seed(20250727); runif(20, -2, 2)`), so the 20 items sit at
identical locations and the misfitting items are the same three — V8, V12, V11
at 0.04, -1 and -2 logits. Person locations, the inter-dimensional correlation
(0.15) and the misfit mechanism are also unchanged. **The only difference from
the dichotomous study is the response format**: four ordered categories (0-3)
with thresholds spread `+/- 1` logit around each item location, instead of a
binary response.

Keeping k = 20 means this tests item *type* while holding item *count* fixed.
The separate question of whether `1 - w^k` tracks a different k is better
answered with a cheap dichotomous run at k = 10 than by changing two things at
once here.

Data are generated with the package's own PCM generator
(`sim_partial_score()`), the same one `RMitemInfitCutoff()` uses internally;
misfitting items are drawn from the second theta dimension, fitting items from
the first.

### A note on effect size

Polytomous items carry more information per item, so the same inter-dimensional
correlation produces **much stronger** misfit than in the dichotomous study: at
n = 500 the misfitting items show infit around 1.75-2.03 against 0.74-0.93 for
fitting items. Detection therefore saturates, and the informative contrast
between rules is on the false-positive side (the overfit cascade) rather than
on detection. The primary question here — whether `1 - w^k` holds under a true
model — is unaffected, since it is measured at k = 0.

### Design

- 20 polytomous items, 4 categories; locations and misfitting items identical
  to the dichotomous study
- n in {250, 1000}; B in {100, 400, 2000}; w in {.90, .95, .99, .995, .999}
- number of misfitting items in {0, 3}
- independent bootstrap per `(rep, k, n, B)` cell
- `R2` starts at 100; runs are resumable, so raising it computes only the new
  replications

```{r setup}
suppressMessages({
  pkgload::load_all("..", quiet = TRUE)   # sim_partial_score() is internal
  library(ggdist)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

NITEMS   <- 20
NCAT     <- 4
WIDTHS   <- c(0.90, 0.95, 0.99, 0.995, 0.999)
NS       <- c(250, 1000)
BS       <- c(100, 400, 2000)
NMISFITS <- c(0, 3)
R2       <- 100          # raise and re-render to extend
NCORES   <- 10
ALPHA    <- 0.05
CHUNK    <- 40
RES      <- "infit_width_iterations_poly.rds"
```

```{r runner}
run_resumable <- function(res_file, grid, keys, fn,
                          ncores = NCORES, chunk_size = CHUNK,
                          extra = list()) {
  key_of <- function(g) {
    if (nrow(g) == 0L) return(character(0))
    do.call(paste, c(unname(as.list(g[keys])), sep = "|"))
  }
  acc <- if (file.exists(res_file)) readRDS(res_file) else
    c(list(res = list(), grid = grid[0, , drop = FALSE], elapsed_min = 0), extra)
  todo <- which(!(key_of(grid) %in% key_of(acc$grid)))
  if (length(todo) == 0L) {
    message("cached: ", nrow(acc$grid), " cells, nothing to run"); return(acc)
  }
  message("running ", length(todo), " new cells (", nrow(acc$grid), " cached)")
  for (s in seq(1, length(todo), by = chunk_size)) {
    idx <- todo[s:min(s + chunk_size - 1L, length(todo))]
    t0 <- Sys.time()
    out <- mclapply(idx, function(i) fn(grid[i, , drop = FALSE]),
                    mc.cores = ncores)
    acc$res <- c(acc$res, out)
    acc$grid <- rbind(acc$grid, grid[idx, , drop = FALSE])
    acc$elapsed_min <- acc$elapsed_min + as.numeric(Sys.time() - t0, "mins")
    acc <- utils::modifyList(acc, extra)
    saveRDS(acc, res_file)
    message("  checkpoint ", nrow(acc$grid), " / ", nrow(grid),
            "  (", round(acc$elapsed_min, 1), " min)")
  }
  acc
}
```

### Data generation

```{r gen}
set.seed(20250727)                 # identical to the dichotomous study
LOCS <- runif(NITEMS, -2, 2)
MISFIT <- integer(0)
for (tg in c(0, -1, -2)) {
  MISFIT <- c(MISFIT, setdiff(order(abs(LOCS - tg)), MISFIT)[1])
}
DELTAS <- lapply(LOCS, function(L) L + seq(-1, 1, length.out = NCAT - 1))
SIGMA <- matrix(c(1.5^2, .15 * 1.5^2, .15 * 1.5^2, 1.5^2), 2, 2)

sim_poly <- function(n, k, seed) {
  set.seed(seed)
  th <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = SIGMA)
  mis <- MISFIT[seq_len(k)]
  d <- matrix(0L, n, NITEMS)
  fit_items <- setdiff(seq_len(NITEMS), mis)
  if (length(fit_items)) d[, fit_items] <- sim_partial_score(DELTAS[fit_items], th[, 1])
  if (length(mis))       d[, mis]       <- sim_partial_score(DELTAS[mis], th[, 2])
  d <- as.data.frame(d)
  names(d) <- paste0("V", seq_len(NITEMS))
  d
}

data.frame(item = paste0("V", MISFIT), intended = c(0, -1, -2),
           actual = round(LOCS[MISFIT], 2))
```

```{r helpers}
bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}
flag_code <- function(msq, lo, hi) ifelse(msq > hi, 2L, ifelse(msq < lo, 1L, 0L))
```

### Run

```{r run}
one_rep <- function(g) {
  rep <- g$rep; k <- g$k; n <- g$n
  d <- sim_poly(n, k, seed = 8e7 + 1e6 * k + 1e3 * n + rep)
  items <- names(d)
  obs <- try(RMitemInfit(d, output = "dataframe"), silent = TRUE)
  if (inherits(obs, "try-error")) return(NULL)
  msq <- obs$Infit_MSQ[match(items, obs$Item)]
  hd <- list(); pv <- list()
  for (j in seq_along(BS)) {
    co <- try(RMitemInfitCutoff(d, iterations = BS[j], parallel = FALSE,
                                seed = rep * 100 + j, verbose = FALSE),
              silent = TRUE)
    if (inherits(co, "try-error")) next
    hd[[length(hd) + 1L]] <- do.call(rbind, lapply(WIDTHS, function(w) {
      b <- bounds_at(co$results, w, items)
      data.frame(B = BS[j], w = w, item = seq_along(items),
                 code = flag_code(msq, b[1, ], b[2, ]))
    }))
    fw <- try(suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                  alpha = ALPHA, output = "dataframe")), silent = TRUE)
    bh <- try(suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fdr_bh",
                  alpha = ALPHA, output = "dataframe")), silent = TRUE)
    if (!inherits(fw, "try-error") && !inherits(bh, "try-error")) {
      i1 <- match(items, fw$Item); i2 <- match(items, bh$Item)
      pv[[length(pv) + 1L]] <- data.frame(
        B = BS[j], item = seq_along(items),
        p_marg = fw$p_infit[i1], padj_fwer = fw$padj_infit[i1],
        padj_bh = bh$padj_infit[i2])
    }
  }
  list(hdci = do.call(rbind, hd), pval = do.call(rbind, pv))
}

grid <- expand.grid(rep = 1:R2, n = NS, k = NMISFITS)
s <- run_resumable(RES, grid, c("rep", "k", "n"), one_rep,
                   extra = list(misfit = MISFIT, locs = LOCS))

bind_part <- function(part) {
  do.call(rbind, lapply(seq_len(nrow(s$grid)), function(i) {
    r <- s$res[[i]][[part]]
    if (is.null(r)) return(NULL)
    cbind(rep = s$grid$rep[i], k = s$grid$k[i], n = s$grid$n[i], r)
  }))
}
tag <- function(x) x |>
  mutate(is_misfit = mapply(function(k, it) it %in% s$misfit[seq_len(k)], k, item))
H <- bind_part("hdci") |> tag() |>
  mutate(flag = as.integer(code > 0), w = sprintf("%.3f", w))
P <- bind_part("pval") |> tag()
cat(nrow(s$grid), "cells |", round(s$elapsed_min, 1), "min |",
    length(unique(s$grid$rep)), "reps\n")
```

## Does `1 - w^k` hold for polytomous items?

Predicted family-wise error at k = 20:

```{r theory}
data.frame(w = sprintf("%.3f", WIDTHS),
           predicted = round(100 * (1 - WIDTHS^NITEMS), 1))
```

Measured under a true model:

```{r fwe}
H |>
  filter(k == 0) |>
  group_by(w, B, n, rep) |>
  summarise(any = as.integer(sum(flag) > 0), .groups = "drop") |>
  group_by(w, B, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(w, B)
```

## Detection and false positives (3 misfitting items)

```{r det}
H |>
  filter(k == 3) |>
  mutate(type = ifelse(is_misfit, "det", "fp")) |>
  group_by(w, B, n, type) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = c(type, n), values_from = v) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(w, B)
```

## p-value rules

```{r pv}
PL <- P |>
  mutate(fwer = as.integer(padj_fwer < ALPHA),
         bh = as.integer(padj_bh < ALPHA),
         uncorr = as.integer(p_marg < ALPHA)) |>
  pivot_longer(c(fwer, bh, uncorr), names_to = "rule", values_to = "flag")

PL |>
  filter(k == 0) |>
  group_by(rule, B, n, rep) |>
  summarise(any = as.integer(sum(flag, na.rm = TRUE) > 0), .groups = "drop") |>
  group_by(rule, B, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rule, B)
```

```{r pvdet}
PL |>
  filter(k == 3) |>
  mutate(type = ifelse(is_misfit, "det", "fp")) |>
  group_by(rule, B, n, type) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = c(type, n), values_from = v) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rule, B)
```

## Head-to-head at matched family-wise error

```{r h2h}
bind_rows(
  H |> filter(k == 3, B == 2000, w == "0.999") |>
    group_by(n) |>
    summarise(rule = "HDCI w=.999 B=2000",
              det = round(100 * mean(flag[is_misfit]), 1),
              fp = round(100 * mean(flag[!is_misfit]), 1), .groups = "drop"),
  PL |> filter(k == 3, B == 2000, rule == "fwer") |>
    group_by(n) |>
    summarise(rule = "FWER B=2000",
              det = round(100 * mean(flag[is_misfit], na.rm = TRUE), 1),
              fp = round(100 * mean(flag[!is_misfit], na.rm = TRUE), 1),
              .groups = "drop")
) |> arrange(n, rule)
```
