---
title: "Interval width and bootstrap iterations for conditional infit MSQ"
subtitle: "Simulation material for a follow-up to Johansson (2025)"
author: "Magnus Johansson"
date: today
---

## Scope

Simulation material only — no manuscript text. Three questions:

- **Study 1 (convergence).** The HDCI endpoint at width *w* is an extreme order
  statistic, so its stability is governed by the expected number of bootstrap
  draws in each tail, `(1 - w)/2 * B`. Where does the interval stop changing,
  for each width, up to B = 10 000?
- **Study 2 (error rates).** Detection and false positive rates over the full
  `(w, B)` grid.
- **Study 3 (multiplicity).** Marginal, FWER (Westfall-Young step-down) and BH
  p-values, read off the same bootstraps as Study 2.

Every `(rep, k, n, B)` cell uses an **independent** bootstrap; no draws are
reused across B conditions.

### Extending a run

Replication counts are set by `R1` and `R2` below. Runs are **resumable**:
raise `R2` from 250 to 500, re-render, and only replications 251-500 are
computed — the cached cells are reused. Seeds are a deterministic function of
`(rep, k, n, B)`, so a resumed run is identical to having run the larger number
from the start. Results are checkpointed to disk after every chunk, so an
interrupted run also resumes from where it stopped rather than from zero.

Delete the `.rds` files to force a full re-run.

```{r setup}
suppressMessages({
  library(easyRasch2)
  library(ggdist)
  library(eRm)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

WIDTHS   <- c(0.90, 0.95, 0.99, 0.995, 0.999)
NS       <- c(150, 250, 500, 1000)
NCORES   <- 10
ALPHA    <- 0.05
CHUNK    <- 100          # cells per checkpoint

B1 <- c(100, 200, 400, 1000, 2000, 5000, 10000)
R1 <- 30
RES_1 <- "infit_width_iterations_S1.rds"

B2 <- c(100, 200, 400, 1000, 2000)
R2 <- 250                # raise to 500 and re-render to extend
NMISFITS <- c(0, 1, 3)
RES_2 <- "infit_width_iterations_S2.rds"
```

### Resumable runner

```{r runner}
run_resumable <- function(res_file, grid, keys, fn,
                          ncores = NCORES, chunk_size = CHUNK,
                          extra = list()) {
  key_of <- function(g) {
    if (nrow(g) == 0L) return(character(0))
    do.call(paste, c(unname(as.list(g[keys])), sep = "|"))
  }
  if (file.exists(res_file)) {
    acc <- readRDS(res_file)
  } else {
    acc <- c(list(res = list(), grid = grid[0, , drop = FALSE],
                  elapsed_min = 0), extra)
  }
  todo <- which(!(key_of(grid) %in% key_of(acc$grid)))
  if (length(todo) == 0L) {
    message("cached: ", nrow(acc$grid), " cells, nothing to run")
    return(acc)
  }
  message("running ", length(todo), " new cells (",
          nrow(acc$grid), " already cached)")
  for (s in seq(1, length(todo), by = chunk_size)) {
    idx <- todo[s:min(s + chunk_size - 1L, length(todo))]
    t0 <- Sys.time()
    out <- mclapply(idx, function(i) fn(grid[i, , drop = FALSE]),
                    mc.cores = ncores)
    acc$res <- c(acc$res, out)
    acc$grid <- rbind(acc$grid, grid[idx, , drop = FALSE])
    acc$elapsed_min <- acc$elapsed_min +
      as.numeric(Sys.time() - t0, "mins")
    acc <- utils::modifyList(acc, extra)
    saveRDS(acc, res_file)
    message("  checkpoint ", nrow(acc$grid), " / ", nrow(grid),
            "  (", round(acc$elapsed_min, 1), " min)")
  }
  acc
}
```

### Data generation

Follows Johansson (2025): 20 dichotomous items, locations
`runif(20, -2, 2)`, person locations bivariate normal with SD 1.5 and
inter-dimensional correlation 0.15, misfit induced by loading items on a second
dimension, misfitting items at approximately 0, -1 and -2 logits.

```{r gen}
make_master <- function(n_misfit, N = 10000, seed = 20250727) {
  set.seed(seed)
  locs <- runif(20, -2, 2)
  misfit_idx <- integer(0)
  for (tg in c(0, -1, -2)) {
    cand <- setdiff(order(abs(locs - tg)), misfit_idx)[1]
    misfit_idx <- c(misfit_idx, cand)
  }
  use <- misfit_idx[seq_len(n_misfit)]
  W <- cbind(rep(1, 20), rep(0, 20))
  if (n_misfit > 0) W[use, ] <- matrix(c(0, 1), nrow = length(use), ncol = 2,
                                       byrow = TRUE)
  Sigma <- matrix(c(1.5^2, .15 * 1.5^2, .15 * 1.5^2, 1.5^2), 2, 2)
  set.seed(seed + n_misfit)
  th <- MASS::mvrnorm(N, mu = c(0, 0), Sigma = Sigma)
  d <- as.data.frame(eRm::sim.xdim(persons = th, items = locs,
                                   Sigma = Sigma, weightmat = W))
  names(d) <- paste0("V", 1:20)
  list(data = d, locs = locs, misfit = use)
}

masters <- lapply(NMISFITS, make_master)
names(masters) <- as.character(NMISFITS)
misfit_all <- masters[["3"]]$misfit
locs <- masters[["3"]]$locs

data.frame(item = paste0("V", misfit_all), intended = c(0, -1, -2),
           actual = round(locs[misfit_all], 2))
```

```{r helpers}
bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}

# 0 = not flagged, 1 = overfit (below interval), 2 = underfit (above)
flag_code <- function(msq, lo, hi) {
  ifelse(msq > hi, 2L, ifelse(msq < lo, 1L, 0L))
}
```

## Study 1: convergence of the interval

Null data only (k = 0), independent bootstrap per B.

```{r s1}
one_rep_S1 <- function(g) {
  rep <- g$rep; n <- g$n
  set.seed(5e7 + 1e3 * n + rep)
  d <- masters[["0"]]$data[sample(10000, n), ]
  items <- names(d)
  out <- vector("list", length(B1))
  for (j in seq_along(B1)) {
    co <- try(RMitemInfitCutoff(d, iterations = B1[j], parallel = FALSE,
                                seed = rep * 100 + j, verbose = FALSE),
              silent = TRUE)
    if (inherits(co, "try-error")) next
    out[[j]] <- do.call(rbind, lapply(WIDTHS, function(w) {
      b <- bounds_at(co$results, w, items)
      data.frame(B = B1[j], w = w, item = seq_along(items),
                 width = b[2, ] - b[1, ])
    }))
  }
  do.call(rbind, out)
}

grid1 <- expand.grid(rep = 1:R1, n = NS)
s1 <- run_resumable(RES_1, grid1, c("rep", "n"), one_rep_S1)

long1 <- do.call(rbind, lapply(seq_len(nrow(s1$grid)), function(i) {
  r <- s1$res[[i]]
  if (is.null(r)) return(NULL)
  cbind(rep = s1$grid$rep[i], n = s1$grid$n[i], r)
}))
cat("Study 1:", nrow(s1$grid), "cells |", round(s1$elapsed_min, 1), "min\n")
```

### Median interval width by B and width

```{r s1tab}
long1 |>
  group_by(n, w, B) |>
  summarise(v = median(width), .groups = "drop") |>
  pivot_wider(names_from = B, values_from = v, names_prefix = "B=") |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  arrange(n, w)
```

### Relative to the B = 10 000 reference

Values below 1 mean the interval is too **narrow**, which inflates the false
positive rate.

```{r s1rel}
long1 |>
  group_by(n, w, B) |>
  summarise(v = median(width), .groups = "drop") |>
  group_by(n, w) |>
  mutate(rel = round(v / v[B == max(B)], 3)) |>
  select(-v) |>
  pivot_wider(names_from = B, values_from = rel, names_prefix = "B=") |>
  arrange(n, w)
```

### Draws expected in each tail

```{r s1tail}
expand.grid(w = WIDTHS, B = B1) |>
  mutate(per_tail = (1 - w) / 2 * B) |>
  pivot_wider(names_from = B, values_from = per_tail, names_prefix = "B=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

```{r s1fig, fig.width = 9, fig.height = 5}
long1 |>
  group_by(n, w, B) |>
  summarise(v = median(width), .groups = "drop") |>
  group_by(n, w) |>
  mutate(rel = v / v[B == max(B)]) |>
  ggplot(aes(B, rel, colour = factor(w))) +
  geom_hline(yintercept = 1, linetype = 2) +
  geom_line() + geom_point(size = 1) +
  scale_x_log10() +
  facet_wrap(~n, labeller = label_both) +
  labs(x = "Bootstrap iterations (log scale)",
       y = "Interval width relative to largest B",
       colour = "hdci_width") +
  theme_minimal(base_size = 11)
```

## Study 2 and 3: error rates and multiplicity

One pass produces both: HDCI flags at every width, and the three p-value rules.

```{r s2}
one_rep_S2 <- function(g) {
  rep <- g$rep; k <- g$k; n <- g$n
  set.seed(6e7 + 1e6 * k + 1e3 * n + rep)
  d <- masters[[as.character(k)]]$data[sample(10000, n), ]
  items <- names(d)
  obs <- RMitemInfit(d, output = "dataframe")
  msq <- obs$Infit_MSQ[match(items, obs$Item)]
  hd <- vector("list", length(B2)); pv <- vector("list", length(B2))
  for (j in seq_along(B2)) {
    co <- try(RMitemInfitCutoff(d, iterations = B2[j], parallel = FALSE,
                                seed = rep * 100 + j, verbose = FALSE),
              silent = TRUE)
    if (inherits(co, "try-error")) next
    hd[[j]] <- do.call(rbind, lapply(WIDTHS, function(w) {
      b <- bounds_at(co$results, w, items)
      data.frame(B = B2[j], w = w, item = seq_along(items),
                 code = flag_code(msq, b[1, ], b[2, ]))
    }))
    fw <- try(suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                  alpha = ALPHA, output = "dataframe")), silent = TRUE)
    bh <- try(suppressWarnings(
      RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fdr_bh",
                  alpha = ALPHA, output = "dataframe")), silent = TRUE)
    if (!inherits(fw, "try-error") && !inherits(bh, "try-error")) {
      i1 <- match(items, fw$Item); i2 <- match(items, bh$Item)
      pv[[j]] <- data.frame(B = B2[j], item = seq_along(items),
                            p_marg = fw$p_infit[i1],
                            padj_fwer = fw$padj_infit[i1],
                            padj_bh = bh$padj_infit[i2])
    }
  }
  list(hdci = do.call(rbind, hd), pval = do.call(rbind, pv))
}

grid2 <- expand.grid(rep = 1:R2, n = NS, k = NMISFITS)
s2 <- run_resumable(RES_2, grid2, c("rep", "k", "n"), one_rep_S2,
                    extra = list(misfit = misfit_all))

bind_part <- function(part) {
  do.call(rbind, lapply(seq_len(nrow(s2$grid)), function(i) {
    r <- s2$res[[i]][[part]]
    if (is.null(r)) return(NULL)
    cbind(rep = s2$grid$rep[i], k = s2$grid$k[i], n = s2$grid$n[i], r)
  }))
}
tag <- function(x) x |>
  mutate(is_misfit = mapply(function(k, it) it %in% s2$misfit[seq_len(k)],
                            k, item))
hdci2 <- bind_part("hdci") |> tag() |> mutate(flag = as.integer(code > 0))
pval2 <- bind_part("pval") |> tag()
cat("Study 2/3:", nrow(s2$grid), "cells |", round(s2$elapsed_min, 1), "min |",
    length(unique(s2$grid$rep)), "reps\n")
```

### Family-wise false positive rate under a true model (k = 0)

```{r s2fwer}
hdci2 |>
  filter(k == 0) |>
  group_by(w, B, n, rep) |>
  summarise(any = as.integer(sum(flag) > 0), .groups = "drop") |>
  group_by(w, B, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(w, B)
```

### Detection and false positives (k = 3)

```{r s2det}
hdci2 |>
  filter(k == 3) |>
  mutate(type = ifelse(is_misfit, "detection", "false_pos")) |>
  group_by(w, B, n, type) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(type, w, B)
```

### Expected flags per analysis (k = 3)

```{r s2exp}
hdci2 |>
  filter(k == 3) |>
  group_by(w, B, n, rep) |>
  summarise(t = sum(flag[is_misfit]), f = sum(flag[!is_misfit]),
            .groups = "drop") |>
  group_by(w, B, n) |>
  summarise(true = round(mean(t), 2), false = round(mean(f), 2),
            .groups = "drop") |>
  pivot_wider(names_from = n, values_from = c(true, false)) |>
  arrange(w, B)
```

### Direction of false flags under a true model

```{r s2dir}
hdci2 |>
  filter(k == 0, code > 0) |>
  mutate(dir = ifelse(code == 1, "overfit", "underfit")) |>
  group_by(w, B, dir) |>
  summarise(ct = n(), .groups = "drop") |>
  group_by(w, B) |>
  mutate(pct = round(100 * ct / sum(ct), 1)) |>
  ungroup() |> select(-ct) |>
  pivot_wider(names_from = dir, values_from = pct, values_fill = 0)
```

## Study 3: FWER vs BH

### Per-item rates

```{r s3rates}
pv_long <- pval2 |>
  mutate(fwer = as.integer(padj_fwer < ALPHA),
         bh = as.integer(padj_bh < ALPHA),
         uncorr = as.integer(p_marg < ALPHA)) |>
  pivot_longer(c(fwer, bh, uncorr), names_to = "rule", values_to = "flag")

pv_long |>
  filter(k %in% c(0, 3)) |>
  mutate(type = case_when(k == 0 ~ "type_I", is_misfit ~ "detection",
                          TRUE ~ "false_pos")) |>
  group_by(rule, B, n, type) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(type, rule, B)
```

### Family-wise error rate under a true model

The quantity FWER is meant to bound. BH bounds the false discovery rate
instead, so it is not expected to sit at alpha here — the comparison shows what
each actually delivers.

```{r s3fwe}
pv_long |>
  filter(k == 0) |>
  group_by(rule, B, n, rep) |>
  summarise(any = as.integer(sum(flag, na.rm = TRUE) > 0), .groups = "drop") |>
  group_by(rule, B, n) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rule, B)
```

### Realised false discovery proportion (k = 3)

Among items flagged, the proportion that were in fact fitting. This is what BH
targets.

```{r s3fdp}
pv_long |>
  filter(k == 3, rule %in% c("fwer", "bh")) |>
  group_by(rule, B, n, rep) |>
  summarise(flagged = sum(flag, na.rm = TRUE),
            false = sum(flag[!is_misfit], na.rm = TRUE), .groups = "drop") |>
  filter(flagged > 0) |>
  group_by(rule, B, n) |>
  summarise(fdp = round(100 * mean(false / flagged), 1), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = fdp, names_prefix = "n=") |>
  arrange(rule, B)
```

### The BH reachability floor

The smallest attainable BH-adjusted p-value is `m / (B + 1)` for *m*
comparisons, so BH cannot flag a single item unless `B >= m/alpha - 1`. With
m = 20 and alpha = .05 that is B >= 399. FWER's studentised-max p-value has no
such floor — it needs only `B >= 1/alpha - 1 = 19`.

```{r s3floor}
data.frame(B = B2) |>
  mutate(marginal_floor = round(1 / (B + 1), 5),
         bh_min_adj_p = round(20 / (B + 1), 4),
         bh_can_flag_one_item = 20 / (B + 1) < ALPHA)
```

Observed minima, as a check:

```{r s3floorobs}
pval2 |>
  group_by(B) |>
  summarise(min_p_marg = round(min(p_marg, na.rm = TRUE), 5),
            min_padj_fwer = round(min(padj_fwer, na.rm = TRUE), 5),
            min_padj_bh = round(min(padj_bh, na.rm = TRUE), 5),
            .groups = "drop")
```

```{r s3fig, fig.width = 9, fig.height = 4.5}
pv_long |>
  filter(k == 3) |>
  mutate(type = ifelse(is_misfit, "Detection", "False positives")) |>
  group_by(rule, B, n, type) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(factor(B), v, colour = rule, group = rule)) +
  geom_line() + geom_point() +
  facet_grid(type ~ n, scales = "free_y", labeller = label_both) +
  labs(x = "Bootstrap iterations (B)", y = "Percent", colour = "Rule") +
  theme_minimal(base_size = 10)
```

## Notes

Self-contained apart from exported easyRasch2 functions; relative paths
throughout, so the whole set can be relocated to a reproducibility folder
without modification.
