---
title: "Decision stability under reseeding, at realistic misfit"
subtitle: "How often does the same data give a different answer?"
author: "Magnus Johansson"
date: today
---

## Scope

Error rates averaged over datasets can look fine while individual decisions
remain unstable. This study holds the **dataset fixed** and varies only the
bootstrap seed, isolating Monte-Carlo noise from sampling noise: if an analyst
re-runs `RMitemInfitCutoff()` on the same data with a different seed, how often
does the flag change?

An earlier version of this exists in `infit_fwer_iterations.qmd` (Study B), but
it has two limitations that this run addresses:

1. It used **dichotomous** items at `rho = 0.15` — misfit roughly 60 percent
   stronger than anything observed in real phq9 data, which we now know
   saturates detection.
2. It covered only n in {250, 500} and B <= 1000.

This matters because the current recommendation of **B >= 1000 for FWER rests
entirely on that single measurement** — stability, not calibration, is what
drives it. Calibration is already nominal from B ~= 400.

### Prediction being tested

Instability should be **worse** at realistic misfit, because when the signal is
weaker more items sit near the decision boundary, and a boundary-adjacent item
is exactly the one a different seed can push either way. Three levels are
included so this is a dose-response test rather than a two-point comparison:
`rho = 0.15` (the unrealistically strong condition used in the earlier study),
`rho = 0.50` (intermediate), and `rho = 0.85` (calibrated to real phq9 misfit,
injected infit ~1.25 against q3 = 1.23 and q8 = 1.26).

### Design

- Same two arms as `infit_width_iterations_poly_weak.qmd`: **sim15** (15
  stylised polytomous items) and **phq9** (9 items, CML parameters from the
  packaged real responses)
- `rho` in {0.15, 0.50, 0.85} for the misfit conditions; k = 0 has no `rho`
- Misfitting items: the maximum per arm (3 for sim15, 2 for phq9), i.e. the
  strongest cascade, plus a k = 0 null baseline
- n in {250, 1000}
- **3 fixed datasets** per condition, each analysed with **25 bootstrap seeds**
  (seeds resolve instability better than datasets do — an item flagged with
  probability 0.1 is caught as non-unanimous 93 percent of the time with 25
  seeds, only 79 percent with 15)
- B in {100, 400, 1000, 2000}; widths {.90, .95, .99, .995, .999} read off the
  same draws
- FWER and BH p-values recorded alongside

Outcomes: the share of items whose flag is not unanimous across seeds, the
probability that two randomly chosen seeds disagree, and the spread of the
adjusted p-value and interval bounds.

```{r setup}
suppressMessages({
  pkgload::load_all("..", quiet = TRUE)
  library(ggdist)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

WIDTHS   <- c(0.90, 0.95, 0.99, 0.995, 0.999)
NS       <- c(250, 1000)
BS       <- c(100, 400, 1000, 2000)
RHOS     <- c(0.15, 0.50, 0.85)
N_DATA   <- 6     # raised from 3 on 2026-07-31; datasets 4-6 resume
N_SEEDS  <- 25    # bootstrap seeds per dataset
NCORES   <- 10
ALPHA    <- 0.05
CHUNK    <- 30    # frequent checkpoints: this run may be stopped mid-way
RES      <- "infit_decision_stability.rds"
```

```{r runner}
run_resumable <- function(res_file, grid, keys, fn,
                          ncores = NCORES, chunk_size = CHUNK,
                          extra = list()) {
  key_of <- function(g) {
    if (nrow(g) == 0L) return(character(0))
    do.call(paste, c(unname(as.list(g[keys])), sep = "|"))
  }
  acc <- if (file.exists(res_file)) readRDS(res_file) else
    c(list(res = list(), grid = grid[0, , drop = FALSE], elapsed_min = 0), extra)
  todo <- which(!(key_of(grid) %in% key_of(acc$grid)))
  if (length(todo) == 0L) {
    message("cached: ", nrow(acc$grid), " cells, nothing to run"); return(acc)
  }
  message("running ", length(todo), " new cells (", nrow(acc$grid), " cached)")
  for (s in seq(1, length(todo), by = chunk_size)) {
    idx <- todo[s:min(s + chunk_size - 1L, length(todo))]
    t0 <- Sys.time()
    out <- mclapply(idx, function(i) fn(grid[i, , drop = FALSE]), mc.cores = ncores)
    acc$res <- c(acc$res, out)
    acc$grid <- rbind(acc$grid, grid[idx, , drop = FALSE])
    acc$elapsed_min <- acc$elapsed_min + as.numeric(Sys.time() - t0, "mins")
    acc <- utils::modifyList(acc, extra)
    saveRDS(acc, res_file)
    message("  checkpoint ", nrow(acc$grid), " / ", nrow(grid),
            "  (", round(acc$elapsed_min, 1), " min)")
  }
  acc
}
```

### Arms (identical to the poly_weak study)

```{r arms}
S15_LOCS <- seq(-2, 2, length.out = 15)
S15_THR  <- lapply(S15_LOCS, function(L) L + seq(-1, 1, length.out = 3))
S15_MIS  <- vapply(c(0, -1, -2), function(tg) which.min(abs(S15_LOCS - tg)),
                   integer(1))

load("../data/phq9.rda")
PH_OBS <- as.matrix(phq9[, paste0("q", 1:9)])
PH_OBS <- PH_OBS[complete.cases(PH_OBS), ]
PH_THR  <- .fit_cml_thresholds(PH_OBS)
PH_LOCS <- vapply(PH_THR, mean, numeric(1))
.g <- seq(-6, 6, length.out = 81)
PH_SD <- .estimate_prior_sd(
  .grid_loglik(PH_OBS, .logp_tables(PH_THR, .g), .g), .g, 0)
PH_MIS <- c(5L, 8L)

ARMS <- list(
  sim15 = list(thr = S15_THR, sd = 1.5,   mis = S15_MIS, locs = S15_LOCS,
               nm = paste0("V", 1:15), kmax = 3L),
  phq9  = list(thr = PH_THR,  sd = PH_SD, mis = PH_MIS,  locs = PH_LOCS,
               nm = paste0("q", 1:9),  kmax = 2L)
)

sim_arm <- function(arm, n, k, rho, seed) {
  a <- ARMS[[arm]]
  set.seed(seed)
  S <- matrix(c(a$sd^2, rho * a$sd^2, rho * a$sd^2, a$sd^2), 2, 2)
  th <- MASS::mvrnorm(n, mu = c(0, 0), Sigma = S)
  ni <- length(a$thr); mis <- a$mis[seq_len(k)]
  d <- matrix(0L, n, ni); fit_items <- setdiff(seq_len(ni), mis)
  if (length(fit_items)) d[, fit_items] <- sim_partial_score(a$thr[fit_items], th[, 1])
  if (length(mis))       d[, mis]       <- sim_partial_score(a$thr[mis], th[, 2])
  d <- as.data.frame(d); names(d) <- a$nm; d
}

bounds_at <- function(results, width, items) {
  vapply(items, function(it) {
    v <- results$InfitMSQ[results$Item == it]
    as.numeric(ggdist::hdci(v, .width = width))[1:2]
  }, numeric(2))
}
flag_code <- function(msq, lo, hi) ifelse(msq > hi, 2L, ifelse(msq < lo, 1L, 0L))
```

### Run

Each cell is one `(dataset, seed)` pair: the dataset is regenerated from a seed
that depends only on `(arm, k, n, rho, dataset)`, so it is **identical across
all bootstrap seeds**, while the bootstrap seed varies.

```{r run}
one_cell <- function(g) {
  out <- try({
    arm <- as.character(g$arm); k <- g$k; n <- g$n; rho <- g$rho
    dat <- g$dataset; sd_seed <- g$seed
    d <- sim_arm(arm, n, k, rho,
                 seed = 5e8 + 1e7 * match(arm, names(ARMS)) + 1e6 * k +
                   1e4 * round(rho * 100) + 1e3 * n + dat)
    items <- names(d)
    obs <- RMitemInfit(d, output = "dataframe")
    msq <- obs$Infit_MSQ[match(items, obs$Item)]
    hd <- list(); pv <- list()
    for (j in seq_along(BS)) {
      co <- try(RMitemInfitCutoff(d, iterations = BS[j], parallel = FALSE,
                                  seed = sd_seed * 1000 + j, verbose = FALSE),
                silent = TRUE)
      if (inherits(co, "try-error")) next
      hd[[length(hd) + 1L]] <- do.call(rbind, lapply(WIDTHS, function(w) {
        b <- bounds_at(co$results, w, items)
        data.frame(B = BS[j], w = w, item = seq_along(items),
                   code = flag_code(msq, b[1, ], b[2, ]),
                   lo = b[1, ], hi = b[2, ])
      }))
      fw <- try(suppressWarnings(
        RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fwer",
                    alpha = ALPHA, output = "dataframe")), silent = TRUE)
      bh <- try(suppressWarnings(
        RMitemInfit(d, cutoff = co, p_value = TRUE, correction = "fdr_bh",
                    alpha = ALPHA, output = "dataframe")), silent = TRUE)
      if (!inherits(fw, "try-error") && !inherits(bh, "try-error")) {
        i1 <- match(items, fw$Item); i2 <- match(items, bh$Item)
        pv[[length(pv) + 1L]] <- data.frame(
          B = BS[j], item = seq_along(items),
          padj_fwer = fw$padj_infit[i1], padj_bh = bh$padj_infit[i2])
      }
    }
    list(hdci = do.call(rbind, hd), pval = do.call(rbind, pv))
  }, silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

blocks <- list()
for (a in names(ARMS)) {
  # null baseline: rho is inert with no misfitting items
  blocks[[length(blocks) + 1L]] <- expand.grid(
    arm = a, k = 0L, rho = 0, n = NS, dataset = 1:N_DATA, seed = 1:N_SEEDS,
    stringsAsFactors = FALSE)
  blocks[[length(blocks) + 1L]] <- expand.grid(
    arm = a, k = ARMS[[a]]$kmax, rho = RHOS, n = NS,
    dataset = 1:N_DATA, seed = 1:N_SEEDS, stringsAsFactors = FALSE)
}
grid <- do.call(rbind, blocks)[, c("arm", "k", "rho", "n", "dataset", "seed")]

# Explicit priority order, so a run stopped early still answers the main
# question. Blocks cost roughly the same (~2.3 h each on 10 cores):
#   1. rho = 0.85  -- the calibrated, realistic condition (the point of the run)
#   2. k = 0       -- instability under a true model, the reference baseline
#   3. rho = 0.15  -- the old unrealistically strong condition, for contrast
#   4. rho = 0.50  -- intermediate; fills in the dose-response, least critical
prio <- with(grid, ifelse(k == 0L, 2L,
                   ifelse(rho == 0.85, 1L,
                   ifelse(rho == 0.15, 3L, 4L))))
grid <- grid[order(prio, grid$n), ]
rownames(grid) <- NULL
cat("cells to run:", nrow(grid), "\n")
print(table(priority = sort(prio)))

s <- run_resumable(RES, grid, c("arm", "k", "rho", "n", "dataset", "seed"),
                   one_cell,
                   extra = list(arms = lapply(ARMS, function(a)
                     a[c("mis", "locs", "sd", "nm", "kmax")])))

bind_part <- function(part) {
  do.call(rbind, lapply(seq_len(nrow(s$grid)), function(i) {
    r <- s$res[[i]][[part]]
    if (is.null(r)) return(NULL)
    cbind(s$grid[i, c("arm", "k", "rho", "n", "dataset", "seed")], r)
  }))
}
H <- bind_part("hdci") |> mutate(flag = as.integer(code > 0),
                                 w = sprintf("%.3f", w),
                                 rho = sprintf("%.2f", rho))
P <- bind_part("pval") |> mutate(rho = sprintf("%.2f", rho))
cat(nrow(s$grid), "cells |", round(s$elapsed_min, 1), "min |",
    "failed:", sum(sapply(s$res, is.null)), "\n")
```

## Instability of the HDCI rule

Share of items whose flag is **not unanimous** across the 25 seeds, averaged
over the 5 datasets.

```{r hdci_unstable}
H |>
  group_by(arm, rho, k, n, w, B, dataset, item) |>
  summarise(rate = mean(flag), .groups = "drop") |>
  group_by(arm, rho, k, n, w, B) |>
  summarise(unstable = round(100 * mean(rate > 0 & rate < 1), 1), .groups = "drop") |>
  filter(w == "0.999") |>
  pivot_wider(names_from = B, values_from = unstable, names_prefix = "B=") |>
  arrange(arm, rho, n)
```

## Instability of the FWER and BH rules

```{r p_unstable}
P |>
  mutate(fwer = as.integer(padj_fwer < ALPHA),
         bh = as.integer(padj_bh < ALPHA)) |>
  pivot_longer(c(fwer, bh), names_to = "rule", values_to = "flag") |>
  group_by(arm, rho, k, n, rule, B, dataset, item) |>
  summarise(rate = mean(flag, na.rm = TRUE), .groups = "drop") |>
  group_by(arm, rho, k, n, rule, B) |>
  summarise(unstable = round(100 * mean(rate > 0 & rate < 1), 1), .groups = "drop") |>
  pivot_wider(names_from = B, values_from = unstable, names_prefix = "B=") |>
  arrange(rule, arm, rho, n)
```

## Probability that two analysts disagree

For an item flagged with probability *p* across seeds, two independently seeded
runs disagree with probability `2p(1-p)`. Averaged over items, this is the
chance that two analysts running the same data reach a different conclusion
about at least one item.

```{r disagree}
P |>
  mutate(fwer = as.integer(padj_fwer < ALPHA)) |>
  group_by(arm, rho, k, n, B, dataset, item) |>
  summarise(p = mean(fwer, na.rm = TRUE), .groups = "drop") |>
  mutate(disagree = 2 * p * (1 - p)) |>
  group_by(arm, rho, k, n, B) |>
  summarise(any_item = round(100 * (1 - prod(1 - disagree)^(1 / n_distinct(dataset))), 1),
            per_item = round(100 * mean(disagree), 2), .groups = "drop") |>
  pivot_wider(names_from = B, values_from = c(per_item, any_item)) |>
  arrange(arm, rho, n)
```

## Spread of the adjusted p-value

```{r sd_padj}
P |>
  group_by(arm, rho, k, n, B, dataset, item) |>
  summarise(sd_p = sd(padj_fwer, na.rm = TRUE), .groups = "drop") |>
  group_by(arm, rho, k, n, B) |>
  summarise(median_sd = round(median(sd_p, na.rm = TRUE), 4), .groups = "drop") |>
  pivot_wider(names_from = B, values_from = median_sd, names_prefix = "B=") |>
  arrange(arm, rho, n)
```

## Spread of the HDCI bounds

```{r sd_bounds}
H |>
  filter(w == "0.999") |>
  group_by(arm, rho, k, n, B, dataset, item) |>
  summarise(sd_hi = sd(hi), .groups = "drop") |>
  group_by(arm, rho, k, n, B) |>
  summarise(median_sd_upper = round(median(sd_hi), 4), .groups = "drop") |>
  pivot_wider(names_from = B, values_from = median_sd_upper, names_prefix = "B=") |>
  arrange(arm, rho, n)
```

```{r fig, fig.width = 9, fig.height = 5.5}
P |>
  mutate(fwer = as.integer(padj_fwer < ALPHA)) |>
  group_by(arm, rho, k, n, B, dataset, item) |>
  summarise(rate = mean(fwer, na.rm = TRUE), .groups = "drop") |>
  group_by(arm, rho, n, B) |>
  summarise(unstable = 100 * mean(rate > 0 & rate < 1), .groups = "drop") |>
  ggplot(aes(factor(B), unstable, colour = rho, group = rho)) +
  geom_line() + geom_point() +
  facet_grid(arm ~ n, labeller = label_both) +
  labs(x = "Bootstrap iterations (B)",
       y = "% of items with a non-unanimous FWER flag",
       colour = "rho") +
  theme_minimal(base_size = 10)
```
