---
title: "Q3 cutoffs under partial missingness"
subtitle: "Null simulation of the resample DGP before and after the missingness fix"
author: "Magnus Johansson"
date: today
---

## Purpose

`RMlocdepQ3()` computes the observed $Q_3$ as a pairwise-complete correlation
of residuals, so with partial missingness each item pair rests on the
respondents who answered both items. Up to the development version after
1.3.1, `RMlocdepQ3Cutoff()` with the default `dgp = "resample"` simulated
**complete** data for every respondent. Each simulated pair therefore had more
respondents than the observed pair, and the null distribution was too narrow.

The fix makes each resampled respondent keep their missingness pattern. This
study measures what the old behaviour cost and whether the fix restores the
nominal rate.

Two things should differ between the arms:

1. **Spread.** The SD of a correlation grows roughly with $1/\sqrt{n}$, so at
   30 percent missingness per cell (about half of each pair's respondents
   left) the observed $Q_3$ varies about 1.4 times as much as the old null
   assumed. As a rough estimate, a one-sided 5 percent test then rejects about
   13 percent of null pairs.
2. **Location.** Respondents with missing items get WLE estimates from fewer
   items, so their residuals carry a stronger negative part-whole bias. The
   old null did not reproduce that shift in the mean $Q_3$.

## Design

- **Items.** 9 polytomous items with four categories, the item set from the
  item-restscore studies. θ ~ N(0, 1.5²), n = 300.
- **Missingness patterns.**
  - `none`: complete data, the reference. The old and new resample arms are
    identical here by construction, so only the new arm is run.
  - `mcar10`, `mcar30`: each cell missing completely at random with
    probability .10 or .30.
  - `item50`: item I5 missing for a random half of the respondents, all other
    items complete. Only the 8 pairs involving I5 are affected, which tests
    whether the fix works per pair and not just on average.
- **Arms.**
  - `old`: the resample DGP without the missingness mask, emulated from the
    package internals with the same seeds as `new`, so the comparison is
    paired.
  - `new`: `RMlocdepQ3Cutoff(dgp = "resample")` as it now ships.
  - `conditional`: `RMlocdepQ3Cutoff(dgp = "conditional")`, which always kept
    the observed missingness pattern.
- **Iterations.** B = 200 per dataset. The marginal per-pair p-value is the
  main outcome, and it is well resolved at 200 (smallest value 1/201).
- **Replications.** 200 datasets per pattern.
- **Outcomes.**
  - Per-pair marginal rejection, one-sided upper at α = .05, as
    `RMlocdepQ3()` computes it. Nominal 5 percent.
  - The same for the I5 pairs and the other pairs separately under `item50`.
  - Family-wise rate with the Westfall-Young step-down over the 36 pairs.
  - The global rule, observed max minus mean $Q_3$ above the 95th percentile
    of its simulated distribution.

**Cost.** About 0.3 s per iteration at n = 300 (CML), so 60 s per dataset and
arm. 10 pattern-arm combinations × 200 datasets is about 35 CPU-hours, or
3.5 to 4 hours on 10 cores. Results are saved per pattern and arm, so an
interrupted render resumes.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

R_DATASETS <- 200
B          <- 200
N          <- 300
THETA_SD   <- 1.5
NCORES     <- 10
RES_DIR    <- "q3_missingness_null"
dir.create(RES_DIR, showWarnings = FALSE)

thr <- lapply(seq(-1.5, 1.5, length.out = 9), function(l) l + c(-1.2, 0, 1.2))
item_names <- paste0("I", seq_along(thr))

design <- data.frame(
  pattern = c("none", rep(c("mcar10", "mcar30", "item50"), each = 3)),
  arm = c("new", rep(c("old", "new", "conditional"), 3)),
  stringsAsFactors = FALSE
)
design
```

### Data generation

```{r gen}
sim_pcm <- function(theta, thr) {
  n <- length(theta)
  sapply(thr, function(d) {
    m <- length(d)
    eta <- outer(theta, 0:m) -
      matrix(c(0, cumsum(d)), n, m + 1, byrow = TRUE)
    p <- exp(eta - apply(eta, 1, max))
    p <- p / rowSums(p)
    cp <- p %*% upper.tri(diag(m + 1), diag = TRUE)
    rowSums(runif(n) > cp)
  })
}

# Complete data with every category observed, then the missingness pattern.
# The dataset depends only on (pattern, rep), so all arms see the same data.
make_data <- function(pattern, rep) {
  set.seed(1e5 * match(pattern, unique(design$pattern)) + rep)
  repeat {
    d <- sim_pcm(rnorm(N, 0, THETA_SD), thr)
    if (all(apply(d, 2, function(x) all(tabulate(x + 1, 4) > 0)))) break
  }
  colnames(d) <- item_names
  if (pattern %in% c("mcar10", "mcar30")) {
    rate <- if (pattern == "mcar10") 0.10 else 0.30
    d[matrix(runif(length(d)) < rate, nrow(d))] <- NA
  } else if (pattern == "item50") {
    d[sample(N, N / 2), "I5"] <- NA
  }
  d <- d[rowSums(!is.na(d)) > 0, , drop = FALSE]
  as.data.frame(d)
}
```

### The old resample DGP

This rebuilds the simulation list the way `RMlocdepQ3Cutoff()` does, with the
same per-iteration seeds, but without the missingness mask. It reproduces the
pre-fix code path, which generated complete data.

```{r oldarm}
old_cutoff <- function(d, B, seed) {
  dm <- as.matrix(d)
  pool <- .wle_theta_pool(dm)
  dl <- list(
    dgp = "resample",
    type = "polytomous",
    thr_list = pool$thr_list,
    n_items = ncol(dm),
    sample_n = nrow(dm),
    item_names = colnames(dm),
    estimator = "CML",
    thetas = pool$thetas,
    deltaslist = pool$thr_list
  )
  set.seed(seed)
  seeds <- sample.int(.Machine$integer.max, B)
  raw <- lapply(seeds, run_single_q3_sim, data_list = dl)
  ok <- Filter(is.list, raw)
  list(
    results = data.frame(
      mean = vapply(ok, `[[`, numeric(1), "mean"),
      max = vapply(ok, `[[`, numeric(1), "max")
    ),
    pair_results = do.call(rbind, lapply(seq_along(ok), function(i) {
      cbind(ok[[i]]$pair_q3, iteration = i)
    })),
    actual_iterations = length(ok)
  )
}
```

The emulation is checked against the package on complete data, where the two
must agree exactly.

```{r oldcheck}
d0 <- make_data("none", 1)
pkg <- suppressMessages(RMlocdepQ3Cutoff(d0, iterations = 5, parallel = FALSE,
                                          seed = 11))
emu <- old_cutoff(d0, 5, 11)
stopifnot(isTRUE(all.equal(pkg$pair_results$Q3, emu$pair_results$Q3)))
"old-arm emulation matches the package on complete data"
```

### One dataset

The p-values are computed the way `RMlocdepQ3()` computes them: the observed
pairwise-complete $Q_3$ against each pair's simulated distribution, one-sided
upper, studentised by the bootstrap mean and SD.

```{r onedataset}
one_dataset <- function(rep, pattern, arm) {
  d <- make_data(pattern, rep)
  seed <- 1e5 * match(pattern, unique(design$pattern)) + rep
  co <- tryCatch({
    if (arm == "old") {
      old_cutoff(d, B, seed)
    } else {
      suppressMessages(RMlocdepQ3Cutoff(
        d, iterations = B, parallel = FALSE, seed = seed,
        cutoff_method = "quantile", dgp = if (arm == "new") "resample" else arm
      ))
    }
  }, error = function(e) NULL)
  if (is.null(co)) return(NULL)

  obs <- .q3_residual_matrix(d, estimator = "CML")
  pr <- co$pair_results
  pr$key <- paste(pr$Item1, pr$Item2, sep = "___")
  sim_mat <- tapply(pr$Q3, list(pr$iteration, pr$key), function(x) x[1L])
  pairs <- unique(pr[, c("Item1", "Item2", "key")])
  observed <- stats::setNames(
    mapply(function(a, b) obs[a, b], pairs$Item1, pairs$Item2),
    pairs$key
  )
  p_none <- .bootstrap_pvalues(observed, sim_mat, "none", tail = "upper")
  p_wy <- .bootstrap_pvalues(observed, sim_mat, "fwer", tail = "upper")

  obs_diff <- max(obs, na.rm = TRUE) - mean(obs, na.rm = TRUE)
  sim_diff <- co$results$max - co$results$mean

  data.frame(
    rep = rep,
    Item1 = pairs$Item1,
    Item2 = pairs$Item2,
    observed = as.numeric(observed),
    null_mean = colMeans(sim_mat, na.rm = TRUE)[pairs$key],
    null_sd = apply(sim_mat, 2, stats::sd, na.rm = TRUE)[pairs$key],
    p = p_none$p[match(pairs$key, p_none$name)],
    padj_wy = p_wy$padj[match(pairs$key, p_wy$name)],
    global_flag = obs_diff > stats::quantile(sim_diff, 0.95),
    iterations = co$actual_iterations,
    n = nrow(d),
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}
```

### Run

```{r run}
for (i in seq_len(nrow(design))) {
  f <- file.path(RES_DIR, paste0(design$pattern[i], "_", design$arm[i], ".rds"))
  if (file.exists(f)) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS),
                  function(r) one_dataset(r, design$pattern[i], design$arm[i]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  saveRDS(list(res = res,
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          f)
}

long <- do.call(rbind, lapply(seq_len(nrow(design)), function(i) {
  f <- file.path(RES_DIR, paste0(design$pattern[i], "_", design$arm[i], ".rds"))
  x <- readRDS(f)
  ok <- Filter(Negate(is.null), x$res)
  cbind(pattern = design$pattern[i], arm = design$arm[i],
        do.call(rbind, ok), elapsed_min = x$elapsed_min,
        failed = length(x$res) - length(ok))
}))
long$pattern <- factor(long$pattern, levels = unique(design$pattern))
long$arm <- factor(long$arm, levels = c("old", "new", "conditional"))
long$i5_pair <- long$Item1 == "I5" | long$Item2 == "I5"

long |>
  group_by(pattern, arm) |>
  summarise(datasets = n_distinct(rep),
            failed = first(failed),
            median_iterations = median(iterations[!duplicated(rep)]),
            minutes = round(first(elapsed_min), 1),
            .groups = "drop")
```

## Results

### Per-pair rejection

Marginal one-sided p-value below .05. Nominal 5 percent.

```{r perpair}
long |>
  group_by(pattern, arm) |>
  summarise(reject = round(100 * mean(p < .05), 2), .groups = "drop") |>
  pivot_wider(names_from = arm, values_from = reject)
```

Under `item50`, the pairs involving I5 and the other pairs separately:

```{r item50}
long |>
  filter(pattern == "item50") |>
  mutate(pairs = ifelse(i5_pair, "with I5", "without I5")) |>
  group_by(pairs, arm) |>
  summarise(reject = round(100 * mean(p < .05), 2), .groups = "drop") |>
  pivot_wider(names_from = arm, values_from = reject)
```

### Spread and location of the null

`sd_ratio` is the empirical SD of the observed $Q_3$ across datasets divided by
the average simulated SD, per pair, averaged over pairs. `mean_gap` is the
average observed $Q_3$ minus the simulated mean. A calibrated null gives 1 and
0.

```{r spread}
long |>
  group_by(pattern, arm, Item1, Item2) |>
  summarise(sd_ratio = sd(observed) / mean(null_sd),
            mean_gap = mean(observed - null_mean),
            .groups = "drop") |>
  group_by(pattern, arm) |>
  summarise(sd_ratio = round(mean(sd_ratio), 3),
            mean_gap = round(mean(mean_gap), 4),
            .groups = "drop")
```

### Family-wise rates

At least one pair flagged by the Westfall-Young step-down, and the global
max minus mean rule at its 95th percentile. Both nominal 5 percent. The Monte
Carlo SE at 200 datasets is 1.5 points.

```{r fwer}
long |>
  group_by(pattern, arm, rep) |>
  summarise(wy = any(padj_wy < .05), global = first(global_flag),
            .groups = "drop") |>
  group_by(pattern, arm) |>
  summarise(wy = round(100 * mean(wy), 1),
            global = round(100 * mean(global), 1),
            .groups = "drop")
```

### Distribution of p-values

```{r phist, fig.height = 6}
long |>
  ggplot(aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "steelblue",
                 colour = "white") +
  facet_grid(arm ~ pattern, scales = "free_y") +
  labs(x = "Marginal one-sided p-value", y = "Count") +
  theme_minimal()
```

### Both tails and the shift in null SD units

The per-pair rejection above uses the upper tail only. `upper_05` repeats
it, `lower_05` is the share of p-values above .95 (the lower tail), and
`gap_sd` is the average of (observed minus simulated mean) divided by the
simulated SD, so the location error is on the same scale as the test.

```{r tails}
long |>
  group_by(pattern, arm) |>
  summarise(upper_05 = round(100 * mean(p < .05), 2),
            lower_05 = round(100 * mean(p > .95), 2),
            gap_sd = round(mean((observed - null_mean) / null_sd), 3),
            .groups = "drop")

long |>
  filter(pattern == "item50") |>
  mutate(pairs = ifelse(i5_pair, "with I5", "without I5")) |>
  group_by(arm, pairs, Item1, Item2) |>
  summarise(sd_ratio = sd(observed) / mean(null_sd),
            gap_sd = mean((observed - null_mean) / null_sd),
            .groups = "drop") |>
  group_by(arm, pairs) |>
  summarise(sd_ratio = round(mean(sd_ratio), 3),
            gap_sd = round(mean(gap_sd), 3), .groups = "drop")
```

## Interpretation

**1. The fix restores the nominal rate.** With the missingness mask, the
resample DGP rejected 4.6 to 4.9 percent of null pairs at α = .05 under all
three missingness patterns, against 4.5 percent with complete data. The
simulated null has the right spread (SD ratio 0.99 to 1.00) and the right
centre (mean gap below 0.02 null SDs). The Westfall-Young family-wise rate
was 5.5 to 7.0 percent and the global max-minus-mean rule 3.5 to 5.0 percent,
within Monte Carlo error of 5 percent (SE 1.5 points at 200 datasets). The
conditional DGP, which always kept the observed missingness, gives the same
picture.

**2. The old null was wrong in two ways that cancel in the per-pair test.**
At 30 percent MCAR the old null was too narrow (SD ratio 1.42) and centred
too high (the observed $Q_3$ fell 0.82 null SDs below the simulated mean on
average). The first error pushes the upper-tail rejection up and the second
pushes it down, and at α = .05 they cancelled to 4.3 percent. The lower tail
shows what was hidden: 29.7 percent of p-values fell above .95. The
family-wise rules do not cancel, because the maximum over 36 pairs is driven
by spread and the max-minus-mean statistic largely removes the shift. Under the old
null, WY flagged 16.5 percent of datasets and the global rule 47.0 percent.
At 10 percent MCAR the same pattern is weaker (WY 8.5, global 15.0 percent).

**3. The shift comes from the respondents with missing items.** Under
`item50` the two errors separate. Pairs involving I5 got the narrow null
(SD ratio 1.41) without a shift and rejected 12.25 percent, close to the 13
percent predicted in the Purpose section from spread alone. The other pairs
got the shift without the extra spread (−0.16 null SDs) and rejected 3.4
percent. Respondents who skipped I5 have WLE estimates based on eight items,
so their residuals on the other items carry a stronger negative part-whole
bias, which lowers $Q_3$ for pairs that do not include I5. The old null, built
on complete data, had neither feature.

**4. The fix is slower with many missingness patterns.** For the same 200
datasets and 200 iterations, the new resample arm took 21.7 minutes at
10 percent MCAR and 51.8 at 30 percent, against 2.9 and 3.3 for the old
arm. Under `item50`, with only two patterns, it took 3.8 minutes against 2.7.
The cause has not been measured. The likely one is the CML refit, which is
fast on complete data and slows with the number of distinct missingness
patterns.

**Implications.** The bug fix in NEWS is warranted, and the size of the old
error (a global rule rejecting in nearly half of null datasets at 30 percent
MCAR) is worth stating in the entry. Users with substantial MCAR missingness
will see a several-fold longer run time, which may need a sentence in the
help page.
