---
title: "An exact null for the Rasch change index"
subtitle: "Why the RCI critical value is not 1.96, and how to compute it without simulating"
author: "Magnus Johansson"
date: today
bibliography: rci-refs.bib
---

```{r setup}
#| include: false
if (requireNamespace("devtools", quietly = TRUE) && file.exists("../DESCRIPTION")) {
  suppressMessages(devtools::load_all("..", quiet = TRUE))
} else {
  library(easyRasch2)
}
library(knitr)
```

## Purpose

`RMpersonChange()` includes an option for `critical = "simulate"` rather than the conventional
1.96, because the Rasch change index

$$RCI = \frac{\hat\theta_2 - \hat\theta_1}{\sqrt{SE(\hat\theta_1)^2 + SE(\hat\theta_2)^2}}$$

is not standard normal under the null. The departure is largest on short
scales and fades only slowly, so that even at 60 items with four response
categories the critical value is still around 1.93. This note shows
that the null does not need simulating at all: for complete data it can be
enumerated exactly. It also records what the exact critical value does across
test length, sample skew and targeting, and one consequence for the RCI as a
statistic.

## Why the null is not N(0, 1)

Two mechanisms, both weakening as test information grows.

**Self-normalisation.** $SE(\hat\theta_1)$ and $SE(\hat\theta_2)$ are
deterministic functions of the realised scores, so the denominator is not
independent of the numerator. A respondent who moves to an extreme score
produces a large numerator and a large denominator together, and the ratio is
damped. Both tails of the null are pulled in, so the critical value falls
**below** 1.96. In a t-test the denominator is independent of the numerator.
Here it is not, which is why the t reference does not help either.

**Discreteness.** With complete data the sum score is sufficient, so
$\hat\theta$ takes only as many values as there are scores and the null is
discrete rather than continuous.

## Exact enumeration

Sufficiency is what makes this tractable. $\hat\theta(r)$ and $SE(r)$ are
deterministic lookups, the ones `RMscoreSE()` builds, and $P(r \mid \theta)$
follows from the Lord-Wingersky recursion. Under the null the two occasions are
independent given $\theta$, so the joint is $P(r_1 \mid \theta)P(r_2 \mid
\theta)$ and the whole null is a discrete distribution over $(k+1)^2$ score
pairs. For six items with four response categories that is 361 cells.

The two ingredients do different jobs and the enumeration needs both.
Sufficiency fixes the **value** of the statistic in every cell, because
$\hat\theta$ and $SE$ are functions of the score alone, and no probability
enters that map. A quantile also needs the **weight** on each cell, which is
$P(r \mid \theta)$, and that is not deterministic. Lord-Wingersky computes it
rather than estimating it: the recursion is the elementary symmetric function
identity in probability form, the same object CML estimation runs on, and it
replaces an enumeration over $\prod_i (m_i + 1)$ response patterns, 4096 here
and around $10^{12}$ at twenty items. Nothing in this note is simulated, which
is what "exact" refers to. Nor is there a closed-form expression for the
critical value. It is read off the enumerated distribution as a quantile.

```{r exact-fns}
# One threshold constructor for the whole note: k items, ncat response
# categories, item locations spread over [-1.2, 1.2].
mk <- function(k, ncat) {
  lapply(seq(-1.2, 1.2, length.out = k),
         function(b) b + seq(-0.9, 0.9, length.out = ncat - 1L))
}

# Lord-Wingersky: exact sum-score distribution given theta under a PCM
score_dist <- function(thr_list, theta) {
  d <- 1
  for (thr in thr_list) {
    p <- easyRasch2:::.pcm_cat_probs(theta, thr)
    nd <- numeric(length(d) + length(p) - 1L)
    for (a in seq_along(d)) {
      nd[a + seq_along(p) - 1L] <- nd[a + seq_along(p) - 1L] + d[a] * p
    }
    d <- nd
  }
  d                                   # index is score + 1
}

# Exact null distribution of the RCI, pooled over a set of person locations
exact_crit <- function(thr_list, thetas, alpha = 0.05) {
  steps <- vapply(thr_list, length, integer(1L))
  lut <- vapply(0:sum(steps), function(r) {
    easyRasch2:::.theta_wle(
      easyRasch2:::.score_pattern(r, steps), thr_list, c(-10, 10)
    )
  }, numeric(2L))
  th <- lut[1L, ]
  se <- lut[2L, ]
  rci <- outer(seq_along(th), seq_along(th),
               function(i, j) (th[j] - th[i]) / sqrt(se[i]^2 + se[j]^2))
  w <- matrix(0, length(th), length(th))
  for (t in thetas) {
    d <- score_dist(thr_list, t)
    w <- w + outer(d, d)
  }
  w <- w / sum(w)
  o <- order(as.numeric(rci))
  v <- as.numeric(rci)[o]
  cw <- cumsum(as.numeric(w)[o])
  c(lower = v[which(cw >= alpha / 2)[1L]],
    upper = v[which(cw >= 1 - alpha / 2)[1L]])
}
```

## Validation against simulation

Run by `dev/` scratch scripts at `sim_iter` 2000 to 4000, n = 200, alpha = .05
two-sided. The simulated and exact values agree to within Monte Carlo error in
every configuration tried, including dichotomous items, items with unequal
category counts, a right-skewed person distribution and a badly off-target one.
Each configuration was run on its own simulated sample, so the values are not
comparable with the length table below, which uses a different person
distribution. The 12-item row sitting near the 20-item value there is an
instance of that, not a property of test length.

| configuration | simulated | exact | difference |
|---|---|---|---|
| 6 items, 4 categories, symmetric theta | −1.695 / +1.695 | ±1.695 | 0.000 |
| 6 items, 4 categories, skewed theta | −1.725 / +1.695 | ±1.695 | 0.030 |
| 6 items, 4 categories, skewed and off target | −1.534 / +1.534 | ±1.522 | 0.012 |
| 20 items, 2 categories, symmetric | −1.798 / +1.798 | ±1.798 | 0.000 |
| 20 items, 2 categories, skewed | −1.798 / +1.798 | ±1.798 | 0.000 |
| 9 items, mixed 2/3/4 categories, symmetric | −1.721 / +1.721 | ±1.721 | 0.000 |
| 9 items, mixed 2/3/4 categories, skewed | −1.721 / +1.721 | ±1.721 | 0.000 |
| 12 items, 4 categories, symmetric | −1.864 / +1.864 | ±1.862 | 0.002 |

The exact values are perfectly symmetric. That is the expected result: under the
null the two occasions are exchangeable, so $\hat\theta_2 - \hat\theta_1$ is
symmetric about zero however skewed $\hat\theta$ itself is. The residual
asymmetry in the simulated column is Monte Carlo noise. Asymmetry becomes real
only when exchangeability fails, which in practice means the two occasions
differing in which items were answered.

## The critical value tracks test information

```{r by-length}
#| echo: true
set.seed(1)
theta <- rnorm(400, 0, 1.4)
tab <- t(vapply(c(4, 6, 10, 20, 40), function(k) {
  c(items = k, exact_crit(mk(k, 4), theta)[["upper"]])
}, numeric(2L)))
kable(tab, digits = 3, col.names = c("Items", "Critical value"))
```

Person locations are a sample of 400 from $N(0, 1.4)$, and the number of
response categories is held at four throughout. The critical value is monotone
in test length, converging on 1.96. Referring the RCI to a normal null is
therefore conservative, and increasingly so the shorter the scale. At six items
and four categories, using 1.96 instead of 1.725 is a materially stricter test.

The convergence is slow. Extending the same setup, integrated over $N(0, 1.4)$
by quadrature rather than sampled, the value is 1.933 at 60 items, 1.949 at 200
and 1.956 at 600, and does not reach 1.96 in that range. It is within 0.03 of
1.96 from about 60 items, where the size of a 1.96 test is roughly .047 and the
difference no longer matters. Dichotomous items converge more slowly and not
monotonically, because of discreteness: 1.734 at 10 items, 1.720 at 20, 1.926
at 100 and 1.953 at 600. Computed in `dev/rci_crit_by_length.R`.

Response categories have to be reported alongside item count, because a
six-item scale with three categories and one with seven are not comparable
instruments. They do not act on the two quantities in the same way:

```{r by-cats}
#| echo: true
cats <- t(vapply(c(2, 3, 4, 5, 7), function(nc) {
  thr <- mk(6, nc)
  names(thr) <- paste0("I", seq_along(thr))
  d <- as.data.frame(easyRasch2:::sim_partial_score(thr, theta))
  colnames(d) <- names(thr)
  rel <- RMreliabilityCurve(d, item_params = thr, output = "dataframe")
  c(categories = nc,
    critical = exact_crit(thr, theta)[["upper"]],
    marginal = attr(rel, "marginal_ratio"))
}, numeric(3L)))
kable(cats, digits = 3,
      col.names = c("Categories", "Critical value", "Marginal reliability"))
```

Reliability climbs steeply with categories, from about .60 at two to about .91
at seven. The critical value does not follow it. Almost all of its movement is
the step from dichotomous to polytomous, after which it is close to flat. So
categories add precision without much changing how far the null lies from
1.96, while items move both. There is a section below with further discussion on reliability and conditional reliability.

## Skew and targeting

Sample skew on its own barely moves the pooled critical value: the skewed and
symmetric rows above agree to the third decimal in five of seven configurations.
Targeting is what matters. Pushing the same skewed distribution 2.2 logits off
target, so that 39.5% of respondents are at an extreme score at one occasion or
the other, drops the critical value from 1.695 to 1.522. The mechanism is the
self-normalisation above: a floor-bound sample is located primarily where the
statistic is most heavily damped.

This also means the pooled critical value is a property of the sample as well as
the items, which is worth stating wherever it is reported.

Recalibrating on each sample, which is what the function does, keeps the test
close to its nominal level even when targeting is poor. **Carrying a critical
value across samples does not.** Applying a value calibrated on a well-targeted
sample to one sitting two logits off target gave a realised size of .028 against
a nominal .05 on six items, where recalibrating on that sample held .042. Quote
the value with the sample it came from, and recalibrate rather than reusing a
published figure.

## Can the sample dependence be removed?

Conditioning on the total score is what removed the latent distribution from the
subtest problem in the companion note, and the same argument applies here. Under
the null the two occasions are independent given $\theta$ and use the same
items, so

$$P(r_1 = a \mid r_1 + r_2 = s) = \frac{\gamma(a)\,\gamma(s-a)}{\gamma_2(s)}$$

where $\gamma_2$ is the elementary symmetric function of the item set counted
twice. The ratio is free of $\theta$, verified to between 2.2e-16 and 3.9e-16
across four configurations including dichotomous and mixed-category scales. A
per-respondent null is therefore available with no latent density and no plug-in
person estimate.

It is not worth using. Conditioning on the total leaves few attainable pairs, so
the discrete test falls short of its nominal level. Mean achievable size at
$\alpha = .05$, weighted by the two-occasion score distribution of an $N(0, 1)$
sample:

| Scale | strict p | mid-p | cannot reject (%) |
|---|---|---|---|
| 6 items, 4 cat | 0.0174 | 0.0493 | 1.4 |
| 12 items, 4 cat | 0.0260 | 0.0524 | 0.2 |
| 20 items, 4 cat | 0.0312 | 0.0526 | 0.0 |
| 20 items, 2 cat | 0.0261 | 0.0341 | 1.2 |
| 40 items, 4 cat | 0.0335 | 0.0482 | 0.0 |

A mid-p correction recovers the level, overshooting slightly on the middle
scales. The degeneracy is minor: respondents whose total admits no rejection
whatever are at most 1.4 percent of an on-target sample, because extreme totals
are rare.

What settles it is that neither version improves on what the function already
does. The case with no sample to pool over, and the one this construction was
meant to serve, is a single respondent measured against external item
parameters. Exact size given that respondent's true location, six items with
four categories:

| true $\theta$ | plug-in | strict | mid-p |
|---|---|---|---|
| -3 | .0195 | .0080 | .0265 |
| -2 | .0492 | .0196 | .0355 |
| -1 | .0473 | .0163 | .0489 |
| 0 | .0428 | .0184 | .0540 |
| +2 | .0492 | .0196 | .0355 |
| +3 | .0195 | .0080 | .0265 |

The plug-in holds between .043 and .051 through the middle of the scale and
turns conservative only in the tails, where the scale runs out of scores. Exact
power at a true location of zero on twenty items tells the same story: at a
change of one logit it is .618 for the plug-in, .626 for mid-p and .537 for
strict, and mid-p's small edge tracks its slightly larger size rather than any
real gain.

The negative result is worth recording because it accounts for something this
note otherwise reports without explanation. The conditional null given the total
does not depend on $\theta$ at all, so an error in a plug-in $\hat\theta$
reaches the critical value only through the weighting. That is why
`conditional_crit = TRUE` performs as well as it does from a single respondent's
own two estimates, and why the pooled default survives poor targeting once it is
recalibrated. The default is unchanged. Worked out in
`dev/conditional_total_null.R`.

## A consequence for the RCI as a statistic

Because the denominator is a function of the same scores as the numerator, the
RCI does not order respondents by how much they changed. A concrete pair from
the six-item scale above:

| | score change | change in logits | RCI |
|---|---|---|---|
| A | 0 → 18 | 7.24 | 3.53 |
| B | 3 → 15 | 3.11 | 3.60 |

Respondent A moved the full width of the scale, 4.14 logits further than B, and
scores lower on the RCI. Both are far past any critical value so nothing turns
on it here, but it shows that the RCI is a standardised distance from the null
rather than a measure of change, and should not be read as the latter.

The constructive version: since the critical value has to be computed from the
item parameters anyway, the normalisation adds nothing that a location-specific
threshold on the raw change would not. The same enumeration gives the null of
$\hat\theta_2 - \hat\theta_1$ directly, so a conditional critical **change**, in
logits, is available at no extra cost and is monotone in what happened. That is
the same object as a conditional minimal detectable change, and connects to the
score-level MDC table parked as a possible sibling function.

## The no-change band in the figure

The scatter plot of occasion 1 against occasion 2 draws a band covering the
changes that would not be flagged. That band was built from a smooth
approximation, half-width $c \cdot \sqrt{2/I(\theta)}$, which places **both**
occasions at the baseline location. The test does not do that. It uses
$SE(r_1)$ and $SE(r_2)$ at two different scores.

On 300 respondents the two disagreed for 33 of them, 11 percent: 27 sat outside
the drawn band without being flagged, and 6 sat inside it while flagged. The
figure contradicted the table it accompanied.

The mechanism shows in the worst case. A respondent moving from $\theta_1 =
1.23$ to $\theta_2 = 3.62$ had a band built on $SE_{diff} = 0.79$, because
information is high at 1.23, while the test used 1.56, because the follow-up
score is at the ceiling. The band was about half the width it should have
been, in precisely the case the figure exists to display: a large change landing
near a boundary.

The band is now built from the same lookup the test uses. For each attainable
baseline score, the follow-up locations that would not be flagged are read off
directly, so a point lies outside the band if and only if that respondent was
flagged. Verified at zero disagreements out of 300 and pinned by a regression
test. Because attainable scores are discrete the band is a step function, which
also makes the discreteness visible rather than hiding it behind a smooth curve.

```{r figure}
#| echo: true
#| fig-width: 7
#| fig-height: 7
#| fig-cap: "Default `RMpersonChange()` figure. Three hundred respondents on a
#|   scale of six items with four response categories, generated with a true
#|   shift of 0.7 logits
#|   between occasions. The grey band covers the follow-up locations that would
#|   not be flagged from each baseline score, so a point lies outside it if and
#|   only if that respondent was flagged. Its steps are the attainable scores,
#|   and it is widest away from the centre."
#| fig-alt: "Scatter plot of person location at occasion 1 against occasion 2,
#|   both in logits from about -4 to 4. Points fall on a discrete grid because
#|   attainable scores are discrete. A grey stepped band runs along the
#|   diagonal, narrow near the centre of the scale and wider towards each end
#|   before narrowing again at the extremes. Points inside the band are grey
#|   and labelled as no detected change, points above it are green and labelled
#|   as an increase, and one point below it is orange and labelled as a
#|   decrease. Most points sit above the diagonal, reflecting the simulated
#|   improvement."
set.seed(1)
thr_demo <- mk(6, 4)
names(thr_demo) <- paste0("I", seq_along(thr_demo))

theta_demo <- rnorm(300, 0, 1.6)
t1 <- as.data.frame(easyRasch2:::sim_partial_score(thr_demo, theta_demo))
t2 <- as.data.frame(easyRasch2:::sim_partial_score(thr_demo, theta_demo + 0.7))
colnames(t1) <- colnames(t2) <- names(thr_demo)

# Item parameters are estimated from the data here, as a reader would, so that
# this figure and the reliability output below share one calibration.
RMpersonChange(t1, t2, critical = "exact", output = "ggplot")
```

The band table below gives the same object numerically.

```{r band}
#| echo: true
thr6 <- mk(6, 4)
crit <- exact_crit(thr6, theta)

band_rows <- function(thr_list, lo, hi) {
  steps <- vapply(thr_list, length, integer(1L))
  lut <- vapply(0:sum(steps), function(r) {
    easyRasch2:::.theta_wle(
      easyRasch2:::.score_pattern(r, steps), thr_list, c(-10, 10)
    )
  }, numeric(2L))
  th <- lut[1L, ]
  se <- lut[2L, ]
  do.call(rbind, lapply(seq_along(th), function(i) {
    rci <- (th - th[i]) / sqrt(se[i]^2 + se^2)
    keep <- rci <= hi & rci >= lo
    data.frame(baseline = th[i], lower = min(th[keep]), upper = max(th[keep]))
  }))
}

b <- band_rows(thr6, crit[["lower"]], crit[["upper"]])
b$width <- b$upper - b$lower
kable(b, digits = 2, row.names = FALSE,
      col.names = c("Baseline theta", "Lower", "Upper", "Width"))
```

The width is not monotone in distance from the centre, which is worth knowing
before reading a figure. It peaks off centre, where the standard error is
already large but the scale has not run out of room, and narrows again at the
boundary itself because there are no further attainable scores to move to. For
this scale, six items with four categories, the widest band is 3.62 logits at a
baseline near $\pm 1.2$, against 2.47 at the centre and 2.39 at the extremes.

Two cases remain indicative rather than exact, and are documented as such: with
mixed missingness the band is drawn for the most common pair of answered-item
sets, and under `conditional_crit = TRUE` it uses the median critical values.

## What the change figure is resting on

A change verdict is only as good as the precision behind it, and the reader of
the figure above has no way to judge that from the figure alone. The same scale,
six items with four response categories, summarised the two ways the package
offers.

```{r reliability-table}
#| echo: true
RMreliability(t1, draws = 200, rmu_iter = 10, seed = 1, boot = TRUE)
```

Marginal reliability near .85 is respectable for six items with four response
categories, and the gap down to PSI is the ordinary one between a model-based
coefficient integrated over the fitted latent density and a sample-based one
computed from the observed spread of the estimates.

The "Marginal (curve mean)" row and the dashed line in the curve below are the
same number, the latent-density-weighted mean of the conditional reliability
curve. `RMreliabilityCurve()` returns it as the `marginal_ratio` attribute, and
the two agree to machine precision whenever both are given the same data and the
same calibration.

A single coefficient hides where that precision lies, which is the same point
the stepped band makes in the change figure.

```{r reliability-curve}
#| echo: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Conditional reliability across the scale for the same six items
#|   with four response categories, with the region reaching .8 shaded and the
#|   respondent distribution behind it."
#| fig-alt: "Conditional reliability plotted against person location in logits.
#|   The curve is an inverted U, peaking a little above 0.85 near the centre of
#|   the scale and falling below 0.5 at both extremes. A shaded vertical band
#|   marks the region where reliability reaches 0.8, covering roughly the middle
#|   two thirds of the plotted range. A grey distribution along the bottom shows
#|   where respondents fall, concentrated near the centre."
RMreliabilityCurve(t1, statistic = "reliability", benchmark = 0.8)
```

Read Figure 1 and 2 together. Where this curve is high the band in the change
figure is narrow and a modest change is detectable. Where it falls away the band
widens and only a large change clears it. The band narrowing again at the very
extremes is not precision returning, it is the scale running out of scores.

## Response dependence between occasions

Everything above assumes the two occasions are independent given $\theta$. That
is a conditional statement, and it is compatible with the two estimates being
strongly correlated marginally, which they are. Both measure the same person,
and that shared $\theta$ is what makes their difference interpretable. The
assumption concerns what is left once $\theta$ is accounted for.

What it rules out is carry-over: a time-2 answer depending on the time-1 answer
to the same item beyond what $\theta$ explains. Administering the occasions
separately does not deliver conditional independence on its own.
Marais [@marais2009] models the failure by shifting the time-2 difficulty of item $j$ by
$(1 - 2x_{j1})d$, so a positive response at time 1 makes that item easier at
time 2 and a negative one makes it harder. Marais and Andrich [@marais2008a] separate this
**response dependence** from **trait dependence**, which is multidimensionality.
Only the second is a failure of unidimensionality, and the two call for
different remedies.

This matters here in a way it does not for the companion note on the PCA/t-test
protocol. There, conditional independence between the two item subsets is the
null hypothesis, so a failure of it is what the test exists to detect. Here it
is an auxiliary assumption: the null is that $\theta$ did not change, and
carry-over can be present while $\theta$ is fixed.

```{r carryover-setup}
#| echo: true
# Marais's model, generalised to polytomous items by shifting every threshold
# of item j by (1 - 2 * x_j1 / m_j) * d, which reduces to her dichotomous form
# when m_j = 1. Shifting all thresholds by s is the same as shifting theta by
# -s, so the shift is applied as an effective theta.
rpcm <- function(theta_eff, thr) {
  m <- length(thr)
  num <- matrix(0, length(theta_eff), m + 1L); cs <- cumsum(thr)
  for (k in seq_len(m)) num[, k + 1L] <- k * theta_eff - cs[k]
  num <- exp(num - apply(num, 1L, max))
  cp <- t(apply(num / rowSums(num), 1L, cumsum))
  max.col(cp >= runif(length(theta_eff)), ties.method = "first") - 1L
}

sim_pair <- function(thr_list, theta1, theta2, d = 0) {
  k <- length(thr_list); n <- length(theta1)
  mj <- vapply(thr_list, length, integer(1L))
  x1 <- vapply(seq_len(k), function(j) rpcm(theta1, thr_list[[j]]), integer(n))
  sh <- if (d == 0) matrix(0, n, k) else (1 - 2 * sweep(x1, 2L, mj, "/")) * d
  x2 <- vapply(seq_len(k), function(j)
          rpcm(theta2 - sh[, j], thr_list[[j]]), integer(n))
  list(t1 = x1, t2 = x2)
}

# Wider item spread than mk() above, following Marais's [-2.5, 2.5]
mk_wide <- function(k, ncat) {
  lapply(seq(-2.5, 2.5, length.out = k),
         function(b) b + seq(-0.9, 0.9, length.out = ncat - 1L))
}

score_lut <- function(thr_list) {
  steps <- vapply(thr_list, length, integer(1L))
  vapply(0:sum(steps), function(r)
    easyRasch2:::.theta_wle(easyRasch2:::.score_pattern(r, steps),
                            thr_list, c(-10, 10)), numeric(2L))[1L, ]
}
```

The mechanism is easiest to see with no true change at all. Holding every
respondent at one location and generating both occasions from it, the mean
observed change should be zero at every location, and is when $d = 0$. Twenty
items, four categories, item locations over [-2.5, 2.5], 4000 respondents per
cell, change in logits.

```{r drift}
#| echo: true
set.seed(808)
thr_d <- mk_wide(20, 4); th_lut <- score_lut(thr_d)
grid_t <- seq(-2.5, 2.5, by = 0.5)
drift <- sapply(c(0, 0.5, 1, 1.5), function(d)
  sapply(grid_t, function(t0) {
    p <- sim_pair(thr_d, rep(t0, 4000), rep(t0, 4000), d = d)
    mean(th_lut[rowSums(p$t2) + 1L] - th_lut[rowSums(p$t1) + 1L])
  }))
dimnames(drift) <- list(sprintf("%.1f", grid_t),
                        paste0("d = ", c(0, 0.5, 1, 1.5)))
kable(drift, digits = 3)
```

Carry-over produces an antisymmetric **outward drift**: zero at the item mean,
pushing respondents further out the further out they already are, and reaching
almost half a logit at the ends by $d = 1.5$. This is the individual-level
mechanism behind the group-level result Marais reports [@marais2009], where easy items
put the sample above the item mean, everyone drifts up, and a group mean shows
change that did not happen.

### What it does to the test

The remaining results come from `dev/carryover_dependence.R`, which is not run
at render time. That script also validates the generator against Marais's
published Table 4, reproducing her mean same-item residual correlations of
-0.01, 0.14, 0.31 and 0.47 as -0.02, 0.13, 0.28 and 0.44.

With no true change, and critical values calibrated under independence as this
note recommends, the test becomes conservative rather than liberal. The
standard deviation of the RCI falls from about 0.97 to between 0.63 and 0.73,
so the exact critical values are too wide for the null actually in force.
Type I error against a nominal .05, on-target sample:

| Scale | d = 0 | d = 0.5 | d = 1 | d = 1.5 |
|---|---|---|---|---|
| 20 items, 4 cat | .0487 | .0267 | .0133 | .0074 |
| 6 items, 4 cat | .0509 | .0261 | .0121 | .0056 |
| 25 items, 2 cat | .0488 | .0262 | .0103 | .0032 |

The direction of the false positives is the part that does damage. At $d = 0$
they split evenly, as they must. As dependence grows they polarise, because the
drift carries each respondent away from the item mean and the sign of that
movement is fixed by where the respondent already was. Twenty items with four
categories, on target:

| d | share "increase" among those above the items | share "decrease" among those below |
|---|---|---|
| 0 | 0.51 | 0.49 |
| 0.5 | 0.75 | 0.75 |
| 1.0 | 0.88 | 0.89 |
| 1.5 | 0.97 | 1.00 |

With a real change the same drift manufactures a severity gradient. Simulating
a uniform improvement of 0.5 logits for every respondent on twenty items with
four categories, the estimated change in logits comes out as:

| d | overall | below items | near items | above items | above - below |
|---|---|---|---|---|---|
| 0 | 0.498 | 0.497 | 0.498 | 0.496 | -0.001 |
| 0.5 | 0.436 | 0.349 | 0.431 | 0.544 | 0.195 |
| 1.0 | 0.374 | 0.225 | 0.368 | 0.551 | 0.326 |
| 1.5 | 0.324 | 0.138 | 0.316 | 0.547 | 0.409 |

A flat treatment effect reads as a 0.41 logit gradient in baseline severity, and
a reader with no access to $d$ has no way to tell that apart from a real
interaction between treatment and baseline. Power falls at the same time, from
.189 to .043 on twenty items and from .083 to .013 on six.

### Why `null = "retest"` does not help

`retest_sd` adds an independent occasion deviation at each administration. That
is symmetric extra variance, and it widens an already conservative null. The
drift above is an antisymmetric shift in the conditional mean, which extra
variance does not touch. The two estimands in `RMpersonChange()` are both
defined under conditional independence, so neither covers this.

### Detecting it

Racking the two occasions into one analysis of $2k$ items and screening the
same-item pairs against `RMlocdepQ3Cutoff()` finds it well below the level that
does damage. Twenty items, four categories, $n = 500$:

| d | pairs flagged of 20 | runs flagging any |
|---|---|---|
| 0 | 0.0 | 0% |
| 0.5 | 3.3 | 100% |
| 1.0 | 14.2 | 100% |

Marais [@marais2009] judged $d = 0.5$ likely to pass unnoticed and $d = 1$
possibly so, working from raw residual correlations and from differential item
functioning by occasion, neither of which carried a calibrated reference.
Against a simulation-based cutoff the lowest level she studied is caught in all
25 runs, with a clean zero under the null. Her conclusion about detectability
reflects the diagnostics available at the time rather than a limit of the data.

Two details decide whether this works in practice. **The statistic has to be
centred on the cross-occasion pairs**, not Yen-corrected against every pair.
Subtracting the mean of all off-diagonal $Q3$ mixes within-occasion pairs into
the reference and costs power: at $d = 0.5$ the two centrings detect in 96 and
100 percent of runs respectively, and the gap widens when real change is also
present.

**Heterogeneous true change does not masquerade as carry-over**, which was the
objection worth settling before recommending the screen at all. Both act on the
cross-occasion block, so it is not obvious on paper that they separate. They do,
because carry-over is item-specific and change is not:

| condition | same-item | other cross-occasion | contrast | within-occasion |
|---|---|---|---|---|
| no change, no carry-over | -0.004 | 0.000 | -0.004 | 0.000 |
| uniform change +0.5 | -0.004 | 0.000 | -0.004 | 0.000 |
| change SD 0.6 | -0.039 | -0.034 | -0.005 | 0.036 |
| carry-over d = 0.5 | 0.147 | -0.005 | 0.152 | -0.004 |
| carry-over d = 1.0 | 0.274 | -0.008 | 0.281 | -0.007 |
| both, d = 1 and SD 0.6 | 0.224 | -0.032 | 0.255 | 0.020 |

Uniform change is absorbed into the time-2 item locations and leaves no trace,
which is the property Marais relies on for the rack analysis. Change that varies
between respondents is a second person dimension: it lifts the within-occasion
pairs, depresses the cross-occasion block as a whole, and leaves the contrast
flat. Carry-over lifts the same-item pairs and nothing else. When both are
present the two signatures add without interfering. Screened against the
simulated cutoff, no run out of five flagged a single pair under heterogeneous
change at SD 0.6 or SD 1.0, so the screen does not confuse the two.

## Scope and limits

* **Complete data only.** Missing data breaks sufficiency of a single sum score.
  Each distinct missingness pattern has its own score distribution and its own
  lookup, so the enumeration still works but per pattern.
* **`null = "retest"` needs quadrature.** The occasion deviations are
  continuous, so the joint becomes a double integral over $u_1, u_2$ rather than
  a pure sum. Still deterministic and cheap at, say, 21 nodes a side.
* **EAP is fine.** On complete data an EAP estimate is also a function of the
  sum score.
* Validated here for WLE with CML thresholds. Not yet checked against MML item
  parameters.
* **Conditional independence between occasions is assumed, not tested.** The
  enumeration is correct under it and says nothing about carry-over. See the
  section above, and screen for it before relying on a change verdict.

## Implemented

Shipped as `critical = "exact"`, now the default in `RMpersonChange()`, in
`R/person_change_exact.R`. Measured against a 4000-iteration simulation on the
same data: critical values agree to within Monte Carlo error, p-values correlate
at 0.9998, classifications are identical, and the exact path takes 0.08 s
against 13.7 s. Incomplete data is enumerated within each missingness pattern,
`null = "retest"` integrates the occasion deviations out by quadrature (they are
independent, so the joint factorises and the retest null costs no more than the
measurement null), and EAP is supported by holding the prior at the one the
observed respondents were scored under.

Having an exact reference immediately paid for itself twice. It exposed a bug in
the simulation path, which had been re-estimating the EAP prior from each null
dataset rather than holding it at the observed one. That is now fixed, and the
two paths agree to four decimals under EAP. It also supplied the lookup needed
to fix the figure, described in the next section.

It also removes the one real objection to `conditional_crit = TRUE`: a
per-respondent critical value from simulation rests on `sim_iter` draws, roughly
25 of which land in each tail at alpha = .05, whereas $P(r \mid \theta_i)$ is
available per respondent for free.

It would also improve the planned simulation study, whose first aim is the
realised Type I error of the simulated critical value. Much of the noise in that
comparison disappears if the reference is exact.

## Reproduction

Scratch scripts used for the simulated column are not part of the package. The
exact computation is self-contained in the `exact-fns` chunk above and needs
only `easyRasch2:::.pcm_cat_probs()`, `:::.theta_wle()` and `:::.score_pattern()`.
