---
title: "An exact null for the Rasch change index"
subtitle: "Why the RCI critical value is not 1.96, and how to compute it without simulating"
author: "Magnus Johansson"
date: today
---

```{r setup}
#| include: false
if (requireNamespace("devtools", quietly = TRUE) && file.exists("../DESCRIPTION")) {
  suppressMessages(devtools::load_all("..", quiet = TRUE))
} else {
  library(easyRasch2)
}
library(knitr)
```

## Purpose

`RMpersonChange()` ships `critical = "simulate"` rather than the conventional
1.96, because the Rasch change index

$$RCI = \frac{\hat\theta_2 - \hat\theta_1}{\sqrt{SE(\hat\theta_1)^2 + SE(\hat\theta_2)^2}}$$

is not standard normal under the null when the scale is short. This note shows
that the null does not need simulating at all: for complete data it can be
enumerated exactly. It also records what the exact critical value does across
test length, sample skew and targeting, and one consequence for the RCI as a
statistic.

## Why the null is not N(0, 1)

Two mechanisms, both weakening as test information grows.

**Self-normalisation.** $SE(\hat\theta_1)$ and $SE(\hat\theta_2)$ are
deterministic functions of the realised scores, so the denominator is not
independent of the numerator. A respondent who moves to an extreme score
produces a large numerator and a large denominator together, and the ratio is
damped. Both tails of the null are pulled in, so the critical value sits
**below** 1.96. In a t-test the denominator is independent of the numerator;
here it is not, which is why the t reference does not help either.

**Discreteness.** With complete data the sum score is sufficient, so
$\hat\theta$ takes only as many values as there are scores and the null is
discrete rather than continuous.

## Exact enumeration

Sufficiency is what makes this tractable. $\hat\theta(r)$ and $SE(r)$ are
deterministic lookups, the ones `RMscoreSE()` builds, and $P(r \mid \theta)$
follows from the Lord-Wingersky recursion. Under the null the two occasions are
independent given $\theta$, so the joint is $P(r_1 \mid \theta)P(r_2 \mid
\theta)$ and the whole null is a discrete distribution over $(k+1)^2$ score
pairs. For a six-item partial credit model that is 361 cells.

```{r exact-fns}
# Lord-Wingersky: exact sum-score distribution given theta under a PCM
score_dist <- function(thr_list, theta) {
  d <- 1
  for (thr in thr_list) {
    p <- easyRasch2:::.pcm_cat_probs(theta, thr)
    nd <- numeric(length(d) + length(p) - 1L)
    for (a in seq_along(d)) {
      nd[a + seq_along(p) - 1L] <- nd[a + seq_along(p) - 1L] + d[a] * p
    }
    d <- nd
  }
  d                                   # index is score + 1
}

# Exact null distribution of the RCI, pooled over a set of person locations
exact_crit <- function(thr_list, thetas, alpha = 0.05) {
  steps <- vapply(thr_list, length, integer(1L))
  lut <- vapply(0:sum(steps), function(r) {
    easyRasch2:::.theta_wle(
      easyRasch2:::.score_pattern(r, steps), thr_list, c(-10, 10)
    )
  }, numeric(2L))
  th <- lut[1L, ]
  se <- lut[2L, ]
  rci <- outer(seq_along(th), seq_along(th),
               function(i, j) (th[j] - th[i]) / sqrt(se[i]^2 + se[j]^2))
  w <- matrix(0, length(th), length(th))
  for (t in thetas) {
    d <- score_dist(thr_list, t)
    w <- w + outer(d, d)
  }
  w <- w / sum(w)
  o <- order(as.numeric(rci))
  v <- as.numeric(rci)[o]
  cw <- cumsum(as.numeric(w)[o])
  c(lower = v[which(cw >= alpha / 2)[1L]],
    upper = v[which(cw >= 1 - alpha / 2)[1L]])
}
```

## Validation against simulation

Run by `dev/` scratch scripts at `sim_iter` 2000 to 4000, n = 200, alpha = .05
two-sided. The simulated and exact values agree to within Monte Carlo error in
every configuration tried, including dichotomous items, items with unequal
category counts, a right-skewed person distribution and a badly off-target one.

| configuration | simulated | exact | difference |
|---|---|---|---|
| 6 polytomous items, symmetric theta | −1.695 / +1.695 | ±1.695 | 0.000 |
| 6 polytomous items, skewed theta | −1.725 / +1.695 | ±1.695 | 0.030 |
| 6 polytomous items, skewed and off target | −1.534 / +1.534 | ±1.522 | 0.012 |
| 20 dichotomous items, symmetric | −1.798 / +1.798 | ±1.798 | 0.000 |
| 20 dichotomous items, skewed | −1.798 / +1.798 | ±1.798 | 0.000 |
| mixed 2/3/4 categories, symmetric | −1.721 / +1.721 | ±1.721 | 0.000 |
| mixed 2/3/4 categories, skewed | −1.721 / +1.721 | ±1.721 | 0.000 |
| 12 polytomous items, symmetric | −1.864 / +1.864 | ±1.862 | 0.002 |

The exact values are perfectly symmetric. That is the expected result: under the
null the two occasions are exchangeable, so $\hat\theta_2 - \hat\theta_1$ is
symmetric about zero however skewed $\hat\theta$ itself is. The residual
asymmetry in the simulated column is Monte Carlo noise. Asymmetry becomes real
only when exchangeability fails, which in practice means the two occasions
differing in which items were answered.

## The critical value tracks test information

```{r by-length}
#| echo: true
set.seed(1)
theta <- rnorm(400, 0, 1.4)
tab <- t(vapply(c(4, 6, 10, 20, 40), function(k) {
  thr <- lapply(seq(-1.2, 1.2, length.out = k), function(b) b + c(-0.8, 0, 0.8))
  c(items = k, exact_crit(thr, theta)[["upper"]])
}, numeric(2L)))
kable(tab, digits = 3, col.names = c("Items", "Critical value"))
```

Monotone in test length, converging on 1.96. Referring the RCI to a normal null
is therefore conservative, and increasingly so the shorter the scale. On a
six-item scale, using 1.96 instead of 1.695 is a materially stricter test.

## Skew and targeting

Sample skew on its own barely moves the pooled critical value: the skewed and
symmetric rows above agree to the third decimal in five of seven configurations.
Targeting is what matters. Pushing the same skewed distribution 2.2 logits off
target, so that 39.5% of respondents sit at an extreme score at one occasion or
the other, drops the critical value from 1.695 to 1.522. The mechanism is the
self-normalisation above: a floor-bound sample spends more of its time where the
statistic is most heavily damped.

This also means the pooled critical value is a property of the sample as well as
the items, which is worth stating wherever it is reported.

## A consequence for the RCI as a statistic

Because the denominator is a function of the same scores as the numerator, the
RCI does not order respondents by how much they changed. A concrete pair from
the six-item scale above:

| | score change | change in logits | RCI |
|---|---|---|---|
| A | 0 → 18 | 7.24 | 3.53 |
| B | 3 → 15 | 3.11 | 3.60 |

Respondent A moved the full width of the scale, 4.14 logits further than B, and
scores lower on the RCI. Both are far past any critical value so nothing turns
on it here, but it shows that the RCI is a standardised distance from the null
rather than a measure of change, and should not be read as the latter.

The constructive version: since the critical value has to be computed from the
item parameters anyway, the normalisation buys nothing that a location-specific
threshold on the raw change would not. The same enumeration gives the null of
$\hat\theta_2 - \hat\theta_1$ directly, so a conditional critical **change**, in
logits, is available at no extra cost and is monotone in what happened. That is
the same object as a conditional minimal detectable change, and connects to the
score-level MDC table parked as a possible sibling function.

## Scope and limits

* **Complete data only.** Missing data breaks sufficiency of a single sum score.
  Each distinct missingness pattern has its own score distribution and its own
  lookup, so the enumeration still works but per pattern.
* **`null = "retest"` needs quadrature.** The occasion deviations are
  continuous, so the joint becomes a double integral over $u_1, u_2$ rather than
  a pure sum. Still deterministic and cheap at, say, 21 nodes a side.
* **EAP is fine.** On complete data an EAP estimate is also a function of the
  sum score.
* Validated here for WLE with CML thresholds. Not yet checked against MML item
  parameters.

## Implemented

Shipped as `critical = "exact"`, now the default in `RMpersonChange()`, in
`R/person_change_exact.R`. Measured against a 4000-iteration simulation on the
same data: critical values agree to within Monte Carlo error, p-values correlate
at 0.9998, classifications are identical, and the exact path takes 0.08 s
against 13.7 s. Incomplete data is enumerated within each missingness pattern,
`null = "retest"` integrates the occasion deviations out by quadrature (they are
independent, so the joint factorises and the retest null costs no more than the
measurement null), and EAP is supported by holding the prior at the one the
observed respondents were scored under.

Having an exact reference immediately paid for itself: it exposed a bug in the
simulation path, which had been re-estimating the EAP prior from each null
dataset rather than holding it at the observed one. That is now fixed, and the
two paths agree to four decimals under EAP.

It also removes the one real objection to `conditional_crit = TRUE`: a
per-respondent critical value from simulation rests on `sim_iter` draws, roughly
25 of which land in each tail at alpha = .05, whereas $P(r \mid \theta_i)$ is
available per respondent for free.

It would also sharpen the planned simulation study, whose first aim is the
realised Type I error of the simulated critical value. Much of the noise in that
comparison disappears if the reference is exact.

## Reproduction

Scratch scripts used for the simulated column are not part of the package. The
exact computation is self-contained in the `exact-fns` chunk above and needs
only `easyRasch2:::.pcm_cat_probs()`, `:::.theta_wle()` and `:::.score_pattern()`.
