---
title: "Seed stability of item-restscore decisions"
subtitle: "B = 400 against B = 1000 for RMitemRestscoreCutoff()"
author: "Magnus Johansson"
date: today
---

## Purpose

The bootstrap item-restscore test holds its family-wise rate at B = 400
(`restscore_cutoff_validation.qmd`). The number of iterations still matters
for two things that the error rate does not show:

1. **Seed dependence.** Two analysts who run the same data with different
   seeds can reach different decisions about an item near the boundary. For
   infit this happened for at least one item in about 10 percent of analyses
   at B = 400 and 4 percent at B = 2000 (Johansson, 2026). Restscore has not
   been measured.
2. **BH resolution.** With k items, BH flags anything only if the smallest
   p-value is at most α/k. At B = 400 the smallest attainable p-value is
   1/401 = .0025, which is α/k for k = 20, so BH with many items depends on
   the observed value lying beyond every simulated one.

This study measures both at B = 400 and B = 1000 with the conditional DGP,
the package default.

**Candidate tests.** `restscore_eti_null.qmd` and
`restscore_power_conditional.qmd` tested an equal-tailed alternative to the
shipped |t|: two one-sided Westfall-Young step-downs, one per direction at
α/2, with the mean and SD computed from all B + 1 values (`split_pooled`).
Its critical value is roughly the 10th largest of 400 simulated maxima
instead of the 20th, so it may depend more on the seed. It is computed here
from the same cutoff simulation, together with |t| with pooled moments
(`wy_pooled`), so seed stability can be compared at no extra run time. Both
are Westfall-Young only.

## Design

**Datasets.** Five cells, 20 datasets each, regenerated from the seeds of
the earlier studies so the data are the same:

| cell | source | why |
|---|---|---|
| `underfit_1item_t0_n400` | power study | one underfitting item, near the boundary at this n |
| `underfit_2item_t0_n400` | power study | two items masking each other, many borderline decisions |
| `overfit_1item_t0_n150` | power study | overfit at small n |
| `overfit_1item_tm2_n400` | power study | overfit off target |
| `dich_tgt_300` | validation study | null, 20 dichotomous items, where BH resolution binds |

**Seeds.** Five cutoff seeds per dataset at each of B = 400 and B = 1000.

**Outcomes**, separately for WY, BH and the two candidates:

- **Two-analyst disagreement:** over the 10 pairs of seeds per dataset, the
  share of pairs that disagree on the flag of at least one item.
- **Unstable items:** the share of items whose flag (none, overfit or
  underfit) is not the same in all five seeds.
- **Flags per dataset**, averaged over seeds, to show whether B shifts
  decisions in one direction.

**Cost.** About 0.1 s per iteration, so 7,000 iterations per dataset and
about 100 datasets, roughly 20 CPU-hours or 2 hours on 10 cores. Results are
saved per cell, so an interrupted render resumes. Set `R_DATASETS` lower for
a quick run.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
})
options(rgl.useNULL = TRUE)

stopifnot(
  "conditional is not the RMitemRestscoreCutoff() default" =
    identical(eval(formals(RMitemRestscoreCutoff)$dgp)[1], "conditional")
)

R_DATASETS <- 20
N_SEEDS    <- 5
BS         <- c(400, 1000)
NCORES     <- 10
RES_DIR    <- "restscore_stability"
dir.create(RES_DIR, showWarnings = FALSE)

source("restscore_power_generator.R")
RHO   <- 0.85
SLOPE <- 1.6
```

### Data

The power-study seed is `1e6 * (row in its 32-cell grid) + rep`, and the
validation-study seed is `1e6 * (row in its 4-cell grid) + rep`, so both
grids are rebuilt in their original order.

```{r data}
power_grid <- expand.grid(
  n = c(150, 250, 400, 600),
  target = c(0, -2),
  n_misfit = c(1, 2),
  direction = c("underfit", "overfit"),
  stringsAsFactors = FALSE
)
power_grid$cell <- with(power_grid, sprintf(
  "%s_%ditem_t%s_n%d", direction, n_misfit, ifelse(target == 0, "0", "m2"), n
))

val_cells <- c("dich_mis_150", "dich_tgt_300", "poly_tgt_150", "poly_tgt_1000")
sim_pcm <- function(theta, thr) {
  n <- length(theta)
  sapply(thr, function(d) {
    m <- length(d)
    eta <- outer(theta, 0:m) -
      matrix(c(0, cumsum(d)), n, m + 1, byrow = TRUE)
    p <- exp(eta - apply(eta, 1, max))
    p <- p / rowSums(p)
    cp <- p %*% upper.tri(diag(m + 1), diag = TRUE)
    rowSums(runif(n) > cp)
  })
}

cells <- c("underfit_1item_t0_n400", "underfit_2item_t0_n400",
           "overfit_1item_t0_n150", "overfit_1item_tm2_n400", "dich_tgt_300")

make_data <- function(cell, rep) {
  if (cell == "dich_tgt_300") {
    thr <- as.list(seq(-2, 2, length.out = 20))
    set.seed(1e6 * match(cell, val_cells) + rep)
    repeat {
      d <- as.data.frame(sim_pcm(rnorm(300, 0, 1.5), thr))
      names(d) <- paste0("I", seq_along(thr))
      if (all(vapply(d, function(x) all(tabulate(x + 1, 2) > 0), logical(1)))) {
        break
      }
    }
    return(list(data = d, misfit = rep(FALSE, 20)))
  }
  g <- power_grid[power_grid$cell == cell, ]
  row <- which(power_grid$cell == cell)
  d <- sim_power_data(n = g$n, n_misfit = g$n_misfit, target = g$target,
                      direction = g$direction, rho = RHO, slope = SLOPE,
                      seed = 1e6 * row + rep)
  list(data = d, misfit = seq_len(7) %in% item_setup(g$n_misfit, g$target)$misfit)
}
```

### One run

One cutoff for one dataset, B and seed, scored with WY and BH, and with the
two candidates from the same simulated values. `wy_local` recomputes the
shipped WY with the arithmetic of `.bootstrap_pvalues()`, as a check that the
candidates are computed on the same footing.

```{r onerun}
# Flags of the shipped |t| (local copy), |t| with pooled moments, and the
# pooled split test, from one dataset's observed differences and simulated
# null values. As split_variants() in restscore_power_conditional.qmd.
candidate_flags <- function(obs, sim, alpha = 0.05) {
  sim <- sim[, names(obs), drop = FALSE]
  B <- nrow(sim)
  stud <- function(ref) {
    m <- colMeans(ref, na.rm = TRUE)
    s <- apply(ref, 2L, stats::sd, na.rm = TRUE)
    s[!is.finite(s) | s < 1e-8] <- NA_real_
    list(o = (obs - m) / s, s = sweep(sweep(sim, 2L, m, `-`), 2L, s, `/`))
  }
  two <- function(z) {
    padj <- .wy_stepdown(abs(z$o), abs(z$s), B)
    ifelse(!is.na(padj) & padj < alpha,
           ifelse(z$o > 0, "overfit", "underfit"), "")
  }
  z_sim <- stud(sim)
  z_pool <- stud(rbind(obs, sim))
  hi <- .wy_stepdown(z_pool$o, z_pool$s, B)
  lo <- .wy_stepdown(-z_pool$o, -z_pool$s, B)
  data.frame(
    item = names(obs),
    wy_local = two(z_sim),
    wy_pooled = two(z_pool),
    split_pooled = ifelse(!is.na(hi) & hi < alpha / 2, "overfit",
                          ifelse(!is.na(lo) & lo < alpha / 2, "underfit", "")),
    stringsAsFactors = FALSE
  )
}

one_run <- function(task) {
  dd <- make_data(task$cell, task$rep)
  d <- dd$data
  seed <- 1e6 * task$rep + 1e3 * task$B + task$seed
  tryCatch({
    co <- suppressWarnings(suppressMessages(RMitemRestscoreCutoff(
      d, iterations = task$B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = "conditional"
    )))
    restscore <- function(correction) {
      suppressWarnings(suppressMessages(RMitemRestscore(
        d, cutoff = co, correction = correction, output = "dataframe"
      )))
    }
    flags <- function(out) out$Flagged[match(names(d), out$Item)]
    fw <- restscore("fwer")
    sim_mat <- tapply(co$results$Difference,
                      list(co$results$iteration, co$results$Item),
                      function(x) x[1L])
    cand <- candidate_flags(stats::setNames(fw$Difference, fw$Item), sim_mat)
    cand <- cand[match(names(d), cand$item), ]
    data.frame(
      cell = task$cell, rep = task$rep, B = task$B, seed = task$seed,
      item = names(d), misfit = dd$misfit,
      wy = flags(fw), bh = flags(restscore("fdr_bh")),
      wy_pooled = cand$wy_pooled, split_pooled = cand$split_pooled,
      wy_local = cand$wy_local,
      iterations = co$actual_iterations,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    data.frame(cell = task$cell, rep = task$rep, B = task$B, seed = task$seed,
               error = conditionMessage(e))
  })
}
```

### Run

The runs within a cell are spread over the cores, the longest (B = 1000)
first.

```{r run}
for (cl in cells) {
  f <- file.path(RES_DIR, paste0(cl, ".rds"))
  if (file.exists(f)) next
  tasks <- expand.grid(cell = cl, rep = seq_len(R_DATASETS),
                       seed = seq_len(N_SEEDS), B = sort(BS, decreasing = TRUE),
                       stringsAsFactors = FALSE)
  t0 <- Sys.time()
  res <- mclapply(seq_len(nrow(tasks)), function(i) one_run(tasks[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  saveRDS(list(res = res,
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          f)
}

raw <- lapply(cells, function(cl) readRDS(file.path(RES_DIR, paste0(cl, ".rds"))))
all_rows <- do.call(rbind, lapply(raw, function(x) do.call(bind_rows, x$res)))
errors <- if ("error" %in% names(all_rows)) {
  filter(all_rows, !is.na(error))
} else {
  NULL
}
runs <- if (is.null(errors)) all_rows else filter(all_rows, is.na(error))
runs$cell <- factor(runs$cell, levels = cells)

data.frame(
  cell = cells,
  runs = vapply(cells, function(cl) {
    nrow(distinct(filter(runs, cell == cl), rep, B, seed))
  }, integer(1)),
  errors = vapply(cells, function(cl) {
    if (is.null(errors)) 0L else sum(errors$cell == cl)
  }, integer(1)),
  minutes = vapply(raw, function(x) round(x$elapsed_min, 1), numeric(1)),
  row.names = NULL
)
if (!is.null(errors)) count(errors, cell, error)

runs |>
  distinct(cell, rep, B, seed, iterations) |>
  group_by(cell, B) |>
  summarise(median_iterations = median(iterations),
            min_iterations = min(iterations), .groups = "drop")

# The local copy must reproduce the package's WY flags in every run.
sum(runs$wy_local != runs$wy)

methods <- c(WY = "wy", BH = "bh", WY_pooled = "wy_pooled",
             split_pooled = "split_pooled")
```

## Results

### Two-analyst disagreement

For each dataset, every pair of seeds at the same B is compared. A pair
disagrees when at least one item gets a different flag.

```{r pairs}
pair_disagreement <- function(runs, method) {
  runs |>
    select(cell, rep, B, seed, item, flag = all_of(method)) |>
    group_by(cell, rep, B) |>
    group_modify(function(d, key) {
      w <- tidyr::pivot_wider(d, names_from = seed, values_from = flag)
      s <- setdiff(names(w), "item")
      pr <- utils::combn(s, 2)
      data.frame(disagree = apply(pr, 2, function(p) {
        any(w[[p[1]]] != w[[p[2]]])
      }))
    }) |>
    ungroup()
}

dis <- bind_rows(lapply(names(methods), function(m) {
  mutate(pair_disagreement(runs, methods[[m]]), method = m)
})) |>
  mutate(method = factor(method, levels = names(methods)))
dis |>
  group_by(cell, method, B) |>
  summarise(pairs = n(), disagree = round(100 * mean(disagree), 1),
            .groups = "drop") |>
  pivot_wider(names_from = B, values_from = c(pairs, disagree),
              names_glue = "{.value}_B{B}")
dis |>
  group_by(method, B) |>
  summarise(pairs = n(), disagree = round(100 * mean(disagree), 1),
            .groups = "drop")
```

### Unstable items

Items whose flag is not the same in all five seeds.

```{r unstable}
runs |>
  pivot_longer(all_of(unname(methods)), names_to = "method",
               values_to = "flag") |>
  mutate(method = factor(method, levels = methods, labels = names(methods))) |>
  group_by(cell, method, B, rep, item) |>
  summarise(unanimous = n_distinct(flag) == 1, .groups = "drop") |>
  group_by(cell, method, B) |>
  summarise(unstable_items = round(100 * mean(!unanimous), 1),
            .groups = "drop") |>
  pivot_wider(names_from = B, values_from = unstable_items,
              names_prefix = "B")
```

### Flags per dataset

Average number of items flagged per run, misfitting items in the correct
direction and other flags separately. Under the null cell every flag is
false.

```{r flags}
runs |>
  left_join(
    tibble::tibble(cell = cells,
                   direction = c("underfit", "underfit", "overfit",
                                 "overfit", "none")),
    by = "cell"
  ) |>
  pivot_longer(all_of(unname(methods)), names_to = "method",
               values_to = "flag") |>
  mutate(method = factor(method, levels = methods, labels = names(methods))) |>
  group_by(cell, method, B, rep, seed) |>
  summarise(correct = sum(misfit & flag == direction),
            other = sum(flag != "" & !(misfit & flag == direction)),
            .groups = "drop") |>
  group_by(cell, method, B) |>
  summarise(correct = round(mean(correct), 2), other = round(mean(other), 2),
            .groups = "drop") |>
  arrange(cell, method, B)
```

## Interpretation

All 1,000 runs ran without errors (17 to 23 minutes per cell), and the local
copy of the WY arithmetic reproduced the package's flags in every run, so
the candidates were scored on the same simulated values as the shipped test.

The results rest on 20 datasets per cell, and the 10 seed pairs within a
dataset are not independent, so differences of a few points between methods
at the same B are within noise. The drop from B = 400 to B = 1000 is
consistent across methods and cells.

**1. At B = 400, about one analysis in ten depends on the seed.** Over all
cells, two seeds disagreed on the flag of at least one item in 10.0 percent
of pairs with WY, close to the 10 percent found for infit at B = 400
(Johansson, 2026). The disagreement is concentrated where decisions are
hard: 22 percent with two underfitting items masking each other and 15
percent with overfit at n = 150, against 0 with one clear underfitting item
at n = 400.

**2. B = 1000 reduces it to about 6 percent.** WY disagreement fell from
10.0 to 6.1 percent and the share of unstable items roughly halved in the
hard cells (7.1 to 3.6 percent with two underfitting items). B did not shift
decisions in one direction: correct and other flags per dataset were nearly
the same at both values for every method.

**3. The split test is as stable as |t|.** Its critical value is a more
extreme quantile of the simulated maxima, which could have made it more
seed-dependent at B = 400. It did not: `split_pooled` disagreed in 10.3
percent of pairs against 10.0 for WY, with 0 to 20.5 percent by cell against
0 to 22, and 0 to 6.4 percent unstable items against 0 to 7.1. At B = 1000
it was the most stable method, 4.6 percent overall and 0 in the null cell
`dich_tgt_300`. Pooled moments did not change stability (`WY_pooled` 9.9
and 6.3). With two underfitting items `split_pooled` detected fewer per run
(0.45 against 0.50), the underfit cost seen in
`restscore_power_conditional.qmd`.

**4. BH is the least stable and gains little from more iterations.** BH
disagreed in 13.5 percent of pairs at B = 400 and 11.4 at B = 1000, more
than WY at both. Its discrete p-values and the α/k threshold for the
smallest one leave it sensitive to where the observed value falls among the
simulated ones even at B = 1000.

**Implications.** Seed stability does not count against the split test, so
the choice between it and |t| rests on the trade measured in the power
study. Separately, the halving of seed dependence at B = 1000 bears on the
default number of iterations, which is 400 now.
