---
title: "Item-restscore and infit under misfit"
subtitle: "Preliminary power and flag-direction study, 7 polytomous items"
author: "Magnus Johansson"
date: today
---

## Purpose

Two observations from practice, comparing conditional infit with
simulation-based cutoffs against item-restscore with BH correction:

1. Underfitting items are flagged later by item-restscore than by infit.
2. Overfitting items are flagged more often by item-restscore than by infit.

`restscore_asymptotic_null.qmd` explains part of both under a true Rasch
model. The asymptotic item-restscore test rejects too rarely in the underfit
tail and, at small n, too often in the overfit tail, and 78 to 100 percent of
its BH false flags are overfit. That leaves open what happens when items
genuinely misfit. The overfit flags seen in practice could come from three
places:

- **Null miscalibration**, which the bootstrap null in
  `RMitemRestscoreCutoff()` is designed to remove.
- **The cascade.** When one item underfits, the others look relatively
  over-discriminating, and the reverse.
- **Sensitivity.** Gamma may respond more strongly than infit to genuine
  over-discrimination.

This study plants items that under- or over-discriminate and records who
flags what, and in which direction. Running the asymptotic and the bootstrap
item-restscore tests on the same datasets separates the first explanation
from the other two.

## Design

- **Scale.** 7 polytomous items, 5 response categories, thresholds at the
  item location ± 0.5 and ± 1.5. θ ~ N(0, 1.5²), as in the null studies.
- **Underfit** is generated by multidimensionality, as in Johansson (2025)
  and the FWER paper: the misfitting items respond to a second dimension
  correlated ρ = .85 with the first. With two misfitting items they share
  that dimension. The shared residual makes them locally dependent on each
  other, which the FWER paper found makes them mask each other. It may also
  raise their association with the rest score, which bears directly on how
  item-restscore labels them.
- **Overfit** is generated by a discrimination above 1, with slope *a* = 1.6
  under the generalised partial credit model. All other items have *a* = 1.
- **Calibration.** Both values are anchored on the infit scale, as in the
  FWER paper, against the phq9 items flagged at n = 600: underfit 1.23 to
  1.31 (q3, q8, q9) and overfit 0.78 to 0.83 (q2, q4).
  `restscore_power_calibration.R` scanned one misfitting item at n = 600
  (80 replications per value). ρ = .85 gave infit 1.32 on target and 1.27 at
  −2 logits. *a* = 1.6 gave 0.79 and 0.80. The parameter is not portable
  across scale lengths (the FWER paper needed ρ = .75 at five items), so it
  was scanned for this scale rather than carried over.
- **Number of misfitting items.** 1 (item I1) or 2 (I1 and I2).
- **Targeting of the misfitting item(s).** Location 0 (the person mean) or −2.
  With two items they sit at the target ± 0.25. The fitting items are evenly
  spread over [−1.5, 1.5].
- **Sample size.** 150, 250, 400, 600.
- **Replications.** 100 datasets per cell, 32 cells. The Monte Carlo SE of a
  detection rate is at most 5 points. That is enough for preliminary results,
  not for fine distinctions.

**Methods**, all at α = .05, on the same dataset:

| label | method |
|---|---|
| `infit_wy` | `RMitemInfit()` with `RMitemInfitCutoff()`, Westfall-Young (package default) |
| `infit_bh` | the same cutoff object, `correction = "fdr_bh"` |
| `rest_boot_wy` | `RMitemRestscore()` with `RMitemRestscoreCutoff()`, Westfall-Young |
| `rest_boot_bh` | the same cutoff object, `correction = "fdr_bh"` |
| `rest_asy_bh` | `RMitemRestscore()` without a cutoff, asymptotic BH (package default) |

Both cutoff functions use the package defaults: 400 iterations and the
resample DGP.

**Outcomes.**

- **Detection:** a misfitting item flagged in the correct direction.
- **Wrong direction:** a misfitting item flagged in the opposite direction.
- **False flags on fitting items**, split into overfit and underfit. This is
  where the cascade shows.
- **Clean decision:** every misfitting item correctly flagged and no fitting
  item flagged.

**Cost.** About 0.1 s per bootstrap iteration for the two cutoff functions
together, so 40 s per dataset. 3,200 datasets is about 36 CPU-hours, or 3.5 to
4 hours on 10 cores. Results are saved per cell, so an interrupted render
resumes.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

R_DATASETS <- 100
B          <- 400
RHO        <- 0.85   # underfit, from restscore_power_calibration.R
SLOPE      <- 1.6    # overfit, from restscore_power_calibration.R
NCORES     <- 10
RES_DIR    <- "restscore_infit_power"
dir.create(RES_DIR, showWarnings = FALSE)

cells <- expand.grid(
  n = c(150, 250, 400, 600),
  target = c(0, -2),
  n_misfit = c(1, 2),
  direction = c("underfit", "overfit"),
  stringsAsFactors = FALSE
)
cells$cell <- with(cells, sprintf("%s_%ditem_t%s_n%d", direction, n_misfit,
                                  ifelse(target == 0, "0", "m2"), n))
nrow(cells)
```

### Data generation

The generator is shared with the calibration script, so the calibrated
values mean the same thing in both.

```{r items}
source("restscore_power_generator.R")
item_setup(2, -2)
```

### One dataset

A dataset with an unobserved category is redrawn, as in the validation study.

```{r onedataset}
one_dataset <- function(rep, cell_row) {
  setup <- item_setup(cell_row$n_misfit, cell_row$target)
  seed <- 1e6 * match(cell_row$cell, cells$cell) + rep
  d <- sim_power_data(
    n = cell_row$n, n_misfit = cell_row$n_misfit, target = cell_row$target,
    direction = cell_row$direction, rho = RHO, slope = SLOPE, seed = seed
  )

  tryCatch({
    ci <- suppressMessages(RMitemInfitCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile"
    ))
    cr <- suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile"
    ))
    flags <- function(tab) tab$Flagged[match(names(d), tab$Item)]
    q <- function(expr) suppressWarnings(suppressMessages(expr))
    data.frame(
      rep = rep,
      item = names(d),
      misfit = seq_len(7) %in% setup$misfit,
      infit_wy = flags(q(RMitemInfit(d, cutoff = ci, output = "dataframe"))),
      infit_bh = flags(q(RMitemInfit(d, cutoff = ci, correction = "fdr_bh",
                                     output = "dataframe"))),
      rest_boot_wy = flags(q(RMitemRestscore(d, cutoff = cr,
                                             output = "dataframe"))),
      rest_boot_bh = flags(q(RMitemRestscore(d, cutoff = cr,
                                             correction = "fdr_bh",
                                             output = "dataframe"))),
      rest_asy_bh = flags(q(RMitemRestscore(d, output = "dataframe"))),
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
}
```

### Run

```{r run}
for (i in seq_len(nrow(cells))) {
  f <- file.path(RES_DIR, paste0(cells$cell[i], ".rds"))
  if (file.exists(f)) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS), function(r) one_dataset(r, cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  saveRDS(list(res = res,
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          f)
}

methods <- c("infit_wy", "infit_bh", "rest_boot_wy", "rest_boot_bh",
             "rest_asy_bh")

long <- do.call(rbind, lapply(seq_len(nrow(cells)), function(i) {
  x <- readRDS(file.path(RES_DIR, paste0(cells$cell[i], ".rds")))
  ok <- Filter(Negate(is.null), x$res)
  cbind(cells[i, c("direction", "n_misfit", "target", "n")],
        do.call(rbind, ok), failed = length(x$res) - length(ok),
        row.names = NULL)
})) |>
  pivot_longer(all_of(methods), names_to = "method", values_to = "flag") |>
  mutate(method = factor(method, levels = methods),
         flag = ifelse(is.na(flag), "", flag),
         targeting = ifelse(target == 0, "misfit at 0", "misfit at -2"),
         misfit_items = paste(n_misfit, ifelse(n_misfit == 1, "item", "items")))

long |>
  distinct(direction, n_misfit, target, n, rep, failed) |>
  group_by(direction, n_misfit, target, n) |>
  summarise(datasets = n(), failed = first(failed), .groups = "drop") |>
  filter(failed > 0)
```

## Results

### Detection of the misfitting items

Proportion of misfitting items flagged in the correct direction.

```{r detection, fig.height = 7}
det <- long |>
  filter(misfit) |>
  group_by(direction, misfit_items, targeting, n, method) |>
  summarise(correct = 100 * mean(flag == direction),
            wrong = 100 * mean(flag != "" & flag != direction),
            .groups = "drop")

ggplot(det, aes(n, correct, colour = method)) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_grid(direction + misfit_items ~ targeting) +
  scale_x_continuous(breaks = c(150, 250, 400, 600)) +
  labs(x = "n", y = "Misfitting items detected (%)", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")

det |>
  select(-wrong) |>
  pivot_wider(names_from = method, values_from = correct) |>
  mutate(across(where(is.double) & !n, \(x) round(x, 0)))
```

Misfitting items flagged in the wrong direction, all cells combined:

```{r wrongdir}
det |>
  group_by(direction, method) |>
  summarise(wrong = round(mean(wrong), 1), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = wrong)
```

### False flags on the fitting items, by direction

This is where the cascade shows. With an underfitting item the fitting items
look relatively over-discriminating, so false overfit flags are expected to
dominate, and the reverse with an overfitting item.

```{r fp, fig.height = 7}
fp <- long |>
  filter(!misfit) |>
  group_by(direction, misfit_items, targeting, n, method) |>
  summarise(false_overfit = 100 * mean(flag == "overfit"),
            false_underfit = 100 * mean(flag == "underfit"),
            .groups = "drop")

fp |>
  pivot_longer(c(false_overfit, false_underfit), names_to = "type",
               values_to = "pct") |>
  ggplot(aes(n, pct, colour = method, linetype = type)) +
  geom_line() +
  facet_grid(direction + misfit_items ~ targeting) +
  scale_x_continuous(breaks = c(150, 250, 400, 600)) +
  labs(x = "n", y = "Fitting items flagged (%)", colour = NULL,
       linetype = NULL,
       title = "Facet rows name the planted misfit") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.box = "vertical")

fp |>
  group_by(direction, method) |>
  summarise(false_overfit = round(mean(false_overfit), 1),
            false_underfit = round(mean(false_underfit), 1),
            .groups = "drop")
```

### Clean decisions

Every misfitting item flagged in the correct direction and no fitting item
flagged.

```{r clean}
long |>
  group_by(direction, misfit_items, targeting, n, method, rep) |>
  summarise(clean = all(flag[misfit] == direction[1]) & all(flag[!misfit] == ""),
            .groups = "drop") |>
  group_by(direction, misfit_items, targeting, n, method) |>
  summarise(clean = round(100 * mean(clean), 0), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = clean)
```

### Paired comparison of restscore and infit

Both tests ran on the same datasets, so the comparison of bootstrap
restscore WY with infit WY is paired at the item level. The table counts
misfitting items detected by one and not the other, pooled over sample size
and the number of misfitting items.

```{r paireddet}
long |>
  filter(misfit, method %in% c("infit_wy", "rest_boot_wy")) |>
  mutate(hit = flag == direction) |>
  select(direction, n_misfit, targeting, n, rep, item, method, hit) |>
  pivot_wider(names_from = method, values_from = hit) |>
  group_by(direction, targeting) |>
  summarise(items = n(),
            restscore_only = sum(rest_boot_wy & !infit_wy),
            infit_only = sum(!rest_boot_wy & infit_wy),
            difference_pts = round(100 * (mean(rest_boot_wy) - mean(infit_wy)),
                                   1),
            .groups = "drop")
```

### Failed datasets

A dataset is lost when either cutoff function stops with "All simulation
iterations failed". The check below regenerates every dataset (no bootstrap)
and records whether some item has a category observed only among
respondents with a zero or perfect total score. CML drops those respondents,
so the category is empty in the estimation data, its threshold diverges, and
no simulated dataset contains it.

```{r failures}
extreme_only <- function(d) {
  tot <- rowSums(d)
  inner <- tot > 0 & tot < 4 * ncol(d)
  any(vapply(d, function(v) any(tabulate(v[inner] + 1, 5) == 0), logical(1)))
}
fail_check <- do.call(rbind, lapply(seq_len(nrow(cells)), function(i) {
  x <- readRDS(file.path(RES_DIR, paste0(cells$cell[i], ".rds")))
  failed <- vapply(x$res, is.null, logical(1))
  eo <- vapply(seq_along(x$res), function(r) {
    extreme_only(sim_power_data(
      n = cells$n[i], n_misfit = cells$n_misfit[i], target = cells$target[i],
      direction = cells$direction[i], rho = RHO, slope = SLOPE,
      seed = 1e6 * i + r
    ))
  }, logical(1))
  data.frame(cell = cells$cell[i], failed = sum(failed),
             failed_extreme_only = sum(failed & eo),
             ok_extreme_only = sum(!failed & eo))
}))
fail_check |> filter(failed > 0 | ok_extreme_only > 0)
colSums(fail_check[, -1])
```

## Interpretation

These are preliminary results at 100 datasets per cell, so differences of a
few points between methods within one cell are not reliable. The statements
below rest on patterns that repeat across cells or on the pooled paired
comparison.

**1. Restscore's late underfit flags come from the asymptotic test.** The
first observation from practice is reproduced by the asymptotic BH rule and
by nothing else. With one underfitting item on target at n = 150, infit WY
detected 66 percent, bootstrap restscore WY 63 percent and the asymptotic
test 16 percent. Pooled over cells, infit was 2 points ahead of bootstrap
restscore on target (40 against 16 discordant items) and restscore 3 points
ahead off target (102 against 65). The two tests are about equally sensitive
to underfit once restscore has a calibrated null.

**2. Restscore's extra overfit flags are partly real sensitivity.** The
bootstrap restscore test detected overfitting items more often than infit,
by 5 points on target (78 against 13 discordant items) and by 12 points off
target (169 against 32). One overfitting item at −2 logits and n = 400 is a
typical case, 78 percent against 58. Moving the
items off target cost infit more power than it cost restscore. At n = 150 the
asymptotic test was the most sensitive of the five methods in all four
overfit cells, which fits its liberal overfit tail under the null.

**3. The cascade produces false flags in the opposite direction, and they
grow with n.** With two overfitting items planted, the bootstrap methods
flagged 8 (WY) to 12 (BH) percent of the fitting items as underfitting, and
with underfit planted they flagged 2 to 5 percent as overfitting. Infit and
bootstrap restscore behave alike here. The asymptotic test shows almost no
underfit cascade (1.3 percent), because its underfit tail hardly ever
rejects. That is also why it has the most clean decisions when the planted
misfit is overfit (67 and 71 percent with two items at n = 600, against 12 to
39 percent for the bootstrap methods). The same conservatism costs it most of
its power against underfit, so this is not an argument for the asymptotic
test.

**4. Two underfitting items mask each other under every method.** With two
items sharing the second dimension, detection on target rose from 14 percent
at n = 150 to 62 percent at n = 600 (infit WY), and off target to 28 percent
(infit WY) and 40 percent (restscore WY). This replicates the masking found
in the FWER paper and is a property of the design, not of either test.

**5. No misfitting item was flagged in the wrong direction** by any method
in any cell.

**6. Failed datasets follow from CML, not from the methods.** 57 of 3,200
datasets were lost, up to 14 of 100 in the off-target cells at n = 150. All
57 had an item category observed only among respondents with a zero or
perfect total score (as did 2 datasets that ran). All five methods drop the
same datasets, so the comparisons stay paired, but the affected cells are
conditioned on data the cutoff functions can fit. For users the error
message ("All simulation iterations failed. Check your data.") does not say
what is wrong, although the cause can be detected before the bootstrap
starts.

**Limitations.** Both cutoff functions used the resample DGP. The validation
study found that for restscore it rejects about 1.4 points above nominal
under the null, so the two tests are not exactly level-matched. Restscore's
advantage against overfit (5 to 12 points) is larger than that, but the
smaller differences against underfit are not.
