---
title: "Calibration of the asymptotic item-restscore test"
subtitle: "Null simulation of iarm::item_restscore() under a true Rasch model"
author: "Magnus Johansson"
date: today
---

## Purpose

`RMitemRestscore()` flags items using the p-value from `iarm::item_restscore()`.
The test statistic is

$$
z_i = \frac{\hat\gamma_i^{\text{obs}} - \hat\gamma_i^{\text{exp}}}{\text{ASE}(\hat\gamma_i^{\text{obs}})},
$$

referred to the standard normal, two-sided, then adjusted over the $k$ items
(BH by default). The observed gamma is Goodman-Kruskal gamma for the
item × restscore table. The expected gamma is computed from the CML item
parameters and the observed score distribution. The ASE is the delta-method
standard error that `vcdExtra::GKgamma()` returns for the observed table.

Two features of this construction give reasons to check calibration directly
instead of assuming it:

1. **The expected gamma is treated as known.** It is a function of estimated
   thresholds, and it is estimated from the same data as the observed gamma.
   The ASE ignores both the estimation uncertainty and the correlation between
   the two terms. The sign of the resulting bias cannot be worked out without
   simulating it.
2. **The ASE is the delta-method SE computed under the alternative.** van der
   Ark and van Aert (2015) found it had the worst coverage of four CI types for
   bivariate gamma, but only for tables up to 5 × 5 at N ≤ 100. The
   item × restscore table is larger (4 × 28 for nine four-category items), with
   sparse cells in the tails of the restscore.

The partial gamma LD and DIF functions already have simulation-based cutoffs.
Item-restscore does not. `RMitemRestscoreBoot()` resamples the data and counts
how often each item is flagged, which describes stability and not calibration.

A hint already exists in `sx2_itemfit_sim.qmd`. Its $k = 0$ condition
(20 dichotomous items, 200 replications) gives per-item false positive rates of
0.07 to 0.73 percent, which that study reported as well calibrated. Recomputed
as the proportion of replications with **at least one** BH-flagged item, the
same results are 12.5, 5.5, 5.0, 2.0 and 1.5 percent at n = 150, 250, 500, 1000
and 2000. Under a complete null, BH controls this family-wise rate at .05, so
n = 150 is above nominal and n ≥ 1000 is below. With 200 replications the Monte
Carlo SE is about 1.5 points, so this is a lead to follow up, not a finding.

## Design

- **Item sets.** (a) 20 dichotomous items with locations evenly spaced on
  [-2, 2], matching the S-X2 study. (b) 9 polytomous items with four categories,
  locations evenly spaced on [-1.5, 1.5], and thresholds at location ± 1.2 and
  location + 0, as a PHQ-9-sized scale.
- **Persons.** θ ~ N(μ, 1.5²), with μ = 0 (targeted) or μ = 1.5 (mistargeted),
  since sparse restscore tails are a plausible source of miscalibration.
- **Sample sizes.** 150, 300, 600, 1000, 2000.
- **Replications.** 4000 per cell. The Monte Carlo SE of a 5 percent rate is
  then 0.34 points.
- **Fitting.** Same as the package: `psychotools::pcmodel()`, then
  `iarm::item_restscore()`. A replication with an unobserved category (or an
  item with zero variance) is discarded and counted, since the package would
  not fit it either.

`iarm` returns its results rounded to four decimals, so $z$ is recomputed from
the rounded observed, expected and SE columns and the p-values from $z$. With SEs
around .02 to .05 the rounding moves $z$ by at most about .003.

For each item and replication the observed gamma, expected gamma and ASE are
stored. Calibration is summarised in three ways.

- Per-item rejection at two-sided α = .05, and separately in each tail at .025.
- Family-wise rejection (any item flagged) with no adjustment, BH as in the
  package default, and Bonferroni.
- A decomposition of the SE. If the ASE were correct for the test statistic,
  the empirical SD of $\hat\gamma^{\text{obs}} - \hat\gamma^{\text{exp}}$ across
  replications would equal the root mean squared ASE. The ratio
  $\text{SD}(\hat\gamma^{\text{obs}}) / \text{RMS(ASE)}$ checks the ASE as an SE
  for the observed gamma alone. Comparing the two ratios separates point 1 from
  point 2 above.

```{r setup}
suppressMessages({
  library(psychotools)
  library(iarm)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

R_REPS   <- 4000
NS       <- c(150, 300, 600, 1000, 2000)
MUS      <- c(0, 1.5)
THETA_SD <- 1.5
NCORES   <- 10
RES_FILE <- "restscore_asymptotic_null.rds"
```

### Item parameters

```{r items}
item_sets <- list(
  dich = as.list(seq(-2, 2, length.out = 20)),
  poly = lapply(seq(-1.5, 1.5, length.out = 9), function(l) l + c(-1.2, 0, 1.2))
)

data.frame(
  set = c("dich", "poly"),
  items = lengths(item_sets),
  categories = c(2, 4),
  location_range = c("-2 to 2", "-1.5 to 1.5")
)
```

### Data generation

Partial credit model, vectorised over persons. `thr` is one threshold vector per
item. A dichotomous item is a PCM item with one threshold.

```{r sim}
sim_pcm <- function(theta, thr) {
  n <- length(theta)
  sapply(thr, function(d) {
    m <- length(d)
    eta <- outer(theta, 0:m) -
      matrix(c(0, cumsum(d)), n, m + 1, byrow = TRUE)
    p <- exp(eta - apply(eta, 1, max))
    p <- p / rowSums(p)
    cp <- p %*% upper.tri(diag(m + 1), diag = TRUE)
    rowSums(runif(n) > cp)
  })
}
```

### One replication

```{r onerep}
one_rep <- function(rep, set, n, mu) {
  set.seed(1e7 * match(set, names(item_sets)) + 1e5 * match(mu, MUS) +
             10 * n + rep)
  thr <- item_sets[[set]]
  d <- as.data.frame(sim_pcm(rnorm(n, mu, THETA_SD), thr))
  names(d) <- paste0("I", seq_along(thr))

  # every category observed in every item, as the CML fit requires
  full <- all(mapply(function(x, m) all(tabulate(x + 1, m + 1) > 0),
                     d, lengths(thr)))
  if (!full) return(NULL)

  res <- tryCatch({
    fit <- psychotools::pcmodel(d, hessian = FALSE)
    r <- iarm::item_restscore(fit, p.adj = "none")
    matrix(as.numeric(r[seq_along(thr), c("observed", "expected", "se")]),
           ncol = 3, dimnames = list(NULL, c("obs", "exp", "se")))
  }, error = function(e) NULL)
  res
}
```

### Run

Results are cached. Delete `r RES_FILE` to force a re-run.

```{r run}
grid <- expand.grid(rep = seq_len(R_REPS), n = NS, mu = MUS,
                    set = names(item_sets), stringsAsFactors = FALSE)

if (file.exists(RES_FILE)) {
  sim <- readRDS(RES_FILE)
} else {
  t0 <- Sys.time()
  res <- mclapply(seq_len(nrow(grid)),
                  function(i) one_rep(grid$rep[i], grid$set[i],
                                      grid$n[i], grid$mu[i]),
                  mc.cores = NCORES, mc.preschedule = TRUE)
  sim <- list(res = res, grid = grid, item_sets = item_sets,
              elapsed_min = as.numeric(Sys.time() - t0, units = "mins"))
  saveRDS(sim, RES_FILE)
}
cat("elapsed:", round(sim$elapsed_min, 1), "min\n")
```

```{r tidy}
ok <- !vapply(sim$res, is.null, logical(1))

sim$grid |>
  mutate(ok = ok) |>
  group_by(set, mu, n) |>
  summarise(discarded = sum(!ok), .groups = "drop") |>
  filter(discarded > 0)

item_loc <- lapply(sim$item_sets, function(s) vapply(s, mean, numeric(1)))

long <- do.call(rbind, lapply(which(ok), function(i) {
  g <- sim$grid[i, ]
  m <- sim$res[[i]]
  data.frame(set = g$set, mu = g$mu, n = g$n, rep = g$rep,
             item = seq_len(nrow(m)),
             location = item_loc[[g$set]], m)
})) |>
  mutate(diff = obs - exp,
         z = diff / se,
         p = 2 * pnorm(-abs(z)))
```

## Results

### Per-item rejection rates

Nominal 5.0 percent two-sided, 2.5 percent in each tail. `over` is observed
gamma above expected (the package labels this overfit), `under` is below.

An observed gamma of exactly ±1 has an ASE of zero, which gives an infinite $z$
and p = 0. The package flags such an item, so it counts as a rejection here, but
it is left out of the mean and SD of $z$. `se_zero` counts these cases.

```{r peritem}
long |>
  group_by(set, mu, n) |>
  summarise(two_sided = 100 * mean(p < .05),
            over = 100 * mean(z > qnorm(.975)),
            under = 100 * mean(z < qnorm(.025)),
            mean_z = mean(z[is.finite(z)]),
            sd_z = sd(z[is.finite(z)]),
            se_zero = sum(se == 0),
            .groups = "drop") |>
  mutate(across(two_sided:under, \(x) round(x, 2)),
         across(c(mean_z, sd_z), \(x) round(x, 3)))
```

### Family-wise rejection

Proportion of replications with at least one flagged item. Under a complete
null, BH and Bonferroni both control this at 5 percent. The unadjusted column
shows what the per-item test implies across 9 or 20 items.

```{r fwer}
fw <- long |>
  group_by(set, mu, n, rep) |>
  summarise(none = any(p < .05),
            bh = any(p.adjust(p, "BH") < .05),
            bonferroni = any(p.adjust(p, "bonferroni") < .05),
            .groups = "drop") |>
  group_by(set, mu, n) |>
  summarise(reps = n(),
            across(c(none, bh, bonferroni), \(x) round(100 * mean(x), 2)),
            .groups = "drop")
fw
```

### Where the SE goes wrong

`sd_diff / rms_se` near 1 means the ASE is the right SE for the test statistic.
`sd_obs / rms_se` near 1 means it is the right SE for the observed gamma alone.
`cor_obs_exp` is the correlation between observed and expected gamma across
replications, averaged over items.

```{r sedecomp}
se_tab <- long |>
  group_by(set, mu, n, item) |>
  summarise(sd_diff = sd(diff), sd_obs = sd(obs), sd_exp = sd(exp),
            rms_se = sqrt(mean(se^2)), cor_obs_exp = cor(obs, exp),
            .groups = "drop") |>
  group_by(set, mu, n) |>
  summarise(sd_diff_over_rms_se = mean(sd_diff / rms_se),
            sd_obs_over_rms_se = mean(sd_obs / rms_se),
            sd_exp_over_rms_se = mean(sd_exp / rms_se),
            cor_obs_exp = mean(cor_obs_exp),
            .groups = "drop") |>
  mutate(across(where(is.double) & !c(mu), \(x) round(x, 3)))
se_tab
```

### Per-item rates by item location

If sparse restscore tails matter, items far from the sample mean should be the
worst calibrated, and more so under mistargeting.

```{r byitem, fig.height = 6}
long |>
  group_by(set, mu, n, item, location) |>
  summarise(rate = 100 * mean(p < .05), .groups = "drop") |>
  mutate(rel_location = location - mu,
         mu = paste("μ =", mu)) |>
  ggplot(aes(rel_location, rate, colour = factor(n))) +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line() +
  geom_point(size = 1) +
  facet_grid(set ~ mu, scales = "free_x") +
  labs(x = "Item location relative to person mean (logits)",
       y = "Per-item rejection rate (%)", colour = "n") +
  theme_minimal()
```

### Distribution of z

Quantile plot against the standard normal, pooled over items, for the smallest
and largest n. Points on the diagonal mean the reference distribution is right.

```{r qq, fig.height = 5}
long |>
  filter(n %in% range(NS)) |>
  group_by(set, mu, n) |>
  mutate(theoretical = qnorm(ppoints(n()))[rank(z, ties.method = "first")]) |>
  slice_sample(n = 5000) |>
  ungroup() |>
  mutate(cell = paste0(set, ", μ = ", mu)) |>
  ggplot(aes(theoretical, z, colour = factor(n))) +
  geom_abline(linetype = "dashed", colour = "grey40") +
  geom_point(size = 0.4, alpha = 0.4) +
  facet_wrap(~cell) +
  coord_equal() +
  labs(x = "N(0, 1) quantile", y = "Observed z", colour = "n") +
  theme_minimal()
```

## Interpretation

The asymptotic item-restscore test is not calibrated. Two separate distortions
act on it, and neither vanishes as n grows.

**1. The ASE is right for the observed gamma but wrong for the test statistic.**
The SD of the observed gamma across replications matches the root mean squared
ASE to within 1 percent in every cell, so the delta-method SE is not the
problem. The problem is that the expected gamma varies across samples and moves
with the observed gamma (correlation about .41 for the dichotomous set and .55
for the polytomous set). Its variance roughly equals the covariance, so the
difference varies less than the observed gamma alone. The SD of the difference
is 0.91 times the ASE for dichotomous items and 0.83 times for polytomous items,
nearly constant across sample size and targeting. Dividing by the ASE therefore
shrinks $z$ by a fixed factor, and the test is conservative at every sample
size. This is point 1 from the Purpose section, and point 2 does not show up at
all.

**2. A small-sample bias pushes $z$ upwards.** Mean $z$ is .15 at n = 150 and
falls to .04 at n = 2000, roughly in proportion to $1/\sqrt{n}$. This is what a
bias of order $1/n$ in the gamma difference produces. It inflates the upper tail
(observed gamma above expected, labelled overfit by the package) and depletes
the lower tail. It is largest for items far from the person mean, reaching
14 percent two-sided at n = 150 for the easiest dichotomous item under
mistargeting.

**Net effect on single items.** Nominal is 5 percent two-sided. Dichotomous
items reject 4.7 percent at n = 150 when targeted, falling to 3.1 at n = 2000.
Polytomous items reject 2.9 falling to 1.9. The tails are lopsided. The lower
tail, which is the underfit direction and the usual signature of
multidimensionality, rejects 0.2 to 1.2 percent against a nominal 2.5 in every
cell, a tenth to a half of what it should. The upper tail exceeds 2.5 percent at
small n for dichotomous items. The power cost of the deflated lower tail is not
measured here, but it follows directly: underfit has to be larger than the
nominal level implies before it is flagged.

**Family-wise rates, which is what a user sees.** With BH over the items, the
probability that a true Rasch model gets at least one item flagged is:

- dichotomous, targeted: 13.5, 6.2, 4.1, 2.7 and 2.3 percent at n = 150 to 2000
- dichotomous, mistargeted: 30.5, 13.9, 7.4, 4.9 and 3.2 percent
- polytomous, targeted: 4.2 falling to 0.9 to 1.1 percent
- polytomous, mistargeted: 5.1 falling to 1.2 to 1.6 percent

So the same test is liberal for dichotomous data at small n,
especially when mistargeted, and conservative for polytomous data at every n.
No single correction to the nominal level fixes both. The 13.5 percent at
n = 150 reproduces the 12.5 percent recomputed from the S-X2 study.
BH and Bonferroni give nearly identical rates, as expected under a complete
null.

**Discarded replications.** 28 percent of the mistargeted polytomous
replications at n = 150 and 6 percent at n = 300 had an unobserved category and
were discarded. Those two cells describe samples where every category is used,
which are also the only samples the package can fit.

**For easyRasch2.** Both distortions come from treating the expected gamma as a
fixed reference, and both depend on the item set, sample size and targeting. A
parametric bootstrap that simulates from the fitted model, **re-estimates the
thresholds in every replication**, and recomputes the observed minus expected
difference would absorb both, in the same way as `RMitemInfitCutoff()`. A
bootstrap that holds the thresholds fixed would reproduce distortion 1. The
item-restscore statistic is cheap (about 0.1 s per replication including the
CML fit), so an `RMitemRestscoreCutoff()` at the same iteration counts as the
infit cutoffs is affordable. This simulation says nothing about power. That
comparison, against infit with simulation-based cutoffs, is the natural next
step.

## Reference

van der Ark, L. A., & van Aert, R. C. M. (2015). Comparing confidence intervals
for Goodman and Kruskal's gamma coefficient. *Journal of Statistical Computation
and Simulation, 85*(12), 2491-2505.
