---
title: "Validation of RMitemRestscoreCutoff()"
subtitle: "Nested null simulation of the bootstrap item-restscore test"
author: "Magnus Johansson"
date: today
---

## Purpose

`restscore_asymptotic_null.qmd` showed that the asymptotic item-restscore test
in `iarm::item_restscore()` is miscalibrated under a true Rasch model. With BH
over the items, the chance of at least one false flag ranged from 30.5 percent
(20 dichotomous items, n = 150, mistargeted by 1.5 logits) down to about
1 percent (9 polytomous items, n ≥ 1000). Underfit was flagged too rarely in
every cell.

`RMitemRestscoreCutoff()` replaces the asymptotic reference distribution with a
parametric bootstrap that refits the model in every iteration. Studentising the
observed minus expected difference by the bootstrap mean and SD should remove
both causes identified in the null study, the small-sample bias and the
correlation between observed and expected gamma. This study tests that claim
in the cells where the asymptotic test did worst, and compares the two
data-generating processes (`dgp = "resample"` and `"conditional"`).

## Design

Each simulated dataset is generated under a true Rasch model, exactly as in
the null study, and then analysed the way a user would:

1. `RMitemRestscoreCutoff(d, iterations = 400, dgp = ...)`, the package
   default number of iterations.
2. `RMitemRestscore(d, cutoff = co)` with `correction = "fwer"` (the
   default), `"fdr_bh"` and `"none"`.
3. `RMitemRestscore(d)` without a cutoff, the asymptotic BH rule, on the same
   dataset for a paired comparison.

**Cells.** Chosen from the null study to span its worst liberal and most
conservative results.

| cell | items | μ | n | asymptotic BH family-wise rate |
|---|---|---|---|---|
| `dich_mis_150` | 20 dichotomous | 1.5 | 150 | 30.5% |
| `dich_tgt_300` | 20 dichotomous | 0 | 300 | 6.2% |
| `poly_tgt_150` | 9 × 4 categories | 0 | 150 | 4.2% |
| `poly_tgt_1000` | 9 × 4 categories | 0 | 1000 | 0.9% |

**Arms.** Both DGPs in every cell, on the same datasets.

**Replications.** 500 datasets per cell. The Monte Carlo SE of a 5 percent
family-wise rate is then 0.97 points, so rates between 3.1 and 6.9 percent are
consistent with nominal.

**Outcomes.**

- Family-wise rate: at least one item flagged, for WY (FWER), BH and the
  asymptotic rule.
- Per-item marginal rejection at α = .05, split by direction (overfit and
  underfit, nominal 2.5 percent each). This tests whether the asymmetry of the
  asymptotic test is gone.
- The distribution of marginal p-values, which should be uniform.

**Cost.** About 0.1 s per bootstrap iteration, so 40 s per dataset and arm.
4 cells × 2 arms × 500 datasets is about 45 CPU-hours, or 4.5 hours on
10 cores. Results are saved per cell and arm, so an interrupted render resumes
where it stopped. Set `R_DATASETS` lower for a quick run.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

R_DATASETS <- 500
B          <- 400
THETA_SD   <- 1.5
NCORES     <- 10
RES_DIR    <- "restscore_cutoff_validation"
dir.create(RES_DIR, showWarnings = FALSE)
```

### Cells and data generation

The item sets and the generator are the same as in the null study.

```{r cells}
item_sets <- list(
  dich = as.list(seq(-2, 2, length.out = 20)),
  poly = lapply(seq(-1.5, 1.5, length.out = 9), function(l) l + c(-1.2, 0, 1.2))
)

cells <- data.frame(
  cell = c("dich_mis_150", "dich_tgt_300", "poly_tgt_150", "poly_tgt_1000"),
  set  = c("dich", "dich", "poly", "poly"),
  mu   = c(1.5, 0, 0, 0),
  n    = c(150, 300, 150, 1000),
  stringsAsFactors = FALSE
)
arms <- c("resample", "conditional")
cells

sim_pcm <- function(theta, thr) {
  n <- length(theta)
  sapply(thr, function(d) {
    m <- length(d)
    eta <- outer(theta, 0:m) -
      matrix(c(0, cumsum(d)), n, m + 1, byrow = TRUE)
    p <- exp(eta - apply(eta, 1, max))
    p <- p / rowSums(p)
    cp <- p %*% upper.tri(diag(m + 1), diag = TRUE)
    rowSums(runif(n) > cp)
  })
}
```

### One dataset

A dataset with an unobserved category is regenerated rather than discarded,
since the question is how the bootstrap behaves on data the package can fit.
The number of redraws is recorded.

```{r onedataset}
one_dataset <- function(rep, cell_row, dgp) {
  thr <- item_sets[[cell_row$set]]
  seed <- 1e6 * match(cell_row$cell, cells$cell) + rep
  set.seed(seed)
  redraws <- 0L
  repeat {
    d <- as.data.frame(sim_pcm(rnorm(cell_row$n, cell_row$mu, THETA_SD), thr))
    names(d) <- paste0("I", seq_along(thr))
    full <- all(mapply(function(x, m) all(tabulate(x + 1, m + 1) > 0),
                       d, lengths(thr)))
    if (full) break
    redraws <- redraws + 1L
  }

  out <- tryCatch({
    co <- suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = dgp
    ))
    run <- function(correction) {
      suppressMessages(RMitemRestscore(
        d, cutoff = co, correction = correction, output = "dataframe"
      ))
    }
    fw <- run("fwer")
    bh <- run("fdr_bh")
    nc <- run("none")
    asy <- suppressMessages(RMitemRestscore(d, output = "dataframe"))
    data.frame(
      rep = rep,
      item = fw$Item,
      difference = fw$Difference,
      p = fw$p_restscore,
      flag_marginal = nc$Flagged,
      padj_fwer = fw$padj_restscore,
      padj_bh = bh$padj_restscore,
      flag_asy = asy$Flagged,
      actual_iterations = co$actual_iterations,
      redraws = redraws,
      stringsAsFactors = FALSE
    )
  }, error = function(e) NULL)
  out
}
```

### Run

One file per cell and arm. Delete a file to re-run that part.

```{r run}
timing <- list()
for (i in seq_len(nrow(cells))) {
  for (dgp in arms) {
    f <- file.path(RES_DIR, paste0(cells$cell[i], "_", dgp, ".rds"))
    if (file.exists(f)) next
    t0 <- Sys.time()
    res <- mclapply(seq_len(R_DATASETS),
                    function(r) one_dataset(r, cells[i, ], dgp),
                    mc.cores = NCORES, mc.preschedule = FALSE)
    saveRDS(list(res = res,
                 elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
            f)
  }
}

long <- do.call(rbind, lapply(seq_len(nrow(cells)), function(i) {
  do.call(rbind, lapply(arms, function(dgp) {
    f <- file.path(RES_DIR, paste0(cells$cell[i], "_", dgp, ".rds"))
    x <- readRDS(f)
    ok <- Filter(Negate(is.null), x$res)
    if (length(ok) == 0) return(NULL)
    cbind(cell = cells$cell[i], dgp = dgp, do.call(rbind, ok),
          elapsed_min = x$elapsed_min, failed = length(x$res) - length(ok))
  }))
}))
long$cell <- factor(long$cell, levels = cells$cell)

long |>
  group_by(cell, dgp) |>
  summarise(datasets = n_distinct(rep),
            failed = first(failed),
            redraws = sum(redraws[!duplicated(rep)]),
            min_iterations = min(actual_iterations),
            minutes = round(first(elapsed_min), 1),
            .groups = "drop")
```

## Results

### Family-wise rate

Proportion of datasets with at least one flagged item, nominal 5 percent. The
asymptotic column is the same in both arms, since it does not use the
bootstrap.

```{r fwer}
fw <- long |>
  group_by(cell, dgp, rep) |>
  summarise(wy = any(padj_fwer < .05),
            bh = any(padj_bh < .05),
            asymptotic = any(flag_asy != ""),
            .groups = "drop") |>
  group_by(cell, dgp) |>
  summarise(datasets = n(),
            across(c(wy, bh, asymptotic), \(x) round(100 * mean(x), 1)),
            .groups = "drop")
fw
```

### Per-item rejection by direction

Marginal bootstrap p-value below .05 (`correction = "none"`), split by the
direction label, against the same split for the asymptotic test with BH. The
bootstrap should give 2.5 percent in each direction.

```{r tails}
long |>
  group_by(cell, dgp) |>
  summarise(boot_over = 100 * mean(flag_marginal == "overfit"),
            boot_under = 100 * mean(flag_marginal == "underfit"),
            asy_over = 100 * mean(flag_asy == "overfit"),
            asy_under = 100 * mean(flag_asy == "underfit"),
            .groups = "drop") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

The asymptotic columns are BH-adjusted and so not directly comparable in
level with the marginal bootstrap columns. What matters is the ratio of the
two tails within each method.

### Marginal p-values

Under the null the marginal p-values should be uniform. With B = 400 the
smallest attainable value is 1/401, so the histogram is coarse at the left
edge.

```{r phist, fig.height = 5}
long |>
  ggplot(aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "steelblue",
                 colour = "white") +
  geom_hline(aes(yintercept = n / 20),
             data = long |> count(cell, dgp),
             linetype = "dashed", colour = "grey40") +
  facet_grid(dgp ~ cell) +
  labs(x = "Marginal bootstrap p-value", y = "Count") +
  theme_minimal()
```

### Additional checks

**Paired DGP comparison.** Both arms analyse the same datasets, so the
difference between them is a paired comparison. The table counts datasets
flagged by WY under one DGP and not the other.

```{r paired}
fw_rep <- long |>
  group_by(cell, dgp, rep) |>
  summarise(wy = any(padj_fwer < .05), bh = any(padj_bh < .05),
            asymptotic = any(flag_asy != ""), .groups = "drop")
fw_rep |>
  select(cell, dgp, rep, wy) |>
  pivot_wider(names_from = dgp, values_from = wy) |>
  filter(!is.na(resample), !is.na(conditional)) |>
  group_by(cell) |>
  summarise(resample_only = sum(resample & !conditional),
            conditional_only = sum(!resample & conditional),
            both = sum(resample & conditional), .groups = "drop")
fw_rep |>
  group_by(dgp) |>
  summarise(datasets = n(), wy = round(100 * mean(wy), 2),
            bh = round(100 * mean(bh), 2), .groups = "drop")
```

**The `dich_tgt_300` draw.** The asymptotic rate in this cell is 10.0
percent against 6.2 percent over 4000 datasets in the null study. The
package path and a direct `iarm::item_restscore()` call agree on every one
of the 500 datasets, and the null study's seeds give 6.4 percent through the
package path (checked outside this document), so the method is the same and
the 500 datasets are a high draw. Whether that carries over to the bootstrap
arms shows in the WY rate split by the asymptotic decision on the same
dataset.

```{r carryover}
fw_rep |>
  group_by(cell, dgp, asymptotic) |>
  summarise(datasets = n(), wy = round(100 * mean(wy), 1),
            .groups = "drop") |>
  pivot_wider(names_from = asymptotic, values_from = c(datasets, wy),
              names_glue = "{.value}_asy_{asymptotic}")
```

**Skew of the null.** Under the null the observed difference across the
500 datasets is a draw from the same distribution the bootstrap estimates,
so its per-item skewness shows the shape the two-sided |t| test folds. The
second number is the correlation over items between skewness and the
underfit-minus-overfit marginal rejection rate.

```{r skew}
skew <- function(x) { x <- x - mean(x); mean(x^3) / mean(x^2)^1.5 }
sk <- long |>
  filter(dgp == "conditional") |>
  group_by(cell, item) |>
  summarise(skew = skew(difference),
            asym = 100 * (mean(flag_marginal == "underfit") -
                            mean(flag_marginal == "overfit")),
            .groups = "drop")
sk |>
  group_by(cell) |>
  summarise(mean_skew = round(mean(skew), 3),
            mean_asym = round(mean(asym), 2), .groups = "drop")
round(cor(sk$skew, sk$asym), 2)
```

**Iterations kept.** WY rate by whether all 400 iterations succeeded.

```{r iterations}
long |>
  group_by(dgp, rep, cell) |>
  summarise(wy = any(padj_fwer < .05),
            full = first(actual_iterations) == B, .groups = "drop") |>
  group_by(dgp, full) |>
  summarise(datasets = n(), wy = round(100 * mean(wy), 2), .groups = "drop")
```

## Interpretation

**1. The bootstrap removes the miscalibration of the asymptotic test.** In
the worst cell of the null study, 20 dichotomous items mistargeted by 1.5
logits at n = 150, the asymptotic BH rule flagged at least one item in 29.6
percent of datasets. On the same datasets the bootstrap gave 4.2 percent
(WY) and 2.8 percent (BH) with the conditional DGP. In the most conservative
cell, 9 polytomous items at n = 1000, the asymptotic rate of 0.8 percent rose
to 5.6 percent (WY). The marginal p-values are close to uniform in every cell
and arm, and the underfit tail, which the asymptotic test almost never
rejected (0.00 to 0.03 percent per item), now rejects 2.6 to 3.4 percent.

**2. The conditional DGP is calibrated and the resample DGP is somewhat
liberal.** Pooled over the four cells (2000 datasets per arm), the WY
family-wise rate was 5.0 percent with the conditional DGP and 6.35 percent
with resample, and BH gave 4.2 and 5.45 percent. Resample was higher in all
four cells, and the paired comparison leaves little room for chance: 43
datasets were flagged by WY under resample only and 16 under conditional only.
The marginal rejection rates show the same pattern (resample 4.9 to 5.9
percent, conditional 4.6 to 5.3). The mechanism has not been measured. A
likely candidate is that resampled WLE locations do not reproduce the score
distribution the observed gamma is conditioned on, which the conditional DGP
holds fixed by construction.

**3. `dich_tgt_300` is a high draw, not a calibration failure.** The 500
datasets in this cell gave the asymptotic test 10.0 percent against a
long-run 6.2 percent, and datasets the asymptotic test flagged were also
flagged by the bootstrap WY in 30 percent of cases, against 4.2 percent for
the rest. The bootstrap rates in this cell (6.8 conditional, 9.2 resample)
are therefore inflated by the same draw. Reweighting the two groups to the
long-run asymptotic rate gives roughly 5.8 and 8.3 percent, a rough figure
that leaves the conditional arm within Monte Carlo error and the resample
arm still above it.

**4. The two directions do not get equal shares of α.** The marginal test
rejects at about the nominal total, but the underfit tail takes more than
half of it (2.6 to 3.4 percent against 1.8 to 2.7 for overfit). The null
distribution of the difference is left-skewed, since gamma is bounded at 1,
and a two-sided test on |t| splits α evenly only under symmetry. The
asymmetry follows the skewness across cells (largest for mistargeted
dichotomous items, smallest for polytomous items at n = 1000) and across
items (correlation −0.79). An equal-tailed p-value would remove it. That
change touches `.bootstrap_pvalues()`, which the other cutoff functions
share, and is not decided.

**5. Failed iterations did not affect the rate.** Some datasets kept fewer
than 400 iterations (down to 258 in `poly_tgt_150`), because simulated
datasets with an unobserved category are rejected. The WY rate was not
higher in datasets with fewer iterations.

**Implications.** The bootstrap test should replace the asymptotic test as
the recommended path, and the conditional DGP is the better default for
`RMitemRestscoreCutoff()`, at 1.2 to 1.6 times the run time of resample. The
study used 400 iterations throughout, so it supports 400 as adequate but says
nothing about fewer.
