---
title: "Equal-tailed item-restscore p-values under the null"
subtitle: "Null check of RMitemRestscore() after the switch from |t| to equal tails"
author: "Magnus Johansson"
date: today
---

## Purpose

`restscore_cutoff_validation.qmd` validated the bootstrap item-restscore
test with p-values from |t|, the difference studentised by the bootstrap
mean and SD and folded at the mean. The marginal level was right, but the
null of the difference is left-skewed, so the underfit tail took 2.6 to 3.4
percent per item and the overfit tail 1.8 to 2.7. `RMitemRestscore()` now
uses equal-tailed p-values instead:

- **Marginal:** twice the smaller one-sided Monte Carlo p-value, so the
  smallest attainable value is 2/(B + 1).
- **Westfall-Young:** each item's difference is mapped through its own
  simulated distribution to a normal score, and the max step-down runs on
  the absolute normal scores. This is a rank-based (min-p) form of the
  step-down, so it should hold the family-wise rate under the same
  exchangeability that the |t| version relies on.

This study checks that claim by simulation, and that the two directions now
get α/2 each.

## Design

The four cells and 500 datasets per cell of the validation study, with the
conditional DGP (now the default) and B = 400. The seeds are the same, so
each dataset and each bootstrap simulation is identical to the conditional
arm of the validation study. Both p-values are computed from the same
simulated values:

- `eq`: equal-tailed, as `RMitemRestscore()` now computes it.
- `abs`: |t|, as before the change. It should reproduce the validation
  study's cached results exactly, which checks that nothing else changed.

**Outcomes.** Family-wise rate (WY and BH) for both, nominal 5 percent with
a Monte Carlo SE of 0.97 points at 500 datasets. Per-item marginal rejection
by direction, nominal 2.5 percent each. The distribution of marginal
p-values.

**Cost.** The conditional arm of the validation study took 132 minutes on
10 cores over the four cells. Results are saved per cell, so an interrupted
render resumes. Set `R_DATASETS` lower for a quick run.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

# The run needed tail = "equal" in .bootstrap_pvalues(), which has since
# been withdrawn. The cached results below come from that version; rendering
# without the cache would fail in one_dataset().

R_DATASETS <- 500
B          <- 400
THETA_SD   <- 1.5
NCORES     <- 10
RES_DIR    <- "restscore_eti_null"
VAL_DIR    <- "restscore_cutoff_validation"
dir.create(RES_DIR, showWarnings = FALSE)
```

### Cells and data generation

Copied from the validation study. The seed of a dataset is
`1e6 * (row of its cell) + rep`, so the cell order must not change.

```{r cells}
item_sets <- list(
  dich = as.list(seq(-2, 2, length.out = 20)),
  poly = lapply(seq(-1.5, 1.5, length.out = 9), function(l) l + c(-1.2, 0, 1.2))
)

cells <- data.frame(
  cell = c("dich_mis_150", "dich_tgt_300", "poly_tgt_150", "poly_tgt_1000"),
  set  = c("dich", "dich", "poly", "poly"),
  mu   = c(1.5, 0, 0, 0),
  n    = c(150, 300, 150, 1000),
  stringsAsFactors = FALSE
)
cells

sim_pcm <- function(theta, thr) {
  n <- length(theta)
  sapply(thr, function(d) {
    m <- length(d)
    eta <- outer(theta, 0:m) -
      matrix(c(0, cumsum(d)), n, m + 1, byrow = TRUE)
    p <- exp(eta - apply(eta, 1, max))
    p <- p / rowSums(p)
    cp <- p %*% upper.tri(diag(m + 1), diag = TRUE)
    rowSums(runif(n) > cp)
  })
}
```

### One dataset

```{r onedataset}
one_dataset <- function(rep, cell_row) {
  thr <- item_sets[[cell_row$set]]
  seed <- 1e6 * match(cell_row$cell, cells$cell) + rep
  set.seed(seed)
  repeat {
    d <- as.data.frame(sim_pcm(rnorm(cell_row$n, cell_row$mu, THETA_SD), thr))
    names(d) <- paste0("I", seq_along(thr))
    full <- all(mapply(function(x, m) all(tabulate(x + 1, m + 1) > 0),
                       d, lengths(thr)))
    if (full) break
  }

  tryCatch({
    co <- suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = "conditional"
    ))
    # The package path, equal-tailed.
    run <- function(correction) {
      suppressMessages(RMitemRestscore(
        d, cutoff = co, correction = correction, output = "dataframe"
      ))
    }
    eq_wy <- run("fwer")
    eq_bh <- run("fdr_bh")
    eq_none <- run("none")

    # |t| from the same simulated values, as RMitemRestscore() computed it
    # before the change.
    sim_mat <- tapply(co$results$Difference,
                      list(co$results$iteration, co$results$Item),
                      function(x) x[1L])
    obs <- stats::setNames(eq_wy$Difference, eq_wy$Item)
    abs_p <- function(correction) {
      pv <- .bootstrap_pvalues(obs, sim_mat, correction = correction,
                               tail = "two.sided")
      pv[match(names(obs), pv$name), ]
    }
    a_wy <- abs_p("fwer")
    a_bh <- abs_p("fdr_bh")
    null_mean <- colMeans(sim_mat, na.rm = TRUE)[names(obs)]

    data.frame(
      rep = rep,
      item = eq_wy$Item,
      difference = eq_wy$Difference,
      p_eq = eq_wy$p_restscore,
      padj_eq_wy = eq_wy$padj_restscore,
      padj_eq_bh = eq_bh$padj_restscore,
      flag_eq_marginal = eq_none$Flagged,
      p_abs = a_wy$p,
      padj_abs_wy = a_wy$padj,
      padj_abs_bh = a_bh$padj,
      flag_abs_marginal = ifelse(
        a_wy$p < .05, ifelse(obs > null_mean, "overfit", "underfit"), ""
      ),
      actual_iterations = co$actual_iterations,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    data.frame(rep = rep, error = conditionMessage(e))
  })
}
```

Errors are kept as rows, so a failure shows in the results instead of
looking like a missing dataset.

### Run

```{r run}
for (i in seq_len(nrow(cells))) {
  f <- file.path(RES_DIR, paste0(cells$cell[i], ".rds"))
  if (file.exists(f)) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS),
                  function(r) one_dataset(r, cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  saveRDS(list(res = res,
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          f)
}

raw <- lapply(cells$cell, function(cl) {
  x <- readRDS(file.path(RES_DIR, paste0(cl, ".rds")))
  is_ok <- vapply(x$res, function(r) "item" %in% names(r), logical(1))
  list(ok = cbind(cell = cl, do.call(rbind, x$res[is_ok])),
       err = if (any(!is_ok)) cbind(cell = cl, do.call(rbind, x$res[!is_ok])),
       minutes = round(x$elapsed_min, 1))
})
long <- do.call(rbind, lapply(raw, `[[`, "ok"))
long$cell <- factor(long$cell, levels = cells$cell)
errors <- do.call(rbind, lapply(raw, `[[`, "err"))

data.frame(
  cell = cells$cell,
  datasets = vapply(raw, function(x) length(unique(x$ok$rep)), integer(1)),
  errors = vapply(raw, function(x) if (is.null(x$err)) 0L else nrow(x$err),
                  integer(1)),
  minutes = vapply(raw, `[[`, numeric(1), "minutes")
)
if (!is.null(errors)) count(errors, cell, error)
```

### Reproduction of the validation study

The |t| p-values should equal the cached conditional arm of the validation
study for every dataset both have.

```{r reproduce}
val <- do.call(rbind, lapply(cells$cell, function(cl) {
  x <- readRDS(file.path(VAL_DIR, paste0(cl, "_conditional.rds")))
  cbind(cell = cl, do.call(rbind, Filter(Negate(is.null), x$res)))
}))
chk <- inner_join(
  long |> mutate(cell = as.character(cell)) |>
    select(cell, rep, item, difference, p_abs, padj_abs_wy, padj_abs_bh),
  val |> select(cell, rep, item, difference_val = difference, p_val = p,
                padj_fwer_val = padj_fwer, padj_bh_val = padj_bh),
  by = c("cell", "rep", "item")
)
chk |>
  group_by(cell) |>
  summarise(items = n(),
            max_diff_difference = max(abs(difference - difference_val)),
            max_diff_p = max(abs(p_abs - p_val)),
            max_diff_padj_wy = max(abs(padj_abs_wy - padj_fwer_val)),
            max_diff_padj_bh = max(abs(padj_abs_bh - padj_bh_val)),
            .groups = "drop")
```

## Results

### Family-wise rate

Datasets with at least one item flagged at α = .05.

```{r fwer}
fw <- long |>
  group_by(cell, rep) |>
  summarise(eq_wy = any(padj_eq_wy < .05, na.rm = TRUE),
            abs_wy = any(padj_abs_wy < .05, na.rm = TRUE),
            eq_bh = any(padj_eq_bh < .05, na.rm = TRUE),
            abs_bh = any(padj_abs_bh < .05, na.rm = TRUE),
            .groups = "drop")
fw |>
  group_by(cell) |>
  summarise(datasets = n(),
            across(c(eq_wy, abs_wy, eq_bh, abs_bh), \(x) round(100 * mean(x), 1)),
            .groups = "drop")
fw |>
  summarise(datasets = n(),
            across(c(eq_wy, abs_wy, eq_bh, abs_bh), \(x) round(100 * mean(x), 2)))
```

Paired: datasets flagged by WY under one p-value and not the other.

```{r fwpaired}
fw |>
  group_by(cell) |>
  summarise(eq_only = sum(eq_wy & !abs_wy), abs_only = sum(!eq_wy & abs_wy),
            both = sum(eq_wy & abs_wy), .groups = "drop")
```

### Per-item rejection by direction

Marginal p < .05, split by direction, nominal 2.5 percent each.
`se` is the Monte Carlo SE with datasets as clusters.

```{r tails}
long |>
  group_by(cell, rep) |>
  summarise(eq_over = mean(flag_eq_marginal == "overfit"),
            eq_under = mean(flag_eq_marginal == "underfit"),
            abs_over = mean(flag_abs_marginal == "overfit"),
            abs_under = mean(flag_abs_marginal == "underfit"),
            .groups = "drop") |>
  group_by(cell) |>
  summarise(across(c(eq_over, eq_under, abs_over, abs_under),
                   list(pct = \(x) round(100 * mean(x), 2),
                        se = \(x) round(100 * sd(x) / sqrt(n()), 2))),
            .groups = "drop")
```

### Marginal p-values

The equal-tailed p-value takes the values 2k/(B + 1), so its histogram is
coarser than the |t| one. Both should be flat.

```{r phist, fig.height = 5}
long |>
  select(cell, p_eq, p_abs) |>
  pivot_longer(c(p_eq, p_abs), names_to = "pvalue", values_to = "p") |>
  ggplot(aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "steelblue",
                 colour = "white") +
  facet_grid(pvalue ~ cell, scales = "free_y") +
  labs(x = "Marginal bootstrap p-value", y = "Count") +
  theme_minimal()
```

### Synthetic nulls

The family-wise excess could come from the restscore bootstrap or from the
step-down itself. To separate the two, the variants below are run on nulls
where the observed vector is exchangeable with the simulated ones by
construction: B + 1 independent rows, the first taken as observed. Each
column is a left-skewed variable, `-exp(s * Z)`, with `Z` standard normal
and correlated `rho` across columns. 4000 replications per condition
(Monte Carlo SE about 0.35 points at 5 percent), B = 400.

The equal-tailed code has been removed from the package, so the variants
are defined here.

- `first`: the version that ran above. Simulated values are scored by their
  mid-rank among the simulated values, the observed value by its mid-rank
  against them.
- `pooled`: observed and simulated values ranked together, B + 1 per item,
  `qnorm((rank - 0.5) / (B + 1))`.
- `pooled_sym`: as `pooled`, but |z| computed from the rank counted from the
  nearer end, so the two tails give identical magnitudes.
- `two_piece`: continuous. The difference from the simulated median, scaled
  by (median − 2.5th percentile) below it and (97.5th percentile − median)
  above it.
- `abs_t`: the package's two-sided |t|.

```{r synthetic}
SYN_FILE <- file.path(RES_DIR, "synthetic.rds")
if (!file.exists(SYN_FILE)) {
  z_first <- function(o, sim) {
    sc <- function(x, ref) {
      ref <- sort(ref)
      b <- findInterval(x, ref, left.open = TRUE)
      a <- findInterval(x, ref)
      stats::qnorm((0.5 * (b + a) + 0.5) / (length(ref) + 1))
    }
    list(o = sc(o, sim), s = sc(sim, sim))
  }
  z_pooled <- function(o, sim, sym) {
    r <- rank(c(o, sim))
    n <- length(r)
    z <- if (sym) {
      sign(r - (n + 1) / 2) * -stats::qnorm((pmin(r, n + 1 - r) - 0.5) / n)
    } else {
      stats::qnorm((r - 0.5) / n)
    }
    list(o = z[1], s = z[-1])
  }
  z_two_piece <- function(o, sim) {
    q <- stats::quantile(sim, c(.025, .5, .975), names = FALSE)
    f <- function(v) ifelse(v < q[2], (v - q[2]) / (q[2] - q[1]),
                            (v - q[2]) / (q[3] - q[2]))
    list(o = f(o), s = f(sim))
  }
  z_abs_t <- function(o, sim) {
    m <- mean(sim); sdv <- stats::sd(sim)
    list(o = (o - m) / sdv, s = (sim - m) / sdv)
  }
  variants <- list(
    first = z_first,
    pooled = function(o, sim) z_pooled(o, sim, FALSE),
    pooled_sym = function(o, sim) z_pooled(o, sim, TRUE),
    two_piece = z_two_piece,
    abs_t = z_abs_t
  )
  one_syn <- function(r, k, rho, skew, B = 400) {
    set.seed(r)
    common <- rnorm(B + 1)
    x <- sapply(seq_len(k), function(j) {
      -exp(skew * (sqrt(rho) * common + sqrt(1 - rho) * rnorm(B + 1)))
    })
    x <- matrix(x, B + 1, k)
    unlist(lapply(variants, function(v) {
      zz <- lapply(seq_len(k), function(j) v(x[1, j], x[-1, j]))
      zo <- vapply(zz, `[[`, numeric(1), "o")
      zs <- sapply(zz, `[[`, "s")
      zs <- matrix(zs, B, k)
      padj <- .wy_stepdown(abs(zo), abs(zs), B)
      c(any = any(padj < .05), low = sum(padj < .05 & zo < 0),
        high = sum(padj < .05 & zo > 0))
    }))
  }
  grid <- expand.grid(k = c(1, 9, 20), rho = c(0, 0.5), skew = c(0.5, 0.15))
  syn <- do.call(rbind, lapply(seq_len(nrow(grid)), function(g) {
    res <- do.call(rbind, mclapply(seq_len(4000), function(r) {
      one_syn(r, grid$k[g], grid$rho[g], grid$skew[g])
    }, mc.cores = NCORES))
    cbind(grid[g, ], as.data.frame(t(colMeans(res[, grepl("any", colnames(res))]) * 100)),
          as.data.frame(t(colSums(res[, !grepl("any", colnames(res))]))))
  }))
  saveRDS(syn, SYN_FILE)
}
syn <- readRDS(SYN_FILE)
names(syn) <- sub("\\.", "_", names(syn))
syn |>
  select(skew, k, rho, ends_with("_any")) |>
  mutate(across(ends_with("_any"), \(x) round(x, 2)))
syn |>
  transmute(skew, k, rho,
            first = paste0(first_low, "/", first_high),
            pooled = paste0(pooled_low, "/", pooled_high),
            pooled_sym = paste0(pooled_sym_low, "/", pooled_sym_high),
            two_piece = paste0(two_piece_low, "/", two_piece_high),
            abs_t = paste0(abs_t_low, "/", abs_t_high))
```

The first table is the family-wise rate. The second counts WY flags in the
lower and upper tail, summed over replications.

## Interpretation

**1. The marginal equal-tailed test works.** Per item, the equal-tailed
p-value rejected 2.2 to 2.6 percent in each tail in all four cells, against
1.8 to 3.0 for |t|. Splitting α evenly between the tails is achieved at the
item level.

**2. Its Westfall-Young step-down, as first built, was liberal.** At least
one item was flagged in 8.85 percent of null datasets (8.8 to 11.0 by cell),
against 5.0 percent for |t| on the same simulated values, and the
equal-tailed test flagged 95 datasets that |t| did not, against 18 the other
way. The |t| results reproduced the validation study exactly, so the
difference lies in the p-value alone.

**3. The cause is in the step-down, not in the restscore bootstrap.** On the
synthetic nulls the first version is nominal with one item and liberal with
several. Simulated values were scored among themselves, so they can never
fall outside the simulated range, while the observed value can. An observed
value beyond every simulated value in its own column then beats every
simulated maximum, which happens with probability about 2/(B + 1) per item.
Ranking observed and simulated values together (`pooled`) restores
exchangeability.

**4. Ranks cannot give a usable step-down at B = 400.** Once the two tails
get identical magnitudes (`pooled_sym`), the test with 20 items almost never
rejects. With pooled ranks, each item has one value at each end of its B + 1
values, so roughly 2k simulated rows share the largest possible |z|, and the
smallest attainable adjusted p-value is about 2k/(B + 1), near .10 for 20
items. A rank-based step-down needs B well above 2k/α, 800 iterations for 20
items. The `pooled` version appeared to work only because `qnorm()` gives
the two ends magnitudes that differ in the last bits, which broke those ties
in favour of the upper tail.

**5. A continuous statistic holds the level but not the balance.** The
two-piece studentisation is continuous, so ties do not arise, and it held
the family-wise rate (4.8 to 5.9 percent). But its flags are still lopsided
toward the long tail, although less than with |t| at the milder skew (267
against 110 lower and upper flags with 20 correlated items, against 364 and
6 for |t|), which makes it the most promising lead for later work.
Matching the tails at the 2.5th and 97.5th percentiles does not match them
at the far more extreme quantiles where the step-down decides, and
estimating those needs many more iterations.

**Decision.** The equal-tailed p-value was removed from the package on
2026-10-02, and `RMitemRestscore()` keeps the two-sided |t| test validated in
`restscore_cutoff_validation.qmd`, with the tail imbalance stated in its help
page. An equal-tailed family-wise test is left as an open problem.

## Follow-up: two one-sided step-downs

Added 2026-10-02, after the withdrawal, as the first candidate in
`HANDOFF-restscore-equal-tails.md`. WY runs twice, once on t (large values
in the overfit direction) and once on −t, each at α/2, and an item is
flagged if either run rejects it. The statistic is continuous, so ties do
not arise, and observed and simulated values get the same transform, as in
|t|. Each direction is calibrated against its own one-sided maximum, so its
family-wise rate should be α/2 whatever the skew, and the total at most α.
The critical value is about the 10th largest of the 400 simulated maxima, so
B = 400 is enough.

Variants, all on the same replications:

- `abs_t`: the shipped two-sided |t| at α. Same seeds and code as the
  `synthetic` chunk, so it must reproduce the cached results.
- `split_t`: t in the upper run, −t in the lower run, each at α/2.
- `split_two_piece`: as `split_t`, but each run scales by its own tail, the
  upper run by (97.5th percentile − median) and the lower run by
  (median − 2.5th percentile), both from the simulated values. This is
  candidate 2 of the handoff. Each run applies one affine transform per item,
  so only the comparison across items changes.
- `marginal_eq`: the equal-tailed marginal test, for the k = 1 check.

In the `synthetic` grid all columns have the same distribution, so scaling
per item changes almost nothing and the two split variants should agree. They
can differ only when items differ in skew. The `mixed` arm adds that: the
skew parameter rises from 0.02 to 0.3 across items (absolute skewness about
0.06 to 0.95, which brackets the mean of 0.28 in the most skewed restscore
cell). Its flags are split by thirds of the items, from least to most skewed.

**What to check.**

1. `abs_t` reproduces the cached synthetic results.
2. With k = 1, `split_t`, `split_two_piece` and `marginal_eq` give the same
   flag counts.
3. The family-wise rate per direction is 2.5 percent, ± 0.5 (two Monte Carlo
   SEs at 4000 replications), and the total at or a little below 5.
4. In the `mixed` arm, flags in each direction spread evenly over the thirds.
   `split_t` should favour the most skewed items in the lower run and the
   least skewed in the upper run, and `split_two_piece` should reduce that.

About 10 to 15 minutes on 10 cores. Results are cached in
`restscore_eti_null/synthetic_split.rds`, so delete it to rerun.

```{r split-run}
SPLIT_FILE <- file.path(RES_DIR, "synthetic_split.rds")
REPS <- 4000
if (!file.exists(SPLIT_FILE)) {
  ALPHA <- 0.05
  # Per item, a statistic for each direction, oriented so that large values
  # are extreme in that direction.
  split_variants <- list(
    split_t = function(o, sim) {
      m <- mean(sim); sdv <- stats::sd(sim)
      to <- (o - m) / sdv; ts <- (sim - m) / sdv
      list(o_hi = to, s_hi = ts, o_lo = -to, s_lo = -ts)
    },
    split_two_piece = function(o, sim) {
      q <- stats::quantile(sim, c(.025, .5, .975), names = FALSE)
      up <- q[3] - q[2]; dn <- q[2] - q[1]
      list(o_hi = (o - q[2]) / up, s_hi = (sim - q[2]) / up,
           o_lo = (q[2] - o) / dn, s_lo = (q[2] - sim) / dn)
    }
  )
  wy_split <- function(v, x, B) {
    k <- ncol(x)
    zz <- lapply(seq_len(k), function(j) v(x[1, j], x[-1, j]))
    obs <- function(nm) vapply(zz, `[[`, numeric(1), nm)
    sims <- function(nm) matrix(sapply(zz, `[[`, nm), B, k)
    list(low = .wy_stepdown(obs("o_lo"), sims("s_lo"), B) < ALPHA / 2,
         high = .wy_stepdown(obs("o_hi"), sims("s_hi"), B) < ALPHA / 2)
  }
  # The shipped |t|, computed as in the `synthetic` chunk.
  wy_abs_t <- function(x, B) {
    k <- ncol(x)
    zz <- lapply(seq_len(k), function(j) {
      m <- mean(x[-1, j]); sdv <- stats::sd(x[-1, j])
      list(o = (x[1, j] - m) / sdv, s = (x[-1, j] - m) / sdv)
    })
    zo <- vapply(zz, `[[`, numeric(1), "o")
    zs <- matrix(sapply(zz, `[[`, "s"), B, k)
    padj <- .wy_stepdown(abs(zo), abs(zs), B)
    list(low = padj < ALPHA & zo < 0, high = padj < ALPHA & zo > 0)
  }
  # Equal-tailed marginal test, twice the smaller one-sided Monte Carlo p.
  marg_eq <- function(x, B) {
    o <- x[1, ]; s <- x[-1, , drop = FALSE]
    p_hi <- (1 + colSums(sweep(s, 2, o, ">="))) / (B + 1)
    p_lo <- (1 + colSums(sweep(s, 2, o, "<="))) / (B + 1)
    list(low = p_lo < ALPHA / 2, high = p_hi < ALPHA / 2)
  }
  # Same generator and seeds as `synthetic`, with skew allowed to vary by item.
  one_rep <- function(r, k, rho, skew, B = 400) {
    set.seed(r)
    common <- rnorm(B + 1)
    s <- rep_len(skew, k)
    x <- sapply(seq_len(k), function(j) {
      -exp(s[j] * (sqrt(rho) * common + sqrt(1 - rho) * rnorm(B + 1)))
    })
    x <- matrix(x, B + 1, k)
    fl <- c(list(abs_t = wy_abs_t(x, B), marginal_eq = marg_eq(x, B)),
            lapply(split_variants, wy_split, x = x, B = B))
    t(vapply(fl, function(f) {
      c(any = any(f$low | f$high), any_low = any(f$low),
        any_high = any(f$high), low = f$low, high = f$high)
    }, numeric(3 + 2 * k)))
  }
  grid <- rbind(
    expand.grid(k = c(1, 9, 20), rho = c(0, 0.5), skew = c("0.5", "0.15"),
                stringsAsFactors = FALSE),
    expand.grid(k = c(9, 20), rho = c(0, 0.5), skew = "mixed",
                stringsAsFactors = FALSE)
  )
  syn_split <- lapply(seq_len(nrow(grid)), function(g) {
    k <- grid$k[g]
    skew <- if (grid$skew[g] == "mixed") {
      seq(0.02, 0.3, length.out = k)
    } else {
      as.numeric(grid$skew[g])
    }
    res <- mclapply(seq_len(REPS), function(r) {
      one_rep(r, k, grid$rho[g], skew)
    }, mc.cores = NCORES)
    list(cond = grid[g, ], counts = Reduce(`+`, res))
  })
  saveRDS(syn_split, SPLIT_FILE)
}
syn_split <- readRDS(SPLIT_FILE)

split_tab <- do.call(rbind, lapply(syn_split, function(x) {
  cnt <- x$counts
  data.frame(
    skew = x$cond$skew, k = x$cond$k, rho = x$cond$rho,
    variant = rownames(cnt),
    any = 100 * cnt[, "any"] / REPS,
    any_low = 100 * cnt[, "any_low"] / REPS,
    any_high = 100 * cnt[, "any_high"] / REPS,
    flags_low = rowSums(cnt[, grep("^low", colnames(cnt)), drop = FALSE]),
    flags_high = rowSums(cnt[, grep("^high", colnames(cnt)), drop = FALSE]),
    row.names = NULL, stringsAsFactors = FALSE
  )
}))
```

### Checks

`abs_t` against the cached `synthetic` results. All three differences
should be 0.

```{r split-reproduce}
cached <- readRDS(SYN_FILE)
names(cached) <- sub("\\.", "_", names(cached))
cached |>
  mutate(skew = as.character(skew)) |>
  select(skew, k, rho, abs_t_any, abs_t_low, abs_t_high) |>
  inner_join(filter(split_tab, variant == "abs_t"),
             by = c("skew", "k", "rho")) |>
  summarise(conditions = n(),
            max_diff_any = max(abs(abs_t_any - any)),
            max_diff_low = max(abs(abs_t_low - flags_low)),
            max_diff_high = max(abs(abs_t_high - flags_high)))
```

With one item, the split variants and the marginal test should give the
same lower/upper flag counts.

```{r split-k1}
split_tab |>
  filter(k == 1, variant != "abs_t") |>
  transmute(skew, rho, variant, flags = paste0(flags_low, "/", flags_high)) |>
  pivot_wider(names_from = variant, values_from = flags)
```

### Family-wise rate and tail balance

```{r split-fwer}
split_tab |>
  filter(variant != "marginal_eq") |>
  transmute(skew, k, rho, variant, any = round(any, 2)) |>
  pivot_wider(names_from = variant, values_from = any)
split_tab |>
  filter(variant != "marginal_eq") |>
  transmute(skew, k, rho, variant,
            rate = paste0(round(any_low, 2), " / ", round(any_high, 2))) |>
  pivot_wider(names_from = variant, values_from = rate)
split_tab |>
  filter(variant != "marginal_eq") |>
  transmute(skew, k, rho, variant,
            flags = paste0(flags_low, "/", flags_high)) |>
  pivot_wider(names_from = variant, values_from = flags)
```

The first table is the family-wise rate, nominal 5. The second is the
family-wise rate per direction, lower / upper, nominal 2.5 each for the
split variants. The third counts flags in the lower and upper tail, summed
over replications, in the format of the `synthetic` table.

### Balance across items in the mixed arm

Lower/upper flags summed over replications, by thirds of the items from
least to most skewed.

```{r split-mixed}
mixed <- do.call(rbind, lapply(
  Filter(function(x) x$cond$skew == "mixed", syn_split),
  function(x) {
    cnt <- x$counts
    third <- cut(seq_len(x$cond$k), 3, labels = c("least", "middle", "most"))
    lo <- cnt[, grep("^low", colnames(cnt)), drop = FALSE]
    hi <- cnt[, grep("^high", colnames(cnt)), drop = FALSE]
    do.call(rbind, lapply(levels(third), function(t3) data.frame(
      k = x$cond$k, rho = x$cond$rho, variant = rownames(cnt),
      skew_third = t3,
      low = rowSums(lo[, third == t3, drop = FALSE]),
      high = rowSums(hi[, third == t3, drop = FALSE]),
      row.names = NULL, stringsAsFactors = FALSE
    )))
  }
))
mixed |>
  filter(variant != "marginal_eq") |>
  transmute(k, rho, variant, skew_third,
            flags = paste0(low, "/", high)) |>
  pivot_wider(names_from = skew_third, values_from = flags) |>
  arrange(k, rho, variant)
```

**Split results.** The checks passed: `abs_t` reproduced the cached
synthetic results, and with one item the split variants matched the marginal
test. Splitting the step-down balanced the two directions. Over the 12 cells
with k ≥ 9, `split_t` flagged at least one item in the lower direction in
2.70 percent of replications and in the upper in 2.68, against 4.94 and 0.09
for |t|. In the worst cell of the handoff (s = .15, k = 20, ρ = .5) the
lower/upper flag counts went from 364/6 to 166/211.

Within a direction the imbalance moved to the items. In the mixed arm with
20 correlated items, `split_t` gave the least, middle and most skewed thirds
17, 53 and 94 lower flags, and 138, 45 and 13 upper flags. Two-piece scaling
roughly halved that gradient, to 38, 52 and 78 lower and 103, 64 and 55
upper. Both split variants were slightly above the nominal rate, which the
next run traces to self-influence.

### Pooled moments and a second seed block

Over the 12 cells with k ≥ 9, `split_t` averaged 5.36 percent and
`split_two_piece` 5.68, against 5.03 for |t| on the same replications. Two
explanations fit. The conditions share seeds, and on these seeds the one-item
marginal test, nominal at 4.99 by construction (2 × 10/401), gave 5.15 and
5.28. Or self-influence: the mean, SD and quantiles come from the simulated
values only, so an extreme simulated value enlarges its own scale and shrinks
its own statistic, which the observed value never does to itself.

The `_pooled` variants compute the mean, SD and quantiles from all B + 1
values. The transform is then the same function of the column whichever row
is taken as observed, so the rows stay exchangeable and each one-sided
step-down rejects with probability 10/401 (2.49 percent) by construction. On
the same seeds, pooled rates near 2.49 per direction point to self-influence,
and pooled rates as high as the simulated-only ones point to seed noise. A
second block of 4000 seeds (4001 to 8000) gives a fresh estimate for every
variant.

The step-down here is a vectorised copy of `.wy_stepdown()`, which keeps the
maximum over the remaining items as a running maximum instead of recomputing
it at each step. Block 1 reruns seeds 1 to 4000, so its simulated-only counts
must equal the cached `split-run` results, which checks the copy against the
package function on every replication.

**What to check.**

1. Block 1 reproduces the `split-run` counts for `abs_t`, `split_t`,
   `split_two_piece` and `marginal_eq`.
2. With k = 1, all split variants and `marginal_eq` give the same flag
   counts.
3. Pooled variants at 2.49 percent per direction. The Monte Carlo SE is
   0.25 points per block and 0.17 over both blocks.
4. Paired, the replications flagged by one version and not the other. Under
   self-influence the simulated-only versions should account for nearly all
   of them.

Results are cached in `restscore_eti_null/synthetic_pooled.rds`.

```{r pooled-run}
POOL_FILE <- file.path(RES_DIR, "synthetic_pooled.rds")
BLOCKS <- list(block1 = 1:4000, block2 = 4001:8000)
if (!file.exists(POOL_FILE)) {
  ALPHA <- 0.05
  # Same result as .wy_stepdown() when nothing is missing. Items are taken
  # from the smallest observed statistic to the largest, and the maximum over
  # the remaining items is updated with pmax().
  wy_fast <- function(stat_obs, stat_sim, B) {
    k <- length(stat_obs)
    ord <- order(stat_obs, decreasing = TRUE)
    m <- rep(-Inf, nrow(stat_sim))
    p <- numeric(k)
    for (j in k:1) {
      m <- pmax(m, stat_sim[, ord[j]])
      p[j] <- (1 + sum(m >= stat_obs[ord[j]])) / (B + 1)
    }
    padj <- numeric(k)
    padj[ord] <- pmin(cummax(p), 1)
    padj
  }
  # The reference sample is the simulated values, or all B + 1 when pooled.
  t_stat <- function(o, sim, pooled) {
    r <- if (pooled) c(o, sim) else sim
    m <- mean(r); sdv <- stats::sd(r)
    list(o = (o - m) / sdv, s = (sim - m) / sdv)
  }
  tp_stat <- function(o, sim, pooled) {
    r <- if (pooled) c(o, sim) else sim
    q <- stats::quantile(r, c(.025, .5, .975), names = FALSE)
    up <- q[3] - q[2]; dn <- q[2] - q[1]
    list(o_hi = (o - q[2]) / up, s_hi = (sim - q[2]) / up,
         o_lo = (q[2] - o) / dn, s_lo = (q[2] - sim) / dn)
  }
  flags_abs <- function(zo, zs, B) {
    padj <- wy_fast(abs(zo), abs(zs), B)
    list(low = padj < ALPHA & zo < 0, high = padj < ALPHA & zo > 0)
  }
  flags_split <- function(o_hi, s_hi, o_lo, s_lo, B) {
    list(low = wy_fast(o_lo, s_lo, B) < ALPHA / 2,
         high = wy_fast(o_hi, s_hi, B) < ALPHA / 2)
  }
  # Same generator and seeds as `synthetic` and `split-run`.
  one_rep <- function(r, k, rho, skew, B = 400) {
    set.seed(r)
    common <- rnorm(B + 1)
    s <- rep_len(skew, k)
    x <- sapply(seq_len(k), function(j) {
      -exp(s[j] * (sqrt(rho) * common + sqrt(1 - rho) * rnorm(B + 1)))
    })
    x <- matrix(x, B + 1, k)
    fl <- list()
    for (pooled in c(FALSE, TRUE)) {
      sfx <- if (pooled) "_pooled" else ""
      tt <- lapply(seq_len(k), function(j) t_stat(x[1, j], x[-1, j], pooled))
      zo <- vapply(tt, `[[`, numeric(1), "o")
      zs <- matrix(sapply(tt, `[[`, "s"), B, k)
      fl[[paste0("abs_t", sfx)]] <- flags_abs(zo, zs, B)
      fl[[paste0("split_t", sfx)]] <- flags_split(zo, zs, -zo, -zs, B)
      tp <- lapply(seq_len(k), function(j) tp_stat(x[1, j], x[-1, j], pooled))
      tp_o <- function(nm) vapply(tp, `[[`, numeric(1), nm)
      tp_s <- function(nm) matrix(sapply(tp, `[[`, nm), B, k)
      fl[[paste0("split_two_piece", sfx)]] <- flags_split(
        tp_o("o_hi"), tp_s("s_hi"), tp_o("o_lo"), tp_s("s_lo"), B
      )
    }
    o <- x[1, ]; sm <- x[-1, , drop = FALSE]
    fl$marginal_eq <- list(
      low = (1 + colSums(sweep(sm, 2, o, "<="))) / (B + 1) < ALPHA / 2,
      high = (1 + colSums(sweep(sm, 2, o, ">="))) / (B + 1) < ALPHA / 2
    )
    t(vapply(fl, function(f) {
      c(any = any(f$low | f$high), any_low = any(f$low),
        any_high = any(f$high), low = f$low, high = f$high)
    }, numeric(3 + 2 * k)))
  }
  grid <- rbind(
    expand.grid(k = c(1, 9, 20), rho = c(0, 0.5), skew = c("0.5", "0.15"),
                stringsAsFactors = FALSE),
    expand.grid(k = c(9, 20), rho = c(0, 0.5), skew = "mixed",
                stringsAsFactors = FALSE)
  )
  syn_pooled <- list()
  for (g in seq_len(nrow(grid))) {
    k <- grid$k[g]
    skew <- if (grid$skew[g] == "mixed") {
      seq(0.02, 0.3, length.out = k)
    } else {
      as.numeric(grid$skew[g])
    }
    for (b in names(BLOCKS)) {
      res <- mclapply(BLOCKS[[b]], function(r) {
        one_rep(r, k, grid$rho[g], skew)
      }, mc.cores = NCORES)
      syn_pooled[[length(syn_pooled) + 1]] <- list(
        cond = grid[g, ], block = b, counts = Reduce(`+`, res),
        any = vapply(res, function(m) m[, "any"] == 1,
                     logical(nrow(res[[1]])))
      )
    }
  }
  saveRDS(syn_pooled, POOL_FILE)
}
syn_pooled <- readRDS(POOL_FILE)

pool_tab <- do.call(rbind, lapply(syn_pooled, function(x) {
  cnt <- x$counts
  reps <- length(BLOCKS[[x$block]])
  data.frame(
    skew = x$cond$skew, k = x$cond$k, rho = x$cond$rho, block = x$block,
    variant = rownames(cnt),
    any = 100 * cnt[, "any"] / reps,
    any_low = 100 * cnt[, "any_low"] / reps,
    any_high = 100 * cnt[, "any_high"] / reps,
    flags_low = rowSums(cnt[, grep("^low", colnames(cnt)), drop = FALSE]),
    flags_high = rowSums(cnt[, grep("^high", colnames(cnt)), drop = FALSE]),
    row.names = NULL, stringsAsFactors = FALSE
  )
}))
```

#### Checks

Block 1 against `split-run`, all count cells of the four shared variants.
The difference should be 0.

```{r pooled-reproduce}
b1 <- Filter(function(x) x$block == "block1", syn_pooled)
data.frame(
  conditions = length(b1),
  same_conditions = all(mapply(function(a, b) {
    identical(a$cond$k, b$cond$k) && identical(a$cond$rho, b$cond$rho) &&
      identical(a$cond$skew, b$cond$skew)
  }, syn_split, b1)),
  max_diff = max(mapply(function(a, b) {
    v <- rownames(a$counts)
    max(abs(a$counts[v, ] - b$counts[v, colnames(a$counts)]))
  }, syn_split, b1))
)
```

With one item, both blocks.

```{r pooled-k1}
pool_tab |>
  filter(k == 1, !grepl("^abs_t", variant)) |>
  transmute(skew, rho, block, variant,
            flags = paste0(flags_low, "/", flags_high)) |>
  pivot_wider(names_from = variant, values_from = flags)
```

#### Family-wise rate, simulated-only against pooled

Means over the 12 cells with k ≥ 9, per block, in percent. Nominal per
direction is 2.49 for the split variants.

```{r pooled-summary}
pool_tab |>
  filter(k > 1, variant != "marginal_eq") |>
  group_by(variant, block) |>
  summarise(fwer = round(mean(any), 2), lower = round(mean(any_low), 2),
            upper = round(mean(any_high), 2), .groups = "drop") |>
  pivot_wider(names_from = block, values_from = c(fwer, lower, upper))
```

Per cell, both blocks together (8000 replications), family-wise rate and
lower / upper rate.

```{r pooled-cells}
pool_both <- pool_tab |>
  filter(variant != "marginal_eq") |>
  group_by(skew, k, rho, variant) |>
  summarise(any = mean(any), any_low = mean(any_low),
            any_high = mean(any_high), .groups = "drop") |>
  mutate(skew = factor(skew, levels = c("0.5", "0.15", "mixed"))) |>
  arrange(skew, rho, k)
pool_both |>
  transmute(skew, k, rho, variant, any = round(any, 2)) |>
  pivot_wider(names_from = variant, values_from = any)
pool_both |>
  filter(grepl("^split", variant)) |>
  transmute(skew, k, rho, variant,
            rate = paste0(round(any_low, 2), " / ", round(any_high, 2))) |>
  pivot_wider(names_from = variant, values_from = rate)
```

#### Paired comparison

Replications flagged by the simulated-only version and not the pooled one,
and the reverse, summed over the 12 cells with k ≥ 9 and both blocks.

```{r pooled-paired}
pairs <- c("abs_t", "split_t", "split_two_piece")
do.call(rbind, lapply(syn_pooled, function(x) {
  a <- x$any
  data.frame(
    k = x$cond$k, variant = pairs,
    sim_only = vapply(pairs, function(v) {
      sum(a[v, ] & !a[paste0(v, "_pooled"), ])
    }, numeric(1)),
    pooled_only = vapply(pairs, function(v) {
      sum(!a[v, ] & a[paste0(v, "_pooled"), ])
    }, numeric(1)),
    row.names = NULL
  )
})) |>
  filter(k > 1) |>
  group_by(variant) |>
  summarise(sim_only = sum(sim_only), pooled_only = sum(pooled_only))
```

**Pooled results.** Block 1 reproduced the `split-run` counts in every cell,
so the vectorised step-down agrees with `.wy_stepdown()`. With one item,
block 2 gave 4.98 and 5.05 percent against 5.15 and 5.28 in block 1, so the
first seed block ran slightly high.

**1. The excess is self-influence.** Over 96,000 paired replications (12
cells, two blocks), the simulated-only versions flagged 331 (`abs_t`), 368
(`split_t`) and 636 (`split_two_piece`) replications that their pooled
versions did not, against 0, 0 and 27 the other way. Seed noise would give
discordance in both directions. The size, about 0.35 points for the t-based
statistics, is what one extreme value's pull on its own SD predicts, roughly
(t² − 1)/(2B) or 1 percent of scale at t = 3 and B = 400. Quantile-based
scales move further, since one value shifts the 97.5th percentile by a whole
order statistic.

**2. Pooling restores the level and keeps the balance.** `split_t_pooled`
flagged at least one item in 4.95 and 4.82 percent of replications in the
two blocks, with 2.48 and 2.48 percent in the lower direction and 2.50 and
2.36 in the upper, against 2.49 by construction. `split_two_piece_pooled`
gave 5.02 and 4.61.

**3. Two-piece scaling is the better split.** In the mixed arm, summed over
its four cells and both blocks, `split_t_pooled` gave the most skewed third
of the items 5.9 times the lower flags of the least skewed third, and the
least skewed third 19 times the upper flags of the most skewed. For
`split_two_piece_pooled` both ratios were 2.1. Pooling changed these ratios
little.

**4. The shipped |t| has the same bias.** The package version flagged 331
replications that pooled |t| did not, about 0.35 points. Pooled |t| came out
at 4.70 and 4.81 against 4.99 by construction, within the noise of cells that
share seeds, but worth watching on the real null. The validation study found
the package |t| at 5.0 on the real restscore null, so the bias is small in
practice. Pooled moments in `.bootstrap_pvalues()` would remove it for every
cutoff function, which is a separate decision.

**Decision.** `split_two_piece_pooled` is the candidate, with
`split_t_pooled` alongside since the two differ in item balance and may
differ in power. Both go to the real restscore null.

## Real restscore null for the split candidates

The synthetic nulls are exchangeable by construction. The restscore
bootstrap is only approximately so, since its null is simulated from
parameters estimated on the observed data. This run repeats the real null of
the first half of this document with the split and pooled variants: the four
cells and 500 datasets per cell of the validation study, conditional DGP,
B = 400, and the same seeds. All variants use the same simulated values.

- `abs_t`: the package's |t| from `RMitemRestscore()`. It must reproduce
  `restscore_cutoff_validation/`, and the local copy of its arithmetic
  (`padj_abs`) must equal it.
- `abs_t_pooled`, `split_t`, `split_t_pooled`, `split_two_piece`,
  `split_two_piece_pooled`: as in the synthetic sections, with the moment
  arithmetic of `.bootstrap_pvalues()` (`colMeans()` and `sd()` with
  `na.rm = TRUE`, zero-variance comparisons set to NA) and `.wy_stepdown()`.

The observed differences and simulated matrices of every dataset are saved
in `restscore_split_null/<cell>_sims.rds`, about 35 MB in all, so later
variants can be checked without rerunning the bootstrap.

**Outcomes.** The family-wise rate, nominal 5 percent with a Monte Carlo SE
of 0.97 points per cell and 0.49 over all 2000 datasets. The rate per
direction, nominal 2.5 for the split variants. The paired comparison of
simulated-only and pooled versions. Flags by thirds of the items, ranked by
the skewness of their simulated null. With about 25 flags per cell and
variant, the last table is descriptive.

**Cost.** About 2.2 hours on 10 cores, the time of the validation study's
conditional arm. The run is off by default. Set `RUN_SPLIT_NULL <- TRUE` to
start it. Results are saved per cell, so an interrupted render resumes, and
the result chunks run once all four cells are saved.

```{r split-null-setup}
RUN_SPLIT_NULL <- TRUE
SPLIT_NULL_DIR <- "restscore_split_null"
split_null_files <- file.path(SPLIT_NULL_DIR, paste0(cells$cell, ".rds"))
SPLIT_NULL_READY <- all(file.exists(split_null_files))
```

```{r split-null-functions}
# Adjusted p-values of every variant for one dataset, with the arithmetic of
# .bootstrap_pvalues().
split_null_variants <- function(obs, sim) {
  sim <- sim[, names(obs), drop = FALSE]
  B <- nrow(sim)
  guard <- function(x) {
    x[!is.finite(x) | x < 1e-8] <- NA_real_
    x
  }
  out <- data.frame(item = names(obs), stringsAsFactors = FALSE)
  for (pooled in c(FALSE, TRUE)) {
    sfx <- if (pooled) "_pooled" else ""
    ref <- if (pooled) rbind(obs, sim) else sim
    m <- colMeans(ref, na.rm = TRUE)
    s <- guard(apply(ref, 2L, stats::sd, na.rm = TRUE))
    t_obs <- (obs - m) / s
    t_sim <- sweep(sweep(sim, 2L, m, `-`), 2L, s, `/`)
    if (!pooled) out$null_skew <- colMeans(t_sim^3, na.rm = TRUE)
    q <- apply(ref, 2L, stats::quantile, probs = c(.025, .5, .975),
               na.rm = TRUE, names = FALSE)
    up <- guard(q[3, ] - q[2, ])
    dn <- guard(q[2, ] - q[1, ])
    dev_sim <- sweep(sim, 2L, q[2, ], `-`)
    out[[paste0("t", sfx)]] <- t_obs
    out[[paste0("padj_abs", sfx)]] <- .wy_stepdown(abs(t_obs), abs(t_sim), B)
    out[[paste0("padj_t_hi", sfx)]] <- .wy_stepdown(t_obs, t_sim, B)
    out[[paste0("padj_t_lo", sfx)]] <- .wy_stepdown(-t_obs, -t_sim, B)
    out[[paste0("padj_tp_hi", sfx)]] <- .wy_stepdown(
      (obs - q[2, ]) / up, sweep(dev_sim, 2L, up, `/`), B
    )
    out[[paste0("padj_tp_lo", sfx)]] <- .wy_stepdown(
      (q[2, ] - obs) / dn, sweep(-dev_sim, 2L, dn, `/`), B
    )
  }
  out
}

# Same data, seed and cutoff call as one_dataset() and the validation study.
one_dataset_split <- function(rep, cell_row) {
  thr <- item_sets[[cell_row$set]]
  seed <- 1e6 * match(cell_row$cell, cells$cell) + rep
  set.seed(seed)
  repeat {
    d <- as.data.frame(sim_pcm(rnorm(cell_row$n, cell_row$mu, THETA_SD), thr))
    names(d) <- paste0("I", seq_along(thr))
    full <- all(mapply(function(x, m) all(tabulate(x + 1, m + 1) > 0),
                       d, lengths(thr)))
    if (full) break
  }

  tryCatch({
    co <- suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = "conditional"
    ))
    fw <- suppressMessages(RMitemRestscore(
      d, cutoff = co, correction = "fwer", output = "dataframe"
    ))
    sim_mat <- tapply(co$results$Difference,
                      list(co$results$iteration, co$results$Item),
                      function(x) x[1L])
    obs <- stats::setNames(fw$Difference, fw$Item)
    list(
      table = data.frame(
        rep = rep, difference = obs, p_pkg = fw$p_restscore,
        padj_pkg = fw$padj_restscore, split_null_variants(obs, sim_mat),
        actual_iterations = co$actual_iterations,
        row.names = NULL, stringsAsFactors = FALSE
      ),
      sims = list(rep = rep, obs = obs,
                  sim = sim_mat[, names(obs), drop = FALSE])
    )
  }, error = function(e) {
    list(error = data.frame(rep = rep, error = conditionMessage(e)))
  })
}
```

```{r split-null-run}
#| eval: !expr RUN_SPLIT_NULL
dir.create(SPLIT_NULL_DIR, showWarnings = FALSE)
for (i in seq_len(nrow(cells))) {
  if (file.exists(split_null_files[i])) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS),
                  function(r) one_dataset_split(r, cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  # A worker that dies returns a try-error, kept as an error row.
  res <- lapply(res, function(x) {
    if (is.list(x)) x else list(error = data.frame(rep = NA, error = as.character(x)))
  })
  ok <- vapply(res, function(x) is.null(x$error), logical(1))
  # Simulated values first, so a saved result implies saved simulations.
  saveRDS(lapply(res[ok], `[[`, "sims"),
          file.path(SPLIT_NULL_DIR, paste0(cells$cell[i], "_sims.rds")))
  saveRDS(list(res = lapply(res, function(x) {
                 if (is.null(x$error)) x$table else x$error
               }),
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          split_null_files[i])
}
SPLIT_NULL_READY <- all(file.exists(split_null_files))
```

```{r split-null-load}
#| eval: !expr SPLIT_NULL_READY
split_raw <- lapply(cells$cell, function(cl) {
  x <- readRDS(file.path(SPLIT_NULL_DIR, paste0(cl, ".rds")))
  is_ok <- vapply(x$res, function(r) "item" %in% names(r), logical(1))
  list(ok = cbind(cell = cl, do.call(rbind, x$res[is_ok])),
       err = if (any(!is_ok)) cbind(cell = cl, do.call(rbind, x$res[!is_ok])),
       minutes = round(x$elapsed_min, 1))
})
split_null <- do.call(rbind, lapply(split_raw, `[[`, "ok"))
split_null$cell <- factor(split_null$cell, levels = cells$cell)
split_errors <- do.call(rbind, lapply(split_raw, `[[`, "err"))

data.frame(
  cell = cells$cell,
  datasets = vapply(split_raw, function(x) length(unique(x$ok$rep)),
                    integer(1)),
  errors = vapply(split_raw, function(x) {
    if (is.null(x$err)) 0L else nrow(x$err)
  }, integer(1)),
  minutes = vapply(split_raw, `[[`, numeric(1), "minutes")
)
if (!is.null(split_errors)) count(split_errors, cell, error)
```

### Checks

The package |t| against the validation study's conditional arm, and the
local copy of its arithmetic against the package. All differences should
be 0.

```{r split-null-reproduce}
#| eval: !expr SPLIT_NULL_READY
val_split <- do.call(rbind, lapply(cells$cell, function(cl) {
  x <- readRDS(file.path(VAL_DIR, paste0(cl, "_conditional.rds")))
  cbind(cell = cl, do.call(rbind, Filter(Negate(is.null), x$res)))
}))
inner_join(
  split_null |> mutate(cell = as.character(cell)) |>
    select(cell, rep, item, difference, p_pkg, padj_pkg, padj_abs),
  val_split |> select(cell, rep, item, difference_val = difference,
                      p_val = p, padj_val = padj_fwer),
  by = c("cell", "rep", "item")
) |>
  group_by(cell) |>
  summarise(items = n(),
            max_diff_difference = max(abs(difference - difference_val)),
            max_diff_p = max(abs(p_pkg - p_val), na.rm = TRUE),
            max_diff_padj = max(abs(padj_pkg - padj_val), na.rm = TRUE),
            max_diff_local = max(abs(padj_abs - padj_pkg), na.rm = TRUE),
            .groups = "drop")
```

### Family-wise rate and tail balance

```{r split-null-fwer}
#| eval: !expr SPLIT_NULL_READY
yes <- function(x) !is.na(x) & x
split_variant_levels <- c("abs_t", "abs_t_pooled", "split_t",
                          "split_t_pooled", "split_two_piece",
                          "split_two_piece_pooled")
split_flags <- split_null |>
  transmute(
    cell, rep, item, null_skew,
    abs_t_low = yes(padj_abs < .05 & t < 0),
    abs_t_high = yes(padj_abs < .05 & t > 0),
    abs_t_pooled_low = yes(padj_abs_pooled < .05 & t_pooled < 0),
    abs_t_pooled_high = yes(padj_abs_pooled < .05 & t_pooled > 0),
    split_t_low = yes(padj_t_lo < .025),
    split_t_high = yes(padj_t_hi < .025),
    split_t_pooled_low = yes(padj_t_lo_pooled < .025),
    split_t_pooled_high = yes(padj_t_hi_pooled < .025),
    split_two_piece_low = yes(padj_tp_lo < .025),
    split_two_piece_high = yes(padj_tp_hi < .025),
    split_two_piece_pooled_low = yes(padj_tp_lo_pooled < .025),
    split_two_piece_pooled_high = yes(padj_tp_hi_pooled < .025)
  ) |>
  pivot_longer(-c(cell, rep, item, null_skew),
               names_to = c("variant", "direction"),
               names_pattern = "(.*)_(low|high)", values_to = "flag") |>
  mutate(variant = factor(variant, levels = split_variant_levels))

split_fw <- split_flags |>
  group_by(cell, rep, variant) |>
  summarise(any = any(flag), low = any(flag & direction == "low"),
            high = any(flag & direction == "high"), .groups = "drop")
split_fw_summary <- bind_rows(
  split_fw |> mutate(cell = as.character(cell)),
  split_fw |> mutate(cell = "all")
) |>
  mutate(cell = factor(cell, levels = c(cells$cell, "all"))) |>
  group_by(cell, variant) |>
  summarise(datasets = n(), fwer = round(100 * mean(any), 2),
            rate = paste0(round(100 * mean(low), 2), " / ",
                          round(100 * mean(high), 2)),
            .groups = "drop")
split_fw_summary |>
  select(cell, datasets, variant, fwer) |>
  pivot_wider(names_from = variant, values_from = fwer)
split_fw_summary |>
  select(cell, variant, rate) |>
  pivot_wider(names_from = variant, values_from = rate)
```

The first table is the family-wise rate, nominal 5. The second is the rate
per direction, lower / upper, nominal 2.5 each for the split variants.

### Paired comparison

Datasets flagged by the simulated-only version and not the pooled one, and
the reverse, over all four cells.

```{r split-null-paired}
#| eval: !expr SPLIT_NULL_READY
split_pw <- split_fw |>
  select(cell, rep, variant, any) |>
  pivot_wider(names_from = variant, values_from = any)
split_pairs <- c("abs_t", "split_t", "split_two_piece")
data.frame(
  variant = split_pairs,
  sim_only = vapply(split_pairs, function(v) {
    sum(split_pw[[v]] & !split_pw[[paste0(v, "_pooled")]])
  }, integer(1)),
  pooled_only = vapply(split_pairs, function(v) {
    sum(!split_pw[[v]] & split_pw[[paste0(v, "_pooled")]])
  }, integer(1)),
  row.names = NULL
)
```

### Balance across items

Items are ranked within each cell by the mean skewness of their simulated
null, most negative first, and split into thirds. The first table gives the
range of skewness, the second the lower/upper flags summed over datasets.

```{r split-null-items}
#| eval: !expr SPLIT_NULL_READY
split_thirds <- split_null |>
  group_by(cell, item) |>
  summarise(skew = mean(null_skew, na.rm = TRUE), .groups = "drop") |>
  group_by(cell) |>
  mutate(skew_third = cut(rank(skew), 3,
                          labels = c("most", "middle", "least"))) |>
  ungroup()
split_thirds |>
  group_by(cell) |>
  summarise(items = n(), min_skew = round(min(skew), 2),
            mean_skew = round(mean(skew), 2), max_skew = round(max(skew), 2))
split_flags |>
  inner_join(select(split_thirds, cell, item, skew_third),
             by = c("cell", "item")) |>
  group_by(cell, variant, skew_third, direction) |>
  summarise(n = sum(flag), .groups = "drop") |>
  pivot_wider(names_from = direction, values_from = n) |>
  transmute(cell, variant, skew_third, flags = paste0(low, "/", high)) |>
  pivot_wider(names_from = skew_third, values_from = flags) |>
  select(cell, variant, most, middle, least) |>
  arrange(cell, variant)
```

### Interpretation of the real null

All 2000 datasets ran without errors (25 to 46 minutes per cell). The package
|t| reproduced the validation study, and the local copy of its arithmetic
reproduced the package, with every difference 0 over 29,000 item results.

**1. The split gives equal tails on real data.** Over all 2000 datasets, the
shipped |t| flagged at least one item in the underfit direction in 3.70
percent of datasets and in the overfit direction in 1.55 percent.
`split_t_pooled` gave 2.40 and 2.75, and `split_two_piece_pooled` 2.40 and
2.55 (Monte Carlo SE 0.35).

**2. Both pooled split variants hold the level.** The family-wise rate was
5.0 percent for |t|, 5.0 for `split_t_pooled` and 4.9 for
`split_two_piece_pooled` (SE 0.49). Without pooling, `split_t` reached 5.5
and `split_two_piece` 6.0, and the paired comparison shows the same
self-influence as on the synthetic nulls: the simulated-only versions
flagged 8, 10 and 22 datasets that their pooled versions did not, and the
pooled versions none the other way. Pooling is required for the two-piece
scale.

**3. `dich_tgt_300` is high for every variant.** All six gave 6.4 to 8.0
percent in that cell, and the shipped |t| gave 6.8 there in the validation
study too. The excess comes from the bootstrap in that cell, not from the
step-down, and pooling does not remove it.

**4. Item balance cannot be judged here.** Apart from `dich_mis_150`, the
skewness of the items' simulated nulls barely differs within a cell (−0.18
to −0.12 in `dich_tgt_300`, −0.05 to −0.02 in `poly_tgt_1000`), so there is
little imbalance for two-piece scaling to correct, and each third of the
items has 0 to 13 flags per cell. In `dich_mis_150` (−0.69 to −0.15) both
split variants put most overfit flags on the least skewed items, and
two-piece scaling did not visibly change that. The case for
`split_two_piece_pooled` over `split_t_pooled` rests on the synthetic mixed
arm.

**Next.** Power, in `restscore_power_conditional.qmd`, for both pooled split
variants against the shipped |t|.
