---
title: "Item-restscore power with the conditional DGP"
subtitle: "Follow-up to restscore_infit_power.qmd, reduced design"
author: "Magnus Johansson"
date: today
---

## Purpose

`restscore_infit_power.qmd` compared item-restscore with infit under planted
misfit, with both cutoff functions at their defaults at the time
(`dgp = "resample"`). After that study and the validation study
(`restscore_cutoff_validation.qmd`), `RMitemRestscoreCutoff()` changed in two
ways:

1. The default DGP is now `"conditional"`, which held the nominal
   family-wise rate under the null where `"resample"` was about 1.4 points
   liberal.
2. The bootstrap p-values in `RMitemRestscore()` are equal-tailed. Folding
   |t| at the mean gave the long underfit tail more than α/2.

This study measures what the shipped configuration does under misfit, and
separates the two changes. Each restscore cutoff is simulated once with the
conditional DGP, and the p-values are computed from the same simulation
both ways, equal-tailed (as shipped) and |t| (as before). Comparing

- conditional |t| with the earlier resample |t| isolates the DGP, and
- conditional equal-tailed with conditional |t| isolates the p-value.

Infit is unchanged, so its results are reused from the earlier study, as
is the asymptotic test.

## Design

Identical to `restscore_infit_power.qmd` (7 items, 5 categories, underfit by
a second dimension at ρ = .85, overfit by slope 1.6, 1 or 2 misfitting items
at 0 or −2 logits, 100 datasets per cell, B = 400), with two exceptions:

- **Sample size** 150 and 400 only, 16 of the 32 cells.
- **Methods** run here: `RMitemRestscoreCutoff(dgp = "conditional")`, then
  WY and BH, each with the equal-tailed and the |t| p-value.

The datasets are regenerated from the same seeds, so every comparison with
the earlier study is paired. In the earlier study 57 datasets failed in both
cutoff functions, because a category was used only by respondents with a
zero or perfect total score (all of them in the n = 150 and 250 cells, and
one at n = 400). The conditional DGP reproduces those respondents' patterns,
so the restscore cutoff is expected to run on them. They are counted, but
left out of the comparisons with infit, which has no result for them.

**Cost.** About 30 to 45 s per dataset for the conditional restscore cutoff
alone, from the timings in the validation study. 1,600 datasets is then
roughly 1.5 to 2 hours on 10 cores. Results are saved per cell, so an
interrupted render resumes. Set `R_DATASETS` lower for a quick run.

```{r setup}
suppressMessages({
  devtools::load_all("..", quiet = TRUE)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
options(rgl.useNULL = TRUE)

# The study depends on the conditional default; stop rather than silently
# measure the old behaviour. (It also checked for the equal-tailed p-value,
# which has since been withdrawn; the cached cond_eq_* results came from it.)
stopifnot(
  "conditional is not the RMitemRestscoreCutoff() default" =
    identical(eval(formals(RMitemRestscoreCutoff)$dgp)[1], "conditional")
)

R_DATASETS <- 100
B          <- 400
RHO        <- 0.85
SLOPE      <- 1.6
NCORES     <- 10
NS         <- c(150, 400)
OLD_DIR    <- "restscore_infit_power"
RES_DIR    <- "restscore_power_conditional"
dir.create(RES_DIR, showWarnings = FALSE)

source("restscore_power_generator.R")

# The full grid of the earlier study, in its order: the seed of a dataset is
# 1e6 * (its row in this grid) + rep, so the grid must not be reordered.
cells <- expand.grid(
  n = c(150, 250, 400, 600),
  target = c(0, -2),
  n_misfit = c(1, 2),
  direction = c("underfit", "overfit"),
  stringsAsFactors = FALSE
)
cells$cell <- with(cells, sprintf("%s_%ditem_t%s_n%d", direction, n_misfit,
                                  ifelse(target == 0, "0", "m2"), n))
cells$row <- seq_len(nrow(cells))
run_cells <- cells[cells$n %in% NS, ]
nrow(run_cells)
```

### One dataset

```{r onedataset}
one_dataset <- function(rep, cell_row) {
  setup <- item_setup(cell_row$n_misfit, cell_row$target)
  seed <- 1e6 * cell_row$row + rep
  d <- sim_power_data(
    n = cell_row$n, n_misfit = cell_row$n_misfit, target = cell_row$target,
    direction = cell_row$direction, rho = RHO, slope = SLOPE, seed = seed
  )

  tryCatch({
    cr <- suppressWarnings(suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = "conditional"
    )))
    q <- function(expr) suppressWarnings(suppressMessages(expr))
    flags <- function(tab) tab$Flagged[match(names(d), tab$Item)]
    eq_wy <- q(RMitemRestscore(d, cutoff = cr, output = "dataframe"))
    eq_bh <- q(RMitemRestscore(d, cutoff = cr, correction = "fdr_bh",
                               output = "dataframe"))

    # The |t| p-values from the same simulation, flagged the way
    # RMitemRestscore() did before the equal-tailed change.
    sim_mat <- tapply(cr$results$Difference,
                      list(cr$results$iteration, cr$results$Item),
                      function(x) x[1L])
    obs <- stats::setNames(eq_wy$Difference, eq_wy$Item)
    null_mean <- colMeans(sim_mat, na.rm = TRUE)[names(obs)]
    abs_flags <- function(correction) {
      pv <- .bootstrap_pvalues(obs, sim_mat, correction = correction,
                               tail = "two.sided")
      padj <- pv$padj[match(names(obs), pv$name)]
      f <- ifelse(!is.na(padj) & padj < .05,
                  ifelse(obs > null_mean, "overfit", "underfit"), "")
      f[match(names(d), names(obs))]
    }

    data.frame(
      rep = rep,
      item = names(d),
      misfit = seq_len(7) %in% setup$misfit,
      cond_eq_wy = flags(eq_wy),
      cond_eq_bh = flags(eq_bh),
      cond_abs_wy = abs_flags("fwer"),
      cond_abs_bh = abs_flags("fdr_bh"),
      iterations = cr$actual_iterations,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    data.frame(rep = rep, error = conditionMessage(e))
  })
}
```

Errors are kept as rows rather than returned as `NULL`, so a failure is
visible in the results instead of looking like a missing dataset.

### Run

```{r run}
for (i in seq_len(nrow(run_cells))) {
  f <- file.path(RES_DIR, paste0(run_cells$cell[i], ".rds"))
  if (file.exists(f)) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS),
                  function(r) one_dataset(r, run_cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  saveRDS(list(res = res,
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          f)
}

read_cell <- function(dir, cell) {
  x <- readRDS(file.path(dir, paste0(cell, ".rds")))
  x$res
}

# New results: successful datasets and errors kept apart.
new_raw <- lapply(run_cells$cell, function(cl) {
  res <- read_cell(RES_DIR, cl)
  is_ok <- vapply(res, function(x) is.data.frame(x) && "item" %in% names(x),
                  logical(1))
  list(ok = do.call(rbind, res[is_ok]),
       err = do.call(rbind, res[!is_ok]))
})
new <- do.call(rbind, Map(function(x, cl) cbind(cell = cl, x$ok),
                          new_raw, run_cells$cell))
errors <- do.call(rbind, Map(function(x, cl) {
  if (is.null(x$err)) NULL else cbind(cell = cl, x$err)
}, new_raw, run_cells$cell))

# Earlier results (infit, resample restscore with |t|, asymptotic).
old <- do.call(rbind, lapply(run_cells$cell, function(cl) {
  res <- read_cell(OLD_DIR, cl)
  ok <- Filter(Negate(is.null), res)
  cbind(cell = cl, do.call(rbind, ok))
})) |>
  rename(rest_resample_abs_wy = rest_boot_wy,
         rest_resample_abs_bh = rest_boot_bh)

run_cells |>
  select(cell) |>
  left_join(count(distinct(new, cell, rep), cell, name = "new_ok"),
            by = "cell") |>
  left_join(count(distinct(old, cell, rep), cell, name = "old_ok"),
            by = "cell") |>
  mutate(new_errors = vapply(cell, function(cl) {
    if (is.null(errors)) 0L else sum(errors$cell == cl)
  }, integer(1)),
  minutes = vapply(cell, function(cl) {
    round(readRDS(file.path(RES_DIR, paste0(cl, ".rds")))$elapsed_min, 1)
  }, numeric(1)))

if (!is.null(errors)) count(errors, error)
```

### Datasets that failed in the earlier study

```{r rescued}
rescued <- new |>
  distinct(cell, rep) |>
  anti_join(distinct(old, cell, rep), by = c("cell", "rep"))
nrow(rescued)
count(rescued, cell)
```

### Paired data

Comparisons below use only datasets with results in both studies.

```{r join}
methods <- c("infit_wy", "rest_resample_abs_wy", "cond_abs_wy", "cond_eq_wy",
             "infit_bh", "rest_resample_abs_bh", "cond_abs_bh", "cond_eq_bh",
             "rest_asy_bh")

both <- inner_join(
  old |> select(cell, rep, item, misfit, infit_wy, infit_bh,
                rest_resample_abs_wy, rest_resample_abs_bh, rest_asy_bh),
  new |> select(cell, rep, item, cond_eq_wy, cond_eq_bh, cond_abs_wy,
                cond_abs_bh),
  by = c("cell", "rep", "item")
) |>
  left_join(run_cells |> select(cell, n, target, n_misfit, direction),
            by = "cell")

long <- both |>
  pivot_longer(all_of(methods), names_to = "method", values_to = "flag") |>
  mutate(flag = ifelse(is.na(flag), "", flag),
         method = factor(method, levels = methods),
         targeting = ifelse(target == 0, "misfit at 0", "misfit at -2"),
         misfit_items = paste(n_misfit, ifelse(n_misfit == 1, "item", "items")))
```

## Results

### Detection of the misfitting items

Proportion of misfitting items flagged in the correct direction, WY methods
first, then BH and the asymptotic test.

```{r detection}
det <- long |>
  filter(misfit) |>
  group_by(direction, misfit_items, targeting, n, method) |>
  summarise(correct = round(100 * mean(flag == direction)),
            .groups = "drop")

det |>
  pivot_wider(names_from = method, values_from = correct) |>
  arrange(direction, misfit_items, desc(targeting), n)
```

```{r detplot, fig.height = 7}
det |>
  filter(method %in% c("infit_wy", "rest_resample_abs_wy", "cond_eq_wy",
                       "rest_asy_bh")) |>
  ggplot(aes(n, correct, colour = method)) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_grid(direction + misfit_items ~ targeting) +
  scale_x_continuous(breaks = NS) +
  labs(x = "n", y = "Misfitting items detected (%)", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Paired comparisons

Misfitting items detected by one method and not the other, pooled over
sample size and the number of misfitting items. `difference_pts` is the
first method minus the second.

```{r paired}
paired <- function(a, b) {
  both |>
    filter(misfit) |>
    mutate(hit_a = .data[[a]] == direction, hit_b = .data[[b]] == direction,
           hit_a = !is.na(hit_a) & hit_a, hit_b = !is.na(hit_b) & hit_b) |>
    group_by(direction, targeting = ifelse(target == 0, "at 0", "at -2")) |>
    summarise(items = n(),
              a_only = sum(hit_a & !hit_b),
              b_only = sum(!hit_a & hit_b),
              difference_pts = round(100 * (mean(hit_a) - mean(hit_b)), 1),
              .groups = "drop") |>
    mutate(comparison = paste(a, "vs", b), .before = 1)
}
bind_rows(
  paired("cond_abs_wy", "rest_resample_abs_wy"),  # the DGP
  paired("cond_abs_wy", "infit_wy"),              # shipped restscore vs infit
  paired("cond_abs_wy", "rest_asy_bh"),           # shipped restscore vs iarm
  paired("cond_eq_wy", "cond_abs_wy")             # withdrawn equal-tailed version
)
```

### False flags on the fitting items, by direction

```{r fp}
long |>
  filter(!misfit) |>
  group_by(direction, misfit_items, method) |>
  summarise(false_overfit = round(100 * mean(flag == "overfit"), 1),
            false_underfit = round(100 * mean(flag == "underfit"), 1),
            .groups = "drop") |>
  pivot_wider(names_from = misfit_items,
              values_from = c(false_overfit, false_underfit))
```

### Clean decisions

Every misfitting item flagged in the correct direction and no fitting item
flagged.

```{r clean}
long |>
  group_by(direction, misfit_items, targeting, n, method, rep) |>
  summarise(clean = all(flag[misfit] == direction[1]) &
              all(flag[!misfit] == ""),
            .groups = "drop") |>
  group_by(direction, misfit_items, targeting, n, method) |>
  summarise(clean = round(100 * mean(clean)), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = clean) |>
  arrange(direction, misfit_items, desc(targeting), n)
```

### Pooled detection

```{r pooled}
long |>
  filter(misfit) |>
  group_by(direction, method) |>
  summarise(detected = round(100 * mean(flag == direction), 1),
            .groups = "drop") |>
  pivot_wider(names_from = method, values_from = detected)
```

### Iterations kept

Simulated datasets with an unobserved category are rejected, so a cutoff can
rest on fewer than the requested 400 iterations.

```{r iterations}
new |>
  distinct(cell, rep, iterations) |>
  left_join(run_cells |> select(cell, n, target), by = "cell") |>
  group_by(n, target) |>
  summarise(datasets = n(),
            below_400 = round(100 * mean(iterations < 400), 1),
            below_300 = round(100 * mean(iterations < 300), 1),
            min = min(iterations), .groups = "drop")
```

### Rescued datasets

Detection with the shipped configuration in the datasets that failed in the
earlier study.

```{r rescueddet}
new |>
  semi_join(rescued, by = c("cell", "rep")) |>
  left_join(run_cells |> select(cell, direction), by = "cell") |>
  filter(misfit) |>
  group_by(cell) |>
  summarise(misfit_items = n(),
            detected = round(100 * mean(cond_eq_wy == direction)),
            .groups = "drop")
```

## Interpretation

**The equal-tailed arm is withdrawn.** The `cond_eq_*` columns come from the
first equal-tailed version, whose Westfall-Young step-down turned out to be
liberal (`restscore_eti_null.qmd`), and the approach was then dropped from
the package. They are kept in the tables as a record and not interpreted.
`RMitemRestscore()` ships with the conditional DGP and the two-sided |t|
p-value, which is `cond_abs_wy` here.

The results rest on 100 datasets per cell. Differences of a few points
within one cell are noise, and the statements below rely on the pooled
paired comparisons.

**1. Every dataset ran.** All 1,600 datasets gave a restscore result, with no
errors. The 36 datasets in these cells that failed in the earlier study,
where a category was used only by respondents with a zero or perfect total
score, all ran with the conditional DGP, because it reproduces those
respondents' patterns. They come from the hardest cells, and detection in
them was low (0 to 40 percent in the four cells with five or more
misfitting items).

**2. The DGP changes little, and mainly off target.** With the |t| p-value,
the conditional DGP detected overfitting items at −2 logits 4.5 points less
often than the resample DGP (6 against 31 discordant items). In the other
three direction and targeting combinations the two were within 1.4 points.
This is the direction expected from the null study, where resample was
about 1.4 points liberal.

**3. Against infit, the shipped test is more sensitive to overfit and about
equal for underfit.** Conditional WY detected overfitting items 7.8 points
more often than infit WY off target (71 against 27 discordant items) and 6.9
points more often on target (49 against 8). For underfit the two were close
(+1.4 off target, −2.2 on target). Pooled, overfit detection was 45.0 against
37.7 percent and underfit 36.9 against 37.3. The conclusions of
`restscore_infit_power.qmd` hold with the restscore test at its nominal
level.

**4. Against the asymptotic iarm test, the shipped test recovers underfit.**
Conditional WY detected underfitting items 22.3 points more often than
asymptotic BH off target and 22.7 points more often on target, and iarm
detected no underfitting item that the bootstrap missed (131 and 136 against
0 discordant items). For overfit, iarm was ahead off target by 6.6 points (9
against 46) and level on target. Part of iarm's overfit lead comes from its
liberal overfit tail at small n, which also produces false overfit flags
under the null.

**5. The cascade.** With an underfitting item planted, the conditional test
flagged 1.9 (one item) and 2.4 percent (two items) of the fitting items as
overfit, against 1.1 and 1.2 for infit. With overfitting items planted,
false underfit flags were 1.4 and 5.9 percent, against 1.5 and 6.5 for
infit. In clean decisions restscore is mostly ahead of infit when the planted
misfit is overfit and mostly behind when it is underfit, for example 79
against 86 percent with one underfitting item on target at n = 400. The
asymptotic test has the most clean decisions in the overfit cells, because
its underfit tail almost never rejects, and the fewest in the underfit
cells (7 against 53 percent with one item on target at n = 150).

**6. The conditional DGP loses iterations in small, off-target samples.** At
n = 150 with the misfitting items at −2 logits, 79 percent of datasets kept
fewer than 400 iterations, 20.5 percent fewer than 300, and one only 156.
The simulated datasets that are rejected are those missing a category, which
the conditional DGP produces more often when extreme categories are sparse.
For infit, fewer than 400 iterations made WY mildly liberal. For restscore
this has not been measured.

**Limitations.** Infit still uses the resample DGP, which the FWER paper
validated for infit.

## Follow-up: split one-sided step-downs

Added 2026-10-02. `restscore_eti_null.qmd` tested an equal-tailed
family-wise test built from two one-sided Westfall-Young step-downs, one per
direction, each at α/2. With the mean, SD or quantiles computed from all
B + 1 values (pooled), it held the family-wise rate and split it evenly
between the directions, on synthetic nulls and on the real restscore null.
This section measures its power on the same 1,600 datasets.

Variants, all from one cutoff simulation per dataset:

- `abs_t`: the shipped |t|. It must reproduce `cond_abs_wy` above.
- `abs_t_pooled`: |t| with pooled moments.
- `split_t`, `split_t_pooled`: t in the overfit run, −t in the underfit run.
- `split_two_piece`, `split_two_piece_pooled`: each run scaled by its own
  tail of the null, (97.5th percentile − median) for overfit and
  (median − 2.5th percentile) for underfit.

The pooled split variants are the candidates. The simulated-only versions
are included to show what pooling costs in power. The observed differences
and simulated matrices of every dataset are saved, so later variants can be
checked without rerunning the bootstrap.

**Cost.** About 1.5 hours on 10 cores, as for the run above. The run is off
by default. Set `RUN_SPLIT_POWER <- TRUE` to start it. Results are saved per
cell, and the result chunks run once all 16 cells are saved.

```{r split-power-setup}
RUN_SPLIT_POWER <- TRUE
SPLIT_POWER_DIR <- "restscore_power_split"
split_power_files <- file.path(SPLIT_POWER_DIR,
                               paste0(run_cells$cell, ".rds"))
SPLIT_POWER_READY <- all(file.exists(split_power_files))
split_variant_levels <- c("abs_t", "abs_t_pooled", "split_t",
                          "split_t_pooled", "split_two_piece",
                          "split_two_piece_pooled")
```

```{r split-power-functions}
# Copy of split_null_variants() in restscore_eti_null.qmd: adjusted p-values
# of every variant for one dataset, with the arithmetic of
# .bootstrap_pvalues().
split_variants <- function(obs, sim) {
  sim <- sim[, names(obs), drop = FALSE]
  B <- nrow(sim)
  guard <- function(x) {
    x[!is.finite(x) | x < 1e-8] <- NA_real_
    x
  }
  out <- data.frame(item = names(obs), stringsAsFactors = FALSE)
  for (pooled in c(FALSE, TRUE)) {
    sfx <- if (pooled) "_pooled" else ""
    ref <- if (pooled) rbind(obs, sim) else sim
    m <- colMeans(ref, na.rm = TRUE)
    s <- guard(apply(ref, 2L, stats::sd, na.rm = TRUE))
    t_obs <- (obs - m) / s
    t_sim <- sweep(sweep(sim, 2L, m, `-`), 2L, s, `/`)
    q <- apply(ref, 2L, stats::quantile, probs = c(.025, .5, .975),
               na.rm = TRUE, names = FALSE)
    up <- guard(q[3, ] - q[2, ])
    dn <- guard(q[2, ] - q[1, ])
    dev_sim <- sweep(sim, 2L, q[2, ], `-`)
    out[[paste0("t", sfx)]] <- t_obs
    out[[paste0("padj_abs", sfx)]] <- .wy_stepdown(abs(t_obs), abs(t_sim), B)
    out[[paste0("padj_t_hi", sfx)]] <- .wy_stepdown(t_obs, t_sim, B)
    out[[paste0("padj_t_lo", sfx)]] <- .wy_stepdown(-t_obs, -t_sim, B)
    out[[paste0("padj_tp_hi", sfx)]] <- .wy_stepdown(
      (obs - q[2, ]) / up, sweep(dev_sim, 2L, up, `/`), B
    )
    out[[paste0("padj_tp_lo", sfx)]] <- .wy_stepdown(
      (q[2, ] - obs) / dn, sweep(-dev_sim, 2L, dn, `/`), B
    )
  }
  out
}

# Flags ("overfit", "underfit" or "") per variant, from split_variants().
split_flags <- function(v, alpha = 0.05) {
  two <- function(padj, t) {
    ifelse(!is.na(padj) & padj < alpha,
           ifelse(t > 0, "overfit", "underfit"), "")
  }
  one <- function(lo, hi) {
    ifelse(!is.na(hi) & hi < alpha / 2, "overfit",
           ifelse(!is.na(lo) & lo < alpha / 2, "underfit", ""))
  }
  data.frame(
    abs_t = two(v$padj_abs, v$t),
    abs_t_pooled = two(v$padj_abs_pooled, v$t_pooled),
    split_t = one(v$padj_t_lo, v$padj_t_hi),
    split_t_pooled = one(v$padj_t_lo_pooled, v$padj_t_hi_pooled),
    split_two_piece = one(v$padj_tp_lo, v$padj_tp_hi),
    split_two_piece_pooled = one(v$padj_tp_lo_pooled, v$padj_tp_hi_pooled),
    stringsAsFactors = FALSE
  )
}

# Same data, seed and cutoff call as one_dataset().
one_dataset_split <- function(rep, cell_row) {
  setup <- item_setup(cell_row$n_misfit, cell_row$target)
  seed <- 1e6 * cell_row$row + rep
  d <- sim_power_data(
    n = cell_row$n, n_misfit = cell_row$n_misfit, target = cell_row$target,
    direction = cell_row$direction, rho = RHO, slope = SLOPE, seed = seed
  )

  tryCatch({
    cr <- suppressWarnings(suppressMessages(RMitemRestscoreCutoff(
      d, iterations = B, parallel = FALSE, seed = seed,
      cutoff_method = "quantile", dgp = "conditional"
    )))
    fw <- suppressWarnings(suppressMessages(RMitemRestscore(
      d, cutoff = cr, output = "dataframe"
    )))
    sim_mat <- tapply(cr$results$Difference,
                      list(cr$results$iteration, cr$results$Item),
                      function(x) x[1L])
    obs <- stats::setNames(fw$Difference, fw$Item)
    v <- split_variants(obs, sim_mat)
    fl <- split_flags(v)[match(names(d), v$item), ]
    list(
      table = data.frame(
        rep = rep, item = names(d),
        misfit = seq_len(7) %in% setup$misfit,
        pkg_wy = fw$Flagged[match(names(d), fw$Item)],
        fl, iterations = cr$actual_iterations,
        row.names = NULL, stringsAsFactors = FALSE
      ),
      sims = list(rep = rep, obs = obs,
                  sim = sim_mat[, names(obs), drop = FALSE])
    )
  }, error = function(e) {
    list(error = data.frame(rep = rep, error = conditionMessage(e)))
  })
}
```

```{r split-power-run}
#| eval: !expr RUN_SPLIT_POWER
dir.create(SPLIT_POWER_DIR, showWarnings = FALSE)
for (i in seq_len(nrow(run_cells))) {
  if (file.exists(split_power_files[i])) next
  t0 <- Sys.time()
  res <- mclapply(seq_len(R_DATASETS),
                  function(r) one_dataset_split(r, run_cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  # A worker that dies returns a try-error, kept as an error row.
  res <- lapply(res, function(x) {
    if (is.list(x)) x else list(error = data.frame(rep = NA, error = as.character(x)))
  })
  ok <- vapply(res, function(x) is.null(x$error), logical(1))
  # Simulated values first, so a saved result implies saved simulations.
  saveRDS(lapply(res[ok], `[[`, "sims"),
          file.path(SPLIT_POWER_DIR, paste0(run_cells$cell[i], "_sims.rds")))
  saveRDS(list(res = lapply(res, function(x) {
                 if (is.null(x$error)) x$table else x$error
               }),
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          split_power_files[i])
}
SPLIT_POWER_READY <- all(file.exists(split_power_files))
```

```{r split-power-load}
#| eval: !expr SPLIT_POWER_READY
sp_raw <- lapply(run_cells$cell, function(cl) {
  x <- readRDS(file.path(SPLIT_POWER_DIR, paste0(cl, ".rds")))
  is_ok <- vapply(x$res, function(r) "item" %in% names(r), logical(1))
  list(ok = cbind(cell = cl, do.call(rbind, x$res[is_ok])),
       err = if (any(!is_ok)) cbind(cell = cl, do.call(rbind, x$res[!is_ok])),
       minutes = round(x$elapsed_min, 1))
})
sp <- do.call(rbind, lapply(sp_raw, `[[`, "ok")) |>
  left_join(run_cells |> select(cell, n, target, n_misfit, direction),
            by = "cell")
sp_errors <- do.call(rbind, lapply(sp_raw, `[[`, "err"))
data.frame(
  cell = run_cells$cell,
  datasets = vapply(sp_raw, function(x) length(unique(x$ok$rep)), integer(1)),
  errors = vapply(sp_raw, function(x) {
    if (is.null(x$err)) 0L else nrow(x$err)
  }, integer(1)),
  minutes = vapply(sp_raw, `[[`, numeric(1), "minutes")
)
if (!is.null(sp_errors)) count(sp_errors, error)
```

### Checks

`abs_t` against the package flags of this run and against the cached
`cond_abs_wy`. Both mismatch counts should be 0.

```{r split-power-reproduce}
#| eval: !expr SPLIT_POWER_READY
sp |>
  inner_join(new |> select(cell, rep, item, cond_abs_wy),
             by = c("cell", "rep", "item")) |>
  summarise(items = n(),
            mismatch_package = sum(abs_t != pkg_wy),
            mismatch_cached = sum(abs_t != cond_abs_wy))
```

### Detection of the misfitting items

Proportion of misfitting items flagged in the correct direction, per cell
(100 datasets) and pooled over sample size and the number of misfitting
items.

```{r split-power-detection}
#| eval: !expr SPLIT_POWER_READY
sp_long <- sp |>
  pivot_longer(all_of(split_variant_levels), names_to = "variant",
               values_to = "flag") |>
  mutate(variant = factor(variant, levels = split_variant_levels),
         targeting = ifelse(target == 0, "at 0", "at -2"))
sp_long |>
  filter(misfit) |>
  group_by(direction, n_misfit, targeting, n, variant) |>
  summarise(correct = round(100 * mean(flag == direction)),
            .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = correct) |>
  arrange(direction, n_misfit, desc(targeting), n)
sp_long |>
  filter(misfit) |>
  group_by(direction, targeting, variant) |>
  summarise(correct = round(100 * mean(flag == direction), 1),
            .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = correct)
```

### Paired comparisons

Misfitting items detected by one variant and not the other, pooled over
sample size and the number of misfitting items. `difference_pts` is the
first variant minus the second.

```{r split-power-paired}
#| eval: !expr SPLIT_POWER_READY
sp_paired <- function(a, b) {
  sp |>
    filter(misfit) |>
    mutate(hit_a = .data[[a]] == direction, hit_b = .data[[b]] == direction) |>
    group_by(direction, targeting = ifelse(target == 0, "at 0", "at -2")) |>
    summarise(items = n(), a_only = sum(hit_a & !hit_b),
              b_only = sum(!hit_a & hit_b),
              difference_pts = round(100 * (mean(hit_a) - mean(hit_b)), 1),
              .groups = "drop") |>
    mutate(comparison = paste(a, "vs", b), .before = 1)
}
bind_rows(
  sp_paired("split_t_pooled", "abs_t"),
  sp_paired("split_two_piece_pooled", "abs_t"),
  sp_paired("split_two_piece_pooled", "split_t_pooled"),
  sp_paired("split_t_pooled", "split_t"),
  sp_paired("abs_t_pooled", "abs_t")
)
```

### False flags on the fitting items, by direction

```{r split-power-fp}
#| eval: !expr SPLIT_POWER_READY
sp_long |>
  filter(!misfit) |>
  group_by(direction, n_misfit, variant) |>
  summarise(false_overfit = round(100 * mean(flag == "overfit"), 1),
            false_underfit = round(100 * mean(flag == "underfit"), 1),
            .groups = "drop") |>
  pivot_wider(names_from = n_misfit,
              values_from = c(false_overfit, false_underfit),
              names_prefix = "k")
```

### Clean decisions

Every misfitting item flagged in the correct direction and no fitting item
flagged, pooled over sample size.

```{r split-power-clean}
#| eval: !expr SPLIT_POWER_READY
sp_long |>
  group_by(direction, n_misfit, targeting, n, variant, rep) |>
  summarise(clean = all(flag[misfit] == direction[1]) &
              all(flag[!misfit] == ""), .groups = "drop") |>
  group_by(direction, n_misfit, targeting, variant) |>
  summarise(clean = round(100 * mean(clean)), .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = clean) |>
  arrange(direction, n_misfit, desc(targeting))
```

### Extension to 300 datasets per cell

The cells above rest on 100 datasets, so a single cell's detection rate has
a Monte Carlo SE of up to 5 points. This extension adds datasets 101 to 300
in the same 16 cells, with the same seed scheme (`1e6 * row + rep`), so the
combined 300 datasets per cell are a continuation of the first 100. Infit
and the asymptotic test are not extended, so comparisons with them stay at
100 datasets.

**Cost.** About 3 hours on 10 cores. Off by default. Set
`RUN_SPLIT_POWER_EXT <- TRUE` to start it.

```{r split-ext-setup}
RUN_SPLIT_POWER_EXT <- TRUE
REPS_EXT <- 101:300
SPLIT_EXT_DIR <- "restscore_power_split_ext"
split_ext_files <- file.path(SPLIT_EXT_DIR, paste0(run_cells$cell, ".rds"))
SPLIT_EXT_READY <- SPLIT_POWER_READY && all(file.exists(split_ext_files))
```

```{r split-ext-run}
#| eval: !expr RUN_SPLIT_POWER_EXT
dir.create(SPLIT_EXT_DIR, showWarnings = FALSE)
for (i in seq_len(nrow(run_cells))) {
  if (file.exists(split_ext_files[i])) next
  t0 <- Sys.time()
  res <- mclapply(REPS_EXT,
                  function(r) one_dataset_split(r, run_cells[i, ]),
                  mc.cores = NCORES, mc.preschedule = FALSE)
  res <- lapply(res, function(x) {
    if (is.list(x)) x else list(error = data.frame(rep = NA, error = as.character(x)))
  })
  ok <- vapply(res, function(x) is.null(x$error), logical(1))
  saveRDS(lapply(res[ok], `[[`, "sims"),
          file.path(SPLIT_EXT_DIR, paste0(run_cells$cell[i], "_sims.rds")))
  saveRDS(list(res = lapply(res, function(x) {
                 if (is.null(x$error)) x$table else x$error
               }),
               elapsed_min = as.numeric(Sys.time() - t0, units = "mins")),
          split_ext_files[i])
}
SPLIT_EXT_READY <- SPLIT_POWER_READY && all(file.exists(split_ext_files))
```

```{r split-ext-load}
#| eval: !expr SPLIT_EXT_READY
ext_raw <- lapply(run_cells$cell, function(cl) {
  x <- readRDS(file.path(SPLIT_EXT_DIR, paste0(cl, ".rds")))
  is_ok <- vapply(x$res, function(r) "item" %in% names(r), logical(1))
  list(ok = cbind(cell = cl, do.call(rbind, x$res[is_ok])),
       err = if (any(!is_ok)) cbind(cell = cl, do.call(rbind, x$res[!is_ok])),
       minutes = round(x$elapsed_min, 1))
})
sp300 <- bind_rows(
  sp |> select(-n, -target, -n_misfit, -direction),
  do.call(rbind, lapply(ext_raw, `[[`, "ok"))
) |>
  left_join(run_cells |> select(cell, n, target, n_misfit, direction),
            by = "cell")
ext_errors <- do.call(rbind, lapply(ext_raw, `[[`, "err"))
data.frame(
  cell = run_cells$cell,
  datasets_ext = vapply(ext_raw, function(x) length(unique(x$ok$rep)),
                        integer(1)),
  errors = vapply(ext_raw, function(x) {
    if (is.null(x$err)) 0L else nrow(x$err)
  }, integer(1)),
  minutes = vapply(ext_raw, `[[`, numeric(1), "minutes"),
  datasets_total = vapply(run_cells$cell, function(cl) {
    length(unique(sp300$rep[sp300$cell == cl]))
  }, integer(1))
)
if (!is.null(ext_errors)) count(ext_errors, error)
# The package flags must still agree with the local |t|.
sum(sp300$abs_t != sp300$pkg_wy)
```

#### Detection, 300 datasets per cell

```{r split-ext-detection}
#| eval: !expr SPLIT_EXT_READY
sp300_long <- sp300 |>
  pivot_longer(all_of(split_variant_levels), names_to = "variant",
               values_to = "flag") |>
  mutate(variant = factor(variant, levels = split_variant_levels),
         targeting = ifelse(target == 0, "at 0", "at -2"))
sp300_long |>
  filter(misfit) |>
  group_by(direction, n_misfit, targeting, n, variant) |>
  summarise(correct = round(100 * mean(flag == direction)),
            .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = correct) |>
  arrange(direction, n_misfit, desc(targeting), n)
sp300_long |>
  filter(misfit) |>
  group_by(direction, targeting, variant) |>
  summarise(correct = round(100 * mean(flag == direction), 1),
            .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = correct)
```

#### Paired comparisons, 300 datasets per cell

```{r split-ext-paired}
#| eval: !expr SPLIT_EXT_READY
sp300_paired <- function(a, b) {
  sp300 |>
    filter(misfit) |>
    mutate(hit_a = .data[[a]] == direction, hit_b = .data[[b]] == direction) |>
    group_by(direction, targeting = ifelse(target == 0, "at 0", "at -2")) |>
    summarise(items = n(), a_only = sum(hit_a & !hit_b),
              b_only = sum(!hit_a & hit_b),
              difference_pts = round(100 * (mean(hit_a) - mean(hit_b)), 1),
              se_pts = round(100 * sqrt(a_only + b_only) / n(), 1),
              .groups = "drop") |>
    mutate(comparison = paste(a, "vs", b), .before = 1)
}
bind_rows(
  sp300_paired("split_t_pooled", "abs_t"),
  sp300_paired("split_two_piece_pooled", "abs_t"),
  sp300_paired("split_two_piece_pooled", "split_t_pooled"),
  sp300_paired("split_t_pooled", "split_t"),
  sp300_paired("abs_t_pooled", "abs_t")
)
```

#### False flags and clean decisions, 300 datasets per cell

```{r split-ext-fp}
#| eval: !expr SPLIT_EXT_READY
sp300_long |>
  filter(!misfit) |>
  group_by(direction, n_misfit, variant) |>
  summarise(false_overfit = round(100 * mean(flag == "overfit"), 1),
            false_underfit = round(100 * mean(flag == "underfit"), 1),
            .groups = "drop") |>
  pivot_wider(names_from = n_misfit,
              values_from = c(false_overfit, false_underfit),
              names_prefix = "k")
sp300_long |>
  group_by(direction, n_misfit, targeting, n, variant, rep) |>
  summarise(clean = all(flag[misfit] == direction[1]) &
              all(flag[!misfit] == ""), .groups = "drop") |>
  group_by(direction, n_misfit, targeting, variant) |>
  summarise(clean = round(100 * mean(clean)), .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = clean) |>
  arrange(direction, n_misfit, desc(targeting))
```

### Interpretation of the split follow-up

All 4,800 datasets ran without errors, 100 per cell in the first run and 200
more in the extension. The local |t| reproduced the package flags and the
cached `cond_abs_wy` on every item, so the comparisons below differ only in
the p-value. The figures are from the 300 datasets per cell, 1,800
misfitting items per direction and targeting.

**1. The split costs underfit detection.** `split_t_pooled` detected
underfitting items 2.9 points less often than the shipped |t| off target and
2.2 points less often on target (SE 0.4), and almost never caught an item
that |t| missed (1 against 54 and 0 against 40 discordant items). The first
100 datasets gave 2.8 and 2.2.

**2. It gains no overfit detection.** The differences were +0.2 off target
and +0.5 on target (SE 0.4, 25 against 21 and 30 against 21 discordant
items). The +1.2 on target in the first 100 datasets did not hold up. The α
moved to the overfit tail buys little extra overfit power in these cells.

**3. It makes fewer false underfit flags.** With overfitting items planted,
fitting items were flagged as underfit 1.0 (one item) and 4.7 percent (two
items) of the time, against 1.6 and 6.2 for |t|. False overfit flags with
underfit planted were about equal (1.9 and 2.2 against 2.1 and 2.3).

**4. Clean decisions improve in the overfit cells.** Every misfitting item
flagged in the correct direction and no fitting item flagged: 55 against 51
and 46 against 42 percent with one overfitting item on and off target, 18
against 14 and 12 against 9 with two. In the underfit cells the split was
level or 1 to 3 points behind (45 against 48 with one item off target).

**5. Two-piece scaling does not help.** `split_two_piece_pooled` was behind
`split_t_pooled` in three of four comparisons, by up to 1.6 points (overfit
on target, 22 against 51 discordant items). Together with the narrow spread
of null skewness across items in the real null, this leaves no case for it.

**6. Pooling costs little.** Pooled moments lowered detection by 0.4 to 0.8
points for both |t| and the split, always in the same direction (0 against
7 to 15 discordant items), in return for an exact level under
exchangeability.

**The trade.** Against the shipped |t|, `split_t_pooled` gives equal tails
under the null (2.40 against 2.75 percent of datasets with a flag in each
direction on the real null, where |t| gives 3.70 and 1.55) and fewer false
underfit flags when items overfit. It costs 2 to 3 points of underfit
detection and gains no overfit detection. Whether to ship it is a decision
about which error matters more, not a question the simulations settle.
