---
title: "Generalized S-X2 compared to conditional infit and item-restscore"
subtitle: "Detection of 1-3 misfitting items under multidimensionality"
author: "Magnus Johansson"
date: today
---

## Purpose

Johansson (2025) compared conditional item infit (with simulation-based cutoffs)
and item-restscore for detecting item misfit caused by multidimensionality, and
found that neither method behaves well at large sample sizes: false positive
rates rise sharply from n = 1000 onwards. Kang and Chen (2008) report that the
generalized S-X2 index maintains nominal Type I error rates from n = 500 up to
n = 5000, which raises the question of whether it could fill that gap.

Two things make this worth testing rather than assuming:

1. Kang and Chen only evaluated power in conditions where **every** item was
   misfitting, and where misfit was a clean nested-model contrast (RSM data
   calibrated as PCM, etc.). The realistic case of a few misfitting items among
   fitting ones was explicitly left for future work.
2. The follow-up study on the graded response model found S-X2 power to be
   "much lower when the source of misfit was multidimensionality" than when it
   came from a wrong item response function. Multidimensionality is precisely
   the misfit source used in Johansson (2025), and the dominant concern in
   applied Rasch analysis.

This simulation therefore replicates the Johansson (2025) design and adds
S-X2 as a fourth method, so all methods are evaluated on the *same* datasets.

## Design

Following Johansson (2025):

- 20 dichotomous items, locations drawn from `runif(20, -2, 2)`
- person locations from a bivariate normal, SD 1.5, correlation 0.15 between
  dimensions (lower correlation = stronger multidimensionality)
- misfit induced with `eRm::sim.xdim()` by loading selected items on the second
  dimension
- misfitting items chosen at approximately 0, -1 and -2 logits relative to the
  sample mean, to vary targeting
- master datasets of 10 000 respondents, from which samples are drawn
- sample sizes 150, 250, 500, 1000, 2000

Three misfit conditions are used: **0** misfitting items (a clean null, to
measure Type I error uncontaminated by other misfitting items), **1**
(the on-target item only), and **3**.

Four decision rules are compared:

| Label | Rule |
|---|---|
| `infit` | conditional infit MSQ outside simulation-based cutoffs (`RMitemInfitCutoff`, 100 iterations for n < 500, 200 otherwise) |
| `restscore` | item-restscore gamma, BH-corrected p < .05 |
| `sx2_bh` | generalized S-X2, BH-corrected p < .05 |
| `sx2_rmsea` | RMSEA.S_X2 > .05 (effect size, no significance test) |

Uncorrected S-X2 p-values are also recorded, since infit cutoffs are not
multiplicity-corrected and the comparison should be readable either way.

S-X2 is computed via `mirt::itemfit()`, which implements the Kang and Chen
generalization. Note that this estimates item parameters by MML under a normal
θ prior, whereas the other two methods use CML — see the discussion at the end.

```{r setup}
suppressMessages({
  library(mirt)
  library(easyRasch2)
  library(eRm)
  library(MASS)
  library(parallel)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

R_REPS   <- 200
NS       <- c(150, 250, 500, 1000, 2000)
NMISFITS <- c(0, 1, 3)
NCORES   <- 12   # performance cores only (M4 Max)
RES_FILE <- "sx2_itemfit_sim_results.rds"
```

### Data generation

```{r gen}
make_master <- function(n_misfit, N = 10000, seed = 20250727) {
  set.seed(seed)
  locs <- runif(20, -2, 2)
  # items closest to 0, -1, -2 logits (sample theta mean = 0)
  misfit_idx <- integer(0)
  for (tg in c(0, -1, -2)) {
    cand <- setdiff(order(abs(locs - tg)), misfit_idx)[1]
    misfit_idx <- c(misfit_idx, cand)
  }
  use <- misfit_idx[seq_len(n_misfit)]
  W <- cbind(rep(1, 20), rep(0, 20))
  if (n_misfit > 0) W[use, ] <- matrix(c(0, 1), nrow = length(use), ncol = 2,
                                       byrow = TRUE)
  Sigma <- matrix(c(1.5^2, .15 * 1.5^2, .15 * 1.5^2, 1.5^2), 2, 2)
  set.seed(seed + n_misfit)
  th <- MASS::mvrnorm(N, mu = c(0, 0), Sigma = Sigma)
  d <- as.data.frame(eRm::sim.xdim(persons = th, items = locs,
                                   Sigma = Sigma, weightmat = W))
  names(d) <- paste0("V", 1:20)
  list(data = d, locs = locs, misfit = use, misfit_all = misfit_idx)
}

masters <- lapply(NMISFITS, make_master)
names(masters) <- as.character(NMISFITS)
misfit_all <- masters[["3"]]$misfit
locs <- masters[["3"]]$locs

data.frame(
  item = paste0("V", misfit_all),
  intended_location = c(0, -1, -2),
  actual_location = round(locs[misfit_all], 2)
)
```

### One replication

```{r onerep}
one_rep <- function(rep, k, n) {
  set.seed(1e6 * k + 1e3 * n + rep)
  md <- masters[[as.character(k)]]$data
  d  <- md[sample(nrow(md), n), ]
  iters <- if (n < 500) 100 else 200
  out <- matrix(NA_integer_, nrow = 20, ncol = 5,
                dimnames = list(NULL, c("infit", "restscore", "sx2_raw",
                                        "sx2_bh", "sx2_rmsea")))
  try({
    co <- RMitemInfitCutoff(d, iterations = iters, parallel = FALSE,
                            seed = rep, verbose = FALSE)
    fi <- RMitemInfit(d, cutoff = co, output = "dataframe")
    out[, "infit"] <- as.integer(fi$Flagged[match(names(d), fi$Item)] != "")
  }, silent = TRUE)
  try({
    rs <- RMitemRestscore(d, output = "dataframe", p_adj = "BH")
    out[, "restscore"] <- as.integer(rs$Flagged[match(names(d), rs$Item)] != "")
  }, silent = TRUE)
  try({
    mm <- mirt(d, 1, itemtype = "Rasch", verbose = FALSE)
    f  <- mirt::itemfit(mm, fit_stats = "S_X2")
    i  <- match(names(d), f$item)
    out[, "sx2_raw"]   <- as.integer(f$p.S_X2[i] < .05)
    out[, "sx2_bh"]    <- as.integer(p.adjust(f$p.S_X2, "BH")[i] < .05)
    out[, "sx2_rmsea"] <- as.integer(f$RMSEA.S_X2[i] > .05)
  }, silent = TRUE)
  out
}
```

### Run

Results are cached; delete `r RES_FILE` to force a re-run.

```{r run}
grid <- expand.grid(rep = 1:R_REPS, n = NS, k = NMISFITS)

if (file.exists(RES_FILE)) {
  sim <- readRDS(RES_FILE)
} else {
  t0 <- Sys.time()
  res <- mclapply(seq_len(nrow(grid)),
                  function(i) one_rep(grid$rep[i], grid$k[i], grid$n[i]),
                  mc.cores = NCORES)
  sim <- list(res = res, grid = grid, misfit = misfit_all, locs = locs,
              elapsed_min = as.numeric(Sys.time() - t0, "mins"))
  saveRDS(sim, RES_FILE)
}

long <- do.call(rbind, lapply(seq_len(nrow(sim$grid)), function(i) {
  m <- sim$res[[i]]
  data.frame(k = sim$grid$k[i], n = sim$grid$n[i], rep = sim$grid$rep[i],
             item = 1:20, m)
})) |>
  pivot_longer(c(infit, restscore, sx2_raw, sx2_bh, sx2_rmsea),
               names_to = "method", values_to = "flag") |>
  mutate(is_misfit = mapply(function(k, it) it %in% sim$misfit[seq_len(k)],
                            k, item))

cat("failed method-replications:", sum(is.na(long$flag)), "of", nrow(long), "\n")
```

## Results

### Type I error under a true Rasch model (k = 0)

With no misfitting items present, the flag rate on any item is a false positive.

```{r null}
long |>
  filter(k == 0) |>
  group_by(method, n) |>
  summarise(fp = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = fp, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Detection rate, 1 misfitting item

```{r det1}
long |>
  filter(k == 1, is_misfit) |>
  group_by(method, n) |>
  summarise(det = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = det, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

False positives on the 19 fitting items in the same datasets:

```{r fp1}
long |>
  filter(k == 1, !is_misfit) |>
  group_by(method, n) |>
  summarise(fp = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = fp, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Detection rate, 3 misfitting items

Broken out by item, since targeting matters.

```{r det3}
long |>
  filter(k == 3, is_misfit) |>
  mutate(target = c("0 logits", "-1 logits", "-2 logits")[match(item, sim$misfit)]) |>
  group_by(method, target, n) |>
  summarise(det = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = det, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(factor(target, levels = c("0 logits", "-1 logits", "-2 logits")), method)
```

False positives on the 17 fitting items:

```{r fp3}
long |>
  filter(k == 3, !is_misfit) |>
  group_by(method, n) |>
  summarise(fp = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = fp, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1)))
```

### Figure: detection vs false positives, 3 misfitting items

```{r fig, fig.width = 9, fig.height = 5}
plot_dat <- long |>
  filter(k == 3, method %in% c("infit", "restscore", "sx2_bh", "sx2_rmsea")) |>
  mutate(type = ifelse(is_misfit, "Detection (misfitting items)",
                       "False positives (fitting items)")) |>
  group_by(method, n, type) |>
  summarise(pct = 100 * mean(flag, na.rm = TRUE), .groups = "drop")

ggplot(plot_dat, aes(factor(n), pct, fill = method)) +
  geom_col(position = "dodge") +
  facet_wrap(~type) +
  labs(x = "Sample size", y = "Percent", fill = "Method") +
  theme_minimal(base_size = 12)
```

## Supplementary study: weaker multidimensionality

The main study uses an inter-dimensional correlation of 0.15, which is strong
misfit — detection saturates at 100% for all methods from n = 500 onwards, so
the main study discriminates between methods almost entirely on false
positives.

The `sx2_rmsea` rule performs strikingly well at n >= 500 there, but it applies
a **fixed** threshold of .05 to an effect size. Its performance therefore
depends on how far the true misfit sits above that threshold, and a strong
misfit condition flatters it. This supplement re-runs the k = 3 condition with
inter-dimensional correlations of 0.50 and 0.75 (progressively weaker
multidimensionality) and records the raw `RMSEA.S_X2` values, to check whether
the rule survives when the effect is subtle.

```{r gen2}
make_master_r <- function(rho, n_misfit = 3, N = 10000, seed = 20250727) {
  set.seed(seed)
  locs <- runif(20, -2, 2)
  misfit_idx <- integer(0)
  for (tg in c(0, -1, -2)) {
    cand <- setdiff(order(abs(locs - tg)), misfit_idx)[1]
    misfit_idx <- c(misfit_idx, cand)
  }
  use <- misfit_idx[seq_len(n_misfit)]
  W <- cbind(rep(1, 20), rep(0, 20))
  W[use, ] <- matrix(c(0, 1), nrow = length(use), ncol = 2, byrow = TRUE)
  Sigma <- matrix(c(1.5^2, rho * 1.5^2, rho * 1.5^2, 1.5^2), 2, 2)
  set.seed(seed + round(rho * 100))
  th <- MASS::mvrnorm(N, mu = c(0, 0), Sigma = Sigma)
  d <- as.data.frame(eRm::sim.xdim(persons = th, items = locs,
                                   Sigma = Sigma, weightmat = W))
  names(d) <- paste0("V", 1:20)
  list(data = d, locs = locs, misfit = use)
}

RHOS  <- c(0.50, 0.75)
NS2   <- c(250, 500, 1000, 2000)
R2    <- 150
RES2  <- "sx2_itemfit_sim_weak_results.rds"

masters2 <- lapply(RHOS, make_master_r)
names(masters2) <- as.character(RHOS)

one_rep2 <- function(rep, rho, n) {
  set.seed(1e6 * round(rho * 100) + 1e3 * n + rep)
  md <- masters2[[as.character(rho)]]$data
  d  <- md[sample(nrow(md), n), ]
  iters <- if (n < 500) 100 else 200
  out <- data.frame(item = 1:20, infit = NA_integer_, restscore = NA_integer_,
                    sx2_bh = NA_integer_, sx2_rmsea = NA_integer_,
                    rmsea = NA_real_)
  try({
    co <- RMitemInfitCutoff(d, iterations = iters, parallel = FALSE,
                            seed = rep, verbose = FALSE)
    fi <- RMitemInfit(d, cutoff = co, output = "dataframe")
    out$infit <- as.integer(fi$Flagged[match(names(d), fi$Item)] != "")
  }, silent = TRUE)
  try({
    rs <- RMitemRestscore(d, output = "dataframe", p_adj = "BH")
    out$restscore <- as.integer(rs$Flagged[match(names(d), rs$Item)] != "")
  }, silent = TRUE)
  try({
    mm <- mirt(d, 1, itemtype = "Rasch", verbose = FALSE)
    f  <- mirt::itemfit(mm, fit_stats = "S_X2")
    i  <- match(names(d), f$item)
    out$sx2_bh    <- as.integer(p.adjust(f$p.S_X2, "BH")[i] < .05)
    out$rmsea     <- f$RMSEA.S_X2[i]
    out$sx2_rmsea <- as.integer(f$RMSEA.S_X2[i] > .05)
  }, silent = TRUE)
  out
}

grid2 <- expand.grid(rep = 1:R2, n = NS2, rho = RHOS)

if (file.exists(RES2)) {
  sim2 <- readRDS(RES2)
} else {
  res2 <- mclapply(seq_len(nrow(grid2)),
                   function(i) one_rep2(grid2$rep[i], grid2$rho[i], grid2$n[i]),
                   mc.cores = NCORES)
  sim2 <- list(res = res2, grid = grid2, misfit = masters2[["0.5"]]$misfit)
  saveRDS(sim2, RES2)
}

long2 <- do.call(rbind, lapply(seq_len(nrow(sim2$grid)), function(i)
  cbind(sim2$grid[i, ], sim2$res[[i]]))) |>
  mutate(is_misfit = item %in% sim2$misfit)
```

Detection of the three misfitting items:

```{r det2}
long2 |>
  filter(is_misfit) |>
  pivot_longer(c(infit, restscore, sx2_bh, sx2_rmsea),
               names_to = "method", values_to = "flag") |>
  group_by(rho, method, n) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rho, method)
```

False positives on the 17 fitting items:

```{r fp2}
long2 |>
  filter(!is_misfit) |>
  pivot_longer(c(infit, restscore, sx2_bh, sx2_rmsea),
               names_to = "method", values_to = "flag") |>
  group_by(rho, method, n) |>
  summarise(v = 100 * mean(flag, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n=") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(rho, method)
```

Median `RMSEA.S_X2` for misfitting vs fitting items, against the .05 threshold:

```{r rmsea2}
long2 |>
  group_by(rho, n, type = ifelse(is_misfit, "misfitting", "fitting")) |>
  summarise(median_rmsea = round(median(rmsea, na.rm = TRUE), 3),
            .groups = "drop") |>
  pivot_wider(names_from = n, values_from = median_rmsea, names_prefix = "n=")
```

## Conclusions

**1. S-X2 has no power advantage, and a power disadvantage against subtle
misfit.** With strong multidimensionality (rho = 0.15) all methods are
equivalent and saturate. With rho = 0.75, conditional infit clearly dominates:
23.8 / 36.4 / 69.6 / 94.7 percent detection at n = 250 / 500 / 1000 / 2000,
against 9.3 / 23.3 / 39.6 / 77.8 for S-X2. The concern raised by the Kang and
Chen GRM follow-up — weak power against multidimensionality — does not appear
under strong misfit but is clearly visible once the misfit is subtle.

**2. S-X2 does control the false positive cascade.** This is its real
contribution. With three strongly misfitting items at n = 2000, false positives
on fitting items are 54.2 percent for infit and 80.6 percent for item-restscore,
against 16.9 percent for S-X2. On the practically relevant metric — recovering
all three misfitting items with no false positives — S-X2 achieves 72 percent
at n = 1000-2000 with rho = 0.5, where infit manages 26 and 3.3 percent.

**3. Infit and item-restscore are not miscalibrated.** Under a true Rasch model
with no misfitting items, false positive rates are 0.1-1.6 percent at every
sample size, including n = 2000. The large-sample false positives reported in
Johansson (2025) are therefore entirely the overfit cascade caused by the
misfitting items themselves, not a defect in the simulation-based cutoffs.

**4. A fixed RMSEA threshold of .05 is not usable.** It fails in both
directions. At n = 150 under a true model it flags 26.2 percent of items,
because the null RMSEA has not yet shrunk. At rho = 0.75 the median RMSEA of
genuinely misfitting items is .030-.043 — below the threshold — so detection
falls to 0.4 percent at n = 2000, and gets *worse* as n grows, since the
estimate concentrates around a true value under the cutoff.

**5. The implied design.** RMSEA.S_X2 has the property that is wanted — it does
not inflate with sample size (median .058-.075 for misfitting items across
n = 250-2000 at rho = 0.5) — but the threshold is wrong and is itself
sample-size dependent. Median RMSEA for fitting items is .000-.007 at
n >= 250, far below .05. This is exactly the problem `RMitemInfitCutoff` solves
for infit, which suggests a simulation-based cutoff for RMSEA.S_X2 would set a
much lower and n-appropriate threshold, and would likely recover the rho = 0.75
detection that the fixed rule misses. **This is a hypothesis, not a result — it
has not been tested here.**

Overall: S-X2 is a large-sample complement to conditional infit, not a
replacement for it. Its niche is several strongly misfitting items at
n >= 1000, which is precisely where the existing methods fail.

## Notes on interpretation

`sx2_rmsea` uses a fixed RMSEA cutoff of .05, borrowed from SEM convention.
Item-level RMSEA cutoffs are not well established, and the null RMSEA shrinks
with sample size, so this rule is expected to behave asymmetrically across n.
It is included to separate two questions that the S-X2 significance test
confounds: whether the departure is detectable, and whether it is large.

A caveat on estimation: `mirt::itemfit()` estimates item parameters by MML
under a normal θ prior. For the Rasch family the S-X2 expected proportions are
in principle distribution-free — the θ density cancels between numerator and
denominator, leaving the CML conditional probabilities — but that only holds at
the true parameters. A CML-based implementation would remove the normality
assumption, which matters here since the generating θ distribution is normal
and therefore favourable to the mirt route.

## References

Johansson, M. (2025). Detecting Item Misfit in Rasch Models.
*Educational Methods & Psychometrics, 3*(18). <https://dx.doi.org/10.61186/emp.2025.5>

Kang, T., & Chen, T. T. (2008). Performance of the Generalized S-X2 Item Fit
Index for Polytomous IRT Models. *Journal of Educational Measurement, 45*(4),
391-406.

Kang, T., & Chen, T. T. (2011). Performance of the generalized S-X2 item fit
index for the graded response model. *Asia Pacific Education Review, 12*,
89-96.
