---
title: "Why S-X2 rejects too often under a true Rasch model"
subtitle: "Null diagnostic of mirt::itemfit(fit_stats = \"S_X2\")"
author: "Magnus Johansson"
date: today
---

```{r setup}
#| include: false
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(knitr)
})
sim <- readRDS("sx2_null_diagnostic/results.rds")
```

## Purpose

In an earlier comparison of S-X2 with infit (July 2026, since replaced by `sx2_comparison_v2.qmd`), uncorrected S-X2 from `mirt::itemfit()` rejected 6.9
to 8.8 percent of items under a true Rasch model, against a nominal 5, and BH
over 20 items flagged at least one item in 10.5 to 19 percent of datasets,
rising with n. This note finds the cause. The design is in
`sx2-null-diagnostic-design.md` and the simulation in `sx2_null_diagnostic.R`.

mirt computes the degrees of freedom for each item as the number of score
groups after collapsing, minus the item's own parameters, minus every estimated
latent parameter. For `itemtype = "Rasch"` the latent variance is estimated, so
each item loses one df that Orlando and Thissen's definition does not remove.
The deduction is deliberate. mirt's NEWS lists it under version 1.20.1 as a df
adjustment "for models where the latent trait hyper-parameters have been
estimated". If S-X2 follows a chi-square on df + 1 and is referred to df, the
rejection rate at .05 is 6.9 to 7.5 percent at the df values seen here.

## Design

Null only, with fresh data for every replication. Three item sets: 20
dichotomous items evenly spaced on [-2, 2], the 20 dichotomous locations of the
original study, and 9 items with four categories on [-1.5, 1.5]. Persons from
$N(0, 1.5^2)$, n = 150, 500 and 2000, 2000 replications per cell. The Monte
Carlo SE is about 0.5 points for a 5 percent rate, per item and family-wise.

| Variant | Fit | Reference |
|---|---|---|
| V0 | Rasch, variance estimated | mirt default |
| V1 | same as V0 | p recomputed on df + 1 |
| V2 | same as V0, `mincell = 5` | mirt default |
| V2c | same as V2 | p recomputed on df + 1 |
| V3 | Rasch, variance fixed at the true 2.25 | mirt default, which then has no latent deduction |

V1 tests the df. V3 removes the deduction and the estimation error in the
variance together, so V1 against V3 separates the two. V2 tests stricter
collapsing of sparse cells.

```{r tidy}
ok <- !vapply(sim$res, is.null, logical(1))
long <- do.call(rbind, lapply(which(ok), function(i) {
  g <- sim$grid[i, ]; m <- sim$res[[i]]
  data.frame(set = g$set, n = g$n, rep = g$rep, item = seq_len(nrow(m)), m)
})) |>
  mutate(p_v1  = pchisq(S_v0, df_v0 + 1, lower.tail = FALSE),
         p_v2c = pchisq(S_v2, df_v2 + 1, lower.tail = FALSE))

pv <- long |>
  select(set, n, rep, item, V0 = p_v0, V1 = p_v1, V2 = p_v2, V2c = p_v2c,
         V3 = p_v3) |>
  pivot_longer(V0:V3, names_to = "variant", values_to = "p")

wide <- function(d, value) d |>
  pivot_wider(names_from = n, values_from = all_of(value), names_prefix = "n = ")
```

No replication was discarded. `elapsed` is `r round(sim$elapsed_min, 1)`
minutes on 10 cores, mirt `r sim$mirt_version`.

## Results

### The statistic is chi-square on df + 1

If the reference is right, the mean of S-X2 equals its df and the variance is
twice the df.

```{r moments}
long |>
  group_by(set, n) |>
  summarise(`mean S - df, V0` = mean(S_v0 - df_v0, na.rm = TRUE),
            `mean S - df, V3` = mean(S_v3 - df_v3, na.rm = TRUE),
            `var S / 2(df + 1), V0` = var(S_v0 - df_v0, na.rm = TRUE) /
              (2 * mean(df_v0 + 1, na.rm = TRUE)),
            `mean df, V0` = mean(df_v0, na.rm = TRUE), .groups = "drop") |>
  kable(digits = 2)
```

Under mirt's df, S-X2 exceeds its df by 0.96 to 1.07 on average for
dichotomous items, and its variance matches $2(\text{df} + 1)$ to within 5
percent from n = 500. With the variance fixed, which takes away the deduction,
the mean excess is within 0.07 of zero. The polytomous set shows the same
pattern from n = 500. At n = 150 its excess is 1.66 under V0 and 0.64 under V3,
so something else adds to the statistic there.

### Per-item rejection

```{r peritem}
pv |>
  group_by(set, variant, n) |>
  summarise(rate = 100 * mean(p < .05, na.rm = TRUE), .groups = "drop") |>
  wide("rate") |>
  kable(digits = 2)
```

Adding the missing df brings every cell to nominal: V1 rejects 4.8 to 5.3
percent of items against 6.0 to 7.5 under V0. V1 and V3 agree to within 0.2
points in every cell, so estimating the latent variance costs nothing. The df
is the whole of the per-item excess.

Stricter collapsing on its own (V2) is worse than the default, because
`mincell = 5` merges so many score groups at n = 150 that the df falls to about
2, and a one-df shortfall weighs more at small df. With the df corrected (V2c)
it is close to nominal for dichotomous items, but at n = 150 it leaves too few
score groups to say much. For dichotomous items 239 to 278 of the 40,000
item-replications at n = 150 had no df left.

### The far tail and the family-wise rate

BH over 20 items rejects on p-values near $.05 / 20$, so the family-wise rate
depends on the far tail, not on the 5 percent point.

```{r tail}
pv |>
  group_by(set, variant, n) |>
  summarise(`per 1000` = 1000 * mean(p < .005, na.rm = TRUE), .groups = "drop") |>
  wide("per 1000") |>
  kable(digits = 1, caption = "Items with p < .005, per 1000 (nominal 5)")
```

```{r familywise}
pv |>
  group_by(set, variant, n, rep) |>
  summarise(bh = any(p.adjust(p, "BH") < .05, na.rm = TRUE), .groups = "drop") |>
  group_by(set, variant, n) |>
  summarise(fw = 100 * mean(bh), .groups = "drop") |>
  wide("fw") |>
  kable(digits = 2, caption = "Datasets with at least one BH-flagged item (%)")
```

Correcting the df leaves part of the problem. Under V1, items reach p < .005 at
6.4 to 7.7 per 1000 for dichotomous items against a nominal 5, and BH flags at
least one item in 6.8 to 8.6 percent of datasets. That is down from 10.7 to
12.5 under V0, but still above 5. The polytomous set is close to nominal under
V1, at 4.3 to 6.5 percent. Under V0, V1 and V3, Bonferroni gives the same rates
to within 0.3 points, as expected under a complete null. V2c removes the tail excess
for dichotomous items (3.4 to 5.3 per 1000 and 3.6 to 5.9 percent
family-wise), which points to sparse score groups as the source of what
remains.

### No rise with sample size

Under V0 the family-wise rate is 10.7 to 12.5 percent at every n for both
dichotomous sets, including the original locations. The rise to 19 percent at
n = 2000 in the original study does not reproduce with fresh data. There, every
replication was drawn from one master dataset of 10,000 respondents. Each such
sample is, taken on its own, a random sample from the model, so the design does
not bias the rates. It makes the replications dependent instead. At n = 2000
the samples share about a fifth of their respondents, so any chance departure
from the model in the master dataset appears in every replication, and the rate
in a cell varies between master datasets by more than its Monte Carlo SE
allows. The rise is most likely such a draw, and the same holds for the other
methods in that study and for its per-item spread of 4.4 to 11.7 percent. Here,
under V1 at n = 2000, per-item rates run from 4.0 to 6.0 percent, which is about
what 2000 independent replications give by chance.

## Interpretation

**1. The per-item excess is mirt's df.** Removing the latent variance from
every item's df makes S-X2 reject 6 to 7.5 percent of items at a nominal 5 under
a true Rasch model. Adding the df back, or fixing the variance, gives 4.8 to
5.3. Whether the adjustment is right for other models is not tested here, but
for the Rasch and partial credit models it makes the test liberal.

**2. A smaller tail excess remains, from sparse cells.** With the df corrected,
20 dichotomous items still give 6.8 to 8.6 percent family-wise under BH.
Stricter collapsing removes it, at the cost of most of the score groups at small
n. Nine polytomous items are close to nominal.

**3. The rise with n in the original study is most likely noise from its
design.** Resampling one master dataset makes its replications dependent, so
differences between sample sizes there are less certain than they look. With
fresh data the rate does not rise with n.

**For the revised comparison.** S-X2 should enter with the corrected df. Since
the corrected test is still about 2 to 3.5 points over 5 percent family-wise
for 20 dichotomous items, it should also be reported against its own null in
the same run, or given a simulation-based cutoff, so that it is not compared
with family-wise controlled methods at a higher error rate. Fresh data per
replication is needed for every method.

Whether to raise the df adjustment with the mirt maintainer is a separate
decision.

## Reference

Orlando, M., & Thissen, D. (2000). Likelihood-based item-fit indices for
dichotomous item response theory models. *Applied Psychological Measurement,
24*(1), 50-64.
