---
title: "S-X2 against conditional infit, revised"
subtitle: "Detection and false flags at a common family-wise error rate"
author: "Magnus Johansson"
date: today
---

```{r setup}
#| include: false
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(knitr)
  library(ggplot2)
})
fs <- list.files("sx2_comparison_v2", "\\.rds$", full.names = TRUE)
L <- lapply(fs, readRDS)
```

## Purpose

An earlier study (July 2026, since removed) compared the generalized S-X2 index with conditional
infit and item-restscore and concluded that S-X2 is a large-sample complement
to infit. A review of that study found seven problems. The two that matter most
are that infit was run without multiplicity correction while S-X2 used BH, and
that mirt's S-X2 p-values are liberal under a true Rasch model.
`sx2_null_diagnostic.qmd` traced the second to a df one too small and found a
smaller far-tail excess that remains after correcting it.

This study asks the original question again with every method held to a
family-wise error rate: does S-X2 detect misfit that infit misses, or flag fewer
fitting items when several items misfit? The design and the decision rule were
written before the run, in `sx2-comparison-v2-design.md`. The simulation is
`sx2_comparison_v2.R`. Item-restscore is left out, since its comparison with
infit is covered by `restscore_infit_power.qmd` and
`restscore_power_conditional.qmd`.

## Design

- 20 dichotomous items, locations evenly spaced on [-2, 2].
- Persons from a bivariate normal, both dimensions SD 1.5, correlation $\rho$.
  Fitting items respond to the first dimension, misfitting items to the second.
  Fresh data in every replication.
- k = 1 misfitting item (nearest 0 logits) or k = 3 (nearest 0, -1 and -2).
  $\rho$ = 0.15 is strong multidimensionality, 0.50 and 0.75 are subtle.
- n = 150, 500, 1000 and 2000.
- 500 replications in each null cell (k = 0), 200 in each misfit cell. The
  Monte Carlo SE of a 5 percent family-wise rate is 1.0 point.

| Label | Method |
|---|---|
| `infit_wy` | conditional infit, `RMitemInfitCutoff()` B = 400, Westfall-Young, the package default |
| `infit_bh` | same bootstrap, BH on the marginal p |
| `sx2_mirt` | S-X2, chi-square on mirt's df, BH |
| `sx2_df1` | S-X2, chi-square on df + 1, BH |
| `sx2_boot` | S-X2, parametric bootstrap B = 400, Westfall-Young |
| `rmsea_boot` | RMSEA.S_X2, same bootstrap draws, Westfall-Young |
| `rmsea_05` | RMSEA.S_X2 above a fixed .05 |

```{r tidy}
long <- do.call(rbind, lapply(L, function(x) do.call(rbind, lapply(seq_along(x$res), function(r) {
  y <- x$res[[r]]
  if (is.null(y$res)) return(NULL)
  cbind(k = x$cell$k, rho = x$cell$rho, n = x$cell$n, rep = r, y$res)
}))))

methods <- c("infit_wy", "infit_bh", "sx2_mirt", "sx2_df1", "sx2_boot",
             "rmsea_boot", "rmsea_05")

adj <- long |>
  group_by(k, rho, n, rep) |>
  mutate(infit_wy = padj_infit, infit_bh = p.adjust(p_infit, "BH"),
         sx2_mirt = p.adjust(p_mirt, "BH"), sx2_df1 = p.adjust(p_df1, "BH"),
         sx2_boot = padj_sx2boot, rmsea_boot = padj_rmseaboot) |>
  ungroup()

flags <- adj |>
  mutate(across(c(infit_wy, infit_bh, sx2_mirt, sx2_df1, sx2_boot, rmsea_boot),
                \(p) p < .05),
         rmsea_05 = rmsea > .05) |>
  pivot_longer(all_of(methods), names_to = "method", values_to = "flag") |>
  mutate(flag = coalesce(flag, FALSE),
         method = factor(method, levels = methods))

# kable with rho printed to two decimals, so digits = 1 does not round it
kb <- function(d, ...) {
  d <- mutate(d, across(any_of("rho"), \(x) sprintf("%.2f", x)))
  kable(d, ...)
}
```

Every replication returned results for every method. No bootstrap draw was
dropped. easyRasch2 `r unique(sapply(L, \(x) x$easyRasch2))`, mirt
`r unique(sapply(L, \(x) x$mirt))`, about 7.6 hours on 10 cores.

## Results

Each figure shows one method per colour and point shape, with the same colour
for a method in every figure. The values behind every figure are in the
appendix.

```{r plot-setup}
pal <- c(infit_wy = "#2a78d6", sx2_df1 = "#eb6834", sx2_boot = "#1baf7a",
         sx2_mirt = "#eda100", rmsea_05 = "#e87ba4")
shp <- c(infit_wy = 16, sx2_df1 = 17, sx2_boot = 15, sx2_mirt = 18,
         rmsea_05 = 4)
lab <- c(infit_wy = "Infit, Westfall-Young",
         sx2_df1 = "S-X2, df + 1, BH",
         sx2_boot = "S-X2, bootstrap, Westfall-Young",
         sx2_mirt = "S-X2, mirt df, BH",
         rmsea_05 = "RMSEA > .05")

cell_lab <- function(d) d |>
  mutate(cell = factor(sprintf("k = %d, rho = %.2f", k, rho),
                       levels = c("k = 1, rho = 0.50", "k = 1, rho = 0.75",
                                  "k = 3, rho = 0.15", "k = 3, rho = 0.50",
                                  "k = 3, rho = 0.75")))

rate_plot <- function(d, ylab) {
  ggplot(d, aes(factor(n), v, colour = method, shape = method,
                group = method)) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 2.2) +
    scale_colour_manual(values = pal, labels = lab, name = NULL) +
    scale_shape_manual(values = shp, labels = lab, name = NULL) +
    scale_y_continuous(limits = c(0, 100)) +
    labs(x = "Sample size", y = ylab) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank()) +
    guides(colour = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
}

fig_methods <- c("infit_wy", "sx2_df1", "sx2_boot", "rmsea_05")
```

### Under the null

```{r}
#| label: fig-null
#| fig-width: 7
#| fig-height: 4
#| fig-cap: "Share of datasets with at least one flagged item under a true
#|   Rasch model (k = 0, 500 datasets per n). The dashed line is the nominal 5
#|   percent and the band covers two Monte Carlo SEs around it. RMSEA above .05
#|   is left out, since it flags 99.2 percent of datasets at n = 150."
#| fig-alt: "Line chart of family-wise false positive rate against sample size
#|   for four methods. Infit with Westfall-Young stays at about 4 percent at
#|   every n. The S-X2 bootstrap is about 6 percent at n = 150 and 500 and
#|   about 4 to 5 percent at n = 1000 and 2000, inside or at the edge of the
#|   band around 5 percent. S-X2 with df + 1 is 6 to 9 percent and S-X2 with
#|   mirt's df is 9 to 12 percent, both above the band at every n."
null_fw <- flags |>
  filter(k == 0) |>
  group_by(method, n, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(method, n) |>
  summarise(v = 100 * mean(any), .groups = "drop")

mcse <- 100 * 2 * sqrt(.05 * .95 / 500)
null_fw |>
  filter(method %in% c("infit_wy", "sx2_mirt", "sx2_df1", "sx2_boot")) |>
  ggplot(aes(factor(n), v, colour = method, shape = method, group = method)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 5 - mcse, ymax = 5 + mcse,
           fill = "grey85", alpha = 0.6) +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2.2) +
  scale_colour_manual(values = pal, labels = lab, name = NULL) +
  scale_shape_manual(values = shp, labels = lab, name = NULL) +
  scale_y_continuous(limits = c(0, 14), breaks = seq(0, 14, 2)) +
  labs(x = "Sample size", y = "Datasets with any flag (%)") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank()) +
  guides(colour = guide_legend(nrow = 2), shape = guide_legend(nrow = 2))
```

Infit with Westfall-Young holds 4.2 to 4.4 percent at every n. The S-X2
bootstrap gives 6.2 and 6.6 percent at n = 150 and 500 and 4.8 and 3.8 percent
at n = 1000 and 2000. The first two are 1.2 and 1.6 Monte Carlo SEs above 5, so
it is at most slightly liberal at small n. With the df corrected, asymptotic
S-X2 gives 5.8 to 9.0 percent, in line with the diagnostic. mirt's default gives
8.8 to 12.4 percent, and RMSEA above .05 flags almost every dataset at n = 150
and none at n = 2000 (@tbl-null).

RMSEA with a bootstrap cutoff behaves like the S-X2 bootstrap in every cell of
this study, as expected, since both are computed from the same draws and RMSEA
is close to a monotone function of S-X2. It is left out of the figures. Infit
with BH on the marginal bootstrap p is close to infit with Westfall-Young and is
left out too. Both are in the appendix tables.

### Detection of the misfitting items

```{r}
#| label: fig-detection
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Detection of the misfitting items, averaged over the misfitting
#|   items in each dataset. k is the number of misfitting items and rho the
#|   correlation between the two dimensions, lower meaning stronger misfit."
#| fig-alt: "Five panels, one per combination of number of misfitting items
#|   and dimension correlation, each plotting detection rate from 0 to 100
#|   percent against sample size for four methods. Detection rises with n in
#|   every panel. Infit with Westfall-Young is highest or tied in every panel.
#|   The S-X2 bootstrap and S-X2 with df + 1 lag behind infit most in the two
#|   panels with rho = 0.75, catching up only where all methods reach 100
#|   percent. RMSEA above .05 is high at n = 150 and falls with n in the rho =
#|   0.75 panels."
flags |>
  filter(k > 0, misfit, method %in% fig_methods) |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Detection (%)") +
  facet_wrap(~ cell, nrow = 2)
```

Infit with Westfall-Young detects at least as much as the S-X2 bootstrap in
every cell, and clearly more wherever detection is not saturated. The gap is
largest for subtle misfit. At $\rho$ = 0.75 it is 80 against 47.5 percent with
one misfitting item at n = 500, and 58 against 32 percent with three at
n = 1000. The S-X2 bootstrap only catches up where both methods reach 100
percent (@tbl-detection).

The fixed RMSEA threshold detects more at n = 150 because it flags almost
everything there, and falls to 1.3 percent at $\rho$ = 0.75 and n = 2000, as in
the original study. A bootstrap cutoff for RMSEA recovers that, 68.7 percent in
the same cell, which confirms the hypothesis in conclusion 5 of the original
study. It does no better than the S-X2 bootstrap, though.

### False flags on fitting items

```{r}
#| label: fig-falseflags
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Share of datasets in which at least one fitting item is flagged.
#|   Panels as in @fig-detection."
#| fig-alt: "Five panels plotting the share of datasets with a falsely flagged
#|   fitting item against sample size. With three misfitting items, infit with
#|   Westfall-Young rises steeply with n, reaching 87 to 100 percent at
#|   n = 2000 for rho 0.15 and 0.50, while the S-X2 bootstrap stays below 10
#|   percent except at rho = 0.15 and n = 2000, where it reaches about 48
#|   percent. With one misfitting item all methods except RMSEA stay low.
#|   RMSEA above .05 is near 100 percent at n = 150 in every panel."
flags |>
  filter(k > 0, !misfit, method %in% fig_methods) |>
  group_by(k, rho, n, method, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Datasets with a fitting item flagged (%)") +
  facet_wrap(~ cell, nrow = 2)
```

This is where S-X2 differs. With three misfitting items, infit flags fitting
items in a share of datasets that rises steeply with n: at $\rho$ = 0.50, 16
percent at n = 500, 39.5 at 1000 and 87 at 2000. The S-X2 bootstrap stays at 2
to 7.5 percent in the same cells. At $\rho$ = 0.15 infit flags fitting items in
97 percent of datasets at n = 1000 against 6.5 for the S-X2 bootstrap, and only
at n = 2000 does the S-X2 bootstrap also cascade, to 47.5 percent. With a single
misfitting item the cascade is milder for infit, up to 21 percent at $\rho$ =
0.50 and n = 2000, against 5 percent (@tbl-falseflags).

The fitting items are fitting by construction, since they respond only to the
first dimension. In the unidimensional model fitted to the whole set, the
misfitting items distort the person locations, so fitting items can show misfit
too. This is the cascade described in Johansson (2025). Flagging them is a false
positive with respect to the data generating model.

### Recovering the right items

```{r}
#| label: fig-perfect
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Share of datasets in which every misfitting item is flagged and no
#|   fitting item is. Panels as in @fig-detection."
#| fig-alt: "Five panels plotting perfect recovery against sample size. With
#|   three misfitting items at rho 0.15 and 0.50, the S-X2 bootstrap climbs to
#|   about 93 percent at n = 1000 or 2000 while infit with Westfall-Young falls
#|   towards zero at n = 2000. At rho = 0.75 infit is higher than the S-X2
#|   bootstrap at every n. With one misfitting item both methods are high from
#|   n = 500, and the S-X2 bootstrap is higher at n = 2000 for rho = 0.50."
flags |>
  filter(k > 0, method %in% fig_methods) |>
  group_by(k, rho, n, method, rep) |>
  summarise(v = all(flag[misfit]) & !any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(v), .groups = "drop") |>
  cell_lab() |>
  rate_plot("Perfect recovery (%)") +
  facet_wrap(~ cell, nrow = 2)
```

The two effects trade off. With three misfitting items the S-X2 bootstrap
recovers the right set far more often at n $\geq$ 1000 when misfit is strong or
moderate: 93.5 against 3 percent at $\rho$ = 0.15 and n = 1000, and 92.5 against
13 percent at $\rho$ = 0.50 and n = 2000. Infit is better at small n and for
subtle misfit, where the S-X2 bootstrap misses too many of the misfitting items:
41.5 against 26.5 percent at $\rho$ = 0.75 and n = 2000, and 47 against 30
percent at $\rho$ = 0.50 and n = 500 (@tbl-perfect).

### Detection by item location

```{r}
#| label: fig-location
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Detection of each of the three misfitting items by its location
#|   relative to the sample mean, for the two subtler conditions."
#| fig-alt: "Two rows of three panels, rows for rho 0.50 and 0.75, columns for
#|   misfitting items at about -2, -1 and 0 logits, each plotting detection
#|   against sample size for infit with Westfall-Young and the S-X2 bootstrap.
#|   Detection is lowest for the item at -2 logits in both rows. At rho = 0.75
#|   infit is well above the S-X2 bootstrap for every item, most of all for the
#|   two items nearer the centre."
flags |>
  filter(k == 3, misfit, rho > 0.15, method %in% c("infit_wy", "sx2_boot")) |>
  mutate(location = factor(sprintf("item at %.2f logits", location),
                           levels = sprintf("item at %.2f logits",
                                            sort(unique(location)))),
         rho = sprintf("rho = %.2f", rho)) |>
  group_by(rho, n, location, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  rate_plot("Detection (%)") +
  facet_grid(rho ~ location)
```

Both methods find it hardest to detect the item two logits below the sample
mean, and the S-X2 bootstrap loses most relative to infit on the two items
closer to the centre at $\rho$ = 0.75 (@tbl-location).

### At a matched null error rate

The S-X2 bootstrap is slightly above 5 percent under the null at small n and
infit slightly below, which favours S-X2 in the comparisons above. In
@tbl-matched each method's adjusted p-value threshold is set so that its
family-wise rate in the null cell at the same n is 5 percent, and that threshold
is applied to the misfit cells. The thresholds come from the same design, so
they are an oracle.

Matching the null error rate moves detection by at most 3.3 points. False flags
move more where a threshold loosens, by up to 6.5 points for infit and 13.5 for
the S-X2 bootstrap, both at $\rho$ = 0.15, where the cascade is steepest. Infit
still detects more in every unsaturated cell, and the S-X2 bootstrap still flags
far fewer fitting items when three items misfit.

### Paired comparison

Both methods were run on the same datasets, so their decisions can be compared
item by item. @tbl-paired counts misfitting items detected by one method and
not the other. Where the two methods disagree about a misfitting item, it is
almost always infit that detects it. The largest counts in the other direction,
15 to 17 out of 600, are in cells where infit alone detects 92 to 172.

## Interpretation

**1. Infit with Westfall-Young is calibrated and detects more.** It holds 4.2 to
4.4 percent family-wise under the null and detects at least as much as the S-X2
bootstrap and the df-corrected S-X2 in every cell, at nominal and at matched
error rates. For subtle
misfit and moderate n the difference is large. The original study's
conclusion 1 survives the fair comparison.

**2. The S-X2 bootstrap resists the cascade.** When three items misfit at
n $\geq$ 1000, it rarely flags fitting items, while infit flags them in most
datasets. Its advantage in recovering the right set of items is large where
misfit is strong or moderate and n is large. The original study's conclusion 2
survives, and the gap is wider than it reported. At $\rho$ = 0.15 and n = 2000
infit with Westfall-Young flags 57.1 percent of fitting items, close to the 54.2
percent the original found without multiplicity correction, while the S-X2
bootstrap flags 3.7 percent against 16.9 for the original's asymptotic S-X2.
Multiplicity control does not remove the cascade for infit, since the fitting
items it flags there are far outside their null distribution.

**3. Asymptotic S-X2 is liberal under the null, but resists the cascade.**
With the df corrected it is still liberal under the null, 5.8 to 9.0 percent
with BH and 5.8 to 8.6 with Holm, and with mirt's df it is 8.8 to 12.4. With BH
it also cascades much more than the bootstrap version, to 28 percent at
$\rho$ = 0.50 and n = 2000, but that is the correction, not the statistic. BH
controls the false discovery rate and loosens its threshold once the misfitting
items are rejected. Recomputed from the stored p-values with Holm, which
controls the family-wise rate as Westfall-Young does, asymptotic S-X2 flags
fitting items in 2.5, 4.5 and 9 percent of datasets at $\rho$ = 0.50 and
n = 500, 1000 and 2000 with three misfitting items, against 2, 3.5 and 7.5 for
the bootstrap, at the same or slightly higher detection. At $\rho$ = 0.15 and
n = 1000 it is 11 against 6.5. Infit with Holm still cascades, 90.5 percent at
$\rho$ = 0.15 and n = 1000 and 72.5 at $\rho$ = 0.50 and n = 2000. The
resistance therefore belongs to the S-X2 statistic, and the bootstrap's
contribution is calibration under the null. A likely reason is that the cascade
leaves fitting items with small slope deviations, which infit, close to a
one-df slope contrast, picks up while S-X2 spreads them over 15 to 17 df at
n = 1000 and 2000. The book chapter uses Holm throughout for this reason.

**4. RMSEA adds nothing beyond S-X2.** A bootstrap cutoff makes it usable, which
confirms the original study's hypothesis, but it then behaves like the S-X2
bootstrap. A fixed threshold of .05 remains unusable.

**Against the decision rule.** The rule set before the run was that S-X2 is
worth adding if its bootstrap version holds 5 percent under the null and, in a
cell with $\rho$ = 0.50 or 0.75, detects misfit infit misses or flags clearly
fewer fitting items, without a comparable loss elsewhere. The first two parts
are met: the S-X2 bootstrap is at most slightly liberal at small n, and at
$\rho$ = 0.50 with three misfitting items it flags fitting items in 3.5 against
39.5 percent of datasets at n = 1000 and 7.5 against 87 at n = 2000, with
detection 93.8 against 98 and 100 against 100. The third part is a judgement.
At $\rho$ = 0.75 the S-X2 bootstrap detects much less than infit at every n up
to 1000.

So S-X2 is not a replacement for infit, and it is not a better default. Its use
would be as a second opinion at large n: when infit flags many items, the S-X2
bootstrap separates the items that misfit from those that are flagged because
others do. The cost is a parametric bootstrap with a refit per draw, about 25
to 40 seconds per analysis at B = 400 on one core for 20 items.

## Scope and limits

- Dichotomous items only. The diagnostic found S-X2 close to nominal for
  polytomous items after the df correction, so a polytomous arm is the obvious
  next step if S-X2 goes further.
- One misfit mechanism, a second dimension. Other mechanisms, such as
  discrimination different from the Rasch model, may favour either method.
- S-X2 is computed by mirt under MML with a normal latent distribution, which
  matches the data generation here and so favours it slightly.
- The bootstrap uses mirt's df + 1 to put S-X2 on a common scale across draws.
  The bootstrap reference does not depend on the df otherwise.

## Appendix: result tables

```{r}
#| label: tbl-null
#| tbl-cap: "Datasets with at least one flagged item under the null (%), k = 0."
null_fw |>
  pivot_wider(names_from = n, values_from = v, names_prefix = "n = ") |>
  kb(digits = 1)
```

```{r}
#| label: tbl-detection
#| tbl-cap: "Detection, mean over the misfitting items (%)."
main <- c("infit_wy", "infit_bh", "sx2_df1", "sx2_boot", "rmsea_05")
flags |>
  filter(k > 0, misfit, method %in% main) |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-falseflags
#| tbl-cap: "Datasets with at least one fitting item flagged (%)."
flags |>
  filter(k > 0, !misfit, method %in% main) |>
  group_by(k, rho, n, method, rep) |>
  summarise(any = any(flag), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(any), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-perfect
#| tbl-cap: "Perfect recovery: every misfitting item flagged and no fitting item (%)."
flags |>
  filter(k > 0, method %in% main) |>
  group_by(k, rho, n, method, rep) |>
  summarise(ok = all(flag[misfit]) & !any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(v = 100 * mean(ok), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-location
#| tbl-cap: "Detection by item location, k = 3 (%)."
flags |>
  filter(k == 3, misfit, rho > 0.15, method %in% c("infit_wy", "sx2_boot")) |>
  mutate(location = round(location, 2)) |>
  group_by(rho, n, location, method) |>
  summarise(v = 100 * mean(flag), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = v) |>
  kb(digits = 1)
```

```{r}
#| label: tbl-matched
#| tbl-cap: "Detection and false flags with each method's threshold set to a 5
#|   percent family-wise rate in the null cell at the same n (%)."
thr <- adj |>
  filter(k == 0) |>
  select(n, rep, infit_wy, sx2_df1, sx2_boot) |>
  pivot_longer(c(infit_wy, sx2_df1, sx2_boot), names_to = "method",
               values_to = "padj") |>
  group_by(method, n, rep) |>
  summarise(mn = min(padj, na.rm = TRUE), .groups = "drop") |>
  group_by(method, n) |>
  summarise(c = quantile(mn, .05, type = 1), .groups = "drop")

adj |>
  filter(k > 0) |>
  select(k, rho, n, rep, misfit, infit_wy, sx2_df1, sx2_boot) |>
  pivot_longer(c(infit_wy, sx2_df1, sx2_boot), names_to = "method",
               values_to = "padj") |>
  left_join(thr, by = c("method", "n")) |>
  mutate(flag = coalesce(padj <= c, FALSE)) |>
  group_by(k, rho, n, method, rep) |>
  summarise(det = mean(flag[misfit]), ff = any(flag[!misfit]), .groups = "drop") |>
  group_by(k, rho, n, method) |>
  summarise(det = 100 * mean(det), ff = 100 * mean(ff), .groups = "drop") |>
  pivot_wider(names_from = method, values_from = c(det, ff)) |>
  select(k, rho, n, det_infit_wy, det_sx2_boot, ff_infit_wy, ff_sx2_boot) |>
  kb(digits = 1, col.names = c("k", "rho", "n", "Detection, infit",
                               "Detection, S-X2 boot", "False flags, infit",
                               "False flags, S-X2 boot"))
```

```{r}
#| label: tbl-paired
#| tbl-cap: "Misfitting items detected by one method and not the other, infit
#|   with Westfall-Young against the S-X2 bootstrap, same datasets."
flags |>
  filter(k > 0, misfit, method %in% c("infit_wy", "sx2_boot")) |>
  select(k, rho, n, rep, item, method, flag) |>
  pivot_wider(names_from = method, values_from = flag) |>
  group_by(k, rho, n) |>
  summarise(`infit only` = sum(infit_wy & !sx2_boot),
            `S-X2 boot only` = sum(!infit_wy & sx2_boot),
            both = sum(infit_wy & sx2_boot), .groups = "drop") |>
  kb()
```

## Reference

Johansson, M. (2025). Detecting item misfit in Rasch models. *Educational
Methods & Psychometrics, 3*(18). https://doi.org/10.61186/emp.2025.5
