---
title: "An exact null for the PCA/t-test unidimensionality protocol"
subtitle: "Why the person-level t-test is not a 5% test, and what that does to the binomial step"
author: "Magnus Johansson"
date: today
bibliography: ttest-protocol-refs.bib
---

```{r setup}
#| include: false
if (requireNamespace("devtools", quietly = TRUE) && file.exists("../DESCRIPTION")) {
  suppressMessages(devtools::load_all("..", quiet = TRUE))
} else {
  library(easyRasch2)
}
library(knitr)
library(ggplot2)
```

## Purpose

The PCA/t-test protocol [@smith2002; @tennant2006] splits a scale into two item
sets from a principal component analysis of residuals, estimates a person
location from each set, compares them per respondent with a t-test against
1.96, and refers the resulting proportion of significant tests to a 5%
benchmark through a binomial confidence interval. Hagell [@hagell2014] showed that the
conclusion depends on sample size and on which of five binomial interval methods
is used, and argued the protocol should be read as ordinary hypothesis testing
rather than as a definite test.

That leaves the reference proportion itself unexamined. This note shows that the
1.96 rule is not a 5% test, computes its exact size, and records what that does
to the binomial step. The conclusion is that the choice of interval method is a
second-order question about the shape of a distribution whose center is wrong by
a factor of 1.3 to 3.7. The direction is uniform, so the protocol is biased toward
declaring unidimensionality.

## The statistic is the change index with subtests in place of occasions

The protocol computes

$$t = \frac{\hat\theta_A - \hat\theta_B}{\sqrt{SE(\hat\theta_A)^2 + SE(\hat\theta_B)^2}}$$

which is the Rasch change index of `dev/rci_exact_null.qmd` with item subtests
in place of occasions. Both mechanisms documented there carry over.

**Self-normalisation.** $SE(\hat\theta_A)$ and $SE(\hat\theta_B)$ are
deterministic functions of the realised subtest scores, so the denominator is
not independent of the numerator, both tails of the null are damped, and the
critical value falls below 1.96. The "t" label is a misnomer for the same
reason: a t denominator is independent of its numerator, which is the one
property this denominator does not have.

**Discreteness.** With complete data each subtest score is sufficient, so
$\hat\theta_A$ and $\hat\theta_B$ take only as many values as there are subtest
scores and the null is a distribution over $(k_A+1) \times (k_B+1)$ cells.

Conditional independence, which is what makes the enumeration possible, also
carries over, but it has a different standing in the two settings. In the change
case it is an auxiliary assumption. The null there is that $\theta$ did not
move, and carry-over between two administrations of the same items can be
present while $\theta$ is fixed, so separate administration does not deliver
conditional independence on its own. `dev/rci_exact_null.qmd` has a section on
what happens when it fails. Here the item sets are disjoint, so local
independence gives $P(r_A, r_B \mid \theta) = P(r_A \mid \theta)\,P(r_B \mid
\theta)$, and that statement is the null hypothesis rather than a condition
attached to it. A failure is what the protocol exists to detect, so enumerating
under it is the right move.

What the protocol cannot do is say which kind of failure it found. Marais and
Andrich [@marais2008a] separate **trait dependence**, where a person parameter
beyond $\theta$ enters the responses, from **response dependence**, where the
response to one item governs the response to another. Only the first is
multidimensionality. Two locally dependent items in the same subtest reduce the
information that subtest carries below what the model assumes, so $SE_A$
understates the spread of $\hat\theta_A$ and $|t|$ is inflated, pushing the
protocol toward rejection in the same way a second dimension would. That follows
from the information rather than from anything measured here. It leaves the size
calculations below untouched, since those hold under a null with no dependence
of either kind, but it bears on what a rejection supports: that the item set
fails local independence, not the stronger claim of multidimensionality that the
procedure's name carries.

## Conditioning on the total score removes the latent distribution

The RCI note enumerates the null at a set of person locations and pools over
them, which makes the pooled critical value a property of the sample as well as
the items. The subtest case admits a stronger construction. Under the Rasch
model the conditional distribution of $r_A$ given the total score $r_A + r_B$ is
free of $\theta$, by the same sufficiency that underlies CML estimation:

$$P(r_A = a \mid r_A + r_B = r) = \frac{\gamma_A(a)\,\gamma_B(r-a)}{\gamma(r)}$$

So the null of $t$ can be enumerated per total score with no latent distribution
assumption and no plug-in person estimate. Numerically it is easier to evaluate
$P(r_A \mid \theta)P(r_B \mid \theta)$ at any convenient $\theta$ and normalise
within each total score, which gives the same ratio. Invariance is checked below
rather than assumed.

Sufficiency and the recursion divide the work. Sufficiency is what makes the
ratio free of $\theta$, and it also fixes the value of $t$ in every cell, since
$\hat\theta$ and $SE$ are functions of the subtest score alone. Neither of
those says what the $\gamma$ terms are, and a quantile needs them. Computing
them is what Lord-Wingersky does, as exact combinatorics rather than
estimation.

```{r exact-fns}
# Thresholds: k items, ncat response categories, locations spread over
# [-spread, spread]. Same constructor as dev/rci_exact_null.qmd.
mk <- function(k, ncat, spread = 1.2) {
  lapply(seq(-spread, spread, length.out = k),
         function(b) b + seq(-0.9, 0.9, length.out = ncat - 1L))
}

# Lord-Wingersky: exact sum-score distribution given theta under a PCM
score_dist <- function(thr_list, theta) {
  d <- 1
  for (thr in thr_list) {
    p  <- easyRasch2:::.pcm_cat_probs(theta, thr)
    nd <- numeric(length(d) + length(p) - 1L)
    for (a in seq_along(d)) nd[a + seq_along(p) - 1L] <-
        nd[a + seq_along(p) - 1L] + d[a] * p
    d <- nd
  }
  d                                   # index = score + 1
}

# Score -> (theta, SE) lookup, WLE. The lookup RMscoreSE() builds.
lut <- function(thr_list) {
  steps <- vapply(thr_list, length, integer(1L))
  m <- vapply(0:sum(steps), function(r)
    easyRasch2:::.theta_wle(easyRasch2:::.score_pattern(r, steps),
                            thr_list, c(-10, 10)), numeric(2L))
  list(theta = m[1L, ], se = m[2L, ])
}

# The t grid and the raw logit-gap grid for one split
split_grid <- function(thr_all, idxA) {
  A <- thr_all[idxA]; B <- thr_all[-idxA]
  lA <- lut(A); lB <- lut(B)
  list(A = A, B = B, lA = lA, lB = lB,
       t = outer(seq_along(lA$theta), seq_along(lB$theta),
                 function(i, j) (lA$theta[i] - lB$theta[j]) /
                   sqrt(lA$se[i]^2 + lB$se[j]^2)),
       d = outer(seq_along(lA$theta), seq_along(lB$theta),
                 function(i, j) lA$theta[i] - lB$theta[j]))
}

# Null of t conditional on the total score. Free of theta.
cond_null <- function(g, theta_eval = 0) {
  dA <- score_dist(g$A, theta_eval); dB <- score_dist(g$B, theta_eval)
  maxA <- length(dA) - 1L; maxB <- length(dB) - 1L
  lapply(0:(maxA + maxB), function(r) {
    a <- max(0L, r - maxB):min(maxA, r)
    w <- dA[a + 1L] * dB[r - a + 1L]
    list(a = a, p = w / sum(w),
         t = g$t[cbind(a + 1L, r - a + 1L)],
         d = g$d[cbind(a + 1L, r - a + 1L)])
  })
}

alpha_by_r <- function(cn, crit = 1.96)
  vapply(cn, function(x) sum(x$p[abs(x$t) > crit]), numeric(1L))

gap_by_r <- function(cn) vapply(cn, function(x) sum(x$p * x$d), numeric(1L))

# Total-score distribution under a latent density, by quadrature
score_weights <- function(thr_all, mu = 0, sigma = 1.4, nq = 121L) {
  q <- seq(mu - 5 * sigma, mu + 5 * sigma, length.out = nq)
  w <- dnorm(q, mu, sigma); w <- w / sum(w)
  Reduce(`+`, Map(function(t, wt) wt * score_dist(thr_all, t), q, w))
}

# Pooled exact critical values achieving nominal alpha
crit_pooled <- function(cn, fr, alpha = 0.05) {
  tv <- unlist(lapply(cn, `[[`, "t"))
  pv <- unlist(Map(function(x, f) x$p * f, cn, fr))
  o <- order(tv); tv <- tv[o]; pv <- pv[o] / sum(pv)
  cw <- cumsum(pv)
  c(lower = tv[which(cw >= alpha / 2)[1L]],
    upper = tv[which(cw >= 1 - alpha / 2)[1L]])
}

# Exact Poisson-binomial distribution of the flagged count
pois_binom <- function(p) {
  d <- 1
  for (pi in p) d <- c(d * (1 - pi), 0) + c(0, d * pi)
  d                                   # index = count + 1
}
```

## Validation

Three anchors, following the discipline of the RCI note: reproduce an
identity to machine precision, and check the fast path against simulation.

```{r anchors}
#| echo: true
thr  <- mk(12, 4)
idxA <- seq(1, 12, by = 2)            # interleaved 6 / 6
g    <- split_grid(thr, idxA)

# Anchor 1: is the conditional null actually free of theta?
cn0 <- cond_null(g, 0); cnm <- cond_null(g, -2); cnp <- cond_null(g, 2.5)
dev1 <- max(vapply(seq_along(cn0), function(i)
  max(abs(cn0[[i]]$p - cnm[[i]]$p), abs(cn0[[i]]$p - cnp[[i]]$p)), numeric(1L)))

# Anchor 3: does the exact pooled critical value recover nominal alpha?
fr <- score_weights(thr, 0, 1.4); fr <- fr / sum(fr)
cp <- crit_pooled(cn0, fr, 0.05)
ach <- sum(vapply(seq_along(cn0), function(i) fr[i] *
  sum(cn0[[i]]$p[cn0[[i]]$t < cp[["lower"]] | cn0[[i]]$t > cp[["upper"]]]),
  numeric(1L)))
```

**Anchor 1, theta-invariance.** Evaluating the conditional distribution at
$\theta = -2$, $0$ and $2.5$ agrees to a maximum absolute difference of
`r sprintf("%.1e", dev1)`. The construction is free of $\theta$ rather than
approximately so.

**Anchor 2, against simulation.** Four hundred thousand respondents on the
12-item scale, drawn from $N(0, 1.4)$, scored and classified by the same 1.96
rule. Per-total-score $\alpha$ agrees with the enumeration to a maximum absolute
difference of 0.0040 against a Monte Carlo standard error of 0.0039. The
simulation is not run at render time. It lives in
`dev/ttest_protocol_exact_null.R`.

**Anchor 3, self-consistency.** Feeding the exact pooled critical values back
through the enumeration gives $\alpha =$ `r sprintf("%.4f", ach)` against a
nominal 0.05, the residual being discreteness.

## The 1.96 rule is not a 5% test

```{r sweep}
#| echo: true
run <- function(label, k, ncat, idxA, mu = 0, sigma = 1.4) {
  thr <- mk(k, ncat); g <- split_grid(thr, idxA); cn <- cond_null(g, 0)
  fr <- score_weights(thr, mu, sigma); fr <- fr / sum(fr)
  a <- alpha_by_r(cn, 1.96); gp <- gap_by_r(cn); cp <- crit_pooled(cn, fr, 0.05)
  nz <- a > 0
  data.frame(config = label, subtests = paste0(length(idxA), "/", k - length(idxA)),
             alpha = sum(fr * a), ratio = 0.05 / sum(fr * a),
             pct_zero = 100 * sum(fr[!nz]),
             lo = cp[["lower"]], hi = cp[["upper"]],
             asym = abs(cp[["lower"]]) - cp[["upper"]],
             gap = max(abs(gp[nz])), stringsAsFactors = FALSE)
}

res <- do.call(rbind, list(
  run("12 items, 4 cat",               12, 4, seq(1, 12, by = 2)),
  run("12 items, 4 cat",               12, 4, seq(1, 12, by = 3)),
  run("20 items, 4 cat",               20, 4, seq(1, 20, by = 2)),
  run("20 items, 4 cat",               20, 4, seq(1, 20, by = 3)),
  run("20 items, 4 cat",               20, 4, seq(1, 20, by = 5)),
  run("30 items, 4 cat",               30, 4, seq(1, 30, by = 2)),
  run("40 items, 4 cat",               40, 4, seq(1, 40, by = 2)),
  run("20 items, 2 cat",               20, 2, seq(1, 20, by = 2)),
  run("20 items, 7 cat",               20, 7, seq(1, 20, by = 2)),
  run("20 items, 4 cat, easy/hard",    20, 4, 1:10),
  run("20 items, 4 cat, 2 logits off", 20, 4, seq(1, 20, by = 2), mu = -2)))

kable(data.frame(res$config, res$subtests, round(res$alpha, 4),
                 round(res$ratio, 2), round(res$pct_zero, 1)),
      col.names = c("Scale", "Split", "alpha", "Too lenient",
                    "Unflaggable (%)"),
      row.names = FALSE)
```

The true size of the 1.96 rule runs from 1.4% to 3.9%. It never reaches 5%, and
at 40 items split 20/20 it is still short by a factor of 1.27. Dichotomous items
are the worst, with the polytomous-versus-dichotomous contrast at 0.0137, and a sample two
logits off target is nearly as bad at 0.0177. Both findings match the RCI note,
where response categories moved the critical value mainly in the step from two
to more than two, and where mistargeting dropped the pooled value from 1.695 to
1.522 because a floor-bound sample is located where the statistic is most
damped.

The effective test length here is the subtest, not the scale, which is what
makes the protocol worse than the change case rather than equivalent to it. The
RCI note's length table reads 1.644 at 4 items, 1.725 at 6, 1.815 at 10 and
1.868 at 20, four categories throughout and persons from $N(0, 1.4)$. A 20-item scale splitting 10/10 is
governed by two 10-item instruments. The published recommendation to use the
protocol with at least 12 item response thresholds [@tennant2006; @horton2010]
amounts to four items at four categories, which is the worst row of that
table.

## A large share of respondents cannot be flagged at all

The last column above counts respondents for whom **no attainable pair of
subtest scores produces $|t| > 1.96$**. Their contribution to the significant
count is zero with probability one, not 0.05. At 20 dichotomous items that is
44% of the sample and on an off-target polytomous scale 46%.

```{r fig-alpha}
#| echo: true
#| fig-width: 6
#| fig-height: 3.8
#| fig-cap: "Exact size of the 1.96 rule by total score, under a perfectly
#|   unidimensional Rasch model. Weighted by a $N(0, 1.4)$ latent density to give
#|   the sample means quoted in the text. The dashed line is the 5% the protocol
#|   assumes. Scores at both ends carry a size of zero because no attainable
#|   split of that total clears 1.96."
#| fig-alt: "Three panels, one per scale, plotting exact type I error against
#|   total score. Each panel shows a flat zero region at both ends of the score
#|   range and a ragged sawtooth in between that mostly falls below the dashed
#|   0.05 reference line. The dichotomous panel has the widest zero regions and
#|   the lowest sawtooth."
panel <- function(label, k, ncat, idxA) {
  g <- split_grid(mk(k, ncat), idxA)
  a <- alpha_by_r(cond_null(g, 0), 1.96)
  data.frame(label = label, r = seq_along(a) - 1L, alpha = a)
}
d_fig <- rbind(
  panel("20 items, 4 cat, 10/10", 20, 4, seq(1, 20, by = 2)),
  panel("12 items, 4 cat, 6/6",   12, 4, seq(1, 12, by = 2)),
  panel("20 items, 2 cat, 10/10", 20, 2, seq(1, 20, by = 2)))
d_fig$label <- factor(d_fig$label, levels = unique(d_fig$label))

ggplot(d_fig, aes(r, alpha)) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "grey40") +
  geom_step(direction = "mid") +
  facet_wrap(~ label, scales = "free_x") +
  labs(x = "Total score across all items", y = "Exact type I error of 1.96") +
  theme_bw() + er2_axis_margins()
```

The sawtooth is discreteness. Adjacent total scores differ in how many attainable
splits they have and where those splits fall relative to the cutoff, so the size
of the test alternates rather than varying smoothly. The RCI note found the same
structure in the conditional critical value, which formed an arch across the
score range rather than a constant.

The figure also shows that the conservatism is not a uniform shrinkage. Around
the middle of the scale the size peaks at 0.062, 0.062 and 0.054 in the three
panels, so the test is mildly anti-conservative there. Under the $N(0, 1.4)$
weighting, 20%, 6% and 10% of respondents hold a total score whose size exceeds
0.05. The whole of the deficit comes from the ends, where attainable scores run
out and the size falls to zero. Averaging over the score distribution is what
the protocol does when it pools respondents into one proportion, and that
average is what has to be compared with 5%, but a reader should know that the
pooled figure is combining two opposite behaviours rather than describing a
test that is uniformly too strict.

## Asymmetry and centering

```{r asym}
#| echo: true
sel <- c(1, 2, 5, 10, 3)
kable(data.frame(res$config[sel], res$subtests[sel], round(res$lo[sel], 3),
                 round(res$hi[sel], 3), round(res$asym[sel], 3),
                 round(res$gap[sel], 3)),
      col.names = c("Scale", "Split", "Lower", "Upper", "Asym.",
                    "Gap (logits)"),
      row.names = FALSE)
```

The RCI note corrected an earlier belief that two-sided simulated critical values
were asymmetric: under the change null the two occasions are exchangeable, so the
difference is symmetric about zero however skewed $\hat\theta$ is, and
exchangeability fails only when the occasions differ in missingness. Subtests
break exchangeability by construction, because they differ in length, targeting
and information. A balanced interleaved split is nearly symmetric, so the change
result transfers there. An unbalanced or difficulty-confounded split is not. The
ten easiest items against the ten hardest gives correct two-sided values of
-2.001 and 1.578, so the upper value is stricter than 1.96 while the lower is
more lenient. A principal component of residuals often separates items by
location as well as by content, which makes that row the relevant one.

The last column is $\max_r |E[\hat\theta_A - \hat\theta_B \mid r]|$ under a
perfectly unidimensional model. On a 4/16 split it reaches 0.557 logits, purely
from the difference in WLE bias between a four-item and a sixteen-item subtest.
That is model-implied signal with no multidimensionality behind it. Reporting
quantity and units at every surface, as the RCI note insists, matters here: the
gap is in logits, the t is unitless, and only the second is being compared with
1.96.

## What happens at the count level

```{r count}
#| echo: true
count_null <- function(label, k, ncat, idxA, n, mu = 0, sigma = 1.4) {
  thr <- mk(k, ncat); g <- split_grid(thr, idxA); cn <- cond_null(g, 0)
  fr <- score_weights(thr, mu, sigma); fr <- fr / sum(fr)
  a  <- alpha_by_r(cn, 1.96)
  nr <- round(n * fr); p_i <- rep(a, times = nr)
  d  <- pois_binom(p_i); cnt <- seq_along(d) - 1L; m <- length(p_i)
  mean_p <- sum(d * cnt) / m
  data.frame(config = label, n = m,
             mean_p = 100 * mean_p,
             q95    = 100 * cnt[which(cumsum(d) >= 0.95)[1L]] / m,
             p_over = sum(d[cnt / m > 0.05]),
             sd_pb  = sqrt(sum(d * cnt^2) - sum(d * cnt)^2),
             sd_bin = sqrt(m * mean_p * (1 - mean_p)),
             stringsAsFactors = FALSE)
}

cfgs <- list(list("20 items, 4 cat, 10/10", 20, 4, seq(1, 20, by = 2)),
             list("12 items, 4 cat, 6/6",   12, 4, seq(1, 12, by = 2)),
             list("20 items, 2 cat, 10/10", 20, 2, seq(1, 20, by = 2)),
             list("40 items, 4 cat, 20/20", 40, 4, seq(1, 40, by = 2)))
cts <- do.call(rbind, lapply(cfgs, function(cf)
  do.call(rbind, lapply(c(200, 500, 1000), function(n)
    count_null(cf[[1]], cf[[2]], cf[[3]], cf[[4]], n)))))

kable(data.frame(cts$config, cts$n, round(cts$mean_p, 2), round(cts$q95, 2),
                 round(cts$p_over, 4), round(cts$sd_pb, 2), round(cts$sd_bin, 2)),
      col.names = c("Scale", "n", "Mean (%)", "95th (%)", "P(>5%)",
                    "SD PB", "SD bin"),
      row.names = FALSE)
```

At $n = 1000$, observing more than 5% significant tests is an event of
probability 0.0007, 0.0000, 0.0000 and 0.044 under a perfectly unidimensional
model. The protocol treats 5% as the boundary of acceptability. The exact null
puts it in the upper tail.

**The binomial family is fine.** Conditioning on the observed scores makes the count
Poisson-binomial rather than binomial, since each respondent carries their own
$\alpha_r$. The last two columns show the two standard deviations agree to
within a percent throughout. The heterogeneity in $\alpha_r$ is too small
relative to its mean to move the variance, because
$\mathrm{Var}_{\text{PB}} = n\bar\alpha - \sum \alpha_i^2$ differs from
$n\bar\alpha(1-\bar\alpha)$ by $n\,\mathrm{Var}(\alpha_i)$, which is order
$10^{-4}$ against a mean of order $10^{-2}$. Marginally, with person locations
drawn iid from a latent density, the count is binomial in any case.

So the distributional shape Hagell [@hagell2014] investigates is not where the
protocol goes wrong. All five of the interval methods compared there are being applied to
roughly the right distribution and compared against the wrong number.

```{r fig-count}
#| echo: true
#| fig-width: 6
#| fig-height: 3.8
#| fig-cap: "Exact null distribution of the observed proportion of significant
#|   t-tests at $n = 500$, under a perfectly unidimensional Rasch model with a
#|   $N(0, 1.4)$ latent density. The dashed line is the protocol's 5% benchmark."
#| fig-alt: "Three overlaid distributions of the observed proportion of
#|   significant tests, one per scale, each roughly bell shaped and centered
#|   between one and four percent. All three lie almost entirely to the left of
#|   a dashed vertical line at five percent."
dist_one <- function(label, k, ncat, idxA, n = 500) {
  thr <- mk(k, ncat); g <- split_grid(thr, idxA)
  fr <- score_weights(thr, 0, 1.4); fr <- fr / sum(fr)
  p_i <- rep(alpha_by_r(cond_null(g, 0), 1.96), times = round(n * fr))
  d <- pois_binom(p_i)
  data.frame(label = label, prop = 100 * (seq_along(d) - 1L) / length(p_i),
             dens = d)
}
d_cnt <- rbind(
  dist_one("20 items, 4 cat, 10/10", 20, 4, seq(1, 20, by = 2)),
  dist_one("12 items, 4 cat, 6/6",   12, 4, seq(1, 12, by = 2)),
  dist_one("20 items, 2 cat, 10/10", 20, 2, seq(1, 20, by = 2)))

ggplot(subset(d_cnt, prop <= 8), aes(prop, dens, colour = label)) +
  geom_vline(xintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_step(direction = "mid") +
  labs(x = "Observed significant tests (%)", y = "Exact probability",
       colour = NULL) +
  theme_bw() + er2_axis_margins() +
  theme(legend.position = "bottom")
```

## Extreme-score handling does not rescue it

The lookup above uses WLE, which is finite at extreme scores. Software using
maximum likelihood person estimates has to drop respondents extreme on either
subtest, so the check is whether that changes the size.

```{r extremes}
#| echo: true
joint_w <- function(thr_all, idxA, mu = 0, sigma = 1.4, nq = 121L) {
  A <- thr_all[idxA]; B <- thr_all[-idxA]
  q <- seq(mu - 5 * sigma, mu + 5 * sigma, length.out = nq)
  w <- dnorm(q, mu, sigma); w <- w / sum(w)
  Reduce(`+`, Map(function(t, wt) wt * outer(score_dist(A, t), score_dist(B, t)),
                  q, w))
}
chk <- function(label, k, ncat, idxA, mu = 0) {
  thr <- mk(k, ncat); g <- split_grid(thr, idxA)
  W <- joint_w(thr, idxA, mu); W <- W / sum(W)
  sig <- abs(g$t) > 1.96
  keep <- matrix(TRUE, nrow(W), ncol(W))
  keep[c(1L, nrow(W)), ] <- FALSE; keep[, c(1L, ncol(W))] <- FALSE
  data.frame(config = label, all = sum(W * sig),
             pct_ex = 100 * (1 - sum(W[keep])),
             dropped = sum(W[keep] * sig[keep]) / sum(W[keep]))
}
ex <- rbind(
  chk("12 items, 4 cat, 6/6",       12, 4, seq(1, 12, by = 2)),
  chk("20 items, 4 cat, 10/10",     20, 4, seq(1, 20, by = 2)),
  chk("20 items, 4 cat, 4/16",      20, 4, seq(1, 20, by = 5)),
  chk("40 items, 4 cat, 20/20",     40, 4, seq(1, 40, by = 2)),
  chk("20 items, 2 cat, 10/10",     20, 2, seq(1, 20, by = 2)),
  chk("20 items, 4 cat, off target", 20, 4, seq(1, 20, by = 2), mu = -2))
kable(data.frame(ex$config, round(ex$all, 4), round(ex$pct_ex, 1),
                 round(ex$dropped, 4)),
      col.names = c("Scale", "alpha, all", "Extreme (%)",
                    "alpha, dropped"),
      row.names = FALSE)
```

Dropping extremes moves the size by two or three thousandths. The finding is not
an artifact of the person estimator's behaviour at the boundary.

## What this changes about the protocol

The protocol's decision rule compares an observed proportion against 5%. That
benchmark is too lenient by a factor of 1.3 to 3.7 across everything tried here,
worst on dichotomous items and off-target samples, which are common in health
measurement. The direction is uniform, so the protocol is biased toward
declaring unidimensionality.

This also gives a mechanism for Hagell's central finding [@hagell2014]. Comparing against
a benchmark that is too lenient shifts the whole power curve toward passing, so
conclusions flipping with sample size is ordinary power behaviour against a
biased reference rather than a pathology of the binomial interval. With the
correct reference, mild multidimensionality would be rejected at appreciably
smaller samples, and larger $n$ would behave the way it does in any well posed
test.

The constructive version is cheap. The enumeration needed to state the correct
reference is the same machinery that produced it: one Lord-Wingersky recursion
per subtest and one score-to-measure lookup, both of which the package already
builds for `RMscoreSE()`. From that, either report the exact $\bar\alpha$ and
test the observed count against it, or replace 1.96 with the exact per-score
critical values so that the 5% benchmark becomes true by construction. The
second is closer to what users expect the protocol to mean and needs no change
to how results are reported.

## Not yet done: the split is chosen from the data

Everything above conditions on a fixed, known split. The protocol chooses the
split by principal component analysis of residuals, taking loadings above +0.3
and below -0.3, a cutoff whose rationale Hagell [@hagell2014] already calls unclear.
That is a selection step, and the null conditional on a data-chosen contrast is not
the null for a fixed one. The enumeration therefore understates the size of the
protocol as practised.

The two errors run in opposite directions. Referring to 1.96, together with the
respondents who cannot be flagged at all, pushes the realised proportion below
5% and makes the protocol too lenient. Selecting the most discrepant contrast
pushes it up and makes the protocol too strict. Which dominates depends on item
count, response categories, targeting and how unbalanced the split is, and
nothing here settles it.

Settling it needs a parametric bootstrap of the whole pipeline: generate under
the fitted unidimensional model, run the PCA, take the split, run the t-tests,
count. That is the same shape as `RMdimResidualPCACutoff()` and the infit and Q3
cutoff work, and the exact component above would serve as its reference, which
removes most of the noise from the comparison the way the exact RCI null did for
the change study.

It is worth holding open that the two errors partly cancel. Existing simulation
work reporting the protocol as well behaved under unidimensionality would be
consistent with that, and a cancellation would not be stable across the
conditions in the first table. I have not checked what Smith [@smith2002] or Horton and
Tennant [@horton2010] report on this point, so that reading is a conjecture to verify
against the sources rather than a finding of this note.

## Scope and limits

* **Complete data only**, as in the RCI note. Missing responses break
  sufficiency of a single subtest score, so the enumeration would run per
  missingness pattern.
* **Fixed split.** See the section above.
* **WLE with CML-style thresholds.** RUMM2030's person estimator may differ. It
  would move the centering column more than the size column, since dropping
  extremes barely changed the latter.
* **The latent density enters only through the weighting.** The per-score sizes
  are free of it. Only the sample means and the count distributions depend on
  the assumed $N(\mu, \sigma)$.
* The 1.96 figure is taken as the rule in force. Software referring to a $t$
  distribution with finite degrees of freedom uses a slightly larger value,
  which makes the test more conservative, not less.

## Reproduction

`dev/ttest_protocol_exact_null.R` runs everything here plus the Monte Carlo
anchor, which is not rendered. The enumeration is self-contained in the
`exact-fns` chunk above and needs only `easyRasch2:::.pcm_cat_probs()`,
`:::.theta_wle()` and `:::.score_pattern()`, the same three internals the RCI
note uses.
